### 2017-11-06 paper.tex
extended summary

%% NSS-MIC_Instructions.tex
%% 8/2007
%% By Bo Yu (yu@bnl.gov)
%% based on:
%% bare_jrnl.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************
\documentclass[journal]{IEEEtran}
\usepackage{graphicx}

\begin{document}
\title{Software reliability: Experiences\\
in European scientific research\\
projects and new trends}
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{D.C. Duma, %~\IEEEmembership{Member,~IEEE,}
        E. Ronchieri, %~\IEEEmembership{Fellow,~OSA,}
        P. Orviz Fernandez, %~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
        M. David, %~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
        J. Gomes, %~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
        D. Salomoni, %~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
\thanks{D. C. Duma, E. Ronchieri, and D. Salomoni are with INFN - CNAF, Bologna, Italy.}%
\thanks{P. Orviz Fernandez, is with CSIC, Santander, Spain, (e-mail: orviz@ifca.unican.es)}%
\thanks{M. David, and J. Gomes are with LIP, Lisbon, Portugal}%
}

\maketitle
\pagestyle{empty}
\thispagestyle{empty}

\begin{abstract}
In this work we describe the trends in scientific software reliability by analyzing past European Commission funded software development projects in the context of distributed computing field such as Enabling Grids for E-sciencE and European Middleware Initiative, and the ongoing INDIGO-DataCloud.
\end{abstract}

%\begin{IEEEkeywords}
%IEEEtran, journal, \LaTeX, paper, template.
%\end{IEEEkeywords}

\section{Evolution of Software Reliability}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
%
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
%
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
%
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
%
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{I}{n} the European scientific research arena, the analysis of some of the last decade software development projects depicts a continuous increase of the prominence and robustness of testing and validation procedures. The observed trends are aligned with the evolution of the software engineering techniques (e.g. rise of DevOps practices \cite{devops}) over the last years. This is tied to the parallel advancements in the ICT field (with virtualization and container technologies) and the appearance of automation and event-response tools. In the following we introduce a set of European Commission funded projects, going from 2001 up to now, where the authors of this summary were and are currently involved. For each project we explain how they addressed the challenge in software reliability.

The DataGrid \cite{cordis:datagrid} project (Jan 2001 - Dec 2003) was devoted to develop software to provide basic Grid computing functionalities and related management tools for widely distributed scientific communities, such as Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics. In this project, a big effort was dedicated to define and implement academic software engineering practices due to the unusual context in which DataGrid sat, starting from the enormous geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}. DataGrid developed its own software distribution, EDG (European DataGrid), strongly based on Globus Middleware Services \cite{globus}. An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.

The Enabling Grids for E-sciencE (EGEE, Apr 2004 - Apr 2010) I \cite{cordis:egee}, II \cite{cordis:egee2}, III \cite{cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e-Science, available to scientists 24 hours-a-day. One of the main achievements of these projects was the middleware - a crucial component of any Grid infrastructure as it provides the ‘glue’ to link the hardware resources within the Grid. The gLite middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re-engineering efforts to converge with LHC Computing Grid (LCG-2), Virtual Data Toolkit (VDT) and Condor \cite{condor} software distributions. The overall goal of the middleware engineering group was to provide and maintain selected middleware services of the gLite distribution to satisfy the needs of the scientific communities. During EGEE-I, a software process was designed and put into action for the early releases of the gLite middleware. This release process favoured a development environment of rapidly changing interfaces that were necessary during re-engineering. Starting from EGEE-II, it raised the need to move to a more dynamic release process, and the first steps were taken towards an automated environment. At this stage  no software metrics were collected, while functional and integration testing were manually performed, and only on the last project, EGEE-III, an automatic build system was adopted, the E-Infrastructure for Testing, Integration and Configuration of Software (ETICS) service. The gLite middleware developed within EGEE will evolve further as part of a new project: the European Middleware Initiative (EMI).

The ETICS \cite{cordis:etics} and ETICS 2 \cite{cordis:etics2} (Jan 2006 - Feb 2010) projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualizations to improve quality, reliability and interoperability of large-scale scientific software. Both projects supported mechanisms to measure static software metrics (such as Lines of Code (LOC), Source LOC (SLOC), and Complexity); measure number of faults, number of defects, number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}. ETICS portal was the first automated service for delivering software products in distributed environments like the Grid.

The EMI (May 2010 - Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe - ARC, dCache, gLite and UNICORE. The EMI Quality Model used ISO/IEC 9126 Software Engineering - Product Quality standard \cite{iso9126} to identify a set of characteristics that need to be present in EMI software products and processes to be able to meet EMI quality requirements. For each software characteristic, a set of associated metrics and Key Performance Indicators (KPIs) were identified and defined in detail in the EMI Metrics Specification \cite{emi-metrics}. EMI was one of the first projects in using a Continuous Integration (CI) system for building, packaging and testing the various components. It leveraged the ETICS tool and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.

The EGI-InSPIRE (Integrated Sustainable Pan-European Infrastructure for Researchers in Europe, May 2010 - May 2014) project \cite{cordis:egi-inspire} marks the continuation of EGEE-III project to establish and maintain a sustainable European Grid Infrastructure. EGI-InSPIRE did not develop any software but maintained a production-ready Grid computing middleware distribution called UMD (Unified Middleware Distribution). The role of UMD was to enforce the fulfillment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered (i.e., EMI and Globus) \cite{mario}. UMD is still being used and deployed in the European scientific e-infrastructures as a result of its continuation under a different project known as EGI-Engage (Engaging the EGI Community towards an Open Science Commons) \cite{cordis:egi-engage}. This distribution is currently complemented by a Cloud-specific one called CMD (Cloud Middleware Distribution).

INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global ExplOitation) \cite{cordis:indigo} is a running software development project started on April 1st 2015 with the aim of providing solutions to address the existing gaps in cloud PaaS and SaaS levels, helping developers, e-infrastructures and scientific communities to exploit cloud computing benefits. The project will end on September 2017.

\section{New Trends in Software Reliability}

The INDIGO-Datacloud project is a reflection of the progress made in software quality and reliability aspects throughout the past scientific European experiences. Nevertheless, the project’s quality assurance procedures are also highly influenced by the insights of nowadays big worldwide collaborations of software development. These collaborations have continuously inspired new technologies and procedures to increase the robustness and quality of the software being delivered. In this scenario, the DevOps culture is one of the major outcomes that has been progressively adopted in the INDIGO-DataCloud project.

\subsection{DevOps practices}

The DevOps methods emphasize on exercising quality assurance techniques to avoid infrastructure disruption whenever new developments are deployed into production. The INDIGO-DataCloud project promoted the application of a CI scenario to enforce the quality assurance requirements for any piece of software being produced within the project. Such environment requires an automated ready-to-go infrastructure where the different services involved (source code management platform, automation server) interact with each other to trigger the quality check execution and return back the exit codes, together with their outputs. To accomplish this, the project leveraged tight integrated open source tools such as GitHub and Jenkins, as well as relying on Docker container provisioning to speed up the source code validation. This CI approach is guiding the project’s software development phase throughout the first and second major releases.

As DevOps suggests, frequent releases positively affects the reliability of the software. The software updates of INDIGO-DataCloud products taking place since the second major release are passing through a continuous delivery (CD) pipeline that adds the packaging of the software right after the execution of the quality checks (as part of the CI). The steps defined within the continuous delivery pipeline differ whether the software is to be distributed in the form of Docker images or via operating system’s packages. In the latter case an extra validation step is added, consisting in the product’s deployment using a configuration management (CM) solution. The installation uses the packages being built in the previous step, which are being uploaded to the testing repository. In the case of Docker images, the CM is used to build the image itself, installing and configuring the product before uploading it to the DockerHub repository \cite{indigo-dockerhub}. Figure \ref{fig:fig_CD} showcases the workflow followed by the products distributed as Docker images.

\begin{figure}[!t]
\centering
\includegraphics[scale=0.21]{images/devops.png}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{Continuous delivery workflow for Docker images.}
\label{fig:fig_CD}
\end{figure}

\subsection{Software Quality Procedures}

The software quality procedures \cite{indigo-d31} cover 1) the identification and description of the \emph{quality requirements} that the software need to comply with, and 2) the \emph{quality metrics}.

The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly. The source code is publicly available in GitHub repositories \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption. In this regard, the source code is compliant with community de-facto style standards, selected from the variety of programming languages being used.

\begin{figure}[!t]
\centering
\includegraphics[scale=0.35]{images/codestyle.png}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{Code style standards followed by INDIGO-DataCloud's software products.}
\label{fig:fig_codestyle}
\end{figure}

Tests are triggered in each build to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case. The documentation of the products is treated as source code, using a markup language, automatically rendered and uploaded in online repositories \cite{indigo-gitbook}. Changes in both documentation and source code are human-reviewed as the last step before merging them into the production repository. Last but not least, as already seen in the DevOps section, in order to facilitate the usage of the INDIGO-DataCloud services, they can be deployed automatically using either Ansible \cite{indigo-ansible} or Puppet \cite{indigo-puppet} open-source tools.

\begin{figure}[!t]
\centering
\includegraphics[scale=0.23]{images/unittest.png}
% where an .eps filename suffix will be assumed under latex,
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{Unit testing coverage for the two major INDIGO-DataCloud releases.}
\label{fig:fig_unittest}
\end{figure}

The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering - Product quality standard. These metrics cover the development, release and maintenance phases of the software lifecycle. Development and release metrics are obtained automatically from several sources, such as GitHub and Jenkins CI service, and graphically displayed as GitHub pages using GrimoireLab framework \cite{grimoirelab}. Maintenance and user support metrics are collected from the different channels of communication like GGUS \cite{ggus} and Github Issues.

\subsection{Integration, preview and early adoption}

Two pilot infrastructures are at the disposal of developers and scientific communities involved in the project. The aim of these testbeds is to test the level of integration between the components involved in the INDIGO-DataCloud solution and the use cases validation, by deploying and executing the applications with the last stable version of the software.

The released software is also tested in production environments through the stage rollout process. Selected resource providers are requested to install the most updated stable versions of the software, accessed by their users. The stage rollout process is key to detect and mitigate issues that could only appear in production.

\section{Conclusion}

It is a fact that the European software produced for scientific purposes is evolving to a more sustainable model where the quality and reliability of software is being prioritized. Latest software engineering insights, such as DevOps, are gradually being applied as the open-source collaborative tools evolve, allowing tighter integrations. The focus of the software quality assurance procedures has been put on the development stage, trying to detect and correct issues early in the software lifecycle. Automation and metric analysis are at the base of prompt issue solving . However, post-release validation, such as integration testbeds and the stage rollout process, are also needed to strengthen the software reliability for scientific usage.

%\appendices
%\section{}
%Appendices, if needed, appear before the acknowledgment.

% use section* for acknowledgement
\section*{Acknowledgment}
The authors would like to thanks European Commission with the various funded projects.

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{devops}
M. Walls, \emph{Building a DevOps Culture}, in O'Reilly Media, 1 edition (April 15, 2013)

\bibitem{cordis:datagrid}
\emph{DataGrid project}, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/53665\_en.html

\bibitem{datagrid}
[3] L. Momtahan, A. Martin, \emph{e-Science Experiences: Software Engineering Practice and the EU DataGrid}, in Proc. Asia-Pacific Software Engineering Conference, Gold Coast, Queensland, Australia, 4-6 Dec. 2002, pp. 269-275, IEEE Press

\bibitem{globus}
I. Foster, C. Kesselman, \emph{Globus: a Metacomputing Infrastructure Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2, pp. 115–128, 1997

\bibitem{cordis:egee}
\emph{Enabling Grids for E-sciencE (EGEE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/80149\_en.html

\bibitem{cordis:egee2}
\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/99189\_en.html

\bibitem{cordis:egee3}
\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/87264\_en.html

\bibitem{glite}
E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A., Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1), pp. 33–45, 2006, Scientific Publishers OWN

\bibitem{condor}
D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the grid}, in Grid Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70, 2002, Eds. John Wiley \& Sons Inc.

\bibitem{cordis:etics}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/80138\_en.html

\bibitem{cordis:etics2}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/86604\_en.html

\bibitem{etics}
A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol. 119, N. 4, 2008, IOP Publishing Ltd

\bibitem{cordis:emi}
\emph{European Middleware Initiative (EMI)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/95311\_en.html

\bibitem{iso9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International Organization for Standarization. Available: https://www.iso.org/standard/22749.html

\bibitem{emi-metrics}
\emph{EMI Metrics Specification}. Available: https://goo.gl/LdS6fL

\bibitem{cordis:egi-inspire}
\emph{European Grid Initiative: Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/95923\_en.html

\bibitem{egi-qc}
\emph{EGI Quality Criteriai}. Available: http://egi-qc.github.io/

\bibitem{mario}
M. David et al, \emph{Validation of Grid Middleware for the European Grid Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543–558, 2014, Springer

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/194937\_en.html

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/194882\_en.html

\bibitem{indigo-dockerhub}
\emph{INDIGO-DataCloud DockerHub repository}. Available: https://hub.docker.com/u/indigodatacloud

\bibitem{indigo-d31}
\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1. Available: https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31

\bibitem{indigo-jenkins}
\emph{INDIGO-DataCloud Jenkins CI service}. Available: https://jenkins.indigo-datacloud.eu:8080/

\bibitem{indigo-github}
\emph{INDIGO-DataCloud GitHub Source Code repository}. Available: https://www.github.com/indigo-dc

\bibitem{indigo-gitbook}
\emph{INDIGO-DataCloud GitBook Documentation repository}. Available: https://www.gitbook.com/@indigo-dc

\bibitem{indigo-puppet}
\emph{INDIGO-DataCloud PuppetForge repository}. Available: https://forge.puppet.com/indigodc

\bibitem{indigo-ansible}
\emph{INDIGO-DataCloud Ansible Galaxy repository}. Available: https://galaxy.ansible.com/indigo-dc/

\bibitem{grimoirelab}
\emph{GrimoireLab}. Available: http://grimoirelab.github.io/

\bibitem{ggus}
\emph{Global Grid User Support (GGUS)}. Available: https://www.ggus.eu/

\end{thebibliography}

% that's all folks
\end{document}

### 2017-11-07 paper.tex
revised version, plus ieee class

@@
 % and "HIS" in caps to complete the first word.
+
 \IEEEPARstart{I}{n} the European scientific research arena, the analysis of some of the last decade software development projects depicts a continuous increase of the prominence and robustness of testing and validation procedures.
@@
 rise of DevOps practices \cite{devops}) over the last years.
-This is tied to the parallel advancements in the ICT field (with virtualization and container technologies) and the appearance of automation and event-response tools.
-In the following we introduce a set of European Commission funded projects, going from 2001 up to now, where the authors of this summary were and are currently involved.
+This is tied to the parallel advancements in the ICT field (with virtualization and container technologies) and the advent of automation and event-response tools.
+In the following we introduce a set of European Commission funded projects, starting in 2001 up to now, where the authors were and are currently involved.
 For each project we explain how they addressed the challenge in software reliability.
 
-The DataGrid \cite{cordis:datagrid} project (Jan 2001 - Dec 2003) was devoted to develop software to provide basic Grid computing functionalities and related management tools for widely distributed scientific communities, such as Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics.
-In this project, a big effort was dedicated to define and implement academic software engineering practices due to the unusual context in which DataGrid sat, starting from the enormous geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}.
-DataGrid developed its own software distribution, EDG (European DataGrid), strongly based on Globus Middleware Services \cite{globus}.
+The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted to develop software to provide Grid computing functionalities and related management tools for widely distributed scientific communities, such as the Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics.
+In this project, a large effort was dedicated in the definition and implementation of academic software engineering practices due to the unusual context in which DataGrid aims, starting from the large geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}.
+DataGrid developed its own software distribution, {\sl EDG} (European DataGrid), strongly based on Globus Middleware Services \cite{globus}.
 An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.
 
-The Enabling Grids for E-sciencE (EGEE, Apr 2004 - Apr 2010) I \cite{cordis:egee}, II \cite{cordis:egee2}, III \cite{cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e-Science, available to scientists 24 hours-a-day.
-One of the main achievements of these projects was the middleware - a crucial component of any Grid infrastructure as it provides the ‘glue’ to link the hardware resources within the Grid.
-The gLite middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re-engineering efforts to converge with LHC Computing Grid (LCG-2), Virtual Data Toolkit (VDT) and Condor \cite{condor} software distributions.
-The overall goal of the middleware engineering group was to provide and maintain selected middleware services of the gLite distribution to satisfy the needs of the scientific communities.
-During EGEE-I, a software process was designed and put into action for the early releases of the gLite middleware.
-This release process favoured a development environment of rapidly changing interfaces that were necessary during re-engineering.
-Starting from EGEE-II, it raised the need to move to a more dynamic release process, and the first steps were taken towards an automated environment.
-At this stage  no software metrics were collected, while functional and integration testing were manually performed, and only on the last project, EGEE-III, an automatic build system was adopted, the E-Infrastructure for Testing, Integration and Configuration of Software (ETICS) service.
-The gLite middleware developed within EGEE will evolve further as part of a new project: the European Middleware Initiative (EMI).
+The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science, available 24 hours--a--day.
+One of the main achievements of these projects was the middleware -- a crucial component of any Grid infrastructure as it provides the {\sl glue} that links the hardware resources within such infrastructure.
+The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions.
+The overall goal of the middleware engineering group was to provide and maintain selected middleware services of the {\sl gLite} distribution to satisfy the needs of the scientific communities.
+During EGEE--I, a software process was designed and put into action for the early releases of the {\sl gLite} middleware.
+This release process favoured a development environment of rapidly changing interfaces that were necessary during re--engineering.
+Starting in EGEE--II, there was a need to move to a more dynamic release process, and the first steps were taken towards an automated environment.
+At this stage  no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.
+The {\sl gLite} middleware developed within EGEE will evolve further as part of a new project: the European Middleware Initiative (EMI).
 
-The ETICS \cite{cordis:etics} and ETICS 2 \cite{cordis:etics2} (Jan 2006 - Feb 2010) projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualizations to improve quality, reliability and interoperability of large-scale scientific software.
-Both projects supported mechanisms to measure static software metrics (such as Lines of Code (LOC), Source LOC (SLOC), and Complexity); measure number of faults, number of defects, number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
-ETICS portal was the first automated service for delivering software products in distributed environments like the Grid.
+The ETICS \cite{cordis:etics} and ETICS 2 \cite{cordis:etics2} (Jan 2006 -- Feb 2010) projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualization to improve quality, reliability and interoperability of large--scale scientific software.
+Both projects supported mechanisms to measure static software metrics (such as Lines of Code (LOC), Source LOC (SLOC), and Complexity); to measure the number of faults, the number of defects, number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
+The {\sl ETICS} portal was the first automated service for delivering software products in distributed environments such as the Grid.
 
-The EMI (May 2010 - Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe - ARC, dCache, gLite and UNICORE.
-The EMI Quality Model used ISO/IEC 9126 Software Engineering - Product Quality standard \cite{iso9126} to identify a set of characteristics that need to be present in EMI software products and processes to be able to meet EMI quality requirements.
+The EMI (May 2010 -- Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe -- {\sl ARC}, {\sl dCache}, {\sl gLite} and {\sl UNICORE}.
+The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics that need to be present in EMI software products and processes to be able to meet EMI quality requirements.
 For each software characteristic, a set of associated metrics and Key Performance Indicators (KPIs) were identified and defined in detail in the EMI Metrics Specification \cite{emi-metrics}.
 EMI was one of the first projects in using a Continuous Integration (CI) system for building, packaging and testing the various components.
-It leveraged the ETICS tool and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.
+It leveraged the {\sl ETICS} tool and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.
 
-The EGI-InSPIRE (Integrated Sustainable Pan-European Infrastructure for Researchers in Europe, May 2010 - May 2014) project \cite{cordis:egi-inspire} marks the continuation of EGEE-III project to establish and maintain a sustainable European Grid Infrastructure.
-EGI-InSPIRE did not develop any software but maintained a production-ready Grid computing middleware distribution called UMD (Unified Middleware Distribution).
-The role of UMD was to enforce the fulfillment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered (i.e., EMI and Globus) \cite{mario}.
-UMD is still being used and deployed in the European scientific e-infrastructures as a result of its continuation under a different project known as EGI-Engage (Engaging the EGI Community towards an Open Science Commons) \cite{cordis:egi-engage}.
-This distribution is currently complemented by a Cloud-specific one called CMD (Cloud Middleware Distribution).
+The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire} is the continuation of EGEE--III project, establishing and maintaining a sustainable European Grid Infrastructure.
+EGI--InSPIRE did not develop any software but maintained a production--ready Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution).
+The role of {\sl UMD} was to enforce the fulfilment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered (i.e., {\sl EMI} and {\sl Globus}) \cite{mario}.
+{\sl UMD} is still being used and deployed in the European scientific e--Infrastructures as a result of its continuation under a follow--up project, the EGI--Engage (Engaging the EGI Community towards an Open Science Commons) \cite{cordis:egi-engage}.
+This distribution is currently complemented by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).
 
-INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global ExplOitation) \cite{cordis:indigo} is a running software development project started on April 1st 2015 with the aim of providing solutions to address the existing gaps in cloud PaaS and SaaS levels, helping developers, e-infrastructures and scientific communities to exploit cloud computing benefits.
+INDIGO--DataCloud (INtegrating Distributed data Infrastructures for Global ExplOitation) \cite{cordis:indigo} is an ongoing software development project that started on April $1^{st}$ 2015 with the aim at providing solutions to address the existing gaps in the three cloud levels; IaaS, PaaS and SaaS, helping developers, e--Infrastructures and scientific communities to exploit cloud computing resources.
 The project will end on September 2017.
@@
 
-The INDIGO-Datacloud project is a reflection of the progress made in software quality and reliability aspects throughout the past scientific European experiences.
-Nevertheless, the project’s quality assurance procedures are also highly influenced by the insights of nowadays big worldwide collaborations of software development.
+The INDIGO--Datacloud project is a reflection of the progress made in software quality and reliability aspects throughout the past scientific European experiences.
+Nevertheless, the project’s Quality Assurance procedures are also highly influenced by the insights of current big worldwide collaborations of software development.
 These collaborations have continuously inspired new technologies and procedures to increase the robustness and quality of the software being delivered.
-In this scenario, the DevOps culture is one of the major outcomes that has been progressively adopted in the INDIGO-DataCloud project.
+In this scenario, the DevOps culture is one of the major outcomes that has been progressively adopted in the INDIGO--DataCloud project.
 
@@
 
-The DevOps methods emphasize on exercising quality assurance techniques to avoid infrastructure disruption whenever new developments are deployed into production.
-The INDIGO-DataCloud project promoted the application of a CI scenario to enforce the quality assurance requirements for any piece of software being produced within the project.
-Such environment requires an automated ready-to-go infrastructure where the different services involved (source code management platform, automation server) interact with each other to trigger the quality check execution and return back the exit codes, together with their outputs.
-To accomplish this, the project leveraged tight integrated open source tools such as GitHub and Jenkins, as well as relying on Docker container provisioning to speed up the source code validation.
+The DevOps methods emphasize on exercising Quality Assurance techniques to avoid infrastructure disruption whenever new developments are deployed into production.
+The INDIGO--DataCloud project promoted the application of a CI scenario to enforce the Quality Assurance requirements for any piece of software being produced within the project.
+Such environment requires an automated ready--to--go infrastructure where the different services involved (source code management platform, automation server) interact with each other to trigger the quality check execution and return back the exit codes, together with their outputs.
+To accomplish this, the project leveraged tightly integrated open source tools such as GitHub and Jenkins, as well as relying on Docker container provisioning to speed up the source code validation.
 This CI approach is guiding the project’s software development phase throughout the first and second major releases.
@@
 As DevOps suggests, frequent releases positively affects the reliability of the software.
-The software updates of INDIGO-DataCloud products taking place since the second major release are passing through a continuous delivery (CD) pipeline that adds the packaging of the software right after the execution of the quality checks (as part of the CI).
-The steps defined within the continuous delivery pipeline differ whether the software is to be distributed in the form of Docker images or via operating system’s packages.
-In the latter case an extra validation step is added, consisting in the product’s deployment using a configuration management (CM) solution.
-The installation uses the packages being built in the previous step, which are being uploaded to the testing repository.
+The software updates of INDIGO--DataCloud products taking place since the second major release are passing through a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the execution of the quality checks (as part of the CI).
+The steps defined within the CD pipeline differ whether the software is to be distributed in the form of Docker images or via operating system’s packages.
+In the latter case an extra validation step is added, consisting in the product’s deployment using a Configuration Management (CM) solution.
+The installation uses the packages being built in the previous step, which are being uploaded to a testing repository.
 In the case of Docker images, the CM is used to build the image itself, installing and configuring the product before uploading it to the DockerHub repository \cite{indigo-dockerhub}.
@@
 
-\begin{figure}[!t]
+\begin{figure*}
 \centering
-\includegraphics[scale=0.21]{images/devops.png}
-% where an .eps filename suffix will be assumed under latex,
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-\caption{Continuous delivery workflow for Docker images.}
+\includegraphics[width=\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker images.}
 \label{fig:fig_CD}
-\end{figure}
+\end{figure*}
 
@@
 
-The software quality procedures \cite{indigo-d31} cover 1) the identification and description of the \emph{quality requirements} that the software need to comply with, and 2) the \emph{quality metrics}.
+The software quality procedures \cite{indigo-d31} cover; 1) the identification and description of the \emph{quality requirements} that the software need to comply with, and 2) the \emph{quality metrics}.
 
 The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly.
-The source code is publicly available in GitHub repositories \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
-In this regard, the source code is compliant with community de-facto style standards, selected from the variety of programming languages being used.
+The source code is publicly available in GitHub repositories under an organization called {\sl indigo-dc} \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
+In this regard, the source code is compliant with community de--facto style standards, selected from the variety of programming languages being used.
 
@@
 \centering
-\includegraphics[scale=0.35]{images/codestyle.png}
-% where an .eps filename suffix will be assumed under latex,
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-\caption{Code style standards followed by INDIGO-DataCloud's software products.}
+\includegraphics[width=0.5\textwidth]{images/codestyle.png}
+\caption{Code style standards followed by INDIGO--DataCloud's software products.}
 \label{fig:fig_codestyle}
@@
 The documentation of the products is treated as source code, using a markup language, automatically rendered and uploaded in online repositories \cite{indigo-gitbook}.
-Changes in both documentation and source code are human-reviewed as the last step before merging them into the production repository.
-Last but not least, as already seen in the DevOps section, in order to facilitate the usage of the INDIGO-DataCloud services, they can be deployed automatically using either Ansible \cite{indigo-ansible} or Puppet \cite{indigo-puppet} open-source tools.
+Changes in both documentation and source code are human--reviewed as the last step before merging them into the production repository.
+Last but not least, as already seen in the DevOps section, in order to facilitate the usage of the INDIGO--DataCloud services, they can be deployed automatically using either Ansible \cite{indigo-ansible} or Puppet \cite{indigo-puppet} open--source tools.
 
-\begin{figure}[!t]
+\begin{figure*}
 \centering
-\includegraphics[scale=0.23]{images/unittest.png}
-% where an .eps filename suffix will be assumed under latex,
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-\caption{Unit testing coverage for the two major INDIGO-DataCloud releases.}
+\includegraphics[width=\textwidth]{images/unittest.png}
+\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
 \label{fig:fig_unittest}
-\end{figure}
+\end{figure*}
 
-The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering - Product quality standard.
+The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard.
 These metrics cover the development, release and maintenance phases of the software lifecycle.
 Development and release metrics are obtained automatically from several sources, such as GitHub and Jenkins CI service, and graphically displayed as GitHub pages using GrimoireLab framework \cite{grimoirelab}.
-Maintenance and user support metrics are collected from the different channels of communication like GGUS \cite{ggus} and Github Issues.
+Maintenance and user support metrics are collected from the different sources of data like GGUS \cite{ggus} and Github Issues.
 
@@
 Two pilot infrastructures are at the disposal of developers and scientific communities involved in the project.
-The aim of these testbeds is to test the level of integration between the components involved in the INDIGO-DataCloud solution and the use cases validation, by deploying and executing the applications with the last stable version of the software.
+The aim of these testbeds is to test the level of integration between the components involved in the INDIGO--DataCloud solution and use cases validation, by deploying and executing the applications with the last stable version of the software.
 
-The released software is also tested in production environments through the stage rollout process.
+The released software is also tested in production environments through the stage--rollout process.
 Selected resource providers are requested to install the most updated stable versions of the software, accessed by their users.
-The stage rollout process is key to detect and mitigate issues that could only appear in production.
+The stage--rollout process is key to detect and mitigate issues that could only appear in production.
 
@@
 It is a fact that the European software produced for scientific purposes is evolving to a more sustainable model where the quality and reliability of software is being prioritized.
-Latest software engineering insights, such as DevOps, are gradually being applied as the open-source collaborative tools evolve, allowing tighter integrations.
-The focus of the software quality assurance procedures has been put on the development stage, trying to detect and correct issues early in the software lifecycle.
-Automation and metric analysis are at the base of prompt issue solving .
-However, post-release validation, such as integration testbeds and the stage rollout process, are also needed to strengthen the software reliability for scientific usage.
+Recent software engineering insights, such as DevOps, are gradually being applied as the open--source collaborative tools evolve, allowing tighter integrations.
+The focus of the Software Quality Assurance procedures has been integrated in the development stage, trying to detect and correct issues early in the software lifecycle.
+Automation and metric analysis are at the base of prompt issue solving.
+However, post--release validation, such as integration testbeds and the stage--rollout process, are also needed to strengthen the software reliability for scientific usage.
 

### 2017-11-07 paper.tex

@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker images.}
+\label{fig:fig_CD}
+\end{figure*}
+
 The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted to develop software to provide Grid computing functionalities and related management tools for widely distributed scientific communities, such as the Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics.
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/unittest.png}
+\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
+\label{fig:fig_unittest}
+\end{figure*}
+
 INDIGO--DataCloud (INtegrating Distributed data Infrastructures for Global ExplOitation) \cite{cordis:indigo} is an ongoing software development project that started on April $1^{st}$ 2015 with the aim at providing solutions to address the existing gaps in the three cloud levels; IaaS, PaaS and SaaS, helping developers, e--Infrastructures and scientific communities to exploit cloud computing resources.
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/devops.png}
-\caption{Continuous Delivery workflow for Docker images.}
-\label{fig:fig_CD}
-\end{figure*}
-
 \subsection{Software Quality Procedures}
 
-The software quality procedures \cite{indigo-d31} cover; 1) the identification and description of the \emph{quality requirements} that the software need to comply with, and 2) the \emph{quality metrics}.
+The software quality procedures \cite{indigo-d31} cover: 1) the identification and description of the \emph{quality requirements} that the software need to comply with, and 2) the \emph{quality metrics}.
 
@@
 Last but not least, as already seen in the DevOps section, in order to facilitate the usage of the INDIGO--DataCloud services, they can be deployed automatically using either Ansible \cite{indigo-ansible} or Puppet \cite{indigo-puppet} open--source tools.
-
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/unittest.png}
-\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
-\label{fig:fig_unittest}
-\end{figure*}
 

### 2017-11-08 paper.tex

@@
         J. Gomes, %~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
-        D. Salomoni, %~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
+        D. Salomoni %~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
 \thanks{D. C. Duma, E. Ronchieri, and D. Salomoni are with INFN - CNAF, Bologna, Italy.}%
@@
 \begin{abstract}
-In this work we describe the trends in scientific software reliability by analyzing past European Commission funded software development projects in the context of distributed computing field such as Enabling Grids for E-sciencE and European Middleware Initiative, and the ongoing INDIGO-DataCloud.
+In this work we describe the trends in scientific software reliability by analyzing past European Commission funded software development projects in the context of distributed computing field, such as Enabling Grids for E-sciencE, European Middleware Initiative and the ongoing INDIGO-DataCloud.
 \end{abstract}

### 2017-11-08 paper.tex
shortened to 3 pages

@@
 DataGrid developed its own software distribution, {\sl EDG} (European DataGrid), strongly based on Globus Middleware Services \cite{globus}.
-An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.
+%An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.
 
-The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science, available 24 hours--a--day.
-One of the main achievements of these projects was the middleware -- a crucial component of any Grid infrastructure as it provides the {\sl glue} that links the hardware resources within such infrastructure.
-The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions.
-The overall goal of the middleware engineering group was to provide and maintain selected middleware services of the {\sl gLite} distribution to satisfy the needs of the scientific communities.
-During EGEE--I, a software process was designed and put into action for the early releases of the {\sl gLite} middleware.
-This release process favoured a development environment of rapidly changing interfaces that were necessary during re--engineering.
+The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
+The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering effortsi.
+%to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions.
 Starting in EGEE--II, there was a need to move to a more dynamic release process, and the first steps were taken towards an automated environment.
@@
 The EMI (May 2010 -- Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe -- {\sl ARC}, {\sl dCache}, {\sl gLite} and {\sl UNICORE}.
-The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics that need to be present in EMI software products and processes to be able to meet EMI quality requirements.
-For each software characteristic, a set of associated metrics and Key Performance Indicators (KPIs) were identified and defined in detail in the EMI Metrics Specification \cite{emi-metrics}.
+The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics, each defining the associated metrics and Key Performance Indicators (KPIs).
 EMI was one of the first projects in using a Continuous Integration (CI) system for building, packaging and testing the various components.
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/unittest.png}
-\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
-\label{fig:fig_unittest}
-\end{figure*}
+%\begin{figure*}
+%\centering
+%\includegraphics[width=\textwidth]{images/unittest.png}
+%\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
+%\label{fig:fig_unittest}
+%\end{figure*}
 
@@
 The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly.
-The source code is publicly available in GitHub repositories under an organization called {\sl indigo-dc} \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
+The source code is publicly available as GitHub repositories \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
 In this regard, the source code is compliant with community de--facto style standards, selected from the variety of programming languages being used.
+Tests are triggered for each change to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case.
 
@@
 
-Tests are triggered in each build to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case.
-The documentation of the products is treated as source code, using a markup language, automatically rendered and uploaded in online repositories \cite{indigo-gitbook}.
-Changes in both documentation and source code are human--reviewed as the last step before merging them into the production repository.
-Last but not least, as already seen in the DevOps section, in order to facilitate the usage of the INDIGO--DataCloud services, they can be deployed automatically using either Ansible \cite{indigo-ansible} or Puppet \cite{indigo-puppet} open--source tools.
-
 The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard.
 These metrics cover the development, release and maintenance phases of the software lifecycle.
-Development and release metrics are obtained automatically from several sources, such as GitHub and Jenkins CI service, and graphically displayed as GitHub pages using GrimoireLab framework \cite{grimoirelab}.
-Maintenance and user support metrics are collected from the different sources of data like GGUS \cite{ggus} and Github Issues.
 
@@
 \bibitem{cordis:datagrid}
-\emph{DataGrid project}, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/53665\_en.html
+\emph{DataGrid project}, CORDIS Europe, http://cordis.europa.eu/project/rcn/53665\_en.html
 
 \bibitem{datagrid}
-[3] L.
-Momtahan, A.
-Martin, \emph{e-Science Experiences: Software Engineering Practice and the EU DataGrid}, in Proc.
-Asia-Pacific Software Engineering Conference, Gold Coast, Queensland, Australia, 4-6 Dec.
-2002, pp.
-269-275, IEEE Press
+[3] L. Momtahan et al, \emph{e-Science Experiences: Software Engineering Practice and the EU DataGrid}, in Proc. Asia-Pacific Software Engineering Conference, 2002, pp. 269-275, IEEE Press
 
@@
 \bibitem{cordis:egee}
-\emph{Enabling Grids for E-sciencE (EGEE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/80149\_en.html
+\emph{Enabling Grids for E-sciencE (EGEE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html
 
 \bibitem{cordis:egee2}
-\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/99189\_en.html
+\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/99189\_en.html
 
 \bibitem{cordis:egee3}
-\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/87264\_en.html
+\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/87264\_en.html
 
 \bibitem{glite}
-E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A., Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1), pp. 33–45, 2006, Scientific Publishers OWN
-
-\bibitem{condor}
-D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the grid}, in Grid Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70, 2002, Eds. John Wiley \& Sons Inc.
+E. Laure et al, \emph{Programming the Grid with gLite}, in Computational Methods in Science and Technology, vol. 12(1), pp. 33–45, 2006, Scientific Publishers OWN
 
 \bibitem{cordis:etics}
-\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, European Community Research and Development Information Service (CORDIS).
-Available: http://cordis.europa.eu/project/rcn/80138\_en.html
+\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80138\_en.html
 
 \bibitem{cordis:etics2}
-\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, European Community Research and Development Information Service (CORDIS).
-Available: http://cordis.europa.eu/project/rcn/86604\_en.html
+\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/86604\_en.html
 
 \bibitem{etics}
-A.
-Di Meglio, M.-E.Bégin, P.
-Couvares, E.
-Ronchieri, E.
-Takacs, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol.
-119, N.
-4, 2008, IOP Publishing Ltd
+A. Di Meglio et al, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol. 119, N. 4, 2008, IOP Publishing Ltd
 
 \bibitem{cordis:emi}
-\emph{European Middleware Initiative (EMI)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/95311\_en.html
+\emph{European Middleware Initiative (EMI)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95311\_en.html
 
 \bibitem{iso9126}
-\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International Organization for Standarization. Available: https://www.iso.org/standard/22749.html
-
-\bibitem{emi-metrics}
-\emph{EMI Metrics Specification}. Available: https://goo.gl/LdS6fL
+\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International Organization for Standarization, https://www.iso.org/standard/22749.html
 
 \bibitem{cordis:egi-inspire}
-\emph{European Grid Initiative: Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, European Community Research and Development Information Service (CORDIS).
-Available: http://cordis.europa.eu/project/rcn/95923\_en.html
+\emph{Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95923\_en.html
 
 \bibitem{egi-qc}
-\emph{EGI Quality Criteriai}. Available: http://egi-qc.github.io/
+\emph{EGI Quality Criteria}, http://egi-qc.github.io/
 
@@
 \bibitem{cordis:egi-engage}
-\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, European Community Research and Development Information Service (CORDIS).
-Available: http://cordis.europa.eu/project/rcn/194937\_en.html
+\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194937\_en.html
 
 \bibitem{cordis:indigo}
-\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, European Community Research and Development Information Service (CORDIS).
-Available: http://cordis.europa.eu/project/rcn/194882\_en.html
+\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194882\_en.html
 
 \bibitem{indigo-dockerhub}
-\emph{INDIGO-DataCloud DockerHub repository}. Available: https://hub.docker.com/u/indigodatacloud
+\emph{INDIGO-DataCloud DockerHub}, https://hub.docker.com/u/indigodatacloud
 
 \bibitem{indigo-d31}
-\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1. Available: https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31
+\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1, https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31
 
 \bibitem{indigo-jenkins}
-\emph{INDIGO-DataCloud Jenkins CI service}. Available: https://jenkins.indigo-datacloud.eu:8080/
+\emph{INDIGO-DataCloud Jenkins}, https://jenkins.indigo-datacloud.eu:8080/
 
 \bibitem{indigo-github}
-\emph{INDIGO-DataCloud GitHub Source Code repository}. Available: https://www.github.com/indigo-dc
-
-\bibitem{indigo-gitbook}
-\emph{INDIGO-DataCloud GitBook Documentation repository}. Available: https://www.gitbook.com/@indigo-dc
-
-\bibitem{indigo-puppet}
-\emph{INDIGO-DataCloud PuppetForge repository}. Available: https://forge.puppet.com/indigodc
-
-\bibitem{indigo-ansible}
-\emph{INDIGO-DataCloud Ansible Galaxy repository}. Available: https://galaxy.ansible.com/indigo-dc/
-
-\bibitem{grimoirelab}
-\emph{GrimoireLab}. Available: http://grimoirelab.github.io/
-
-\bibitem{ggus}
-\emph{Global Grid User Support (GGUS)}. Available: https://www.ggus.eu/
+\emph{INDIGO-DataCloud GitHub}, https://www.github.com/indigo-dc
 

### 2017-11-08 paper.tex
remove sloc

@@
 The ETICS \cite{cordis:etics} and ETICS 2 \cite{cordis:etics2} (Jan 2006 -- Feb 2010) projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualization to improve quality, reliability and interoperability of large--scale scientific software.
-Both projects supported mechanisms to measure static software metrics (such as Lines of Code (LOC), Source LOC (SLOC), and Complexity); to measure the number of faults, the number of defects, number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
+Both projects supported mechanisms to measure static software metrics, number of faults, number of defects and number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
 The {\sl ETICS} portal was the first automated service for delivering software products in distributed environments such as the Grid.

### 2017-11-08 paper.tex
fix typo

@@
 The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
-The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering effortsi.
+The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts.
 %to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions.

### 2017-11-08 paper.tex
cutting

@@
 In this project, a large effort was dedicated in the definition and implementation of academic software engineering practices due to the unusual context in which DataGrid aims, starting from the large geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}.
-DataGrid developed its own software distribution, {\sl EDG} (European DataGrid), strongly based on Globus Middleware Services \cite{globus}.
-%An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.
 
-The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
+%The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
 The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts.
-%to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions.
-Starting in EGEE--II, there was a need to move to a more dynamic release process, and the first steps were taken towards an automated environment.
-At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.
-The {\sl gLite} middleware developed within EGEE will evolve further as part of a new project: the European Middleware Initiative (EMI).
+Starting in EGEE--II, there was a need to move to a more dynamic release process.
+At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.i
+The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
+The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts.
+Starting in EGEE--II, there was a need to move to a more dynamic release process.
+At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.i
 
@@
 The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics, each defining the associated metrics and Key Performance Indicators (KPIs).
-EMI was one of the first projects in using a Continuous Integration (CI) system for building, packaging and testing the various components.
-It leveraged the {\sl ETICS} tool and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.
+EMI was one of the first projects in using a Continuous Integration (CI) system, leveraging on {\sl ETICS} tool  for building, packaging and testing the various components.
+%It leveraged the  and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.
 
-The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire} is the continuation of EGEE--III project, establishing and maintaining a sustainable European Grid Infrastructure.
-EGI--InSPIRE did not develop any software but maintained a production--ready Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution).
-The role of {\sl UMD} was to enforce the fulfilment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered (i.e., {\sl EMI} and {\sl Globus}) \cite{mario}.
+The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire} established a sustainable European Grid Infrastructure.
+EGI--InSPIRE maintained a production--ready Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution).
+The role of {\sl UMD} was to enforce the fulfilment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered \cite{mario}.
 {\sl UMD} is still being used and deployed in the European scientific e--Infrastructures as a result of its continuation under a follow--up project, the EGI--Engage (Engaging the EGI Community towards an Open Science Commons) \cite{cordis:egi-engage}.
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=0.5\textwidth]{images/codestyle.png}
-\caption{Code style standards followed by INDIGO--DataCloud's software products.}
-\label{fig:fig_codestyle}
-\end{figure}
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=0.5\textwidth]{images/codestyle.png}
+%\caption{Code style standards followed by INDIGO--DataCloud's software products.}
+%\label{fig:fig_codestyle}
+%\end{figure}
 
-The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard.
+The evaluation of the software quality is performed by measuring the values of the metrics and KPIs defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard.
 These metrics cover the development, release and maintenance phases of the software lifecycle.
@@
 
-\bibitem{globus}
-I. Foster, C. Kesselman, \emph{Globus: a Metacomputing Infrastructure Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2, pp. 115–128, 1997
-
 \bibitem{cordis:egee}
-\emph{Enabling Grids for E-sciencE (EGEE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html
-
-\bibitem{cordis:egee2}
-\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/99189\_en.html
-
-\bibitem{cordis:egee3}
-\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/87264\_en.html
+%\emph{Enabling Grids for E-sciencE (EGEE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html, EGEE-II -- http://cordis.europa.eu/project/rcn/99189\_en.html, EGEE-III -- http://cordis.europa.eu/project/rcn/87264\_en.html
+\emph{EGEE} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html, EGEE-II -- http://cordis.europa.eu/project/rcn/99189\_en.html, EGEE-III -- http://cordis.europa.eu/project/rcn/87264\_en.html
+%
+%\bibitem{cordis:egee}
+%\emph{Enabling Grids for E-sciencE (EGEE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html
+%
+%\bibitem{cordis:egee2}
+%\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/99189\_en.html
+%
+%\bibitem{cordis:egee3}
+%\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/87264\_en.html
 
@@
 \bibitem{cordis:etics}
-\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80138\_en.html
-
-\bibitem{cordis:etics2}
-\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/86604\_en.html
+\emph{ETICS} project, CORDIS Europe, ETICS -- http://cordis.europa.eu/project/rcn/80138\_en.html, ETICS-2 -- http://cordis.europa.eu/project/rcn/86604\_en.html
+%\bibitem{cordis:etics}
+%\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80138\_en.html
+%
+%\bibitem{cordis:etics2}
+%\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/86604\_en.html
 
@@
 \bibitem{cordis:emi}
-\emph{European Middleware Initiative (EMI)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95311\_en.html
+%\emph{European Middleware Initiative (EMI)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95311\_en.html
+\emph{EMI} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95311\_en.html
 
@@
 \bibitem{cordis:egi-inspire}
-\emph{Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95923\_en.html
+%\emph{Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95923\_en.html
+\emph{EGI-InSPIRE} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95923\_en.html
 
@@
 \bibitem{cordis:egi-engage}
-\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194937\_en.html
+%\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194937\_en.html
+\emph{EGI-ENGAGE} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194937\_en.html
 
 \bibitem{cordis:indigo}
-\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194882\_en.html
+%\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194882\_en.html
+\emph{INDIGO-DataCloud} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194882\_en.html
 

### 2017-11-09 paper.tex
removed indigo tools references and some rephrasing

@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/devops.png}
-\caption{Continuous Delivery workflow for Docker images.}
-\label{fig:fig_CD}
-\end{figure*}
-
-The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted to develop software to provide Grid computing functionalities and related management tools for widely distributed scientific communities, such as the Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics.
+The DataGrid \cite{cordis:datagrid} project (2001 -- 2003) was devoted to develop software to provide Grid computing functionalities and related management tools for widely distributed scientific communities, such as the Large Hadron Collider (LHC) experiments, Earth observations and Bioinformatics.
 In this project, a large effort was dedicated in the definition and implementation of academic software engineering practices due to the unusual context in which DataGrid aims, starting from the large geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}.
@@
 At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.i
-The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
+The three phases of Enabling Grids for E--sciencE (EGEE, 2004 -- 2010) \cite{cordis:egee} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
 The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts.
 Starting in EGEE--II, there was a need to move to a more dynamic release process.
-At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.i
+At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.
 
-The ETICS \cite{cordis:etics} and ETICS 2 \cite{cordis:etics2} (Jan 2006 -- Feb 2010) projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualization to improve quality, reliability and interoperability of large--scale scientific software.
+The ETICS and ETICS 2 (2006 -- 2010) \cite{cordis:etics} projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualization to improve quality, reliability and interoperability of large--scale scientific software.
 Both projects supported mechanisms to measure static software metrics, number of faults, number of defects and number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
@@
 
-The EMI (May 2010 -- Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe -- {\sl ARC}, {\sl dCache}, {\sl gLite} and {\sl UNICORE}.
-The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics, each defining the associated metrics and Key Performance Indicators (KPIs).
+The EMI (2010 -- 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe -- {\sl ARC}, {\sl dCache}, {\sl gLite} and {\sl UNICORE}.
+The EMI Quality Model used ISO/IEC 9126 standard \cite{iso9126} to guide the definition of a set of metrics and Key Performance Indicators (KPIs).
 EMI was one of the first projects in using a Continuous Integration (CI) system, leveraging on {\sl ETICS} tool  for building, packaging and testing the various components.
@@
 
-The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire} established a sustainable European Grid Infrastructure.
-EGI--InSPIRE maintained a production--ready Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution).
-The role of {\sl UMD} was to enforce the fulfilment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered \cite{mario}.
+The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for Researchers in Europe, 2010 -- 2014) project \cite{cordis:egi-inspire} maintained a production--ready Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution).
+The role of {\sl UMD} was to enforce the fulfilment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered.
 {\sl UMD} is still being used and deployed in the European scientific e--Infrastructures as a result of its continuation under a follow--up project, the EGI--Engage (Engaging the EGI Community towards an Open Science Commons) \cite{cordis:egi-engage}.
@@
 
+%\begin{figure*}
+%\centering
+%\includegraphics[width=\textwidth]{images/devops.png}
+%\caption{Continuous Delivery workflow for Docker container images.}
+%\label{fig:fig_CD}
+%\end{figure*}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.47\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker container images.}
+\label{fig:fig_CD}
+\end{figure}
+
 INDIGO--DataCloud (INtegrating Distributed data Infrastructures for Global ExplOitation) \cite{cordis:indigo} is an ongoing software development project that started on April $1^{st}$ 2015 with the aim at providing solutions to address the existing gaps in the three cloud levels; IaaS, PaaS and SaaS, helping developers, e--Infrastructures and scientific communities to exploit cloud computing resources.
-The project will end on September 2017.
+The project ends on September 2017.
 
@@
 The INDIGO--Datacloud project is a reflection of the progress made in software quality and reliability aspects throughout the past scientific European experiences.
-Nevertheless, the project’s Quality Assurance procedures are also highly influenced by the insights of current big worldwide collaborations of software development.
+Nevertheless, the project’s Quality Assurance (QA) procedures are also highly influenced by the insights of current big worldwide collaborations of software development.
 These collaborations have continuously inspired new technologies and procedures to increase the robustness and quality of the software being delivered.
@@
 
-The DevOps methods emphasize on exercising Quality Assurance techniques to avoid infrastructure disruption whenever new developments are deployed into production.
-The INDIGO--DataCloud project promoted the application of a CI scenario to enforce the Quality Assurance requirements for any piece of software being produced within the project.
+%The DevOps methods emphasize on exercising QA techniques to avoid infrastructure disruption whenever new developments are deployed into production.
+The INDIGO--DataCloud project promoted the application of a CI scenario to enforce the QA requirements for any piece of software being produced within the project.
 Such environment requires an automated ready--to--go infrastructure where the different services involved (source code management platform, automation server) interact with each other to trigger the quality check execution and return back the exit codes, together with their outputs.
@@
 This CI approach is guiding the project’s software development phase throughout the first and second major releases.
-
-As DevOps suggests, frequent releases positively affects the reliability of the software.
-The software updates of INDIGO--DataCloud products taking place since the second major release are passing through a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the execution of the quality checks (as part of the CI).
-The steps defined within the CD pipeline differ whether the software is to be distributed in the form of Docker images or via operating system’s packages.
-In the latter case an extra validation step is added, consisting in the product’s deployment using a Configuration Management (CM) solution.
-The installation uses the packages being built in the previous step, which are being uploaded to a testing repository.
-In the case of Docker images, the CM is used to build the image itself, installing and configuring the product before uploading it to the DockerHub repository \cite{indigo-dockerhub}.
-Figure \ref{fig:fig_CD} showcases the workflow followed by the products distributed as Docker images.
+The DevOps methods emphasize on exercising QA techniques to avoid infrastructure disruption whenever new developments are deployed into production.
+The INDIGO--DataCloud project promoted from its early stages the application of a CI scenario to enforce the QA requirements during the development phase.
+As the project matured, it evolved towards a Continuous Delivery (CD) implementation, automating the packaging of the software.
+Figure \ref{fig:fig_CD} showcases the workflow followed by the products distributed as container images.
 
@@
 
-The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly.
+%The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly.
+The source code is publicly available as GitHub repositories \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
+In this regard, the source code is compliant with community de--facto style standards, selected from the variety of programming languages being used.
+Tests are triggered for each change to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case.
+The QA requirements define that the source code shall be publicly available to augment its visibility while promoting external contributions.
+The developed software is compliant with community de--facto style standards, automatically checked in the CI infrastructure together with the analysis of the unit and functional testing coverage.
+As the last step, a human-based code review is performed to detect potential incoherences that may appear during the automated process.
+
+%are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly.
 The source code is publicly available as GitHub repositories \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
@@
 
-The evaluation of the software quality is performed by measuring the values of the metrics and KPIs defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard.
+The evaluation of the software quality is performed by measuring the values of the metrics and KPIs defined based upon the ISO/IEC 9126.
 These metrics cover the development, release and maintenance phases of the software lifecycle.
@@
 Recent software engineering insights, such as DevOps, are gradually being applied as the open--source collaborative tools evolve, allowing tighter integrations.
-The focus of the Software Quality Assurance procedures has been integrated in the development stage, trying to detect and correct issues early in the software lifecycle.
+The focus of the Software QA procedures has been integrated in the development stage, trying to detect and correct issues early in the software lifecycle.
 Automation and metric analysis are at the base of prompt issue solving.
@@
 
-\bibitem{etics}
-A. Di Meglio et al, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol. 119, N. 4, 2008, IOP Publishing Ltd
+%\bibitem{etics}
+%A. Di Meglio et al, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol. 119, N. 4, 2008, IOP Publishing Ltd
 
@@
 
-\bibitem{mario}
-M. David et al, \emph{Validation of Grid Middleware for the European Grid Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543–558, 2014, Springer
+%\bibitem{mario}
+%M. David et al, \emph{Validation of Grid Middleware for the European Grid Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543–558, 2014, Springer
 
@@
 
-\bibitem{indigo-dockerhub}
-\emph{INDIGO-DataCloud DockerHub}, https://hub.docker.com/u/indigodatacloud
+%\bibitem{indigo-dockerhub}
+%\emph{INDIGO-DataCloud DockerHub}, https://hub.docker.com/u/indigodatacloud
 
@@
 
-\bibitem{indigo-jenkins}
-\emph{INDIGO-DataCloud Jenkins}, https://jenkins.indigo-datacloud.eu:8080/
-
-\bibitem{indigo-github}
-\emph{INDIGO-DataCloud GitHub}, https://www.github.com/indigo-dc
+%\bibitem{indigo-jenkins}
+%\emph{INDIGO-DataCloud Jenkins}, https://jenkins.indigo-datacloud.eu:8080/
+%
+%\bibitem{indigo-github}
+%\emph{INDIGO-DataCloud GitHub}, https://www.github.com/indigo-dc
 

### 2017-11-09 paper.tex
remove cite

@@
 The ETICS and ETICS 2 (2006 -- 2010) \cite{cordis:etics} projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualization to improve quality, reliability and interoperability of large--scale scientific software.
-Both projects supported mechanisms to measure static software metrics, number of faults, number of defects and number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
+Both projects supported mechanisms to measure static software metrics, number of faults, number of defects and number of bugs; execute unit testing; deploy and test software on a virtual set of resources.
 The {\sl ETICS} portal was the first automated service for delivering software products in distributed environments such as the Grid.

### 2017-11-16 paper.tex
fix typo

--- 2017-11-08 paper.tex
@@
 In this project, a large effort was dedicated in the definition and implementation of academic software engineering practices due to the unusual context in which DataGrid aims, starting from the large geographical distribution of the teams involved in the development tasks, evolving requirements coming from the different application areas and the design of a new technology: the Grid \cite{datagrid}.
+DataGrid developed its own software distribution, {\sl EDG} (European DataGrid), strongly based on Globus Middleware Services \cite{globus}.
+An Architecture Task Force coordinated the overall design and technical consistency of the developments, complemented by the definition of various testing phases: preparation, execution and documentation.
 
-%The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
-The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts.
-Starting in EGEE--II, there was a need to move to a more dynamic release process.
-At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.i
-The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science.
-The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts.
-Starting in EGEE--II, there was a need to move to a more dynamic release process.
-At this stage no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.i
+The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010) \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together scientists and engineers from more than 240 institutions in 45 countries worldwide to provide a seamless Grid infrastructure for e--Science, available 24 hours--a--day.
+One of the main achievements of these projects was the middleware -- a crucial component of any Grid infrastructure as it provides the {\sl glue} that links the hardware resources within such infrastructure.
+The {\sl gLite} middleware \cite{glite} was the ultimate official distribution of EGEE as of 2006, after two years of prototyping and re--engineering efforts to converge with LHC Computing Grid ({\sl LCG-2}), Virtual Data Toolkit ({\sl VDT}) and {\sl Condor} \cite{condor} software distributions.
+The overall goal of the middleware engineering group was to provide and maintain selected middleware services of the {\sl gLite} distribution to satisfy the needs of the scientific communities.
+During EGEE--I, a software process was designed and put into action for the early releases of the {\sl gLite} middleware.
+This release process favoured a development environment of rapidly changing interfaces that were necessary during re--engineering.
+Starting in EGEE--II, there was a need to move to a more dynamic release process, and the first steps were taken towards an automated environment.
+At this stage  no software metrics were collected, while functional and integration testing were performed manually, and only on the last project, EGEE--III, an automatic build system was adopted; the E--Infrastructure for Testing, Integration and Configuration of Software ({\sl ETICS}) service.
+The {\sl gLite} middleware developed within EGEE will evolve further as part of a new project: the European Middleware Initiative (EMI).
 
 The ETICS \cite{cordis:etics} and ETICS 2 \cite{cordis:etics2} (Jan 2006 -- Feb 2010) projects aimed at defining and implementing a framework able to integrate a combination of new technologies in test management, release management tools and resource virtualization to improve quality, reliability and interoperability of large--scale scientific software.
-Both projects supported mechanisms to measure static software metrics, number of faults, number of defects and number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
+Both projects supported mechanisms to measure static software metrics (such as Lines of Code (LOC), Source LOC (SLOC), and Complexity); to measure the number of faults, the number of defects, number of bugs; execute unit testing; deploy and test software on a virtual set of resources \cite{etics}.
 The {\sl ETICS} portal was the first automated service for delivering software products in distributed environments such as the Grid.
@@
 The EMI (May 2010 -- Apr 2013) project \cite{cordis:emi} aimed to deliver a consolidated set of middleware products based on the four major middleware providers in Europe -- {\sl ARC}, {\sl dCache}, {\sl gLite} and {\sl UNICORE}.
-The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics, each defining the associated metrics and Key Performance Indicators (KPIs).
-EMI was one of the first projects in using a Continuous Integration (CI) system, leveraging on {\sl ETICS} tool  for building, packaging and testing the various components.
-%It leveraged the  and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.
+The EMI Quality Model used ISO/IEC 9126 Software Engineering -- Product Quality Standard \cite{iso9126} to identify a set of characteristics that need to be present in EMI software products and processes to be able to meet EMI quality requirements.
+For each software characteristic, a set of associated metrics and Key Performance Indicators (KPIs) were identified and defined in detail in the EMI Metrics Specification \cite{emi-metrics}.
+EMI was one of the first projects in using a Continuous Integration (CI) system for building, packaging and testing the various components.
+It leveraged the {\sl ETICS} tool and its plugin framework that provided not only the ability of collecting metrics during build and test execution, but also to make queries on the collected data and chart them through its chart generation framework.
 
-The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire} established a sustainable European Grid Infrastructure.
-EGI--InSPIRE maintained a production--ready Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution).
-The role of {\sl UMD} was to enforce the fulfilment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered \cite{mario}.
+The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire} is the continuation of EGEE--III project, establishing and maintaining a sustainable European Grid Infrastructure.
+EGI--InSPIRE did not develop any software but maintained a production--ready Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution).
+The role of {\sl UMD} was to enforce the fulfilment of a set of quality criteria definitions \cite{egi-qc} in the software being delivered (i.e., {\sl EMI} and {\sl Globus}) \cite{mario}.
 {\sl UMD} is still being used and deployed in the European scientific e--Infrastructures as a result of its continuation under a follow--up project, the EGI--Engage (Engaging the EGI Community towards an Open Science Commons) \cite{cordis:egi-engage}.
@@
 
-%\begin{figure*}
-%\centering
-%\includegraphics[width=\textwidth]{images/unittest.png}
-%\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
-%\label{fig:fig_unittest}
-%\end{figure*}
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/unittest.png}
+\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
+\label{fig:fig_unittest}
+\end{figure*}
 
@@
 The requirements are applied at early stages in the software development process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any bug or design issue is likely to be detected and corrected promptly.
-The source code is publicly available as GitHub repositories \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
+The source code is publicly available in GitHub repositories under an organization called {\sl indigo-dc} \cite{indigo-github} to augment the visibility of the product catalogue, promoting the external contributions and software adoption.
 In this regard, the source code is compliant with community de--facto style standards, selected from the variety of programming languages being used.
-Tests are triggered for each change to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case.
 
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=0.5\textwidth]{images/codestyle.png}
-%\caption{Code style standards followed by INDIGO--DataCloud's software products.}
-%\label{fig:fig_codestyle}
-%\end{figure}
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.5\textwidth]{images/codestyle.png}
+\caption{Code style standards followed by INDIGO--DataCloud's software products.}
+\label{fig:fig_codestyle}
+\end{figure}
 
-The evaluation of the software quality is performed by measuring the values of the metrics and KPIs defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard.
+Tests are triggered in each build to measure the unit and functional testing coverage, with a recommended value of 70\% of coverage in the former case.
+The documentation of the products is treated as source code, using a markup language, automatically rendered and uploaded in online repositories \cite{indigo-gitbook}.
+Changes in both documentation and source code are human--reviewed as the last step before merging them into the production repository.
+Last but not least, as already seen in the DevOps section, in order to facilitate the usage of the INDIGO--DataCloud services, they can be deployed automatically using either Ansible \cite{indigo-ansible} or Puppet \cite{indigo-puppet} open--source tools.
+
+The evaluation of the software quality is performed by measuring the values of the metrics and Key Performance Indicators (KPIs) defined based upon the ISO/IEC 9126 Software engineering -- Product quality standard.
 These metrics cover the development, release and maintenance phases of the software lifecycle.
+Development and release metrics are obtained automatically from several sources, such as GitHub and Jenkins CI service, and graphically displayed as GitHub pages using GrimoireLab framework \cite{grimoirelab}.
+Maintenance and user support metrics are collected from the different sources of data like GGUS \cite{ggus} and Github Issues.
 
@@
 \bibitem{cordis:datagrid}
-\emph{DataGrid project}, CORDIS Europe, http://cordis.europa.eu/project/rcn/53665\_en.html
+\emph{DataGrid project}, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/53665\_en.html
 
 \bibitem{datagrid}
-[3] L. Momtahan et al, \emph{e-Science Experiences: Software Engineering Practice and the EU DataGrid}, in Proc. Asia-Pacific Software Engineering Conference, 2002, pp. 269-275, IEEE Press
+L.
+Momtahan, A.
+Martin, \emph{e-Science Experiences: Software Engineering Practice and the EU DataGrid}, in Proc.
+Asia-Pacific Software Engineering Conference, Gold Coast, Queensland, Australia, 4-6 Dec.
+2002, pp.
+269-275, IEEE Press
+
+\bibitem{globus}
+I. Foster, C. Kesselman, \emph{Globus: a Metacomputing Infrastructure Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2, pp. 115–128, 1997
 
 \bibitem{cordis:egee}
-%\emph{Enabling Grids for E-sciencE (EGEE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html, EGEE-II -- http://cordis.europa.eu/project/rcn/99189\_en.html, EGEE-III -- http://cordis.europa.eu/project/rcn/87264\_en.html
-\emph{EGEE} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html, EGEE-II -- http://cordis.europa.eu/project/rcn/99189\_en.html, EGEE-III -- http://cordis.europa.eu/project/rcn/87264\_en.html
-%
-%\bibitem{cordis:egee}
-%\emph{Enabling Grids for E-sciencE (EGEE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80149\_en.html
-%
-%\bibitem{cordis:egee2}
-%\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/99189\_en.html
-%
-%\bibitem{cordis:egee3}
-%\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/87264\_en.html
+\emph{Enabling Grids for E-sciencE (EGEE)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/80149\_en.html
+
+\bibitem{cordis:egee2}
+\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/99189\_en.html
+
+\bibitem{cordis:egee3}
+\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/87264\_en.html
 
 \bibitem{glite}
-E. Laure et al, \emph{Programming the Grid with gLite}, in Computational Methods in Science and Technology, vol. 12(1), pp. 33–45, 2006, Scientific Publishers OWN
+E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A., Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1), pp. 33–45, 2006, Scientific Publishers OWN
+
+\bibitem{condor}
+D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the grid}, in Grid Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70, 2002, Eds. John Wiley \& Sons Inc.
 
 \bibitem{cordis:etics}
-\emph{ETICS} project, CORDIS Europe, ETICS -- http://cordis.europa.eu/project/rcn/80138\_en.html, ETICS-2 -- http://cordis.europa.eu/project/rcn/86604\_en.html
-%\bibitem{cordis:etics}
-%\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/80138\_en.html
-%
-%\bibitem{cordis:etics2}
-%\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/86604\_en.html
+\emph{E-Infrastructure for Testing, Integration and Configuration of Software (ETICS)} project, European Community Research and Development Information Service (CORDIS).
+Available: http://cordis.europa.eu/project/rcn/80138\_en.html
+
+\bibitem{cordis:etics2}
+\emph{E-Infrastructure for Testing, Integration and Configuration of Software - Phase 2 (ETICS 2)} project, European Community Research and Development Information Service (CORDIS).
+Available: http://cordis.europa.eu/project/rcn/86604\_en.html
 
 \bibitem{etics}
-A. Di Meglio et al, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol. 119, N. 4, 2008, IOP Publishing Ltd
+A.
+Di Meglio, M.-E.Bégin, P.
+Couvares, E.
+Ronchieri, E.
+Takacs, \emph{ETICS: the international software engineering service for the grid}, in Journal of Physics: Conference Series, Vol.
+119, N.
+4, 2008, IOP Publishing Ltd
 
 \bibitem{cordis:emi}
-%\emph{European Middleware Initiative (EMI)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95311\_en.html
-\emph{EMI} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95311\_en.html
+\emph{European Middleware Initiative (EMI)} project, European Community Research and Development Information Service (CORDIS). Available: http://cordis.europa.eu/project/rcn/95311\_en.html
 
 \bibitem{iso9126}
-\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International Organization for Standarization, https://www.iso.org/standard/22749.html
+\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International Organization for Standarization. Available: https://www.iso.org/standard/22749.html
+
+\bibitem{emi-metrics}
+\emph{EMI Metrics Specification}. Available: https://goo.gl/LdS6fL
 
 \bibitem{cordis:egi-inspire}
-%\emph{Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95923\_en.html
-\emph{EGI-InSPIRE} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/95923\_en.html
+\emph{European Grid Initiative: Integrated Sustainable Pan-European Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, European Community Research and Development Information Service (CORDIS).
+Available: http://cordis.europa.eu/project/rcn/95923\_en.html
 
 \bibitem{egi-qc}
-\emph{EGI Quality Criteria}, http://egi-qc.github.io/
+\emph{EGI Quality Criteria}. Available: http://egi-qc.github.io/
 
@@
 \bibitem{cordis:egi-engage}
-%\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194937\_en.html
-\emph{EGI-ENGAGE} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194937\_en.html
+\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)} project, European Community Research and Development Information Service (CORDIS).
+Available: http://cordis.europa.eu/project/rcn/194937\_en.html
 
 \bibitem{cordis:indigo}
-%\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194882\_en.html
-\emph{INDIGO-DataCloud} project, CORDIS Europe, http://cordis.europa.eu/project/rcn/194882\_en.html
+\emph{INtegrating Distributed data Infrastructures for Global ExplOitation (INDIGO-DataCloud)} project, European Community Research and Development Information Service (CORDIS).
+Available: http://cordis.europa.eu/project/rcn/194882\_en.html
 
 \bibitem{indigo-dockerhub}
-\emph{INDIGO-DataCloud DockerHub}, https://hub.docker.com/u/indigodatacloud
+\emph{INDIGO-DataCloud DockerHub repository}. Available: https://hub.docker.com/u/indigodatacloud
 
 \bibitem{indigo-d31}
-\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1, https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31
+\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1. Available: https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31
 
 \bibitem{indigo-jenkins}
-\emph{INDIGO-DataCloud Jenkins}, https://jenkins.indigo-datacloud.eu:8080/
+\emph{INDIGO-DataCloud Jenkins CI service}. Available: https://jenkins.indigo-datacloud.eu:8080/
 
 \bibitem{indigo-github}
-\emph{INDIGO-DataCloud GitHub}, https://www.github.com/indigo-dc
+\emph{INDIGO-DataCloud GitHub Source Code repository}. Available: https://www.github.com/indigo-dc
+
+\bibitem{indigo-gitbook}
+\emph{INDIGO-DataCloud GitBook Documentation repository}. Available: https://www.gitbook.com/@indigo-dc
+
+\bibitem{indigo-puppet}
+\emph{INDIGO-DataCloud PuppetForge repository}. Available: https://forge.puppet.com/indigodc
+
+\bibitem{indigo-ansible}
+\emph{INDIGO-DataCloud Ansible Galaxy repository}. Available: https://galaxy.ansible.com/indigo-dc/
+
+\bibitem{grimoirelab}
+\emph{GrimoireLab}. Available: http://grimoirelab.github.io/
+
+\bibitem{ggus}
+\emph{Global Grid User Support (GGUS)}. Available: https://www.ggus.eu/
 

### 2017-11-16 bare_jrnl.tex
Using TNS journal template

%% bare_jrnl.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE!
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/pkg/ifpdf
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\usepackage{graphicx}

\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at:
% http://www.ctan.org/pkg/graphicx
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/pkg/epslatex
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics.
%
% Note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/amsmath

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/pkg/algorithms
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/pkg/algorithmicx

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/array

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

% *** SUBFIGURE PACKAGES ***
%\ifCLASSOPTIONcompsoc
%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
%  \usepackage[caption=false,font=footnotesize]{subfig}
%\fi
% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
% for subfigure.sty, the latter of which is no longer maintained and is
% incompatible with some LaTeX packages including fixltx2e. However,
% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
% which will override IEEEtran.cls' handling of captions and this will result
% in non-IEEE style figure/table captions. To prevent this problem, be sure
% and invoke subfig.sty's "caption=false" package option (available since
% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
% handling of captions.
% Note that the Computer Society format requires a larger sans serif font
% than the serif footnote size font used in traditional IEEE formatting
% and thus the need to invoke different subfig.sty package options depending
% on whether compsoc mode has been enabled.
%
% The latest version and documentation of subfig.sty can be obtained at:
% http://www.ctan.org/pkg/subfig

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure.
% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
% corrections already built into the system in which case a warning will
% be issued if an attempt is made to load fixltx2e.sty as it is no longer
% needed.
% The latest version and documentation can be found at:
% http://www.ctan.org/pkg/fixltx2e

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/pkg/stfloats
% Do not use the stfloats baselinefloat ability as the IEEE does not allow
% \baselineskip to stretch. Authors submitting work to the IEEE should note
% that the IEEE rarely uses double column equations and that authors should try
% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
% such ways.
% Do not attempt to use stfloats with fixltx2e as they are incompatible.
% Instead, use Morten Hogholm'a dblfloatfix which combines the features
% of both fixltx2e and stfloats:
%
% \usepackage{dblfloatfix}
% The latest version can be found at:
% http://www.ctan.org/pkg/dblfloatfix

%\ifCLASSOPTIONcaptionsoff
%  \usepackage[nomarkers]{endfloat}
% \let\MYoriglatexcaption\caption
% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\fi
% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and
% Axel Sommerfeldt. This package may be useful when used in conjunction with
% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
% submissions have lists of figures/tables at the end of the paper and that
% figures/tables without any captions are placed on a page by themselves at
% the end of the document. If needed, the draftcls IEEEtran class option or
% \CLASSINPUTbaselinestretch interface can be used to increase the line
% spacing as well. Be sure and use the nomarkers option of endfloat to
% prevent endfloat from "marking" where the figures would have been placed
% in the text. The two hack lines of code above are a slight modification of
% that suggested by in the endfloat docs (section 8.4.1) to ensure that
% the full captions always appear in the list of figures/tables - even if
% the user used the short optional argument of \caption[]{}.
% IEEE papers do not typically make use of \caption[]'s optional argument,
% so this should not be an issue. A similar trick can be used to disable
% captions of packages such as subfig.sty that lack options to turn off
% the subcaptions:
% For subfig.sty:
% \let\MYorigsubfloat\subfloat
% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
% However, the above trick will not work if both optional arguments of
% the \subfloat command are used. Furthermore, there needs to be a
% description of each subfigure *somewhere* and endfloat does not add
% subfigure captions to its list of figures. Thus, the best approach is to
% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
% and instead reference/explain all the subfigures within the main caption.
% The latest version of endfloat.sty and its documentation can obtained at:
% http://www.ctan.org/pkg/endfloat
%
% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
% later in the document, say, to conditionally put the References on a
% page by themselves.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version and documentation can be obtained at:
% http://www.ctan.org/pkg/url
% Basically, \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Delivering reliable software in the European scientific arena: evolution of methodologies and new trend analysis}
%\title{15 years of European scientific software development: evolution and methodologies}

%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{Michael~Shell,~\IEEEmembership{Member,~IEEE,}
        John~Doe,~\IEEEmembership{Fellow,~OSA,}
        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
\thanks{M. Shell was with the Department
of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

% note the % following the last \IEEEmembership and also \thanks -
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
%
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
%
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
Ever since the Grid technology appeared as the new paradigm of distributed
computing to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements was
addressed in Europe by the creation of software development and
e-Infrastructure coordination dedicated projects. This work examines the
evolution of the software engineering methodologies used in past and present
EU-funded software projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the
foundation of the first project in 2001. Concretely, we describe the team
organizational structure, techniques and procedures that guided the Quality
Assurance definitions to deliver quality software. We highlight the challenges
and barriers faced throughout these years and how they were totally or
partially overcomed. To conclude we discuss the future directions of software
Quality Assurance methodologies for the incoming projects.
\end{abstract}

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
software reliability, quality assurance, software metrics, software testing
techniques
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
%
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
%
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
%
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
%
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{I}{n} the European scientific research arena, the analysis of
some of the last decade software development projects depicts a continuous
increase of the prominence and robustness of testing and validation procedures.
The observed trends are aligned with the evolution of the software engineering
techniques (e.g. rise of DevOps practices) over the last years. This is tied to
the parallel advancements in the ICT field (with virtualization and container
technologies) and the appearance of automation and event-response tools. In the
following we introduce a set of European Commission funded projects that go
from 2001 up to now, where the authors of this summary were and are currently
involved. For each project we explain how they addressed the challenge in
software reliability.

\section{Evolution of research on software reliability in EU-projects}
The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted
to develop the first European infrastructure for Grid computing. The project
brought together 21 academic and industry partners, coming from 15 different
countries. There were 3 application areas for requirement gathering i.e.
Particle physics, Bioinformatics and Earth observation. The project delivered
their own software distribution, named EDG (EU DataGrid), strongly based on
Globus Middleware Services \cite{globus}. The development organization
consisted in 5 work packages coordinated by an Architecture Task Force, in
charge of the overall design and technical consistency of the developments.
Mainly due to the fact of having no previous experience in such widely
collaborative projects, a big effort was spent trying to sort out the several
challenges arised \cite{datagrid}, namely the 1) communication overhead, as a
result of the large geographical separation of the parties involved in the
development tasks, 2) the evolution of the requirements coming from the user
communities, and 3) the lack of a body of knowledge for academic software
engineering. Hence agile methodologies were being introduced at that time,
through the Agile manifesto \cite{agile-manifesto}, so the project's
development design suffered from the methods, tools, techniques and best
practices that new discipline software engineering has brought along
\cite{agile}. However, the project focused on the experiences and procedures
from open source software projects, such as Linux and Apache, to try reach a
higher maturity level \cite{cmm}.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{images/devops.png}
\caption{Continuous Delivery workflow for Docker images.}
\label{fig:fig_CD}
\end{figure*}

The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from more than 240 institutions in 45 countries to
provide a seamless Grid infrastructure for e--Science. EGEE--II and EGEE--III
featured the internationalization of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15K researchers and deployed in
over 250 institutions. The {\sl gLite} middleware \cite{glite} was the ultimate
official distribution of EGEE as of 2006, after two years of prototyping and
re--engineering efforts to converge with LHC Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
development team was comprised of more that 80 people from 12 academic and
industrial institutions, that issued more than 10K bugs, 1.7K patches and
defined over 300 tasks tracked by the use of bug/task management tools. The
source code was available at a private, centralized version control system.
The code was not validated by any QA policy, but passed through a manual
certification procedure at the time of the release. This procedure tried to
improve the reliability of the software components by applying a set of
requirements [ref] to be fulfilled at each stage (integration, certification,
pre--production, production), and carried out by different teams. However, from
EGEE--II project, the project adopted the ETICS solution, that aimed to develop
an automatic build system for Grid middleware.

The E-Infrastructure for Testing, Integration and Configuration of Software
\cite{cordis:etics, cordis:etics2} (ETICS, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects like EGEE and its gLite middleware. The framework
integrated different technologies and tools to provide automated configuration,
build and testing capabilities, as well as auto-generated documentation and
software metrics gathering such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics}. ETICS portal was the first automated
service for delivering software products in distributed environments like
Grids.

The European Middleware Initiative (EMI, May 2010 -- Apr 2013)
\cite{cordis:emi} project joined the 4 major Grid middleware providers in
Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
to maintain and evolve the middleware focusing on extending the
interoperability in the Grid and improving the reliability of the services. The
ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
characteristics that needed to be present in the EMI software products and
processes to be able to meet the EMI quality requirements
\cite{emi-quality-model}. For each software characteristic, a set of associated
metrics and Key Performance Indicators (KPIs) were identified and defined in
detail in the EMI Metrics Specification \cite{emi-metrics}. The project
leveraged the {\sl ETICS} service for the development and release management,
as well as for metric tracking, making queries on the collected data to display
them through the chart generation framework.

The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
appeared as the continuation of the EGEE--III project, establishing and
maintaining a sustainable European Grid Infrastructure. EGI--InSPIRE did not
develop any software but maintained a production--ready Grid computing
middleware distribution called {\sl UMD} (Unified Middleware Distribution).
The role of {\sl UMD} was to enforce the fulfilment of a set of quality
criteria definitions \cite{egi-qc} in the software being delivered (i.e.,
{\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the EGI--Engage (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
ExplOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the ultimate
project of software development of the ones reviewed. At the time of writing
the project is succeeding to provide solutions to address the existing gaps in
cloud Platform--as--a--Service (PaaS) and Software-as-a-Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{images/unittest.png}
\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
\label{fig:fig_unittest}
\end{figure*}

\section{New Trends in Software Reliability}

The INDIGO--Datacloud project is a reflection of the progress made in software
quality and reliability aspects throughout the past scientific European
experiences. Nevertheless, the project’s Quality Assurance procedures are also
highly influenced by the insights of current big worldwide collaborations of
software development. These collaborations have continuously inspired new
technologies and procedures to increase the robustness and quality of the
software being delivered. In this scenario, the DevOps culture is one of the
major outcomes that has been progressively adopted in the INDIGO--DataCloud
project.

\subsection{DevOps practices}

The DevOps methods emphasize on exercising Quality Assurance techniques to
avoid infrastructure disruption whenever new developments are deployed into
production. The INDIGO--DataCloud project promoted the application of a CI
scenario to enforce the Quality Assurance requirements for any piece of
software being produced within the project. Such environment requires an
automated ready--to--go infrastructure where the different services involved
(source code management platform, automation server) interact with each other
to trigger the quality check execution and return back the exit codes, together
with their outputs. To accomplish this, the project leveraged tightly
integrated open source tools such as GitHub and Jenkins, as well as relying on
Docker container provisioning to speed up the source code validation. This CI
approach is guiding the project’s software development phase throughout the
first and second major releases.

As DevOps suggests, frequent releases positively affects the reliability of the
software. The software updates of INDIGO--DataCloud products taking place since
the second major release are passing through a Continuous Delivery (CD)
pipeline that adds the packaging of the software right after the execution of
the quality checks (as part of the CI). The steps defined within the CD
pipeline differ whether the software is to be distributed in the form of Docker
images or via operating system’s packages. In the latter case an extra
validation step is added, consisting in the product’s deployment using a
Configuration Management (CM) solution. The installation uses the packages
being built in the previous step, which are being uploaded to a testing
repository. In the case of Docker images, the CM is used to build the image
itself, installing and configuring the product before uploading it to the
DockerHub repository \cite{indigo-dockerhub}. Figure \ref{fig:fig_CD} showcases
the workflow followed by the products distributed as Docker images.

\subsection{Software Quality Procedures}

The software quality procedures \cite{indigo-d31} cover: 1) the identification
and description of the \emph{quality requirements} that the software need to
comply with, and 2) the \emph{quality metrics}.

The requirements are applied at early stages in the software development
process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any
bug or design issue is likely to be detected and corrected promptly. The source
code is publicly available in GitHub repositories under an organization called
{\sl indigo-dc} \cite{indigo-github} to augment the visibility of the product
catalogue, promoting the external contributions and software adoption. In this
regard, the source code is compliant with community de--facto style standards,
selected from the variety of programming languages being used.

\begin{figure}[!t]
\centering
\includegraphics[width=0.5\textwidth]{images/codestyle.png}
\caption{Code style standards followed by INDIGO--DataCloud's software products.}
\label{fig:fig_codestyle}
\end{figure}

Tests are triggered in each build to measure the unit and functional testing
coverage, with a recommended value of 70\% of coverage in the former case. The
documentation of the products is treated as source code, using a markup
language, automatically rendered and uploaded in online repositories
\cite{indigo-gitbook}. Changes in both documentation and source code are
human--reviewed as the last step before merging them into the production
repository. Last but not least, as already seen in the DevOps section, in order
to facilitate the usage of the INDIGO--DataCloud services, they can be deployed
automatically using either Ansible \cite{indigo-ansible} or Puppet
\cite{indigo-puppet} open--source tools.

The evaluation of the software quality is performed by measuring the values of
the metrics and Key Performance Indicators (KPIs) defined based upon the
ISO/IEC 9126 standard. These metrics cover the development, release and
maintenance phases of the software lifecycle. Development and release metrics
are obtained automatically from several sources, such as GitHub and Jenkins CI
service, and graphically displayed as GitHub pages using GrimoireLab framework
\cite{grimoirelab}. Maintenance and user support metrics are collected from the
different sources of data like GGUS \cite{ggus} and Github Issues.

\subsection{Integration, preview and early adoption}

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the INDIGO--DataCloud
solution and use cases validation, by deploying and executing the applications
with the last stable version of the software.

The released software is also tested in production environments through the
stage--rollout process. Selected resource providers are requested to install
the most updated stable versions of the software, accessed by their users. The
stage--rollout process is key to detect and mitigate issues that could only
appear in production.

\section{Conclusion}

It is a fact that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. Recent software engineering insights, such as
DevOps, are gradually being applied as the open--source collaborative tools
evolve, allowing tighter integrations. The focus of the Software Quality
Assurance procedures has been integrated in the development stage, trying to
detect and correct issues early in the software lifecycle. Automation and
metric analysis are at the base of prompt issue solving. However, post--release
validation, such as integration testbeds and the stage--rollout process, are
also needed to strengthen the software reliability for scientific usage.

%\appendices
%\section{Proof of the First Zonklar Equation}
%Appendix one text goes here.
%
%% you can choose not to have a title for an appendix
%% if you want by leaving the argument blank
%\section{}
%Appendix two text goes here.

% use section* for acknowledgment
\section*{Acknowledgment}

The authors would like to thanks European Commission with the various funded
projects.

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{cordis:datagrid}
\emph{DataGrid project}, European Community Research and Development
Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/53665\_en.html

\bibitem{globus}
I. Foster and C. Kesselman, \emph{Globus: a Metacomputing Infrastructure
Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2,
pp. 115–128, 1997.

\bibitem{datagrid}
L. Momtahan and A. Martin, \emph{e-Science Experiences: Software Engineering
Practice and the EU DataGrid}, in Proc. Asia-Pacific Software Engineering
Conference, Gold Coast, Queensland, Australia, pp. 269-275, IEEE Press,
4-6 Dec. 2002.

\bibitem{agile-manifesto}
\emph{Manifesto for Agile Software Development}, http://agilemanifesto.org

\bibitem{agile}
T. Dingsoyr, \emph{A decade of agile methodologies: Towards explaining agile
software development} in The Journal of Systems and Software, Elsevier Inc,
2012.

\bibitem{cmm}
Paulk et al., \emph{Capability maturity model for software}, Software
Engineering Institute, report CMU/SEI-93-TR-24, ESC-TR-93-177, Feb. 1993.

\bibitem{cordis:egee}
\emph{Enabling Grids for E-sciencE (EGEE)} project, European Community
Research and Development Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/80149\_en.html

\bibitem{cordis:egee2}
\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, European Community
Research and Development Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/99189\_en.html

\bibitem{cordis:egee3}
\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, European Community
Research and Development Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/87264\_en.html

\bibitem{glite}
E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A.,
Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1),
pp. 33–45, Scientific Publishers OWN, 2006.

\bibitem{condor}
D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the grid}, in Grid
Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70,
Eds. John Wiley \& Sons Inc., 2002.

\bibitem{cordis:etics}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software
(ETICS)} project, European Community Research and Development Information
Service (CORDIS), http://cordis.europa.eu/project/rcn/80138\_en.html

\bibitem{cordis:etics2}
\emph{E-Infrastructure for Testing, Integration and Configuration of Software -
Phase 2 (ETICS 2)} project, European Community Research and Development
Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/86604\_en.html

\bibitem{etics}
A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
the international software engineering service for the grid}, in Journal of
Physics: Conference Series, Vol. 119, N. 4, IOP Publishing Ltd, 2008.

\bibitem{cordis:emi}
\emph{European Middleware Initiative (EMI)} project, European Community
Research and Development Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/95311\_en.html

\bibitem{iso-9126}
\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International
Organization for Standarization, https://www.iso.org/standard/22749.html

\bibitem{emi-metrics}
\emph{EMI Metrics Specification}, https://goo.gl/CCtY7x

\bibitem{emi-quality-model}
\emph{EMI Quality Model}. Available: https://goo.gl/LdS6fL

\bibitem{cordis:egi-inspire}
\emph{European Grid Initiative: Integrated Sustainable Pan-European
Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, European
Community Research and Development Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/95923\_en.html

\bibitem{egi-qc}
\emph{EGI Quality Criteria}. Available: http://egi-qc.github.io/

\bibitem{mario}
M. David et al, \emph{Validation of Grid Middleware for the European Grid
Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543–558,
Springer, 2014.

\bibitem{cordis:egi-engage}
\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)}
project, European Community Research and Development Information Service
(CORDIS), http://cordis.europa.eu/project/rcn/194937\_en.html

\bibitem{cordis:indigo}
\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
(INDIGO-DataCloud)} project, European Community Research and Development
Information Service (CORDIS),
http://cordis.europa.eu/project/rcn/194882\_en.html

\bibitem{indigo-dockerhub}
\emph{INDIGO-DataCloud DockerHub repository},
https://hub.docker.com/u/indigodatacloud

\bibitem{indigo-d31}
\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1,
https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31

\bibitem{indigo-jenkins}
\emph{INDIGO-DataCloud Jenkins CI service},
https://jenkins.indigo-datacloud.eu:8080/

\bibitem{indigo-github}
\emph{INDIGO-DataCloud GitHub Source Code repository},
https://www.github.com/indigo-dc

\bibitem{indigo-gitbook}
\emph{INDIGO-DataCloud GitBook Documentation repository},
https://www.gitbook.com/@indigo-dc

\bibitem{indigo-puppet}
\emph{INDIGO-DataCloud PuppetForge repository},
https://forge.puppet.com/indigodc

\bibitem{indigo-ansible}
\emph{INDIGO-DataCloud Ansible Galaxy repository},
https://galaxy.ansible.com/indigo-dc/

\bibitem{grimoirelab}
\emph{GrimoireLab}, http://grimoirelab.github.io/

\bibitem{ggus}
\emph{Global Grid User Support (GGUS)}, https://www.ggus.eu/

\end{thebibliography}

% biography section
%
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

\begin{IEEEbiography}{Michael Shell}
Biography text here.
\end{IEEEbiography}

% if you will not have a photo at all:
\begin{IEEEbiographynophoto}{John Doe}
Biography text here.
\end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

\begin{IEEEbiographynophoto}{Jane Doe}
Biography text here.
\end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2017-11-30 paper-ieee-trans.tex
several corrections, modifications

--- 2017-11-16 bare_jrnl.tex
@@
-%% bare_jrnl.tex
-%% V1.4b
-%% 2015/08/26
-%% by Michael Shell
-%% see http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
-%% journal paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/pkg/ieeetran
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE!
-%% User assumes all risk.
-%% In no event shall the IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. The IEEE's font choices and paper sizes can   ***
-% *** trigger bugs that do not appear when using other class files.       ***                          ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
 \documentclass[journal]{IEEEtran}
-%
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[journal]{../sty/IEEEtran}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/pkg/ifpdf
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of the IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off
-% such as if a citation ever needs to be enclosed in parenthesis.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 5.0 (2009-03-20) and later if using hyperref.sty.
-% The latest version can be obtained at:
-% http://www.ctan.org/pkg/cite
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
 \usepackage{graphicx}
-
 \ifCLASSINFOpdf
   % \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
   % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
 \else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
   % \DeclareGraphicsExtensions{.eps}
 \fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation
-% can be obtained at:
-% http://www.ctan.org/pkg/graphicx
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found at:
-% http://www.ctan.org/pkg/epslatex
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). The IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
 
-% *** MATH PACKAGES ***
-%
-%\usepackage{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics.
-%
-% Note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/amsmath
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-%\usepackage{algorithmic}
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as the IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/pkg/algorithms
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/pkg/algorithmicx
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/array
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-% *** SUBFIGURE PACKAGES ***
-%\ifCLASSOPTIONcompsoc
-%  \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
-%\else
-%  \usepackage[caption=false,font=footnotesize]{subfig}
-%\fi
-% subfig.sty, written by Steven Douglas Cochran, is the modern replacement
-% for subfigure.sty, the latter of which is no longer maintained and is
-% incompatible with some LaTeX packages including fixltx2e. However,
-% subfig.sty requires and automatically loads Axel Sommerfeldt's caption.sty
-% which will override IEEEtran.cls' handling of captions and this will result
-% in non-IEEE style figure/table captions. To prevent this problem, be sure
-% and invoke subfig.sty's "caption=false" package option (available since
-% subfig.sty version 1.3, 2005/06/28) as this is will preserve IEEEtran.cls
-% handling of captions.
-% Note that the Computer Society format requires a larger sans serif font
-% than the serif footnote size font used in traditional IEEE formatting
-% and thus the need to invoke different subfig.sty package options depending
-% on whether compsoc mode has been enabled.
-%
-% The latest version and documentation of subfig.sty can be obtained at:
-% http://www.ctan.org/pkg/subfig
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure.
-% Be aware that LaTeX2e kernels dated 2015 and later have fixltx2e.sty's
-% corrections already built into the system in which case a warning will
-% be issued if an attempt is made to load fixltx2e.sty as it is no longer
-% needed.
-% The latest version and documentation can be found at:
-% http://www.ctan.org/pkg/fixltx2e
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/pkg/stfloats
-% Do not use the stfloats baselinefloat ability as the IEEE does not allow
-% \baselineskip to stretch. Authors submitting work to the IEEE should note
-% that the IEEE rarely uses double column equations and that authors should try
-% to avoid such use. Do not be tempted to use the cuted.sty or midfloat.sty
-% packages (also by Sigitas Tolusis) as the IEEE does not format its papers in
-% such ways.
-% Do not attempt to use stfloats with fixltx2e as they are incompatible.
-% Instead, use Morten Hogholm'a dblfloatfix which combines the features
-% of both fixltx2e and stfloats:
-%
-% \usepackage{dblfloatfix}
-% The latest version can be found at:
-% http://www.ctan.org/pkg/dblfloatfix
-
-%\ifCLASSOPTIONcaptionsoff
-%  \usepackage[nomarkers]{endfloat}
-% \let\MYoriglatexcaption\caption
-% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
-%\fi
-% endfloat.sty was written by James Darrell McCauley, Jeff Goldberg and
-% Axel Sommerfeldt. This package may be useful when used in conjunction with
-% IEEEtran.cls'  captionsoff option. Some IEEE journals/societies require that
-% submissions have lists of figures/tables at the end of the paper and that
-% figures/tables without any captions are placed on a page by themselves at
-% the end of the document. If needed, the draftcls IEEEtran class option or
-% \CLASSINPUTbaselinestretch interface can be used to increase the line
-% spacing as well. Be sure and use the nomarkers option of endfloat to
-% prevent endfloat from "marking" where the figures would have been placed
-% in the text. The two hack lines of code above are a slight modification of
-% that suggested by in the endfloat docs (section 8.4.1) to ensure that
-% the full captions always appear in the list of figures/tables - even if
-% the user used the short optional argument of \caption[]{}.
-% IEEE papers do not typically make use of \caption[]'s optional argument,
-% so this should not be an issue. A similar trick can be used to disable
-% captions of packages such as subfig.sty that lack options to turn off
-% the subcaptions:
-% For subfig.sty:
-% \let\MYorigsubfloat\subfloat
-% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
-% However, the above trick will not work if both optional arguments of
-% the \subfloat command are used. Furthermore, there needs to be a
-% description of each subfigure *somewhere* and endfloat does not add
-% subfigure captions to its list of figures. Thus, the best approach is to
-% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
-% and instead reference/explain all the subfigures within the main caption.
-% The latest version of endfloat.sty and its documentation can obtained at:
-% http://www.ctan.org/pkg/endfloat
-%
-% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
-% later in the document, say, to conditionally put the References on a
-% page by themselves.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
-%\usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version and documentation can be obtained at:
-% http://www.ctan.org/pkg/url
-% Basically, \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
 \hyphenation{op-tical net-works semi-conduc-tor}
@@
 \begin{document}
-%
-% paper title
-% Titles are generally capitalized except for words such as a, an, and, as,
-% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
-% not capitalized unless they are the first or last word of the title.
-% Linebreaks \\ can be used within to get better formatting as desired.
-% Do not put math or special symbols in the title.
 \title{Delivering reliable software in the European scientific arena: evolution of methodologies and new trend analysis}
@@
 
-%
-%
-% author names and IEEE memberships
-% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
-
-\author{Michael~Shell,~\IEEEmembership{Member,~IEEE,}
-        John~Doe,~\IEEEmembership{Fellow,~OSA,}
-        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
-\thanks{M. Shell was with the Department
-of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
-GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
-\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
-\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}
-
-% note the % following the last \IEEEmembership and also \thanks -
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-%
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
+\author{E. Ronchieri,
+        P. Orviz Fernandez,
+        M. David,
+		D.C. Duma,
+        J. Gomes,
+        D. Salomoni
+\thanks{D. C. Duma, E. Ronchieri, and D. Salomoni are with INFN - CNAF, Bologna, Italy.}
+\thanks{P. Orviz Fernandez, is with CSIC, Santander, Spain, (e-mail: orviz@ifca.unican.es)}
+\thanks{M. David, and J. Gomes are with LIP, Lisbon, Portugal}%
+}
 
@@
 {Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-%
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
 
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
 \maketitle
 
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
-Ever since the Grid technology appeared as the new paradigm of distributed
+From the advent of Grid technology as the new paradigm of distributed
 computing to the current days of Cloud computing models, the continuous need
@@
 reliability has progressed in Europe over the last 15 years following the
-foundation of the first project in 2001. Concretely, we describe the team
+foundation of the first project in 2001. In particular, we describe the team
 organizational structure, techniques and procedures that guided the Quality
 Assurance definitions to deliver quality software. We highlight the challenges
-and barriers faced throughout these years and how they were totally or
-partially overcomed. To conclude we discuss the future directions of software
+and barriers faced throughout these years and how to overcome them totally or
+partially. To conclude, we discuss the future directions of software
 Quality Assurance methodologies for the incoming projects.
@@
 
-% Note that keywords are not normally used for peerreview papers.
 \begin{IEEEkeywords}
@@
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
 \IEEEpeerreviewmaketitle
@@
 \section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-%
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-%
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by the IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-%
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-%
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
+
 \IEEEPARstart{I}{n} the European scientific research arena, the analysis of
@@
 the parallel advancements in the ICT field (with virtualization and container
-technologies) and the appearance of automation and event-response tools. In the
-following we introduce a set of European Commission funded projects that go
-from 2001 up to now, where the authors of this summary were and are currently
+technologies) and the emergence of automation and event-response tools. In the
+following, we introduce a set of European Commission funded projects that cover
+2001 up to now, where the authors of this summary were and are currently
 involved. For each project we explain how they addressed the challenge in
@@
 \section{Evolution of research on software reliability in EU-projects}
+
 The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted
 to develop the first European infrastructure for Grid computing. The project
-brought together 21 academic and industry partners, coming from 15 different
-countries. There were 3 application areas for requirement gathering i.e.
+brought together 21 academic and industry partners, from 15 different
+countries. There were 3 application areas for requirement gathering;
 Particle physics, Bioinformatics and Earth observation. The project delivered
@@
 Globus Middleware Services \cite{globus}. The development organization
-consisted in 5 work packages coordinated by an Architecture Task Force, in
-charge of the overall design and technical consistency of the developments.
-Mainly due to the fact of having no previous experience in such widely
-collaborative projects, a big effort was spent trying to sort out the several
-challenges arised \cite{datagrid}, namely the 1) communication overhead, as a
+consisted in 5 work packages coordinated by an Architecture Task Force,
+supervising the overall design and technical consistency of the developments.
+
+Due to having no previous experience in such widely
+collaborative projects, a big effort was spent trying to devise solutions for several
+challenges \cite{datagrid}, namely the 1) communication overhead, as a
 result of the large geographical separation of the parties involved in the
@@
 communities, and 3) the lack of a body of knowledge for academic software
-engineering. Hence agile methodologies were being introduced at that time,
-through the Agile manifesto \cite{agile-manifesto}, so the project's
+engineering.
+
+Agile methodologies were being introduced at the time
+through the Agile manifesto \cite{agile-manifesto}, as such the project's
 development design suffered from the methods, tools, techniques and best
-practices that new discipline software engineering has brought along
-\cite{agile}. However, the project focused on the experiences and procedures
-from open source software projects, such as Linux and Apache, to try reach a
+practices emerging from the new discipline of software engineering
+\cite{agile}. Nonetheless, the project focused on the experiences and procedures
+from open source software projects, such as Linux and Apache, trying to reach a
 higher maturity level \cite{cmm}.
@@
 rich set of new services while maintaining a sustainable infrastructure for
-Grid computing, eventually used by more than 15K researchers and deployed in
-over 250 institutions. The {\sl gLite} middleware \cite{glite} was the ultimate
+Grid computing, eventually used by more than 15 thousand researchers and deployed in
+over 250 institutions.
+
+The {\sl gLite} middleware \cite{glite} was the ultimate
 official distribution of EGEE as of 2006, after two years of prototyping and
@@
 development team was comprised of more that 80 people from 12 academic and
-industrial institutions, that issued more than 10K bugs, 1.7K patches and
-defined over 300 tasks tracked by the use of bug/task management tools. The
+industrial partners, that issued more than 10 thousand bugs, 1.7 thousand patches and
+defined over 300 tasks tracked by the use of bug/task management tools.
+
+The
 source code was available at a private, centralized version control system.
-The code was not validated by any QA policy, but passed through a manual
+The code was not validated by any Quality Assurance policy, but passed through a manual
 certification procedure at the time of the release. This procedure tried to
@@
 requirements [ref] to be fulfilled at each stage (integration, certification,
-pre--production, production), and carried out by different teams. However, from
-EGEE--II project, the project adopted the ETICS solution, that aimed to develop
+pre--production, production), carried out by different teams. However, starting in
+EGEE--II, the project adopted the ETICS solution, aiming to develop
 an automatic build system for Grid middleware.
 
-The E-Infrastructure for Testing, Integration and Configuration of Software
+The E--Infrastructure for Testing, Integration and Configuration of Software
 \cite{cordis:etics, cordis:etics2} (ETICS, Jan 2006 -- Feb 2010) project aimed
 at addressing the challenges in producing quality software in distributed,
-collaborative projects like EGEE and its gLite middleware. The framework
+collaborative projects such as EGEE and its gLite middleware. The framework
 integrated different technologies and tools to provide automated configuration,
@@
 software metrics gathering such as Source Lines Of Code (SLOC), complexity and
-number of defects/bugs \cite{etics}. ETICS portal was the first automated
+number of defects/bugs \cite{etics}. The ETICS portal was the first automated
 service for delivering software products in distributed environments like
-Grids.
+the Grids.
 
@@
 processes to be able to meet the EMI quality requirements
-\cite{emi-quality-model}. For each software characteristic, a set of associated
+\cite{emi-quality-model}.
+
+For each software characteristic, a set of associated
 metrics and Key Performance Indicators (KPIs) were identified and defined in
@@
 Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
-appeared as the continuation of the EGEE--III project, establishing and
-maintaining a sustainable European Grid Infrastructure. EGI--InSPIRE did not
+arose as the continuation of the EGEE--III project, establishing and
+maintaining a sustainable European Grid Infrastructure.
+
+EGI--InSPIRE did not
 develop any software but maintained a production--ready Grid computing
@@
 The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
-ExplOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the ultimate
-project of software development of the ones reviewed. At the time of writing
-the project is succeeding to provide solutions to address the existing gaps in
-cloud Platform--as--a--Service (PaaS) and Software-as-a-Service (SaaS) levels,
+ExplOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
+project of software development considered in this review. At the time of writing
+the project is succeeding in providing solutions to address the existing gaps in
+cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
 helping developers, e--Infrastructures and scientific communities to exploit
@@
 
-The INDIGO--Datacloud project is a reflection of the progress made in software
+The INDIGO--Datacloud project reflects the progress made in software
 quality and reliability aspects throughout the past scientific European
@@
 avoid infrastructure disruption whenever new developments are deployed into
-production. The INDIGO--DataCloud project promoted the application of a CI
+production. The INDIGO--DataCloud project promoted the application of a Continuous Integration (CI)
 scenario to enforce the Quality Assurance requirements for any piece of
@@
 to trigger the quality check execution and return back the exit codes, together
-with their outputs. To accomplish this, the project leveraged tightly
+with their outputs. To accomplish this, the project leveraged on tightly
 integrated open source tools such as GitHub and Jenkins, as well as relying on
@@
 pipeline that adds the packaging of the software right after the execution of
-the quality checks (as part of the CI). The steps defined within the CD
+the quality checks (as part of the CI).
+
+The steps defined within the CD
 pipeline differ whether the software is to be distributed in the form of Docker
@@
 Configuration Management (CM) solution. The installation uses the packages
-being built in the previous step, which are being uploaded to a testing
+built in the previous step and uploaded to a testing
 repository. In the case of Docker images, the CM is used to build the image
 itself, installing and configuring the product before uploading it to the
-DockerHub repository \cite{indigo-dockerhub}. Figure \ref{fig:fig_CD} showcases
+DockerHub repository \cite{indigo-dockerhub}. Figure \ref{fig:fig_CD} shows
 the workflow followed by the products distributed as Docker images.
@@
 The requirements are applied at early stages in the software development
-process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so any
+process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
 bug or design issue is likely to be detected and corrected promptly. The source
 code is publicly available in GitHub repositories under an organization called
-{\sl indigo-dc} \cite{indigo-github} to augment the visibility of the product
-catalogue, promoting the external contributions and software adoption. In this
-regard, the source code is compliant with community de--facto style standards,
+{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the product
+catalogue, promoting the external contributions and software adoption. Furthermore,
+the source code is compliant with community de--facto style standards,
 selected from the variety of programming languages being used.

### 2017-11-30 paper-ieee-trans.tex
add text and new image for automatic configuration

@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/unittest.png}
-\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
-\label{fig:fig_unittest}
-\end{figure*}
-
 \section{New Trends in Software Reliability}
@@
 \subsection{DevOps practices}
+\label{sec:devops}
 
@@
 the source code is compliant with community de--facto style standards,
-selected from the variety of programming languages being used.
+selected from the variety of programming languages being used as seen in
+Fig. \ref{fig:fig_codestyle}.
 
@@
 
-Tests are triggered in each build to measure the unit and functional testing
-coverage, with a recommended value of 70\% of coverage in the former case. The
-documentation of the products is treated as source code, using a markup
-language, automatically rendered and uploaded in online repositories
+Unit testing coverage has increased between the first ({\sl INDIGO-1}) and the
+second ({\sl INDIGO-2}) INDIGO--DataCloud releases, reaching an average value of
+66.69\% on the {\sl INDIGO-2} release, very close to the recommended 70\% threshold
+defined by the project’s SQA base criteria \cite{indigo-d31}. A comparison graph
+between the code coverage for the products involved both in {\sl INDIGO-1} and
+{\sl INDIGO-2} releases is shown in Fig. \ref{fig:fig_unittest}. As it can be seen,
+a very small set of products with low coverage values highly impacted the overall
+result, although a significant part of these products are upstream contributions
+or products with codebases not fully under the control of the project. Nevertheless,
+60\% of the products were over this threshold while 80\% of them have values above
+the 50\% coverage.
+
+The documentation of the products is treated as source code, using a markup
+language, automatically rendered and uploaded to online repositories
 \cite{indigo-gitbook}. Changes in both documentation and source code are
 human--reviewed as the last step before merging them into the production
-repository. Last but not least, as already seen in the DevOps section, in order
+repository.
+
+Last but not least, as described in section \ref{sec:devops}, in order
 to facilitate the usage of the INDIGO--DataCloud services, they can be deployed
 automatically using either Ansible \cite{indigo-ansible} or Puppet
-\cite{indigo-puppet} open--source tools.
+\cite{indigo-puppet} open--source tools. The adoption of such Configuration Management
+tools not only eases the consumption of the products developed by external users or
+infrastructures, but also tackles the correct installation and configuration of the
+services and applications. A representative example is the number of roles currently
+hosted in the Ansible Galaxy portal for the project.
+
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/unittest.png}
+\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
+\label{fig:fig_unittest}
+\end{figure*}
+
+Fig. \ref{fig:fig_confman} shows the number of products that offer an automated means
+for deployment. It shows an increase after the {\sl INDIGO-1} release with specially
+in the previous weeks of each of the two major releases.
 
@@
 \cite{grimoirelab}. Maintenance and user support metrics are collected from the
-different sources of data like GGUS \cite{ggus} and Github Issues.
+different sources of data such as GGUS \cite{ggus} and Github Issues.
+
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/confman.png}
+\caption{Trend line showing the adoption of Configuration Management tools throughout the project lifetime.}
+\label{fig:fig_confman}
+\end{figure*}
 
@@
 
-%\appendices
-%\section{Proof of the First Zonklar Equation}
-%Appendix one text goes here.
-%
-%% you can choose not to have a title for an appendix
-%% if you want by leaving the argument blank
-%\section{}
-%Appendix two text goes here.
-
-% use section* for acknowledgment
 \section*{Acknowledgment}
@@
 
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
-\ifCLASSOPTIONcaptionsoff
-  \newpage
-\fi
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
 \begin{thebibliography}{1}
@@
 
-% biography section
-%
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
 \begin{IEEEbiography}{Michael Shell}
@@
 
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-\begin{IEEEbiographynophoto}{Jane Doe}
-Biography text here.
-\end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
-% that's all folks
 \end{document}

### 2017-11-30 paper-ieee-trans.tex
add text and new image for devops for applications

@@
 
+\subsection{DevOps adoption from user communities}
+
+The experience gathered throughout the project with regard to the adoption of
+different DevOps practices is not only useful and suitable for the software related
+to the core services in the INDIGO-DataCloud solution, but also applicable to the
+development and distribution of the applications coming from the user communities.
+
+Two applications from WP2, DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
+integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
+Figure \ref{fig:fig_disvis} shows the pipeline for the Disvis application.
+
+User application developers were provided with both a means to validate the
+source code before merging and the creation of a new versioned Docker image,
+automatically available in the INDIGO-DataCloud’s application repository.
+
+The novelty introduced in the pipeline above is the validation of the application.
+Once the application is deployed as a Docker container, and subsequently uploaded
+to the Dockerhub repository, it is instantiated in a new container to be validated.
+The application is then executed and the results compared with a set of reference outputs.
+Thus this pipeline implementation goes a step forward by testing the application
+execution for the last available Docker image in the catalogue.
+
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/disvis.png}
+\caption{DevOps pipeline to distribute Docker images for Disvis application.}
+\label{fig:fig_disvis}
+\end{figure*}
+
 \subsection{Integration, preview and early adoption}
@@
 solution and use cases validation, by deploying and executing the applications
-with the last stable version of the software.
+with the last stable version of the software. A map of the Pilot Preview
+infrastructure is depicted in Fig. \ref{fig:fig_pilotpreview}, it shows the
+resources providers and the components or services they deployed and supported.
+
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/pilotpreview.png}
+\caption{Resource centers supporting the Pilot Preview testbed and corresponding
+set of deployed INDIGO-DataCloud components or services.}
+\label{fig:fig_confman}
+\end{figure*}
 
@@
 
+\bibitem{disvis} G.C.P. Van Zundert, A.M.J.J. Bonvin,
+DisVis: quantifying and visualizing the accessible interaction space of distance
+restrained biomolecular complexes, Bioinformatics 31 (2015) 3222–3224
+
+\bibitem{powerfit} G.C.P. Van Zundert, A.M.J.J. Bonvin,
+Fast and sensitive rigid-body fitting into cryo-EM density maps with PowerFit,
+AIMS Biophys. 2 (2015) 73–87.
+
 \end{thebibliography}

### 2017-12-01 paper-ieee-trans.tex
added changes

@@
 
-\hyphenation{op-tical net-works semi-conduc-tor}
+%\hyphenation{op-tical net-works semi-conduc-tor}
 
 \begin{document}
-\title{Delivering reliable software in the European scientific arena: evolution of methodologies and new trend analysis}
+\title{Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis}
 %\title{15 years of European scientific software development: evolution and methodologies}
@@
         M. David,
-		D.C. Duma,
+		D. C. Duma,
         J. Gomes,
@@
 % The paper headers
-\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
-{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
+%\markboth{IEEEtran for IEEE Journals}
 
@@
 
-\begin{IEEEbiography}{Michael Shell}
+\begin{IEEEbiography} [{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/Ronchieri}}]{Elisabetta Ronchieri}
+is a Computer Science Engineer at INFN (Istituto Nazionale di Fisica Nucleare - National Institute of Nuclear Physics) CNAF.
+Her main area of interest include software engineering issues such as software quality and software prediction model.
+She is also interested in (big and open) data analysis by using predictive techniques coming from various theories.
+such as Machine Learning.
+\end{IEEEbiography}
+
+P. Orviz Fernandez,
+        M. David,
+		D. C. Duma,
+        J. Gomes,
+        D. Salomoni
+
+% if you will not have a photo at all:
+\begin{IEEEbiography}{Pablo Orviz Fernandez}
 Biography text here.
@@
 
-% if you will not have a photo at all:
-\begin{IEEEbiographynophoto}{John Doe}
+\begin{IEEEbiography}{Mario David}
 Biography text here.
-\end{IEEEbiographynophoto}
+\end{IEEEbiography}
+
+\begin{IEEEbiography}{D. Cristina Duma}
+Biography text here.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}{J. Gomes}
+Biography text here.
+\end{IEEEbiography}
+
+\begin{IEEEbiography}{Davide Salomoni}
+Biography text here.
+\end{IEEEbiography}
 

### 2017-12-01 paper-ieee-trans.tex
added changes

@@
 computing to the current days of Cloud computing models, the continuous need
-of new tools and services to match the scientific community requirements was
+of new tools and services to match the scientific community requirements has been
 addressed in Europe by the creation of software development and
@@
 evolution of the software engineering methodologies used in past and present
-EU-funded software projects to illustrate how the research on software
+European-funded software projects to illustrate how the research on software
 reliability has progressed in Europe over the last 15 years following the
 foundation of the first project in 2001. In particular, we describe the team
-organizational structure, techniques and procedures that guided the Quality
-Assurance definitions to deliver quality software. We highlight the challenges
-and barriers faced throughout these years and how to overcome them totally or
+organizational structure, techniques and procedures that have guided the Quality
+Assurance definitions in order to deliver quality software. We highlight the challenges
+and barriers that have been faced throughout these years and how to overcome them totally or
 partially. To conclude, we discuss the future directions of software
@@
 \begin{IEEEkeywords}
-software reliability, quality assurance, software metrics, software testing
-techniques
+Software Reliability, Quality Assurance, Software Metrics, Software Testing
+Techniques
 \end{IEEEkeywords}
@@
 
-P. Orviz Fernandez,
-        M. David,
-		D. C. Duma,
-        J. Gomes,
-        D. Salomoni
-
 % if you will not have a photo at all:

### 2017-12-01 paper-ieee-trans.tex
added changes

@@
 evolution of the software engineering methodologies used in past and present
-European-funded software projects to illustrate how the research on software
+European Commission-funded software projects to illustrate how the research on software
 reliability has progressed in Europe over the last 15 years following the
@@
 The observed trends are aligned with the evolution of the software engineering
-techniques (e.g. rise of DevOps practices) over the last years. This is tied to
+techniques (e.g. rise of DevOps practices \cite{zhu}) over the last years. This is tied to
 the parallel advancements in the ICT field (with virtualization and container
-technologies) and the emergence of automation and event-response tools. In the
-following, we introduce a set of European Commission funded projects that cover
+technologies) and the emergence of automation and event-response tools.
+
+In this paper, we introduce a set of European Commission (EC)-funded projects that goes from
 2001 up to now, where the authors of this summary were and are currently
-involved. For each project we explain how they addressed the challenge in
-software reliability.
+involved. For each project, we explain how they addressed the challenge in
+software reliability. Furthermore, we detail new trends in software reliability.
 
-\section{Evolution of research on software reliability in EU-projects}
+The reminder of this paper is organized as follows.
+Section \ref{sec:ev} details the evolution of research on software reliability in EC-projects.
+Section \ref{sec:ntsr} provides information about new research trends in software reliability.
+Finally, Section \ref{sec:con} concludes.
+
+\section{Evolution of research on software reliability in EC-projects}
+\label{sec:ev}
 
@@
 
-\section{New Trends in Software Reliability}
+\section{New trends in software reliability}
+\label{sec:ntsr}
 
@@
 \section{Conclusion}
+\label{sec:con}
 
@@
 \begin{thebibliography}{1}
+
+\bibitem{zhu}
+L. Zhu and L. Bass and G. Champlin-Scharff, \emph{DevOps and Its Practices}, 
+IEEE Software, vol. 33, no. 3, pp. 32--34, May-June 2016.
 

### 2017-12-01 paper-ieee-trans.tex
added changes

@@
 \thanks{P. Orviz Fernandez, is with CSIC, Santander, Spain, (e-mail: orviz@ifca.unican.es)}
-\thanks{M. David, and J. Gomes are with LIP, Lisbon, Portugal}%
+\thanks{M. David, and J. Gomes are with LIP, Lisbon, Portugal.}%
 }
@@
 
-\IEEEPARstart{I}{n} the European scientific research arena, the analysis of
-some of the last decade software development projects depicts a continuous
-increase of the prominence and robustness of testing and validation procedures.
-The observed trends are aligned with the evolution of the software engineering
-techniques (e.g. rise of DevOps practices \cite{zhu}) over the last years. This is tied to
-the parallel advancements in the ICT field (with virtualization and container
-technologies) and the emergence of automation and event-response tools.
+\IEEEPARstart{A}{n}alyzing the European Commission-funded software projects in the last decade we have observed a continuous
+increase of the prominence and robustness of testing and validation procedures. The first of many reasons is related to the evolution of the software engineering
+techniques (e.g. rise of DevOps practices \cite{zhu}) together with the parallel advancements in the ICT field with virtualization and container
+technologies. Another one is the emergence of automation and event-response tools.
+
+Software developers involved in these projects have learnt over time how to make their product more reliable in order to satisfy requirements of different communities, such as physics and bio-informatics.
 

### 2017-12-01 paper-ieee-trans.tex

@@
 
-Software developers involved in these projects have learnt over time how to make their product more reliable in order to satisfy requirements of different communities, such as physics and bio-informatics.
+Software developers involved in these projects over time have learned  how to make their final products more reliable in order to satisfy requirements of different communities, such as physics.
 

### 2017-12-01 paper-ieee-trans.tex
added changes

@@
 \usepackage{graphicx}
+\usepackage{array}
+
 \ifCLASSINFOpdf
@@
 
+In the following we provide an overview of the main EC-projects that have dealt with the issue of software reliability.
+Table \ref{tab:eup} shows the various projects with their logo, short name, long name and duration.
+
+\begin{table}[!h]
+%% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{List of EC-projects}
+\label{tab:eup}
+\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
+\hline
+\hline
+\\
+Logo & Short Name & Long Name & Duration\\
+\hline
+\hline
+\begin{minipage}{.3\textwidth}
+\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
+\end{minipage}
+    & DataGrid &
+Research and Technological Development for an International Data GridTwo & 2001--2004\\
+\begin{minipage}{.3\textwidth}
+\includegraphics[width=15mm,height=7.5mm]{images/egee}
+\end{minipage}
+     & EGEE I, II, III &
+Enabling Grids for E-sciencE & 2004--2010\\
+\begin{minipage}{.3\textwidth}
+\includegraphics[width=15mm,height=7.5mm]{images/etics}
+\end{minipage}
+     & ETICS 1, 2 &
+E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
+\begin{minipage}{.3\textwidth}
+\includegraphics[width=15mm,height=7.5mm]{images/emi}
+\end{minipage}
+     & EMI &
+Europeean Middleware Initiative & 2010--2013\\
+\begin{minipage}{.3\textwidth}
+\includegraphics[width=15mm,height=7.5mm]{images/egi}
+\end{minipage}
+     & EGI-Inspire &
+Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
+ & 2010--2014\\
+\begin{minipage}{.3\textwidth}
+\includegraphics[width=15mm,height=7.5mm]{images/indigo}
+\end{minipage}
+     & INDIGO-DataCloud &
+INtegrating Distributed data Infrastructures for Global ExplOitation
+ & 2015--2017\\
+\hline
+\hline
+\end{tabular}
+\end{table}
+
+\subsection{DataGrid}
+
 The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted
@@
 
+\subsection{Enabling Grids for E--sciencE}
+
 The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010)
@@
 
+\subsection{E--Infrastructure for Testing, Integration and Configuration of Software}
+
 The E--Infrastructure for Testing, Integration and Configuration of Software
@@
 the Grids.
+
+\subsection{European Middleware Initiative}
 
@@
 
+\subsection{EGI--Integrated Sustainable Pan--European Infrastructure}
+
 The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for
@@
 by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).
+
+\subsection{INtegrating Distributed data Infrastructures for Global
+ExplOitation}
 
@@
 
-\begin{IEEEbiography}{D. Cristina Duma}
+\begin{IEEEbiography}{Doina Cristina Duma}
 Biography text here.

### 2017-12-19 paper-ieee-trans.tex
fixed fig issue

@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/devops.png}
-\caption{Continuous Delivery workflow for Docker images.}
-\label{fig:fig_CD}
-\end{figure*}
-
 \subsection{Enabling Grids for E--sciencE}
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker images.}
+\label{fig:fig_CD}
+\end{figure*}
+
 \subsection{Software Quality Procedures}
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
-\includegraphics[width=\textwidth]{images/confman.png}
+\includegraphics[width=0.45\textwidth, height=50mm]{images/confman.png}
 \caption{Trend line showing the adoption of Configuration Management tools throughout the project lifetime.}
 \label{fig:fig_confman}
-\end{figure*}
+\end{figure}
 
@@
 set of deployed INDIGO-DataCloud components or services.}
-\label{fig:fig_confman}
+\label{fig:fig_pilotpreview}
 \end{figure*}
@@
 
-\begin{IEEEbiography}{Mario David}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/mario-david}}]{Mario David}
+Mario David is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
+Member of DEEP-HybridDatacloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-Hub project and
+of the Portuguese National Computing Distributed Infrastructure.
+He held a research associate position at Institut de Physique du Globe de Paris (IPGP/CNRS) as Scientific Software Developer for the VERCE project, in particular in the data intensive use cases for seismology.
+He was been actively involved in the Validation, Quality Assurance and testing of middleware, in regional and global operations.
 \end{IEEEbiography}
@@
 
-\begin{IEEEbiography}{J. Gomes}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/jorge-gomes}}]{Jorge Gomes}
+Jorge Gomes is a computing researcher at LIP.
+He worked in the development of advanced data acquisition systems at CERN, and participated in pioneering projects in the domain of digital satellite data communications, IP over ATM, and advanced videoconferencing over IP networks.
+Since 2001 he has participated in numerous projects regarding distributed computing, networks and security in Europe and Latin America.
+He is the head of the LIP Advanced Computing and Digital Infrastructures Group and technical coordinator of the Portuguese National Grid Infrastructure, representative of Portugal in the Council of the European Grid Infrastructure (EGI) and responsible for the Portuguese participation in IBERGRID, that joins Portuguese and Spanish distributed computing infrastructures.
 \end{IEEEbiography}

### 2017-12-20 paper-ieee-trans.tex

@@
 is a Computer Science Engineer at INFN (Istituto Nazionale di Fisica Nucleare - National Institute of Nuclear Physics) CNAF.
-Her main area of interest include software engineering issues such as software quality and software prediction model.
-She is also interested in (big and open) data analysis by using predictive techniques coming from various theories.
-such as Machine Learning.
+She holds a PhD in Automation, Robotics and Bioengineering from the University of Pisa.
+Her main area of interest include software engineering issues, such as software quality and software prediction model.
+She is also interested in (big and open) data analysis by using techniques coming from various theories, such as Machine Learning.
 \end{IEEEbiography}

### 2017-12-21 paper-ieee-trans.tex

@@
 
-\begin{IEEEbiography}{Davide Salomoni}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/DavideSalomoni}}]{Davide Salomoni}
+Davide SAlomoni is Director of Technology at the Italian National Institute for Nuclear Physics (INFN).
+He has 27 years of international experience in private and public environments related to distributed computing and communication technologies.
+He currently leads the Software Development and Distributed Systems department at CNAF, the INFN National Center Dedicated to research and development on IT technologies, located in Bologna, Italy.
+He was Project Coordinator of INDIGO-DataCloud funded by the EC Horizon2020 that developed innovative open source computing and storage solutions.
 \end{IEEEbiography}

### 2017-12-21 paper-ieee-trans.tex
added changes

@@
 He currently leads the Software Development and Distributed Systems department at CNAF, the INFN National Center Dedicated to research and development on IT technologies, located in Bologna, Italy.
-He was Project Coordinator of INDIGO-DataCloud funded by the EC Horizon2020 that developed innovative open source computing and storage solutions.
+He was Project Coordinator of INDIGO-DataCloud, a project funded by the EC Horizon2020, that developed innovative open source computing and storage solutions.
 \end{IEEEbiography}

### 2017-12-21 paper-ieee-trans.tex
Pablo's bio

--- 2017-12-20 paper-ieee-trans.tex
@@
 % if you will not have a photo at all:
-\begin{IEEEbiography}{Pablo Orviz Fernandez}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/pablo-orviz}}]{Pablo Orviz Fernandez}
+is a computing researcher at CSIC (Consejo Superior de Investigaciones
+Cientificas), holding an MSc in Scientific Computing from the University of
+Cantabria. During the past years he has been devoted to software quality
+assurance tasks within different European advanced computing research projects
+such as INDIGO-DataCloud and EGI-Engage. Currently involved in EOSC-hub project
+and participating in the software quality work packages of DEEP-HybridDatacloud
+and eXtreme-DataCloud.
 \end{IEEEbiography}
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/mario-david}}]{Mario David}
-Mario David is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
-Member of DEEP-HybridDatacloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-Hub project and
+is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
+Member of DEEP-HybridDatacloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-hub project and
 of the Portuguese National Computing Distributed Infrastructure.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/jorge-gomes}}]{Jorge Gomes}
-Jorge Gomes is a computing researcher at LIP.
+is a computing researcher at LIP.
 He worked in the development of advanced data acquisition systems at CERN, and participated in pioneering projects in the domain of digital satellite data communications, IP over ATM, and advanced videoconferencing over IP networks.
@@
 
-\begin{IEEEbiography}{Davide Salomoni}
-Biography text here.
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/DavideSalomoni}}]{Davide Salomoni}
+is Director of Technology at the Italian National Institute for Nuclear Physics (INFN).
+He has 27 years of international experience in private and public environments related to distributed computing and communication technologies.
+He currently leads the Software Development and Distributed Systems department at CNAF, the INFN National Center dedicated to research and development on IT technologies, located in Bologna, Italy.
+He was Project Coordinator of INDIGO-DataCloud, a project funded by the EC Horizon2020, that developed innovative open source computing and storage solutions.
 \end{IEEEbiography}

### 2017-12-21 paper-ieee-trans.tex

@@
 
-\IEEEPARstart{A}{n}alyzing the European Commission-funded software projects in the last decade we have observed a continuous
-increase of the prominence and robustness of testing and validation procedures. The first of many reasons is related to the evolution of the software engineering
-techniques (e.g. rise of DevOps practices \cite{zhu}) together with the parallel advancements in the ICT field with virtualization and container
-technologies. Another one is the emergence of automation and event-response tools.
+\IEEEPARstart{W}hen e-Infrastructures in Europe were first conceived, new
+services to access distributed resources -- data, computing, instrumental --
+were initially considered. Major investments were then injected in order to
+develop distributed solutions that provided a seamless and transparent access
+to the resources, available for the researchers. One clear example is Worldwide
+Large Hadron Collider Computing Grid (WLCG) e-Infrastructure, powered by the
+Particle Physics scientific community, built up to satisfy the computing needs
+of processing, storing and analysing all the data coming from the Large Hadron
+Collider particle accelerator.
 
-Software developers involved in these projects over time have learned  how to make their final products more reliable in order to satisfy requirements of different communities, such as physics.
+Years of major developments followed, mainly focused on providing the new
+functionalities, hindering the adoption of software engineering methodologies.
+The software delivered was by then little tested, resulting in constant
+rollbacks and patches applied to production systems. Thereby, the stability and
+reliability of the e-Infrastructures was deeply compromised, affecting
+negatively their exploitation by the scientific communities.
 
-In this paper, we introduce a set of European Commission (EC)-funded projects that goes from
-2001 up to now, where the authors of this summary were and are currently
-involved. For each project, we explain how they addressed the challenge in
-software reliability. Furthermore, we detail new trends in software reliability.
+Soon it become apparent that a balance was needed between new features addition
+and the reliability of the software produced. Thus, a substantial part of the
+software lifecycle should be devoted to deal with the quality of software being
+delivered. At this point, evolving software engineering practices were
+considered and gradually adopted to define the quality procedures, team
+organization and pilot testbeds.
 
-The reminder of this paper is organized as follows.
-Section \ref{sec:ev} details the evolution of research on software reliability in EC-projects.
-Section \ref{sec:ntsr} provides information about new research trends in software reliability.
-Finally, Section \ref{sec:con} concludes.
+Analyzing the European Commission-funded software projects in the last decade
+we have observed a continuous increase of the prominence and robustness of
+testing and validation procedures. The first of many reasons is related to the
+evolution of the software engineering techniques (e.g. creation of the Agile
+Manifesto \cite{agile-manifesto}, rise of DevOps practices \cite{zhu}) together
+with the parallel advancements in the ICT field with virtualization and
+container technologies. Another one is the emergence of automation and
+event-response tools.
+
+In this paper, we introduce a set of European Commission (EC)-funded projects
+that goes from 2001 up to now, where the authors of this summary were involved.
+For each project, we describe the challenges arised, the methodologies used and
+the achievements towards software reliability. The reminder of this paper is
+organized as follows. Section \ref{sec:ev} details the evolution of research on
+software reliability in EC-projects. Section \ref{sec:ntsr} provides
+information about new research trends in software reliability. Finally,
+Section \ref{sec:con} concludes.
 

### 2017-12-21 paper-ieee-trans.tex

@@
 \usepackage{array}
+\usepackage{subfigure}
 
@@
 
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker images.}
+\label{fig:fig_CD}
+\end{figure*}
+
 EGI--InSPIRE did not
@@
 the quality checks (as part of the CI). 
-
 The steps defined within the CD
-pipeline differ whether the software is to be distributed in the form of Docker
-images or via operating system’s packages. In the latter case an extra
+pipeline differ whether the software is to be distributed via Docker
+images (see Figure \ref{fig:fig_CD}) or via operating system’s packages. In the latter case an extra
 validation step is added, consisting in the product’s deployment using a
@@
 itself, installing and configuring the product before uploading it to the
-DockerHub repository \cite{indigo-dockerhub}. Figure \ref{fig:fig_CD} shows
-the workflow followed by the products distributed as Docker images.
-
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/devops.png}
-\caption{Continuous Delivery workflow for Docker images.}
-\label{fig:fig_CD}
-\end{figure*}
+DockerHub repository \cite{indigo-dockerhub}.
 
@@
 selected from the variety of programming languages being used as seen in
-Fig. \ref{fig:fig_codestyle}.
+Figure \ref{fig:fig_codestyle}.
 
-\begin{figure}[!t]
+\begin{figure}[ht]
 \centering
@@
 between the code coverage for the products involved both in {\sl INDIGO-1} and
-{\sl INDIGO-2} releases is shown in Fig. \ref{fig:fig_unittest}. As it can be seen,
+{\sl INDIGO-2} releases is shown in Figure \ref{fig:fig_unittest}. As it can be seen,
 a very small set of products with low coverage values highly impacted the overall
@@
 
-\begin{figure*}
+\begin{figure*}[ht]
 \centering
@@
 
-Fig. \ref{fig:fig_confman} shows the number of products that offer an automated means
+Figure \ref{fig:fig_confman} shows the number of products that offer an automated means
 for deployment. It shows an increase after the {\sl INDIGO-1} release with specially
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \centering
@@
 \end{figure}
+
+\begin{figure*}	
+	\centering
+	\begin{subfigure}
+		\centering
+\includegraphics[width=\textwidth]{images/disvis.png}
+\caption{DevOps pipeline to distribute Docker images for Disvis application.}
+\label{fig:fig_disvis}		
+	\end{subfigure}
+	\quad
+	\begin{subfigure}
+		\centering
+\includegraphics[width=\textwidth]{images/pilotpreview.png}
+\caption{Resource centers supporting the Pilot Preview testbed and corresponding
+set of deployed INDIGO-DataCloud components or services.}
+\label{fig:fig_pilotpreview}
+	\end{subfigure}
+\end{figure*}
 
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/disvis.png}
-\caption{DevOps pipeline to distribute Docker images for Disvis application.}
-\label{fig:fig_disvis}
-\end{figure*}
-
 \subsection{Integration, preview and early adoption}
@@
 with the last stable version of the software. A map of the Pilot Preview
-infrastructure is depicted in Fig. \ref{fig:fig_pilotpreview}, it shows the
+infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
 resources providers and the components or services they deployed and supported.
-
-\begin{figure*}
-\centering
-\includegraphics[width=\textwidth]{images/pilotpreview.png}
-\caption{Resource centers supporting the Pilot Preview testbed and corresponding
-set of deployed INDIGO-DataCloud components or services.}
-\label{fig:fig_pilotpreview}
-\end{figure*}
 

### 2017-12-23 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 
-\begin{IEEEbiography}{Doina Cristina Duma}
-Biography text here.
+\begin{IEEEbiography}\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/cris_foto_good}}]{Doina Cristina Duma}
+is the leader of the Distributed Systems group at the INFN National Center (CNAF), providing the core operational support for the INFN-wide Grid infrastructure and for the CNAF Cloud infrastructure.
+She has a long experience in managing distributed e-infrastructures, being involved since 2003 in major European projects such as DataGrid, EGEE (I, II, III), EGI-Inspire and EGI-Engage.
+She was the Release Manager of the EMI (European Middleware Initiative) and INDIGO - DataCloud European project project and having as main responsibilities to manage all release activities, including release planning, building, deploying, reporting, establishing and maintaining the overall release timeline and milestones.
+She is now the leader of the WP3 – “Software Lifecycle Management, Pilot Infrastructures & Exploitation”
+of the eXtreme - DataCloud European project and leader of the WP3.2 - "Software quality assurance, release, maintenance and support" of the DEEP-HybridDataCloud.
 \end{IEEEbiography}

### 2017-12-23 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 of new tools and services to match the scientific community requirements has been
-addressed in Europe by the creation of software development and
-e-Infrastructure coordination dedicated projects. This work examines the
-evolution of the software engineering methodologies used in past and present
-European Commission-funded software projects to illustrate how the research on software
+addressed in Europe by a various number of European Commission funded software development
+and e-Infrastructures creation, operation and management dedicated projects.
+% the creation of software development and e-Infrastructure coordination dedicated projects. 
+This work examines the evolution of the software engineering methodologies used in past and present
+European Commission-funded projects to illustrate how the research on software
 reliability has progressed in Europe over the last 15 years following the
 foundation of the first project in 2001. In particular, we describe the team
-organizational structure, techniques and procedures that have guided the Quality
+organizational structure (?), techniques and procedures that have guided the Quality
 Assurance definitions in order to deliver quality software. We highlight the challenges
-and barriers that have been faced throughout these years and how to overcome them totally or
-partially. To conclude, we discuss the future directions of software
-Quality Assurance methodologies for the incoming projects.
+and barriers that have been faced throughout these years and how they were overcomed totally or
+partially. To conclude, we discuss possible future directions in software
+Quality Assurance methodologies for the upcoming projects.
 \end{abstract}

### 2017-12-23 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 develop distributed solutions that provided a seamless and transparent access
-to the resources, available for the researchers. One clear example is Worldwide
-Large Hadron Collider Computing Grid (WLCG) e-Infrastructure, powered by the
+to those resources, available for the researchers. One of the most known examples is Worldwide
+Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered by the
 Particle Physics scientific community, built up to satisfy the computing needs
-of processing, storing and analysing all the data coming from the Large Hadron
+of processing, storing and analysing the huge quantity of data coming from the Large Hadron
 Collider particle accelerator.
 
-Years of major developments followed, mainly focused on providing the new
+Years of major developments followed, mainly focused on providing new
 functionalities, hindering the adoption of software engineering methodologies.
-The software delivered was by then little tested, resulting in constant
+The software delivered was by then less tested, resulting in constant
 rollbacks and patches applied to production systems. Thereby, the stability and
-reliability of the e-Infrastructures was deeply compromised, affecting
+reliability of the e-Infrastructures was often compromised, affecting
 negatively their exploitation by the scientific communities.
@@
 Soon it become apparent that a balance was needed between new features addition
-and the reliability of the software produced. Thus, a substantial part of the
-software lifecycle should be devoted to deal with the quality of software being
+and the reliability of the software produced. As a consequence, a substantial part of the
+software lifecycle would have to be devoted to deal with the quality of software being
 delivered. At this point, evolving software engineering practices were

### 2017-12-24 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 foundation of the first project in 2001. In particular, we describe the team
-organizational structure (?), techniques and procedures that have guided the Quality
+organizational structure, %(what does it mean?)
+ techniques and procedures that have guided the Quality
 Assurance definitions in order to deliver quality software. We highlight the challenges
@@
 
-\IEEEPARstart{W}hen e-Infrastructures in Europe were first conceived, new
-services to access distributed resources -- data, computing, instrumental --
-were initially considered. Major investments were then injected in order to
-develop distributed solutions that provided a seamless and transparent access
-to those resources, available for the researchers. One of the most known examples is Worldwide
-Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered by the
-Particle Physics scientific community, built up to satisfy the computing needs
+%\IEEEPARstart{W}hen e-Infrastructures in Europe were first conceived, new
+%services to access distributed resources -- data, computing, instrumental --
+%were initially considered. Major investments were then injected in order to
+%develop distributed solutions that provided a seamless and transparent access
+%to those resources, available for the researchers. One of the most known examples is Worldwide
+%Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered by the
+%Particle Physics scientific community, built up to satisfy the computing needs
+%of processing, storing and analysing the huge quantity of data coming from the Large Hadron
+%Collider particle accelerator.
+
+\IEEEPARstart{D}uring the last 15 years the provision of a European e-Infrastructure has been driven by the 
+needs of user communities to access large scale computing and data analysis infrastructure to 
+support their research needs as part of their pan-European research collaborations. One of the most 
+known examples is Worldwide Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, 
+powered by the Particle Physics scientific community, built up to satisfy the computing needs
 of processing, storing and analysing the huge quantity of data coming from the Large Hadron
 Collider particle accelerator.
+As a consequence a number of important projects have been run in Europe, supported by the European Commission funding of various research programms (Framework Programms 5th, 6th, 7th, Horizon 2020) in order to
+develop distributed solutions that provided a seamless and transparent access to those resources, available for the researchers.
 
-Years of major developments followed, mainly focused on providing new
+Major developments were done throughout the years, mainly focused on providing new
 functionalities, hindering the adoption of software engineering methodologies.
@@
 we have observed a continuous increase of the prominence and robustness of
-testing and validation procedures. The first of many reasons is related to the
+testing and validation procedures. One of many reasons is related to the
 evolution of the software engineering techniques (e.g. creation of the Agile

### 2017-12-28 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 
-\IEEEPARstart{D}uring the last 15 years the provision of a European e-Infrastructure has been driven by the 
-needs of user communities to access large scale computing and data analysis infrastructure to 
-support their research needs as part of their pan-European research collaborations. One of the most 
-known examples is Worldwide Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, 
-powered by the Particle Physics scientific community, built up to satisfy the computing needs
-of processing, storing and analysing the huge quantity of data coming from the Large Hadron
-Collider particle accelerator.
-As a consequence a number of important projects have been run in Europe, supported by the European Commission funding of various research programms (Framework Programms 5th, 6th, 7th, Horizon 2020) in order to
-develop distributed solutions that provided a seamless and transparent access to those resources, available for the researchers.
+\IEEEPARstart{D}uring the last 15 years the provision of a European e-Infrastructure 
+has been driven by the needs of user communities to access large scale computing and 
+data analysis infrastructure to support their research needs as part of their 
+pan-European research collaborations. One of the most known examples is the Worldwide 
+Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered 
+by the Particle Physics scientific community, built up to satisfy the computing needs
+of processing, storing and analysing the huge quantity of data coming from the Large 
+Hadron Collider particle accelerator. As a consequence a number of important projects 
+have been run in Europe, supported by the European Commission funding of various 
+research programms (Framework Programms 5th, 6th, 7th, Horizon 2020) in order to
+develop distributed solutions that provided a seamless and transparent access to those 
+resources, available for the researchers.
 
 Major developments were done throughout the years, mainly focused on providing new
-functionalities, hindering the adoption of software engineering methodologies.
-The software delivered was by then less tested, resulting in constant
+functionalities, leaving secondly the adoption of software engineering methodologies.
+The software delivered was then less tested, resulting in constant
 rollbacks and patches applied to production systems. Thereby, the stability and
@@
 delivered. At this point, evolving software engineering practices were
-considered and gradually adopted to define the quality procedures, team
+considered and gradually adopted to define the quality procedures, teams
 organization and pilot testbeds.
@@
 In this paper, we introduce a set of European Commission (EC)-funded projects
-that goes from 2001 up to now, where the authors of this summary were involved.
+that goes from 2001 up to now, in which the authors of this summary were involved.
 For each project, we describe the challenges arised, the methodologies used and
@@
     & DataGrid &
-Research and Technological Development for an International Data GridTwo & 2001--2004\\
+Research and Technological Development for an International Data Grid & 2001--2004\\
 \begin{minipage}{.3\textwidth}
@@
 \begin{minipage}{.3\textwidth}
-\includegraphics[width=15mm,height=7.5mm]{images/egi}
+\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
 \end{minipage}
@@
  & 2010--2014\\
+\begin{minipage}{.3\textwidth}
+\includegraphics[width=15mm,height=7.5mm]{images/egi-engage}
+\end{minipage}
+     & EGI Engage &
+Engaging the EGI Community towards an Open Science Commons
+ & 2015--2017\\
 \begin{minipage}{.3\textwidth}
@@
 
-The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) was devoted
-to develop the first European infrastructure for Grid computing. The project
-brought together 21 academic and industry partners, from 15 different
-countries. There were 3 application areas for requirement gathering;
-Particle physics, Bioinformatics and Earth observation. The project delivered
-their own software distribution, named EDG (EU DataGrid), strongly based on
-Globus Middleware Services \cite{globus}. The development organization
-consisted in 5 work packages coordinated by an Architecture Task Force,
-supervising the overall design and technical consistency of the developments.
+The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
+had as main goal to develop the software to provide basic grid functionality
+and associated management tools for a large scale testbed for demonstration projects in three
+specific areas of science, Particle physics, Bioinformatics and Earth observation
+%was devoted to develop the first European infrastructure for Grid computing. 
+The project brought together 21 academic and industry partners, from 15 different
+countries. 
+%There were 3 application areas for requirement gathering; Particle physics, Bioinformatics and Earth observation. 
+The project delivered its own software distribution, named EDG (EU DataGrid), strongly 
+based on Globus Middleware Services \cite{globus}. The project was organized in 12 workpages, from 
+which 5 were devoted to software development coordinated by an 
+Architecture Task Force, supervising the overall design and technical consistency 
+of the developments.
 
@@
 development tasks, 2) the evolution of the requirements coming from the user
-communities, and 3) the lack of a body of knowledge for academic software
+communities (50 use cases from the three scientific area), and 3) the lack of a body of knowledge for academic software
 engineering.
@@
 \cite{agile}. Nonetheless, the project focused on the experiences and procedures
-from open source software projects, such as Linux and Apache, trying to reach a
+used by different open source software projects, such as Linux and Apache, trying to reach a
 higher maturity level \cite{cmm}.
@@
 The {\sl gLite} middleware \cite{glite} was the ultimate
-official distribution of EGEE as of 2006, after two years of prototyping and
+official software distribution of EGEE as of 2006, after two years of prototyping and
 re--engineering efforts to converge with LHC Computing Grid (LCG--2), Virtual
@@
 development team was comprised of more that 80 people from 12 academic and
-industrial partners, that issued more than 10 thousand bugs, 1.7 thousand patches and
-defined over 300 tasks tracked by the use of bug/task management tools.
+industrial partners, that issued more than 10 thousand bug fixes, 1.7 thousand patches and
+defined over 300 development tasks tracked by the use of bug/task management tools.
 
-The
-source code was available at a private, centralized version control system.
+The source code was available at a private, centralized version control system.
 The code was not validated by any Quality Assurance policy, but passed through a manual
@@
 improve the reliability of the software components by applying a set of
-requirements [ref] to be fulfilled at each stage (integration, certification,
-pre--production, production), carried out by different teams. However, starting in
-EGEE--II, the project adopted the ETICS solution, aiming to develop
+requirements [ref] to be fulfilled at each stage of the  development lifecycle (integration, certification,
+pre--production, production), carried out by different teams. However, with 
+EGEE--II, the developers started to use the the ETICS system [ref], solution aiming to develop
 an automatic build system for Grid middleware.

### 2017-12-28 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 collaborative projects such as EGEE and its gLite middleware. The framework
-integrated different technologies and tools to provide automated configuration,
+integrated different technologies and tools in order to provide automated configuration,
 build and testing capabilities, as well as auto-generated documentation and
@@
 number of defects/bugs \cite{etics}. The ETICS portal was the first automated
-service for delivering software products in distributed environments like
+service for delivering quality software products in distributed environments like
 the Grids.
@@
 Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
-to maintain and evolve the middleware focusing on extending the
-interoperability in the Grid and improving the reliability of the services. The
+with the goal to maintain and evolve the middleware focusing on extending their
+interoperability and improving the reliability of the services. The
 ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
@@
 detail in the EMI Metrics Specification \cite{emi-metrics}. The project
-leveraged the {\sl ETICS} service for the development and release management,
+leveraged the {\sl ETICS} service for the development, continuous integration and release management,
 as well as for metric tracking, making queries on the collected data to display
@@
 Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
-arose as the continuation of the EGEE--III project, establishing and
+arose as the continuation of the EGEE--III project, with the mission of establishing and
 maintaining a sustainable European Grid Infrastructure.
@@
 
-EGI--InSPIRE did not
-develop any software but maintained a production--ready Grid computing
-middleware distribution called {\sl UMD} (Unified Middleware Distribution).
-The role of {\sl UMD} was to enforce the fulfilment of a set of quality
-criteria definitions \cite{egi-qc} in the software being delivered (i.e.,
+EGI--InSPIRE not only ensured the coordination of the operation of a secure, reliable 
+European-wide production grid infrastructure federated from national grid initiatives 
+that is integrated and interoperates with other grids worldwide but also foreseed the 
+maintenance and development of operational tools, like the Operations Portal [ref], 
+the EGI Helpdesk [ref] or Grid configuration repository (GOCDB) [ref] and the 
+coordination of the of  the external provision of software required by EGI to 
+form the production infrastructure. For the later it maintained a production--ready 
+Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution) and 
+defined general and component specific quality criteria to be applied to software 
+components made available through this repository.
+The role of {\sl UMD} was to enforce the fulfilment of this set of quality
+criteria \cite{egi-qc} in the software being delivered (i.e.,
 {\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
@@
 highly influenced by the insights of current big worldwide collaborations of
-software development. These collaborations have continuously inspired new
+software development. These collaborations have continuously inspired the use new
 technologies and procedures to increase the robustness and quality of the
-software being delivered. In this scenario, the DevOps culture is one of the
-major outcomes that has been progressively adopted in the INDIGO--DataCloud
+software being delivered. In this scenario, the implementation of the DevOps culture 
+is one of the major outcomes, as this paradigm has been progressively adopted by
 project.

### 2017-12-28 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 
-In the following we provide an overview of the main EC-projects that have dealt with the issue of software reliability.
-Table \ref{tab:eup} shows the various projects with their logo, short name, long name and duration.
+In the following we provide an overview of the main EC-projects that have dealt with 
+the issue of software reliability. Table \ref{tab:eup} shows the various projects 
+with their logo, short name, long name and duration.
 
@@
 \subsection{INtegrating Distributed data Infrastructures for Global
-ExplOitation}
+explOitation}
 
 The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
-ExplOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
+explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
 project of software development considered in this review. At the time of writing
@@
 quality and reliability aspects throughout the past scientific European
-experiences. Nevertheless, the project’s Quality Assurance procedures are also
+experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
 highly influenced by the insights of current big worldwide collaborations of
-software development. These collaborations have continuously inspired the use new
+software development. These collaborations have continuously inspired the use of new
 technologies and procedures to increase the robustness and quality of the
 software being delivered. In this scenario, the implementation of the DevOps culture 
-is one of the major outcomes, as this paradigm has been progressively adopted by
+is one of the major outcomes, as this paradigm has been progressively adopted by the
 project.
@@
 avoid infrastructure disruption whenever new developments are deployed into
-production. The INDIGO--DataCloud project promoted the application of a Continuous Integration (CI)
-scenario to enforce the Quality Assurance requirements for any piece of
-software being produced within the project. Such environment requires an
-automated ready--to--go infrastructure where the different services involved
+production. The INDIGO--DataCloud project promoted the application of a 
+Continuous Integration (CI) scenario to enforce the Quality Assurance requirements 
+for any piece of software being produced within the project. Such environment 
+requires an automation ready--to--go infrastructure where the different services involved
 (source code management platform, automation server) interact with each other
@@
 with their outputs. To accomplish this, the project leveraged on tightly
-integrated open source tools such as GitHub and Jenkins, as well as relying on
+integrated open source tools such as GitHub [ref], as source code repository
+and Jenkins [ref], as continuous integration system, as well as relying on
 Docker container provisioning to speed up the source code validation. This CI
@@
 
-As DevOps suggests, frequent releases positively affects the reliability of the
+As DevOps suggests, frequent releases positively affect the reliability of the
 software. The software updates of INDIGO--DataCloud products taking place since
 the second major release are passing through a Continuous Delivery (CD)
-pipeline that adds the packaging of the software right after the execution of
+pipeline that adds the packaging of the software right after the successful execution of
 the quality checks (as part of the CI). 
-The steps defined within the CD
-pipeline differ whether the software is to be distributed via Docker
-images (see Figure \ref{fig:fig_CD}) or via operating system’s packages. In the latter case an extra
+The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
+images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
 validation step is added, consisting in the product’s deployment using a
@@
 
-The software quality procedures \cite{indigo-d31} cover: 1) the identification
+The software quality policies \cite{indigo-d31} cover: 1) the identification
 and description of the \emph{quality requirements} that the software need to
@@
 
-The requirements are applied at early stages in the software development
+The requirements are checked at early stages in the software development
 process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
 bug or design issue is likely to be detected and corrected promptly. The source
-code is publicly available in GitHub repositories under an organization called
-{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the product
+code is publicly available in GitHub repositories under the project organization called
+{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
 catalogue, promoting the external contributions and software adoption. Furthermore,
-the source code is compliant with community de--facto style standards,
-selected from the variety of programming languages being used as seen in
+the source code is checked for compliance with community de--facto style standards,
+selected for each one of the variety of programming languages being used as seen in
 Figure \ref{fig:fig_codestyle}.
@@
 
-Unit testing coverage has increased between the first ({\sl INDIGO-1}) and the
+Unit testing coverage, another project quality metric, has increased between the first ({\sl INDIGO-1}) and the
 second ({\sl INDIGO-2}) INDIGO--DataCloud releases, reaching an average value of
@@
 or products with codebases not fully under the control of the project. Nevertheless,
-60\% of the products were over this threshold while 80\% of them have values above
+60\% of the products were over this threshold while 80\% of them had values above
 the 50\% coverage.
@@
 human--reviewed as the last step before merging them into the production
-repository.
+repository or official documentation.
 
@@
 automatically using either Ansible \cite{indigo-ansible} or Puppet
-\cite{indigo-puppet} open--source tools. The adoption of such Configuration Management
+\cite{indigo-puppet} open--source configuration tools. The adoption of such Configuration Management
 tools not only eases the consumption of the products developed by external users or
@@
 services and applications. A representative example is the number of roles currently
-hosted in the Ansible Galaxy portal for the project.
+hosted in the Ansible Galaxy portal for the project [ref].
 
@@
 Figure \ref{fig:fig_confman} shows the number of products that offer an automated means
-for deployment. It shows an increase after the {\sl INDIGO-1} release with specially
+for deployment. It shows an increase after the {\sl INDIGO-1} release especially
 in the previous weeks of each of the two major releases.
@@
 
-Two applications from WP2, DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
+Two applications from "Definition of Support to Research Communities" workpackage, 
+DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
 integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
@@
 The novelty introduced in the pipeline above is the validation of the application.
-Once the application is deployed as a Docker container, and subsequently uploaded
+Once the application is packaged as a Docker image, and subsequently uploaded
 to the Dockerhub repository, it is instantiated in a new container to be validated.
@@
 Thus this pipeline implementation goes a step forward by testing the application
-execution for the last available Docker image in the catalogue.
+execution for the latest available Docker image in the catalogue.
 
@@
 level of integration between the components involved in the INDIGO--DataCloud
-solution and use cases validation, by deploying and executing the applications
-with the last stable version of the software. A map of the Pilot Preview
+solutions and use cases validation, by deploying and executing the applications
+with the last stable version of the software in environments as similar as 
+possible to production ones. A map of the Pilot Preview
 infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
@@
 
-The released software is also tested in production environments through the
-stage--rollout process. Selected resource providers are requested to install
-the most updated stable versions of the software, accessed by their users. The
-stage--rollout process is key to detect and mitigate issues that could only
-appear in production.
+The released software is tested in production environments through the
+Staged--Rollout process, before inclusion in the EGI's CloudManagement Distribution (CMD). 
+Selected resource providers are requested to install
+the most updated stable versions of the software, and to give access to them to their users. The
+Staged--Rollout process is key to detect and mitigate issues that could only
+appear in production environments.
 
@@
 software is being prioritized. Recent software engineering insights, such as
-DevOps, are gradually being applied as the open--source collaborative tools
-evolve, allowing tighter integrations. The focus of the Software Quality
-Assurance procedures has been integrated in the development stage, trying to
+DevOps approaches, are gradually being applied as the open--source collaborative tools
+evolve, allowing tighter integrations. The Software Quality
+Assurance procedures have been integrated in the development stage, in order to
 detect and correct issues early in the software lifecycle. Automation and
-metric analysis are at the base of prompt issue solving. However, post--release
-validation, such as integration testbeds and the stage--rollout process, are
+metrics analysis are at the base of prompt issue solving. However, post--release
+validation, through the preview testbeds or the Staged--rollout process, are
 also needed to strengthen the software reliability for scientific usage.

### 2017-12-28 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 development tasks, 2) the evolution of the requirements coming from the user
-communities (50 use cases from the three scientific area), and 3) the lack of a body of knowledge for academic software
-engineering.
+communities (50 use cases from the three scientific area), and 3) the lack of a body 
+of knowledge for academic software engineering.
 
@@
 pre--production, production), carried out by different teams. However, with 
-EGEE--II, the developers started to use the the ETICS system [ref], solution aiming to develop
+EGEE--II, the developers started to use the the ETICS system \cite{etics}, solution aiming to develop
 an automatic build system for Grid middleware.
@@
 software metrics gathering such as Source Lines Of Code (SLOC), complexity and
-number of defects/bugs \cite{etics}. The ETICS portal was the first automated
+number of defects/bugs \cite{etics}. The ETICS framework was the first automated
 service for delivering quality software products in distributed environments like
@@
 that is integrated and interoperates with other grids worldwide but also foreseed the 
-maintenance and development of operational tools, like the Operations Portal [ref], 
-the EGI Helpdesk [ref] or Grid configuration repository (GOCDB) [ref] and the 
+maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
+the EGI Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
 coordination of the of  the external provision of software required by EGI to 
@@
 with their outputs. To accomplish this, the project leveraged on tightly
-integrated open source tools such as GitHub [ref], as source code repository
-and Jenkins [ref], as continuous integration system, as well as relying on
+integrated open source tools such as GitHub \cite{github}, as source code repository
+and Jenkins \cite{jenkins}, as continuous integration system, as well as relying on
 Docker container provisioning to speed up the source code validation. This CI
@@
 
+\bibitem{egi-ops}
+\emph{EGI Operations Portal},
+https://operations-portal.egi.eu/
+
+\bibitem{ggus}
+\emph{GGUS} EGI Helpdesk,
+https://ggus.eu/
+
+\bibitem{gocdb}
+\emph{GOCDB} EGI Grid Configuration Repository,
+https://ggus.eu/
+
+\bibitem{github}
+\emph{GITHUB},
+https://github.com/
+
+\bibitem{jenkins}
+\emph{Jenkins} Continuous Integration System,
+https://jenkins-ci.org/
+
 \bibitem{egi-qc}

### 2017-12-29 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 \begin{minipage}{.3\textwidth}
-\includegraphics[width=15mm,height=7.5mm]{images/egi-engage}
+\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
 \end{minipage}
@@
 She was the Release Manager of the EMI (European Middleware Initiative) and INDIGO - DataCloud European project project and having as main responsibilities to manage all release activities, including release planning, building, deploying, reporting, establishing and maintaining the overall release timeline and milestones.
-She is now the leader of the WP3 – “Software Lifecycle Management, Pilot Infrastructures & Exploitation”
-of the eXtreme - DataCloud European project and leader of the WP3.2 - "Software quality assurance, release, maintenance and support" of the DEEP-HybridDataCloud.
 \end{IEEEbiography}

### 2018-01-03 paper-ieee-trans.tex

@@
 
-\begin{IEEEbiography}\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/cris_foto_good}}]{Doina Cristina Duma}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/cris_foto_good}}]{Doina Cristina Duma}
 is the leader of the Distributed Systems group at the INFN National Center (CNAF), providing the core operational support for the INFN-wide Grid infrastructure and for the CNAF Cloud infrastructure.

### 2018-01-09 paper-ieee-trans.tex

@@
 and associated management tools for a large scale testbed for demonstration projects in three
-specific areas of science, Particle physics, Bioinformatics and Earth observation
+specific areas of science, such as Particle physics, Bioinformatics and Earth observation.
 %was devoted to develop the first European infrastructure for Grid computing. 
@@
 pre--production, production), carried out by different teams. However, with 
-EGEE--II, the developers started to use the the ETICS system \cite{etics}, solution aiming to develop
+EGEE--II, the developers started to use the ETICS system \cite{etics}, solution aiming to develop
 an automatic build system for Grid middleware.
@@
 
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=\textwidth]{images/devops.png}
-\caption{Continuous Delivery workflow for Docker images.}
-\label{fig:fig_CD}
-\end{figure*}
-
 EGI--InSPIRE not only ensured the coordination of the operation of a secure, reliable 
@@
 the EGI Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
-coordination of the of  the external provision of software required by EGI to 
+coordination of the external provision of software required by EGI to 
 form the production infrastructure. For the later it maintained a production--ready 
@@
 Cloud computing benefits.
+
+\subsection{Features on software reliability in EC-projects}
+
+In the following we summarize the features on software reliability in the aforementioned EC-projects.
+Table \ref{tab:feat} shows the various projects 
+with their features: the symbol $\surd$ crosses project and feature.
+
+\begin{table*}[!h]
+%% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{Features on software reliability in EC-projects}
+\label{tab:feat}
+\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+\begin{tabular}{llllllll}
+\hline
+\hline
+Feature & DataGrid & EGEE I, II, III & ETICS 1, 2 & EMI & EGI-Inspire & EGI Engage & INDIGO-DataCloud\\
+\hline
+\hline
+Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Agile Methodologies&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Buu/Test Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&&\\
+Automatic Certification Tool&&&$\surd$&&$\surd$&$\surd$&$\surd$\\
+Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Software Product Metrics&$\surd$&$\surd$&$\surd$&&$\surd$&$\surd$&$\surd$\\
+Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+\hline
+\hline
+\end{tabular}
+\end{table*}
 
@@
 DockerHub repository \cite{indigo-dockerhub}.
+
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker images.}
+\label{fig:fig_CD}
+\end{figure*}
 

### 2018-01-15 paper-ieee-trans.tex
Added reference to EGEE's sw lifecycle

@@
 The source code was available at a private, centralized version control system.
-The code was not validated by any Quality Assurance policy, but passed through a manual
-certification procedure at the time of the release. This procedure tried to
-improve the reliability of the software components by applying a set of
-requirements [ref] to be fulfilled at each stage of the  development lifecycle (integration, certification,
-pre--production, production), carried out by different teams. However, with 
-EGEE--II, the developers started to use the ETICS system \cite{etics}, solution aiming to develop
-an automatic build system for Grid middleware.
+The code was passed through a manual certification procedure at the time of the release. 
+This procedure tried to improve the reliability of the software components by applying an
+acceptance criteria plan at each stage of the software release process 
+\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
+production. Starting with EGEE--II, the project adopted automation in the software
+lifecycle process by leveraging the automatic build system for Grid middleware: ETICS 
+\cite{etics} solution.
 
@@
 
+\bibitem{egee:acceptance-criteria}
+\emph{Definition and Documentation of the Revised Software Life-Cycle Process},
+Milestone MSA3.4.2, 2010, EGEE-III project,
+https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1\_4.pdf
+
 \bibitem{glite}

### 2018-01-15 paper-ieee-trans.tex
Added indigo-dc roles in Ansible Galaxy reference as footnote

--- 2018-01-09 paper-ieee-trans.tex
@@
 infrastructures, but also tackles the correct installation and configuration of the
-services and applications. A representative example is the number of roles currently
-hosted in the Ansible Galaxy portal for the project [ref].
+services and applications. A representative example is the \textit{50 roles} being 
+developed within the project and currently hosted in the Ansible Galaxy portal for 
+the project \footnote{https://galaxy.ansible.com/indigo-dc/}.
 

### 2018-01-15 paper-ieee-trans.tex
fix typo

--- 2018-01-15 paper-ieee-trans.tex
@@
 Agile Methodologies&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Buu/Test Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Build/Test Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
 Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&&\\
@@
 infrastructures, but also tackles the correct installation and configuration of the
-services and applications. A representative example is the \textit{50 roles} being 
-developed within the project and currently hosted in the Ansible Galaxy portal for 
-the project \footnote{https://galaxy.ansible.com/indigo-dc/}.
+services and applications. A representative example is the number of roles currently
+hosted in the Ansible Galaxy portal for the project [ref].
 

### 2018-01-15 paper-ieee-trans.tex
disvis image
disvis devops workflow

@@
 		\centering
-\includegraphics[width=\textwidth]{images/disvis.png}
+\includegraphics[width=0.85\textwidth]{images/disvis-flow.png}
 \caption{DevOps pipeline to distribute Docker images for Disvis application.}
@@
 http://cordis.europa.eu/project/rcn/87264\_en.html
+
+\bibitem{egee:acceptance-criteria}
+\emph{Definition and Documentation of the Revised Software Life-Cycle Process},
+Milestone MSA3.4.2, 2010, EGEE-III project,
+https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1\_4.pdf
 

### 2018-01-15 paper-ieee-trans.tex
Table II modified and intro of section II rephrased.

@@
 \usepackage{subfigure}
+\usepackage{adjustbox}
 
@@
 
-In the following we provide an overview of the main EC-projects that have dealt with 
-the issue of software reliability. Table \ref{tab:eup} shows the various projects 
-with their logo, short name, long name and duration.
+The EC projects considered in this text are listed chronologically in Table
+\ref{tab:eup}. In most cases, each project paved the way of the subsequent projects,
+following a logic evolution in which the quality and reliability of software have
+increasingly had a more significant role. Experiences gathered during the 
+course of each project support this fact and, as it is shown in Table \ref{tab:feat},
+latest projects are the most committed in terms of software reliability awareness.
+This table summarizes the topics addressed in the following sections, highlighting 
+the commonalities and specific features on software reliability.
 
@@
 
-\subsection{Features on software reliability in EC-projects}
-
-In the following we summarize the features on software reliability in the aforementioned EC-projects.
-Table \ref{tab:feat} shows the various projects 
-with their features: the symbol $\surd$ crosses project and feature.
-
-\begin{table*}[!h]
+\begin{table}[!h]
 %% increase table row spacing, adjust to taste
@@
 %% than the plain LaTeX2e tabular which is used here.
+\begin{adjustbox}{max width=0.5\textwidth}
 \begin{tabular}{llllllll}
@@
 \hline
-Feature & DataGrid & EGEE I, II, III & ETICS 1, 2 & EMI & EGI-Inspire & EGI Engage & INDIGO-DataCloud\\
+Feature & DataGrid & EGEE I, II, III & ETICS 1, 2 & EMI & EGI-Inspire \& Engage & INDIGO-DataCloud\\
 \hline
 \hline
-Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Agile Methodologies&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Build/Test Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&&\\
-Automatic Certification Tool&&&$\surd$&&$\surd$&$\surd$&$\surd$\\
-Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Software Product Metrics&$\surd$&$\surd$&$\surd$&&$\surd$&$\surd$&$\surd$\\
-Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Source code inspection&&&&$\surd$&&$\surd$\\
+Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Software Product Metrics&$\surd$&$\surd$&$\surd$&&$\surd$&$\surd$\\
+Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
+%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
+Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
+Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
 \hline
@@
 \end{tabular}
-\end{table*}
+\end{adjustbox}
+\end{table}
 

### 2018-01-16 paper-ieee-trans.tex
Added Cris comments

@@
 \hline
-Feature & DataGrid & EGEE I, II, III & ETICS 1, 2 & EMI & EGI-Inspire \& Engage & INDIGO-DataCloud\\
+Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO-DC\\
 \hline
@@
 Source code inspection&&&&$\surd$&&$\surd$\\
-Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Software Product Metrics&$\surd$&$\surd$&$\surd$&&$\surd$&$\surd$\\
+Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
+Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
 Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\

### 2018-01-17 paper-ieee-trans.tex
Rephrase last paragraph of the abstract

@@
 of new tools and services to match the scientific community requirements has been
-addressed in Europe by a various number of European Commission funded software development
-and e-Infrastructures creation, operation and management dedicated projects.
+addressed in Europe by a various number of projects dedicated to software development
+and e-Infrastructure creation, operation and management.
 % the creation of software development and e-Infrastructure coordination dedicated projects. 
-This work examines the evolution of the software engineering methodologies used in past and present
-European Commission-funded projects to illustrate how the research on software
-reliability has progressed in Europe over the last 15 years following the
-foundation of the first project in 2001. In particular, we describe the team
-organizational structure, %(what does it mean?)
- techniques and procedures that have guided the Quality
-Assurance definitions in order to deliver quality software. We highlight the challenges
-and barriers that have been faced throughout these years and how they were overcomed totally or
-partially. To conclude, we discuss possible future directions in software
-Quality Assurance methodologies for the upcoming projects.
+This work examines the evolution of the software engineering methodologies used in past and
+present European Commission-funded projects to illustrate how the research on software
+reliability has progressed in Europe over the last 15 years following the foundation of the
+first project in 2001. In particular, we will summarize the techniques and procedures
+applied to deliver quality software, hightlighting the main challenges and barriers confronted
+at the time, and how they were totally or partially overcomed. The base of knowledge collected
+throughout these years, sustained by the advances in the area of software engineering,
+definitely conformed a significant breakthrough in the reliability of software delivered in the
+European research arena. Latest software development projects founded by the European 
+Commission are good evidences of such insights, where the enforcement of Quality Assurance
+procedures is present since the very early stages of the software development process.
 \end{abstract}

### 2018-01-18 paper-ieee-trans.tex
Rephrase and complete introduction

@@
 
-\IEEEPARstart{D}uring the last 15 years the provision of a European e-Infrastructure 
-has been driven by the needs of user communities to access large scale computing and 
-data analysis infrastructure to support their research needs as part of their 
-pan-European research collaborations. One of the most known examples is the Worldwide 
+\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
+that supports an unified access to large-scale computing and intense data analysis
+applications, has been driven by the needs of scientific communities as part of their
+pan-European research collaborations. One of the well-known examples is the Worldwide 
 Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered 
@@
 of processing, storing and analysing the huge quantity of data coming from the Large 
-Hadron Collider particle accelerator. As a consequence a number of important projects 
-have been run in Europe, supported by the European Commission funding of various 
-research programms (Framework Programms 5th, 6th, 7th, Horizon 2020) in order to
-develop distributed solutions that provided a seamless and transparent access to those 
-resources, available for the researchers.
+Hadron Collider particle accelerator. A significant number of innovation projects and
+technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
+and ultimately by the Horizon 2020 programme \footnote{https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
+develop distributed solutions that provide a seamless and transparent access to those
+computing and data resources, available for the researchers.
 
 Major developments were done throughout the years, mainly focused on providing new
-functionalities, leaving secondly the adoption of software engineering methodologies.
+functionalities, leaving aside at early times the adoption of software engineering methodologies.
 The software delivered was then less tested, resulting in constant
@@
 
-Analyzing the European Commission-funded software projects in the last decade
+Analyzing the European Commission (EC)-funded software projects in the last decade
 we have observed a continuous increase of the prominence and robustness of
 testing and validation procedures. One of many reasons is related to the
-evolution of the software engineering techniques (e.g. creation of the Agile
+evolution of the software engineering methodologies (e.g. creation of the Agile
 Manifesto \cite{agile-manifesto}, rise of DevOps practices \cite{zhu}) together
-with the parallel advancements in the ICT field with virtualization and
-container technologies. Another one is the emergence of automation and
-event-response tools.
+with the parallel advancements in the information and communications technology (ICT)
+field such as virtualization, enabling technology for Cloud computing and 
+operating--system--level virtualization known as containerization. All of which
+complemented by the emergence of automation and event-response tools.
 
-In this paper, we introduce a set of European Commission (EC)-funded projects
-that goes from 2001 up to now, in which the authors of this summary were involved.
+In this paper, we introduce a set of EC-funded projects
+that go from 2001 up to now, in which the authors of this summary were involved.
 For each project, we describe the challenges arised, the methodologies used and
@@
 software reliability in EC-projects. Section \ref{sec:ntsr} provides
-information about new research trends in software reliability. Finally,
-Section \ref{sec:con} concludes.
+information about new research trends in software reliability based upon the 
+experiences from the recent INDIGO-DataCloud project. Finally, Section \ref{sec:con} concludes.
 

### 2018-01-18 paper-ieee-trans.tex
syntax corrections and rephrasing new trends section

@@
 \begin{abstract}
-From the advent of Grid technology as the new paradigm of distributed
-computing to the current days of Cloud computing models, the continuous need
+From the advent of Grid technology -- as the new paradigm of distributed
+computing -- to the current days of Cloud computing models, the continuous need
 of new tools and services to match the scientific community requirements has been
@@
 \IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
-that supports an unified access to large-scale computing and intense data analysis
+which supports an unified access to large-scale computing and intense data analysis
 applications, has been driven by the needs of scientific communities as part of their
@@
 technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
-and ultimately by the Horizon 2020 programme \footnote{https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
+and ultimately by the Horizon 2020 programme\footnote{https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
 develop distributed solutions that provide a seamless and transparent access to those
@@
 
+\begin{table}[!h]
+%% increase table row spacing, adjust to taste
+\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+\caption{Features on software reliability in EC-projects}
+\label{tab:feat}
+\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+\begin{adjustbox}{max width=0.5\textwidth}
+\begin{tabular}{llllllll}
+\hline
+\hline
+Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO-DC\\
+\hline
+\hline
+Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Source code inspection&&&&$\surd$&&$\surd$\\
+Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
+Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
+%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
+Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
+Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
+\hline
+\hline
+\end{tabular}
+\end{adjustbox}
+\end{table}
+
 \subsection{Enabling Grids for E--sciencE}
@@
 
-\begin{table}[!h]
-%% increase table row spacing, adjust to taste
-\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-\caption{Features on software reliability in EC-projects}
-\label{tab:feat}
-\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-\begin{adjustbox}{max width=0.5\textwidth}
-\begin{tabular}{llllllll}
-\hline
-\hline
-Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO-DC\\
-\hline
-\hline
-Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Source code inspection&&&&$\surd$&&$\surd$\\
-Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
-Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
-%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
-Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
-Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
-\hline
-\hline
-\end{tabular}
-\end{adjustbox}
-\end{table}
-
 \section{New trends in software reliability}
 \label{sec:ntsr}
+
+The INDIGO--DataCloud project is the most recent project in the list and, as such, it is 
+the result of the lessons learned from previous experiences in terms of the evaluation and application
+of innovative software engineering methodologies, the management of new technologies to put
+them into practice, and the compilation and analysis of metrics to measure the success of the
+solutions implemented. The expertise gathered is complemented by the fact of being framed in
+collaborative projects with an extensive list of partners.
+
+framed projects
+
+All of this with the added complexity of being framed in major
+collaborations with
+
+procollecting and analyzing metrics relevant for the .. to empower the sw reliability bla bla bla
 

### 2018-01-19 paper-ieee-trans.tex

@@
 pan-European research collaborations. One of the well-known examples is the Worldwide 
-Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered 
-by the Particle Physics scientific community, built up to satisfy the computing needs
+Large Hadron Collider (LHC) Computing Grid (WLCG) global computing infrastructure, powered 
+by the particle physics scientific community, built up to satisfy the computing needs
 of processing, storing and analysing the huge quantity of data coming from the Large 
@@
 technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
-and ultimately by the Horizon 2020 programme\footnote{https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
+and ultimately by the Horizon 2020 programme\footnote{HORIZON 2020 - The EU Framework
+Programme for Research and Innovation https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
 develop distributed solutions that provide a seamless and transparent access to those
@@
 testing and validation procedures. One of many reasons is related to the
-evolution of the software engineering methodologies (e.g. creation of the Agile
-Manifesto \cite{agile-manifesto}, rise of DevOps practices \cite{zhu}) together
-with the parallel advancements in the information and communications technology (ICT)
-field such as virtualization, enabling technology for Cloud computing and 
-operating--system--level virtualization known as containerization. All of which
+evolution of the software engineering methodologies (creation of the Agile
+Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
+with the parallel advancements in the information and communication technology (ICT)
+field (such as virtualization, the enabling technology for Cloud computing and 
+operating--system--level virtualization known as containerization). All of which
 complemented by the emergence of automation and event-response tools.
@@
 latest projects are the most committed in terms of software reliability awareness.
-This table summarizes the topics addressed in the following sections, highlighting 
+Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting 
 the commonalities and specific features on software reliability.
@@
     & DataGrid &
-Research and Technological Development for an International Data Grid & 2001--2004\\
+Research and Technological Development for an International Data Grid & 2001--2003\\
 \begin{minipage}{.3\textwidth}
@@
 The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
-had as main goal to develop the software to provide basic grid functionality
+had as main goal to develop the software to provide basic Grid functionality
 and associated management tools for a large scale testbed for demonstration projects in three
-specific areas of science, such as Particle physics, Bioinformatics and Earth observation.
+specific areas of science, such as particle physics, bioinformatics and earth observation.
 %was devoted to develop the first European infrastructure for Grid computing. 
@@
 countries. 
-%There were 3 application areas for requirement gathering; Particle physics, Bioinformatics and Earth observation. 
 The project delivered its own software distribution, named EDG (EU DataGrid), strongly 
-based on Globus Middleware Services \cite{globus}. The project was organized in 12 workpages, from 
+based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
 which 5 were devoted to software development coordinated by an 
@@
 production. Starting with EGEE--II, the project adopted automation in the software
-lifecycle process by leveraging the automatic build system for Grid middleware: ETICS 
-\cite{etics} solution.
+lifecycle process by leveraging the automatic build system for Grid middleware, that is
+the ETICS \cite{etics} solution.
 
@@
 EGI--InSPIRE not only ensured the coordination of the operation of a secure, reliable 
-European-wide production grid infrastructure federated from national grid initiatives 
-that is integrated and interoperates with other grids worldwide but also foreseed the 
+European-wide production Grid infrastructure federated from National Grid Initiatives (NGIs)
+that is integrated and interoperates with other Grids worldwide but also foreseed the 
 maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
@@
 form the production infrastructure. For the later it maintained a production--ready 
-Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution) and 
+Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution), and 
 defined general and component specific quality criteria to be applied to software 
@@
 project of software development considered in this review. At the time of writing
-the project is succeeding in providing solutions to address the existing gaps in
-cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
+this paper, the project is succeeding in providing solutions to address the existing gaps in
+Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
 helping developers, e--Infrastructures and scientific communities to exploit
@@
 
-The INDIGO--DataCloud project is the most recent project in the list and, as such, it is 
-the result of the lessons learned from previous experiences in terms of the evaluation and application
-of innovative software engineering methodologies, the management of new technologies to put
-them into practice, and the compilation and analysis of metrics to measure the success of the
-solutions implemented. The expertise gathered is complemented by the fact of being framed in
-collaborative projects with an extensive list of partners.
+INDIGO--DataCloud is the most recent project included in this paper thus it
+emerged from the lessons learned in the previous experiences described in
+Section \ref{sec:ev}. The expertise gathered throughout these years in terms
+of the evaluation and application of innovative software engineering
+methodologies, the management of new technologies to put those methodologies
+into practice, and the compilation and analysis of metrics to measure the
+success of the solutions implemented, has been complemented by the added 
+complexity of being framed into large collaborative projects with an extensive
+list of partners.
 
-framed projects
+Three main pillars sustain the project's strategy to seek reliability in the 
+software produced. A solid \textit{Software Quality Assurance (SQA) criteria
+and metrics gathering definition} guides the development of the software in
+a per-change basis, validating the software at the very early stages. The 
+software delivery process is designed as a bottom-up \textit{adoption of DevOps
+principles}, passing through Continuous Integration (CI) and Continuous
+Delivery (CD) approaches. Once the software is released, it follows a 
+\textit{two-factor validation} in the project's pilot infrastructures and
+resource provider's early-adoption.
 
-All of this with the added complexity of being framed in major
-collaborations with
-
-procollecting and analyzing metrics relevant for the .. to empower the sw reliability bla bla bla
-
-The INDIGO--Datacloud project reflects the progress made in software
-quality and reliability aspects throughout the past scientific European
-experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
-highly influenced by the insights of current big worldwide collaborations of
-software development. These collaborations have continuously inspired the use of new
-technologies and procedures to increase the robustness and quality of the
-software being delivered. In this scenario, the implementation of the DevOps culture 
-is one of the major outcomes, as this paradigm has been progressively adopted by the
-project.
-
-\subsection{DevOps practices}
-\label{sec:devops}
-
-The DevOps methods emphasize on exercising Quality Assurance techniques to
-avoid infrastructure disruption whenever new developments are deployed into
-production. The INDIGO--DataCloud project promoted the application of a 
-Continuous Integration (CI) scenario to enforce the Quality Assurance requirements 
-for any piece of software being produced within the project. Such environment 
-requires an automation ready--to--go infrastructure where the different services involved
-(source code management platform, automation server) interact with each other
-to trigger the quality check execution and return back the exit codes, together
-with their outputs. To accomplish this, the project leveraged on tightly
-integrated open source tools such as GitHub \cite{github}, as source code repository
-and Jenkins \cite{jenkins}, as continuous integration system, as well as relying on
-Docker container provisioning to speed up the source code validation. This CI
-approach is guiding the project’s software development phase throughout the
-first and second major releases.
-
-As DevOps suggests, frequent releases positively affect the reliability of the
-software. The software updates of INDIGO--DataCloud products taking place since
-the second major release are passing through a Continuous Delivery (CD)
-pipeline that adds the packaging of the software right after the successful execution of
-the quality checks (as part of the CI). 
-The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
-images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
-validation step is added, consisting in the product’s deployment using a
-Configuration Management (CM) solution. The installation uses the packages
-built in the previous step and uploaded to a testing
-repository. In the case of Docker images, the CM is used to build the image
-itself, installing and configuring the product before uploading it to the
-DockerHub repository \cite{indigo-dockerhub}.
-
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=\textwidth]{images/devops.png}
-\caption{Continuous Delivery workflow for Docker images.}
-\label{fig:fig_CD}
-\end{figure*}
+%The INDIGO--Datacloud project reflects the progress made in software
+%quality and reliability aspects throughout the past scientific European
+%experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
+%highly influenced by the insights of current big worldwide collaborations of
+%software development. These collaborations have continuously inspired the use of new
+%technologies and procedures to increase the robustness and quality of the
+%software being delivered. In this scenario, the implementation of the DevOps culture 
+%is one of the major outcomes, as this paradigm has been progressively adopted by the
+%project.
 
 \subsection{Software Quality Procedures}
+\label{subsec:sqa}
 
@@
 
+\subsection{DevOps practices}
+\label{sec:devops}
+
+The DevOps methods emphasize on exercising SQA techniques to avoid infrastructure
+disruption whenever new developments are deployed into production systems. The 
+following DevOps approaches have been progressively adopted throughout the project's
+lifetime, being one of its major outcomes.
+
+\subsubsection{Continuous Integration}
+The INDIGO--DataCloud project promoted the application of a CI scenario to enforce
+the SQA requirements -- described in Section \ref{subsec:sqa} -- for any piece of
+software being produced within the project. Such environment requires an automation
+ready--to--go infrastructure where the different technologies involved -- source code
+management platform, automation server and containerization -- interact with each
+other to trigger the source code validation pipeline for each candidate change. The
+pipeline is comprised of diverse quality checks, enforcing at least the successful
+execution of code style and minimum unit testing coverage. To take over the 
+implementation of such infrastructure, the project leveraged on tightly integrated open
+source tools such as GitHub \cite{github}, as the online source code repository,
+and Jenkins \cite{jenkins}, the event-response CI system. Docker container provisioning
+is used to provide instant computing power needed for executing the pipeline jobs.
+
+This CI approach is guiding the project’s software development phase throughout the
+first and second major releases.
+
+\subsubsection{Continuous Delivery}
+As DevOps suggests, frequent releases positively affect the reliability of the
+software. The software updates of INDIGO--DataCloud products taking place since
+the second major release are passing through a Continuous Delivery (CD)
+pipeline that adds the packaging of the software right after the successful execution of
+the quality checks (as part of the CI). 
+The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
+images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
+validation step is added, consisting in the product’s deployment using a
+Configuration Management (CM) solution. The installation uses the packages
+built in the previous step and uploaded to a testing
+repository. In the case of Docker images, the CM is used to build the image
+itself, installing and configuring the product before uploading it to the
+DockerHub repository \cite{indigo-dockerhub}.
+
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker images.}
+\label{fig:fig_CD}
+\end{figure*}
+
 \begin{figure}[ht]
@@
 
-\subsection{DevOps adoption from user communities}
+\subsubsection{DevOps adoption from user communities}
 
@@
 \bibitem{condor}
-D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the grid}, in Grid
+D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the Grid}, in Grid
 Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70,
@@
 A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
-the international software engineering service for the grid}, in Journal of
+the international software engineering service for the Grid}, in Journal of
 Physics: Conference Series, Vol. 119, N. 4, IOP Publishing Ltd, 2008.
@@
 such as INDIGO-DataCloud and EGI-Engage. Currently involved in EOSC-hub project
-and participating in the software quality work packages of DEEP-HybridDatacloud
+and participating in the software quality work packages of DEEP-HybridDataCloud
 and eXtreme-DataCloud.
@@
 is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
-Member of DEEP-HybridDatacloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-hub project and
+Member of DEEP-HybridDataCloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-hub project and
 of the Portuguese National Computing Distributed Infrastructure.

### 2018-02-04 paper-ieee-trans.tex
Add brief explaination of SQA requirements

@@
 
-Three main pillars sustain the project's strategy to seek reliability in the 
+In the following section we will describe the three main pillars that sustain the project's strategy to seek reliability in the 
 software produced. A solid \textit{Software Quality Assurance (SQA) criteria
@@
 The software quality policies \cite{indigo-d31} cover: 1) the identification
-and description of the \emph{quality requirements} that the software need to
-comply with, and 2) the \emph{quality metrics}.
+and description of the \emph{SQA requirements} that the produced software need 
+to comply with, and 2) the \emph{quality metrics}, to monitor each software product's 
+behaviour throughout the development, release and production stages.
 
-The requirements are checked at early stages in the software development
-process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
+The SQA requirements are checked automatically at early stages in the software
+development process, following a CI approach described in Section \ref{subsec:ci}, so
+that that any bug or design issue is likely to be detected and corrected promptly.
+
+\subsubsection{Code style}
+The source code is adhered to any style standard and shall be compliant with the de-facto
+guidelines present for the programming language being used.
+\subsubsection{Unit and Functional testing}
+Whenever a new change involves a new functionality addition, it is requested to be tested. 
+Regression testing in this context is done enforcing the successful execution of the previous
+functionality tests already present in the code. On the other hand, unit testing evaluates the
+code's internal design setting a threshold of 70\% coverage for the code developed.
+\subsubsection{Documentation}
+\subsubsection{Code review}
+Represents the last step in the change management pipeline, once the candidate change has 
+successfully passed through the testing methods described previously. It implies the human-based
+revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
+documentation completeness, etc. On approval, the candidate change is definetively merged into
+production.
+\subsubsection{Automated deployment}
+
+integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
 bug or design issue is likely to be detected and corrected promptly. The source
@@
 \subsubsection{Continuous Integration}
+\label{subsec:ci}
 The INDIGO--DataCloud project promoted the application of a CI scenario to enforce

### 2018-02-06 paper-ieee-trans.tex
Extend CI implementation and figure

--- 2018-01-19 paper-ieee-trans.tex
@@
 The software quality policies \cite{indigo-d31} cover: 1) the identification
-and description of the \emph{quality requirements} that the software need to
-comply with, and 2) the \emph{quality metrics}.
+and description of the \emph{SQA requirements} that the produced software need 
+to comply with, and 2) the \emph{quality metrics}, to monitor each software product's 
+behaviour throughout the development, release and production stages.
 
-The requirements are checked at early stages in the software development
-process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
-bug or design issue is likely to be detected and corrected promptly. The source
-code is publicly available in GitHub repositories under the project organization called
-{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
-catalogue, promoting the external contributions and software adoption. Furthermore,
-the source code is checked for compliance with community de--facto style standards,
-selected for each one of the variety of programming languages being used as seen in
-Figure \ref{fig:fig_codestyle}.
+\subsubsection{SQA criteria}
+
+The SQA requirements cover the set of conventions and recommendations that pave the way for
+an adequate development, timely deliver and reliable operation of the software components
+produced. The SQA criteria is in continuous evolution, reviewed periodically to match the
+project's needs.
+
+\begin{itemize}
+\item \textbf{Code style}.
+Code style standards are enforced for every software component in the project stack. Community
+most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
+choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
+improve the readability and reusability of the source code produced under the scope of the
+project.
+
+\item \textbf{Unit and Functional testing}.
+Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
+testing} in this context is done enforcing the successful execution of the previous
+functionality tests already present in the code, so to uncover any malfunction introduced by
+the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
+focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
+70\% coverage for unit testing.
+
+\item \textbf{Code review}.
+Represents the last step in the change management pipeline, once the candidate change has 
+successfully passed through the testing methods described previously. It implies the human-based
+revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
+documentation completeness, etc. On approval, the candidate change is definitely merged into
+production.
+
+\item \textbf{Documentation and Automated Deployment}.
+Last but not least, the SQA criteria sets the path for the adoption of the developed software by
+defining the documentation content, according to the target audience. These requirements also 
+promote the automation, both in terms of documentation creation and service deployment. The
+\textit{documentation is treated as code} so that it is portable and easily reviewed, using the same workflow
+as the code does, validating the changes before being updated into the production repository. On the
+other hand, the software component deployment is requested to be automated using a configuration 
+management tool. These tools allow to have an \textit{infrastructure as code}, managing the component's
+deployment through declarative definitions. The definition files appear as self-documenting 
+code, as they sequentially guide the component's deployment process in multiple platforms.
+\end{itemize}
+
+%integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
+%bug or design issue is likely to be detected and corrected promptly. The source
+%code is publicly available in GitHub repositories under the project organization called
+%{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
+%catalogue, promoting the external contributions and software adoption. Furthermore,
+%the source code is checked for compliance with community de--facto style standards,
+%selected for each one of the variety of programming languages being used as seen in
+%Figure \ref{fig:fig_codestyle}.
 
@@
 following DevOps approaches have been progressively adopted throughout the project's
-lifetime, being one of its major outcomes.
+lifetime, being one of its major outcomes in terms of sofware reliability.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
+\caption{.}
+\label{fig:fig_jenkins_CI_builds}
+\end{figure}
 
 \subsubsection{Continuous Integration}
-The INDIGO--DataCloud project promoted the application of a CI scenario to enforce
-the SQA requirements -- described in Section \ref{subsec:sqa} -- for any piece of
-software being produced within the project. Such environment requires an automation
+\label{subsec:ci}
+The INDIGO--DataCloud project promoted the application of a CI scenario that enforces,
+for any piece of software produced, the testing requirements defined in the SQA criteria,
+as described in Section \ref{subsec:sqa}. Such environment requires an automation
 ready--to--go infrastructure where the different technologies involved -- source code
@@
 other to trigger the source code validation pipeline for each candidate change. The
-pipeline is comprised of diverse quality checks, enforcing at least the successful
-execution of code style and minimum unit testing coverage. To take over the 
-implementation of such infrastructure, the project leveraged on tightly integrated open
-source tools such as GitHub \cite{github}, as the online source code repository,
-and Jenkins \cite{jenkins}, the event-response CI system. Docker container provisioning
-is used to provide instant computing power needed for executing the pipeline jobs.
+pipeline is mainly comprised by the code style validation, unit and functional testing
+coverage. The CI pipeline is complemented with additional quality checks, such as 
+integration tests for the software components where automation of this type of testing
+is applicable, or security linters for the static analysis of suspicious
+constructs that could lead to security risks. Metrics gathering also takes part of
+the CI pipeline so to have a per-change trend evolution of the common source code 
+related metrics, such as the source lines of code (SLOC) or the cyclomatic complexity.
 
-This CI approach is guiding the project’s software development phase throughout the
-first and second major releases.
+The practical implementation of the CI scenario leveraged on tightly integrated open
+source tools such as GitHub \cite{github} -- as the online source code repository --,
+Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
+provisioning to provide instant computing power needed for executing the pipeline jobs. 
+The project defined a source code contribution workflow, based on GitHub Pull Requests 
+(PRs), where each change automatically triggers in Jenkins the associated quality and
+metric tracking checks on PR creation or update. The set of required checks contained
+in the CI pipeline report their exit status back to GitHub, preventing the change to be
+approved if they are not successfully executed.
+
+Figure \ref{fig:fig_jenkins_CI_builds}
+
+The described CI approach is guiding the project’s software development phase throughout
+the first and second major releases. Throughout its lifetime, 
+% FIXME complement with the paragraph above and the references to the figures for code style and unit testing!!!!
+%The implementation of a CI scenario is accomplished by automating the execution of
+%the test types defined in the SQA criteria described in Section \ref{subsec:sqa}. 
+%Each change in the source code will trigger the associated with the software component.
+%As each change is validated, the chances of early detection of defects increases. Within
+%this scenario, the cost of defect solving is dramatically reduced and the reliability of the 
+%software solutions improved, as any bug or design issue is likely to be detected and subsequently
+%corrected at this phase.
+%
 

### 2018-02-06 paper-ieee-trans.tex
Add integration testing requirement to SQA criteria

@@
 improve the readability and reusability of the source code produced under the scope of the
-project.
+project. Figure \ref{fig:fig_codestyle} showcases the code style standards and their popularity
+over the software components contributing to the project.
 
@@
 focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
-70\% coverage for unit testing.
+70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
+for each software component over the two major INDIGO-DataCloud releases, showing an increment in the 
+averaged value of {\sl INDIGO-2} release. As it can be seen, a very small set of products with low coverage
+values highly impacted the overall result, although a significant part of these products are upstream
+contributions or products with codebases not fully under the control of the project. Nevertheless,
+60\% of the products were over this threshold while 80\% of them had values above
+the 50\% coverage.
+
+\item \textbf{Integration testing}. Software components usually interact with other services while on
+operation state. Integration testing deals with the interactions among coupled software components or
+parts of a system that cooperates to achieve a given functionality. This type of testing might be 
+complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
+guarantee the overall operation of the component, focused on the services it intefaces with, whenever
+new functionalities are introduced.
 
@@
 
-\item \textbf{Documentation and Automated Deployment}.
+\item \textbf{Documentation}.
 Last but not least, the SQA criteria sets the path for the adoption of the developed software by
@@
 promote the automation, both in terms of documentation creation and service deployment. The
-\textit{documentation is treated as code} so that it is portable and easily reviewed, using the same workflow
-as the code does, validating the changes before being updated into the production repository. On the
-other hand, the software component deployment is requested to be automated using a configuration 
-management tool. These tools allow to have an \textit{infrastructure as code}, managing the component's
-deployment through declarative definitions. The definition files appear as self-documenting 
-code, as they sequentially guide the component's deployment process in multiple platforms.
+\textit{documentation is treated as code}, using a markup language, automatically rendered and
+uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
+human--reviewed, using the same workflow as the code does, validating the changes before being
+updated into the production repository.
+
+\item \textbf{Automated Deployment}.
+In order to lower the barriers of software adoption, the SQA criteria requires the automated 
+deployment of the products delivered as part of the catalogue. Automation in this context is 
+tackled using a configuration management tool. These tools allow to have an 
+\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
+definitions. The definition files appear as additional source of documentation, 
+self-documenting code, as they sequentially guide the component's deployment process in multiple
+platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
+as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
+such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
+the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
+\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
+It shows an increase after the {\sl INDIGO-1} release especially in the previous weeks of each
+of the two major releases.
 \end{itemize}
-
-%integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
-%bug or design issue is likely to be detected and corrected promptly. The source
-%code is publicly available in GitHub repositories under the project organization called
-%{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
-%catalogue, promoting the external contributions and software adoption. Furthermore,
-%the source code is checked for compliance with community de--facto style standards,
-%selected for each one of the variety of programming languages being used as seen in
-%Figure \ref{fig:fig_codestyle}.
 
@@
 
-Unit testing coverage, another project quality metric, has increased between the first ({\sl INDIGO-1}) and the
-second ({\sl INDIGO-2}) INDIGO--DataCloud releases, reaching an average value of
-66.69\% on the {\sl INDIGO-2} release, very close to the recommended 70\% threshold
-defined by the project’s SQA base criteria \cite{indigo-d31}. A comparison graph
-between the code coverage for the products involved both in {\sl INDIGO-1} and
-{\sl INDIGO-2} releases is shown in Figure \ref{fig:fig_unittest}. As it can be seen,
-a very small set of products with low coverage values highly impacted the overall
-result, although a significant part of these products are upstream contributions
-or products with codebases not fully under the control of the project. Nevertheless,
-60\% of the products were over this threshold while 80\% of them had values above
-the 50\% coverage.
-
-The documentation of the products is treated as source code, using a markup
-language, automatically rendered and uploaded to online repositories
-\cite{indigo-gitbook}. Changes in both documentation and source code are
-human--reviewed as the last step before merging them into the production
-repository or official documentation.
-
-Last but not least, as described in section \ref{sec:devops}, in order
-to facilitate the usage of the INDIGO--DataCloud services, they can be deployed
-automatically using either Ansible \cite{indigo-ansible} or Puppet
-\cite{indigo-puppet} open--source configuration tools. The adoption of such Configuration Management
-tools not only eases the consumption of the products developed by external users or
-infrastructures, but also tackles the correct installation and configuration of the
-services and applications. A representative example is the \textit{50 roles} being 
-developed within the project and currently hosted in the Ansible Galaxy portal for 
-the project \footnote{https://galaxy.ansible.com/indigo-dc/}.
-
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=\textwidth]{images/unittest.png}
-\caption{Unit testing coverage for the two major INDIGO--DataCloud releases.}
-\label{fig:fig_unittest}
-\end{figure*}
-
-Figure \ref{fig:fig_confman} shows the number of products that offer an automated means
-for deployment. It shows an increase after the {\sl INDIGO-1} release especially
-in the previous weeks of each of the two major releases.
+\subsubsection{Quality metrics}
 
@@
 \includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
-\caption{.}
+\caption{Evolution of the number of builds triggered as part of the CI implementation.}
 \label{fig:fig_jenkins_CI_builds}
@@
 in the CI pipeline report their exit status back to GitHub, preventing the change to be
-approved if they are not successfully executed.
+approved if they are not successfully executed. As each change is validated, the chances
+of early detection of defects augments. Within this scenario, the cost of defect solving
+is dramatically reduced and the reliability of the software solutions improved, as any
+bug or design issue is likely to be detected and subsequently corrected at this phase.
 
-Figure \ref{fig:fig_jenkins_CI_builds}
-
-The described CI approach is guiding the project’s software development phase throughout
-the first and second major releases. Throughout its lifetime, 
-% FIXME complement with the paragraph above and the references to the figures for code style and unit testing!!!!
-%The implementation of a CI scenario is accomplished by automating the execution of
-%the test types defined in the SQA criteria described in Section \ref{subsec:sqa}. 
-%Each change in the source code will trigger the associated with the software component.
-%As each change is validated, the chances of early detection of defects increases. Within
-%this scenario, the cost of defect solving is dramatically reduced and the reliability of the 
-%software solutions improved, as any bug or design issue is likely to be detected and subsequently
-%corrected at this phase.
-%
+Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
+INDIGO-DataCloud's first and second releases, of the required test types builds since the
+implementation of the CI infrastructure. Automated functional testing coverage were not
+available for all the software stack, thus the associated number of builds are fewer. The
+executions are not precisely aligned as checks can be manually triggered, through Jenkins
+interface, useful for instance whenever a failed check needs to be rebuilt.
 

### 2018-02-06 paper-ieee-trans.tex
CD description improved

@@
 information about new research trends in software reliability based upon the 
-experiences from the recent INDIGO-DataCloud project. Finally, Section \ref{sec:con} concludes.
+experiences from the recent INDIGO--DataCloud project. Finally, Section \ref{sec:con} concludes.
 
@@
 \end{minipage}
-     & INDIGO-DataCloud &
+     & INDIGO--DataCloud &
 INtegrating Distributed data Infrastructures for Global ExplOitation
@@
 \hline
-Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO-DC\\
+Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
 \hline
@@
 
-The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
+The INDIGO--DataCloud (INtegrating Distributed data Infrastructures for Global
 explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
@@
 70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
-for each software component over the two major INDIGO-DataCloud releases, showing an increment in the 
-averaged value of {\sl INDIGO-2} release. As it can be seen, a very small set of products with low coverage
+for each software component over the two major INDIGO--DataCloud releases, showing an increment in the 
+averaged value of {\sl INDIGO--2} release. As it can be seen, a very small set of products with low coverage
 values highly impacted the overall result, although a significant part of these products are upstream
@@
 \ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
-It shows an increase after the {\sl INDIGO-1} release especially in the previous weeks of each
+It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
 of the two major releases.
@@
 ISO/IEC 9126 standard. These metrics cover the development, release and
-maintenance phases of the software lifecycle. Development and release metrics
-are obtained automatically from several sources, such as GitHub and Jenkins CI
-service, and graphically displayed as GitHub pages using GrimoireLab framework
-\cite{grimoirelab}. Maintenance and user support metrics are collected from the
-different sources of data such as GGUS \cite{ggus} and Github Issues.
+maintenance phases of the software lifecycle.
+
+\textit{Development metrics} are obtained programatically from several sources, namely GitHub
+API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
+framework \cite{grimoirelab}. Per-component weekly reports, including the SQA requirement
+completeness, are issued and individually discussed through the different communication channels
+with the development teams. Issue tracking metrics and KPIs are an essential piece of information
+of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
+development phase.
+
+\textit{Release metric} sources are mainly the online repository servers, namely the 
+Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins 
+server contains as well valuable release data as it is the service where the packages
+and containers are being built before being uploaded to the online repositories. Release
+KPIs primarly focus on the frequency and effiency -- mainly rollbacks -- statistics.
+
+\textit{Maintenance and user support metrics} are key for continuously improving the response
+to issues reported by external users. Feeback is collected from outside helpdesks, such as 
+EGI's GGUS \cite{ggus}, and the INDIGO--DataCloud's GitHub Issues tracker.
 
@@
 Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
-INDIGO-DataCloud's first and second releases, of the required test types builds since the
+INDIGO--DataCloud's first and second releases, of the required test types builds since the
 implementation of the CI infrastructure. Automated functional testing coverage were not
@@
 \subsubsection{Continuous Delivery}
-As DevOps suggests, frequent releases positively affect the reliability of the
-software. The software updates of INDIGO--DataCloud products taking place since
-the second major release are passing through a Continuous Delivery (CD)
-pipeline that adds the packaging of the software right after the successful execution of
-the quality checks (as part of the CI). 
-The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
-images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
-validation step is added, consisting in the product’s deployment using a
-Configuration Management (CM) solution. The installation uses the packages
-built in the previous step and uploaded to a testing
-repository. In the case of Docker images, the CM is used to build the image
-itself, installing and configuring the product before uploading it to the
-DockerHub repository \cite{indigo-dockerhub}.
+As DevOps suggests, frequent releases positively affect the reliability of the software as
+they involve fewer changes and usually more correlated. The software updates of 
+INDIGO--DataCloud products, taking place since the second major release, are passing through
+a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
+successful execution of the CI pipeline previously described. Consequently, CI-validated 
+software components are automatically delivered in online repositories without the need of 
+any human interaction but ultimately requiring expert supervision to validate the results.
+
+INDIGO--DataCloud software stack is delivered in form of Linux software packages (rpms, debs) 
+and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
+package pipeline} intentionally delivers the packages produced in a pre-production -- preview --
+branch. Before that, once the software packages have been created, they are uploaded to a 
+testing branch. The available automated deployment solution -- see \cite{subsec:sqa} -- checks
+the installation and configuration of the component relying on the testing repository. On 
+successful completion, the packages are then automatically signed and subsequently moved to 
+the preview branch, from where they finally be made available in the production branch after
+the release manager supervises the process.
+
+The \textit{Docker container pipeline} uses instead the automated deployment solution to 
+actually install and configure the software component in the container image. Once this is 
+done the recently created image is uploaded to the production DockerHub repository, tagged
+with the label corresponding to the current INDIGO-DataCloud release. Figure 
+\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation, 
+automatically triggered by a change in the source code.
 
@@
 \caption{Resource centers supporting the Pilot Preview testbed and corresponding
-set of deployed INDIGO-DataCloud components or services.}
+set of deployed INDIGO--DataCloud components or services.}
 \label{fig:fig_pilotpreview}
@@
 different DevOps practices is not only useful and suitable for the software related
-to the core services in the INDIGO-DataCloud solution, but also applicable to the
+to the core services in the INDIGO--DataCloud solution, but also applicable to the
 development and distribution of the applications coming from the user communities.
@@
 source code before merging and the creation of a new versioned Docker image,
-automatically available in the INDIGO-DataCloud’s application repository.
+automatically available in the INDIGO--DataCloud’s application repository.
 
@@
 software is being prioritized. Recent software engineering insights, such as
-DevOps approaches, are gradually being applied as the open--source collaborative tools
-evolve, allowing tighter integrations. The Software Quality
+DevOps approaches, are gradually being applied as the open--source collaborative 
+tools evolve, allowing tighter integrations. The Software Quality
 Assurance procedures have been integrated in the development stage, in order to
@@
 \emph{INtegrating Distributed data Infrastructures for Global ExplOitation
-(INDIGO-DataCloud)} project, European Community Research and Development
+(INDIGO--DataCloud)} project, European Community Research and Development
 Information Service (CORDIS),
@@
 \bibitem{indigo-dockerhub}
-\emph{INDIGO-DataCloud DockerHub repository},
+\emph{INDIGO--DataCloud DockerHub repository},
 https://hub.docker.com/u/indigodatacloud
@@
 \bibitem{indigo-d31}
-\emph{Initial Plan for WP3} INDIGO-DataCloud Deliverable 3.1,
+\emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1,
 https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31
@@
 \bibitem{indigo-jenkins}
-\emph{INDIGO-DataCloud Jenkins CI service},
+\emph{INDIGO--DataCloud Jenkins CI service},
 https://jenkins.indigo-datacloud.eu:8080/
@@
 \bibitem{indigo-github}
-\emph{INDIGO-DataCloud GitHub Source Code repository},
+\emph{INDIGO--DataCloud GitHub Source Code repository},
 https://www.github.com/indigo-dc
@@
 \bibitem{indigo-gitbook}
-\emph{INDIGO-DataCloud GitBook Documentation repository},
+\emph{INDIGO--DataCloud GitBook Documentation repository},
 https://www.gitbook.com/@indigo-dc
@@
 \bibitem{indigo-puppet}
-\emph{INDIGO-DataCloud PuppetForge repository},
+\emph{INDIGO--DataCloud PuppetForge repository},
 https://forge.puppet.com/indigodc
@@
 \bibitem{indigo-ansible}
-\emph{INDIGO-DataCloud Ansible Galaxy repository},
+\emph{INDIGO--DataCloud Ansible Galaxy repository},
 https://galaxy.ansible.com/indigo-dc/
@@
 \emph{GrimoireLab}, http://grimoirelab.github.io/
+
+\bibitem{indigo-pkg-repo}
+\emph{INDIGO--DataCloud Linux Package repository}, https://jenkins.indigo-datacloud.eu/
 
@@
 assurance tasks within different European advanced computing research projects
-such as INDIGO-DataCloud and EGI-Engage. Currently involved in EOSC-hub project
+such as INDIGO--DataCloud and EGI-Engage. Currently involved in EOSC-hub project
 and participating in the software quality work packages of DEEP-HybridDataCloud
@@
 He currently leads the Software Development and Distributed Systems department at CNAF, the INFN National Center dedicated to research and development on IT technologies, located in Bologna, Italy.
-He was Project Coordinator of INDIGO-DataCloud, a project funded by the EC Horizon2020, that developed innovative open source computing and storage solutions.
+He was Project Coordinator of INDIGO--DataCloud, a project funded by the EC Horizon2020, that developed innovative open source computing and storage solutions.
 \end{IEEEbiography}

### 2018-02-07 paper-ieee-trans.tex
conclusions extended

@@
 
-It is a fact that the European software produced for scientific purposes is
+After more that a decade of experiences collected throughout the reviewed 
+EU-funded software development and e-Infrastructure management projects, it 
+becomes apparent that the European software produced for scientific purposes is
 evolving to a more sustainable model where the quality and reliability of
-software is being prioritized. Recent software engineering insights, such as
-DevOps approaches, are gradually being applied as the open--source collaborative 
-tools evolve, allowing tighter integrations. The Software Quality
-Assurance procedures have been integrated in the development stage, in order to
-detect and correct issues early in the software lifecycle. Automation and
-metrics analysis are at the base of prompt issue solving. However, post--release
-validation, through the preview testbeds or the Staged--rollout process, are
-also needed to strengthen the software reliability for scientific usage.
+software is being prioritized. The gradual adoption of software engineering insights, such as
+DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
+collaborative tools, extending their capabilities to bring along tight 
+integrations amongst other services, matched the requirements of such practices
+and procedures, facilitating in a great manner their implementation in real scenarios.
+
+Reliability of the software has been substantially improved by applying the SQA
+criteria in the development stage. Consequently, each change in the source code 
+that is meant to be integrated in the production version, is validated according
+to the SQA testing requirements. Thus, defects are detected and corrected early in the 
+software lifecycle, reducing the cost of fixing errors at later stages. Automation is
+key in this process, allowing the implementation of a CI scenario that enforces
+the SQA criteria compliance. The application of more advanced DevOps practices
+resulted in the implementation of CD pipelines to extend automation to the
+delivery of packages. Thereby, safe and frequent software releases are doable, reducing
+the impact on production systems. In this regard, a pre-production validation,
+through the preview testbeds or resource centers early adoption -- handled by the
+Staged--rollout process --, alleviates the likelihood of discovering uncovered
+issues at production.
+
+Software lifecycle demands as well from a continuous measurement and analysis
+process that permits timely reactions to discover and fix defects throughout the
+development, release and post-release or maintenance stages. Conscientious metric
+tracking and response sits at the core of an appropriate software maturity, as 
+ultimate reliability can only be achieved by enforcing and empowering the SQA 
+criteria through each stage of the software lifecycle.
 

### 2018-02-08 paper-ieee-trans.tex
Introduction of section III rephrased

@@
 emerged from the lessons learned in the previous experiences described in
-Section \ref{sec:ev}. The expertise gathered throughout these years in terms
-of the evaluation and application of innovative software engineering
+Section \ref{sec:ev}. The expertise gathered throughout these years was highly
+profitable in terms of the evaluation and application of innovative software engineering
 methodologies, the management of new technologies to put those methodologies
-into practice, and the compilation and analysis of metrics to measure the
-success of the solutions implemented, has been complemented by the added 
-complexity of being framed into large collaborative projects with an extensive
-list of partners.
+into practice, and the compilation and analysis of the appropriate set of metrics to measure the
+success of the solutions implemented. All the above with the additional complexity of
+being framed into large collaborative projects with an extensive list of partners.
 
-In the following section we will describe the three main pillars that sustain the project's strategy to seek reliability in the 
-software produced. A solid \textit{Software Quality Assurance (SQA) criteria
-and metrics gathering definition} guides the development of the software in
-a per-change basis, validating the software at the very early stages. The 
-software delivery process is designed as a bottom-up \textit{adoption of DevOps
-principles}, passing through Continuous Integration (CI) and Continuous
-Delivery (CD) approaches. Once the software is released, it follows a 
-\textit{two-factor validation} in the project's pilot infrastructures and
-resource provider's early-adoption.
+In the following section we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
+software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
+and metrics gathering definition} will provide guidelines to develop quality 
+software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
+by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
+process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
+the foregoing change-based development process and set up advanced delivery mechanisms.
+The last key pillar is constituted by a \textit{two-factor validation} carried out over the project's pilot testbeds and
+e-Infrastructure resource provider's early-adoption.
 
@@
 \begin{itemize}
-\item \textbf{Code style}.
+\item \textit{Code style}.
 Code style standards are enforced for every software component in the project stack. Community
@@
 
-\item \textbf{Unit and Functional testing}.
+\item \textit{Unit and Functional testing}.
 Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
@@
 
-\item \textbf{Integration testing}. Software components usually interact with other services while on
+\item \textit{Integration testing}. Software components usually interact with other services while on
 operation state. Integration testing deals with the interactions among coupled software components or
@@
 
-\item \textbf{Code review}.
+\item \textit{Code review}.
 Represents the last step in the change management pipeline, once the candidate change has 
@@
 
-\item \textbf{Documentation}.
+\item \textit{Documentation}.
 Last but not least, the SQA criteria sets the path for the adoption of the developed software by
@@
 
-\item \textbf{Automated Deployment}.
+\item \textit{Automated Deployment}.
 In order to lower the barriers of software adoption, the SQA criteria requires the automated 

### 2018-02-08 paper-ieee-trans.tex
minor rephrases

@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.5\textwidth]{images/codestyle.png}
+\caption{Code style standards followed by INDIGO--DataCloud's software products.}
+\label{fig:fig_codestyle}
+\end{figure}
+
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=\textwidth]{images/unittest.png}
+\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
+\label{fig:fig_unittest}
+\end{figure*}
+
 \subsection{Software Quality Procedures}
@@
 
-The software quality policies \cite{indigo-d31} cover: 1) the identification
+The software quality policies were initially defined in the first deliverable document of the
+project \cite{indigo-d31} and periodically reviewed to be endowed with applicable insights of 
+software engineering practices, responses to evolving needs emanated from the user communities and 
+feedback obtained from development teams. These policies cover the 1) identification
 and description of the \emph{SQA requirements} that the produced software need 
-to comply with, and 2) the \emph{quality metrics}, to monitor each software product's 
-behaviour throughout the development, release and production stages.
+to comply with, and the 2) \emph{quality metrics}, to monitor each software product's 
+behaviour throughout the development, release and post-release stages.
 
@@
 
-The SQA requirements cover the set of conventions and recommendations that pave the way for
-an adequate development, timely deliver and reliable operation of the software components
-produced. The SQA criteria is in continuous evolution, reviewed periodically to match the
-project's needs.
+The SQA criteria is a collection of conventions and recommendations that pave the way for
+the adequate development, timely deliver and reliable operation of the software components
+produced. It emphasizes the quality requirements and best practices to be applied at the
+development phase, such as style and testing compliance or human code reviews, to protect the
+production source code versions. Besides, the criteria promotes the adoption of the software 
+by providing minimum requirements of documentation content and stressing the usage of automated
+solutions for software deployment.
 
@@
 \item \textit{Code style}.
-Code style standards are enforced for every software component in the project stack. Community
-most-adopted or \textit{de-facto standard} is initially recommended, however the developer teams eventually
-choose the set of guidelines to comply with. The ultimate goal for code style assessment is to
-improve the readability and reusability of the source code produced under the scope of the
-project. Figure \ref{fig:fig_codestyle} showcases the code style standards and their popularity
-over the software components contributing to the project.
+The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
+project. Code style standards are enforced for every software component in the project stack. Community
+most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
+choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
+over the INDIGO--DataCloud software stack.
 
@@
 the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
-focusing on the code's internal design. Project's SQA criteria sets a recommended threshold of
+focusing on the code's internal design. The SQA criteria sets a recommended threshold of
 70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
-for each software component over the two major INDIGO--DataCloud releases, showing an increment in the 
-averaged value of {\sl INDIGO--2} release. As it can be seen, a very small set of products with low coverage
-values highly impacted the overall result, although a significant part of these products are upstream
-contributions or products with codebases not fully under the control of the project. Nevertheless,
-60\% of the products were over this threshold while 80\% of them had values above
-the 50\% coverage.
+for each software component over the two major INDIGO--DataCloud releases, denoting an incremental 
+trend over the course of the project. As depicted in the Figure, 50\% of the components were over the
+70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.
 
@@
 complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
-guarantee the overall operation of the component, focused on the services it intefaces with, whenever
-new functionalities are introduced.
+guarantee the overall operation of the component with regards to the services it interfaces with,
+whenever new functionalities are introduced.
 
@@
 documentation completeness, etc. On approval, the candidate change is definitely merged into
-production.
+production. Human code reviewing is paramount in the software quality assessment, specially 
+important when the SQA testing requirements are fully automated. Secure code reviews are
+done at this stage, assessing common vulnerabilities from inputs coming from automated linters
+and manual dynamic application security testing.
 
 \item \textit{Documentation}.
-Last but not least, the SQA criteria sets the path for the adoption of the developed software by
+The SQA criteria sets the path for the adoption of the developed software by
 defining the documentation content, according to the target audience. These requirements also 
@@
 \end{itemize}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.5\textwidth]{images/codestyle.png}
-\caption{Code style standards followed by INDIGO--DataCloud's software products.}
-\label{fig:fig_codestyle}
-\end{figure}
 
@@
 to issues reported by external users. Feeback is collected from outside helpdesks, such as 
-EGI's GGUS \cite{ggus}, and the INDIGO--DataCloud's GitHub Issues tracker.
+EGI's GGUS \cite{ggus}, and the GitHub Issues tracker.
 
@@
 \includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
-\caption{Evolution of the number of builds triggered as part of the CI implementation.}
+\caption{Evolution of the total number of testing builds triggered automatically as part of the Jenkins CI implementation.}
 \label{fig:fig_jenkins_CI_builds}
@@
 
-\subsubsection{Continuous Integration}
+\subsubsection{Continuous Integration (CI)}
 \label{subsec:ci}
@@
 as described in Section \ref{subsec:sqa}. Such environment requires an automation
-ready--to--go infrastructure where the different technologies involved -- source code
-management platform, automation server and containerization -- interact with each
+ready--to--go infrastructure where the different services and technologies involved interact with each
 other to trigger the source code validation pipeline for each candidate change. The
@@
 integration tests for the software components where automation of this type of testing
-is applicable, or security linters for the static analysis of suspicious
+is applicable, or security linters for the code static analysis of suspicious
 constructs that could lead to security risks. Metrics gathering also takes part of
@@
 implementation of the CI infrastructure. Automated functional testing coverage were not
-available for all the software stack, thus the associated number of builds are fewer. The
-executions are not precisely aligned as checks can be manually triggered, through Jenkins
-interface, useful for instance whenever a failed check needs to be rebuilt.
+available for all the software stack, thus the associated number of builds are fewer.
 
@@
 
-Two applications from "Definition of Support to Research Communities" workpackage, 
+Two applications from "Definition of Support to Research Communities" work package, 
 DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
@@
 The released software is tested in production environments through the
-Staged--Rollout process, before inclusion in the EGI's CloudManagement Distribution (CMD). 
+Staged--Rollout process, before inclusion in the EGI's Cloud Middleware Distribution (CMD). 
 Selected resource providers are requested to install

### 2018-02-08 paper-ieee-trans.tex
cosmetic changes

@@
 
-%\IEEEPARstart{W}hen e-Infrastructures in Europe were first conceived, new
-%services to access distributed resources -- data, computing, instrumental --
-%were initially considered. Major investments were then injected in order to
-%develop distributed solutions that provided a seamless and transparent access
-%to those resources, available for the researchers. One of the most known examples is Worldwide
-%Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered by the
-%Particle Physics scientific community, built up to satisfy the computing needs
-%of processing, storing and analysing the huge quantity of data coming from the Large Hadron
-%Collider particle accelerator.
-
 \IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
@@
 pan-European research collaborations. One of the well-known examples is the Worldwide 
-Large Hadron Collider (LHC) Computing Grid (WLCG) global computing infrastructure, powered 
+Large Hadron Collider (LHC) Computing Grid infrastructure, powered 
 by the particle physics scientific community, built up to satisfy the computing needs
@@
 Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
-with the parallel advancements in the information and communication technology (ICT)
+with the parallel advancements in the information and communication technology
 field (such as virtualization, the enabling technology for Cloud computing and 
@@
 information about new research trends in software reliability based upon the 
-experiences from the recent INDIGO--DataCloud project. Finally, Section \ref{sec:con} concludes.
+experiences from the recent {\sl INDIGO--DataCloud} project. Finally, Section \ref{sec:con} concludes.
 
@@
 
-The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
+The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
 had as main goal to develop the software to provide basic Grid functionality
@@
 countries. 
-The project delivered its own software distribution, named EDG (EU DataGrid), strongly 
+The project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
 based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
@@
 
-The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010)
+The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
 \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
 scientists and engineers from more than 240 institutions in 45 countries to
-provide a seamless Grid infrastructure for e--Science. EGEE--II and EGEE--III
+provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
 featured the internationalization of the project, embracing worldwide research
@@
 The {\sl gLite} middleware \cite{glite} was the ultimate
-official software distribution of EGEE as of 2006, after two years of prototyping and
-re--engineering efforts to converge with LHC Computing Grid (LCG--2), Virtual
+official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
+re--engineering efforts to converge with {\sl LHC} Computing Grid (LCG--2), Virtual
 Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
@@
 \cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
-production. Starting with EGEE--II, the project adopted automation in the software
+production. Starting with {\sl EGEE--II}, the project adopted automation in the software
 lifecycle process by leveraging the automatic build system for Grid middleware, that is
-the ETICS \cite{etics} solution.
+the {\sl ETICS} \cite{etics} solution.
 
@@
 The E--Infrastructure for Testing, Integration and Configuration of Software
-\cite{cordis:etics, cordis:etics2} (ETICS, Jan 2006 -- Feb 2010) project aimed
+\cite{cordis:etics, cordis:etics2} ({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
 at addressing the challenges in producing quality software in distributed,
-collaborative projects such as EGEE and its gLite middleware. The framework
+collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
 integrated different technologies and tools in order to provide automated configuration,
@@
 software metrics gathering such as Source Lines Of Code (SLOC), complexity and
-number of defects/bugs \cite{etics}. The ETICS framework was the first automated
+number of defects/bugs \cite{etics}. The {\sl ETICS} framework was the first automated
 service for delivering quality software products in distributed environments like
@@
 
-The European Middleware Initiative (EMI, May 2010 -- Apr 2013)
+The European Middleware Initiative ({\sl EMI}, May 2010 -- Apr 2013)
 \cite{cordis:emi} project joined the 4 major Grid middleware providers in
@@
 ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
-characteristics that needed to be present in the EMI software products and
-processes to be able to meet the EMI quality requirements
+characteristics that needed to be present in the {\sl EMI} software products and
+processes to be able to meet the {\sl EMI} quality requirements
 \cite{emi-quality-model}.
@@
 metrics and Key Performance Indicators (KPIs) were identified and defined in
-detail in the EMI Metrics Specification \cite{emi-metrics}. The project
+detail in the {\sl EMI} Metrics Specification \cite{emi-metrics}. The project
 leveraged the {\sl ETICS} service for the development, continuous integration and release management,
@@
 
-The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for
+The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
 Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
-arose as the continuation of the EGEE--III project, with the mission of establishing and
+arose as the continuation of the {\sl EGEE--III} project, with the mission of establishing and
 maintaining a sustainable European Grid Infrastructure.
 
-EGI--InSPIRE not only ensured the coordination of the operation of a secure, reliable 
+{\sl EGI--InSPIRE} not only ensured the coordination of the operation of a secure, reliable 
 European-wide production Grid infrastructure federated from National Grid Initiatives (NGIs)
@@
 maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
-the EGI Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
-coordination of the external provision of software required by EGI to 
+the {\sl EGI} Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
+coordination of the external provision of software required by {\sl EGI} to 
 form the production infrastructure. For the later it maintained a production--ready 
@@
 deployed in the European scientific e--Infrastructures under a follow--up
-project, the EGI--Engage (Engaging the EGI Community towards an Open Science
+project, the {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
 Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
@@
 
-The INDIGO--DataCloud (INtegrating Distributed data Infrastructures for Global
+The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
 explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
@@
 
-INDIGO--DataCloud is the most recent project included in this paper thus it
+{\sl INDIGO--DataCloud} is the most recent project included in this paper thus it
 emerged from the lessons learned in the previous experiences described in
@@
 
-%The INDIGO--Datacloud project reflects the progress made in software
+%The {\sl INDIGO--DataCloud} project reflects the progress made in software
 %quality and reliability aspects throughout the past scientific European
@@
 \includegraphics[width=0.5\textwidth]{images/codestyle.png}
-\caption{Code style standards followed by INDIGO--DataCloud's software products.}
+\caption{Code style standards followed by {\sl INDIGO--DataCloud}'s software products.}
 \label{fig:fig_codestyle}
@@
 choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
-over the INDIGO--DataCloud software stack.
+over the {\sl INDIGO--DataCloud} software stack.
 
@@
 70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
-for each software component over the two major INDIGO--DataCloud releases, denoting an incremental 
+for each software component over the two major {\sl INDIGO--DataCloud} releases, denoting an incremental 
 trend over the course of the project. As depicted in the Figure, 50\% of the components were over the
@@
 self-documenting code, as they sequentially guide the component's deployment process in multiple
-platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
+platforms. In this regard, {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
 as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
@@
 \label{subsec:ci}
-The INDIGO--DataCloud project promoted the application of a CI scenario that enforces,
+The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
 for any piece of software produced, the testing requirements defined in the SQA criteria,
@@
 Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
-INDIGO--DataCloud's first and second releases, of the required test types builds since the
+{\sl INDIGO--DataCloud}'s first and second releases, of the required test types builds since the
 implementation of the CI infrastructure. Automated functional testing coverage were not
@@
 they involve fewer changes and usually more correlated. The software updates of 
-INDIGO--DataCloud products, taking place since the second major release, are passing through
+{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
 a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
@@
 
-INDIGO--DataCloud software stack is delivered in form of Linux software packages (rpms, debs) 
+{\sl INDIGO--DataCloud} software stack is delivered in form of Linux software packages (rpms, debs) 
 and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
@@
 done the recently created image is uploaded to the production DockerHub repository, tagged
-with the label corresponding to the current INDIGO-DataCloud release. Figure 
+with the label corresponding to the current {\sl INDIGO--DataCloud} release. Figure 
 \ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation, 
@@
 \caption{Resource centers supporting the Pilot Preview testbed and corresponding
-set of deployed INDIGO--DataCloud components or services.}
+set of deployed {\sl INDIGO--DataCloud} components or services.}
 \label{fig:fig_pilotpreview}
@@
 different DevOps practices is not only useful and suitable for the software related
-to the core services in the INDIGO--DataCloud solution, but also applicable to the
+to the core services in the {\sl INDIGO--DataCloud} solution, but also applicable to the
 development and distribution of the applications coming from the user communities.
@@
 source code before merging and the creation of a new versioned Docker image,
-automatically available in the INDIGO--DataCloud’s application repository.
+automatically available in the {\sl INDIGO--DataCloud}’s application repository.
 
@@
 communities involved in the project. The aim of these testbeds is to test the
-level of integration between the components involved in the INDIGO--DataCloud
+level of integration between the components involved in the {\sl INDIGO--DataCloud}
 solutions and use cases validation, by deploying and executing the applications

### 2018-02-08 paper-ieee-trans.tex
remove footnotes

@@
 \thanks{D. C. Duma, E. Ronchieri, and D. Salomoni are with INFN - CNAF, Bologna, Italy.}
-\thanks{P. Orviz Fernandez, is with CSIC, Santander, Spain, (e-mail: orviz@ifca.unican.es)}
+\thanks{P. Orviz Fernandez, is with CSIC, Santander, Spain.}
 \thanks{M. David, and J. Gomes are with LIP, Lisbon, Portugal.}%
@@
 Software Reliability, Quality Assurance, Software Metrics, Software Testing
-Techniques
+Techniques, DevOps
 \end{IEEEkeywords}
@@
 technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
-and ultimately by the Horizon 2020 programme\footnote{HORIZON 2020 - The EU Framework
-Programme for Research and Innovation https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
+and ultimately by the Horizon 2020 programme \cite{h2020} -- are being held to coordinate and 
 develop distributed solutions that provide a seamless and transparent access to those
@@
 such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
-the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
+the Ansible Galaxy portal \cite{indigo-galaxy}. Figure 
 \ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
@@
 
+\bibitem{h2020}
+\emph{HORIZON 2020}, The EU Framework Programme for Research and Innovation [Online]. Available: https://ec.europa.eu/programmes/horizon2020/
+
 \bibitem{zhu}
@@
 
+\bibitem{indigo-galaxy}
+\emph{INDIGO--DataCloud Ansible Galaxy repository},
+https://galaxy.ansible.com/indigo-dc/
+
 \bibitem{indigo-ansible}

### 2018-02-08 paper-ieee-trans.tex
online references aligned

@@
 \emph{DataGrid project}, European Community Research and Development
-Information Service (CORDIS),
-http://cordis.europa.eu/project/rcn/53665\_en.html
+Information Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/53665\_en.html
 
@@
 \bibitem{agile-manifesto}
-\emph{Manifesto for Agile Software Development}, http://agilemanifesto.org
+\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org
 
@@
 \emph{Enabling Grids for E-sciencE (EGEE)} project, European Community
-Research and Development Information Service (CORDIS),
+Research and Development Information Service (CORDIS) [Online]. Available: 
 http://cordis.europa.eu/project/rcn/80149\_en.html
@@
 \emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, European Community
-Research and Development Information Service (CORDIS),
+Research and Development Information Service (CORDIS) [Online]. Available: 
 http://cordis.europa.eu/project/rcn/99189\_en.html
@@
 \emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, European Community
-Research and Development Information Service (CORDIS),
+Research and Development Information Service (CORDIS) [Online]. Available: 
 http://cordis.europa.eu/project/rcn/87264\_en.html
@@
 \emph{Definition and Documentation of the Revised Software Life-Cycle Process},
-Milestone MSA3.4.2, 2010, EGEE-III project,
-https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1\_4.pdf
+Milestone MSA3.4.2, 2010, EGEE-III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1\_4.pdf
 
@@
 (ETICS)} project, European Community Research and Development Information
-Service (CORDIS), http://cordis.europa.eu/project/rcn/80138\_en.html
+Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/80138\_en.html
 
@@
 Phase 2 (ETICS 2)} project, European Community Research and Development
-Information Service (CORDIS),
-http://cordis.europa.eu/project/rcn/86604\_en.html
+Information Service (CORDIS), [Online]. Available: http://cordis.europa.eu/project/rcn/86604\_en.html
 
@@
 \emph{European Middleware Initiative (EMI)} project, European Community
-Research and Development Information Service (CORDIS),
+Research and Development Information Service (CORDIS) [Online]. Available: 
 http://cordis.europa.eu/project/rcn/95311\_en.html
@@
 \emph{ISO/IEC 9126 Software Engineering - Product Quality}, International
-Organization for Standarization, https://www.iso.org/standard/22749.html
+Organization for Standarization [Online]. Available: https://www.iso.org/standard/22749.html
 
 \bibitem{emi-metrics}
-\emph{EMI Metrics Specification}, https://goo.gl/CCtY7x
+\emph{EMI Metrics Specification} [Online]. Available: https://goo.gl/CCtY7x
 
 \bibitem{emi-quality-model}
-\emph{EMI Quality Model}. Available: https://goo.gl/LdS6fL
+\emph{EMI Quality Model} [Online]. Available: https://goo.gl/LdS6fL
 
@@
 Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, European
-Community Research and Development Information Service (CORDIS),
+Community Research and Development Information Service (CORDIS) [Online]. Available: 
 http://cordis.europa.eu/project/rcn/95923\_en.html
@@
 \bibitem{egi-ops}
-\emph{EGI Operations Portal},
+\emph{EGI Operations Portal} [Online]. Available: 
 https://operations-portal.egi.eu/
@@
 \bibitem{ggus}
-\emph{GGUS} EGI Helpdesk,
+\emph{GGUS} EGI Helpdesk [Online]. Available: 
 https://ggus.eu/
@@
 \bibitem{gocdb}
-\emph{GOCDB} EGI Grid Configuration Repository,
+\emph{GOCDB} EGI Grid Configuration Repository [Online]. Available: 
 https://ggus.eu/
@@
 \bibitem{github}
-\emph{GITHUB},
+\emph{GITHUB} [Online]. Available: 
 https://github.com/
@@
 \bibitem{jenkins}
-\emph{Jenkins} Continuous Integration System,
+\emph{Jenkins} Continuous Integration System [Online]. Available: 
 https://jenkins-ci.org/
@@
 \bibitem{egi-qc}
-\emph{EGI Quality Criteria}. Available: http://egi-qc.github.io/
+\emph{EGI Quality Criteria} [Online]. Available: http://egi-qc.github.io/
 
@@
 project, European Community Research and Development Information Service
-(CORDIS), http://cordis.europa.eu/project/rcn/194937\_en.html
+(CORDIS), [Online]. Available:  http://cordis.europa.eu/project/rcn/194937\_en.html
 
@@
 (INDIGO--DataCloud)} project, European Community Research and Development
-Information Service (CORDIS),
+Information Service (CORDIS) [Online]. Available: 
 http://cordis.europa.eu/project/rcn/194882\_en.html
@@
 \bibitem{indigo-dockerhub}
-\emph{INDIGO--DataCloud DockerHub repository},
+\emph{INDIGO--DataCloud DockerHub repository} [Online]. Available: 
 https://hub.docker.com/u/indigodatacloud
@@
 \bibitem{indigo-d31}
-\emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1,
+\emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1 [Online]. Available: 
 https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31
@@
 \bibitem{indigo-jenkins}
-\emph{INDIGO--DataCloud Jenkins CI service},
+\emph{INDIGO--DataCloud Jenkins CI service} [Online]. Available: 
 https://jenkins.indigo-datacloud.eu:8080/
@@
 \bibitem{indigo-github}
-\emph{INDIGO--DataCloud GitHub Source Code repository},
+\emph{INDIGO--DataCloud GitHub Source Code repository} [Online]. Available: 
 https://www.github.com/indigo-dc
@@
 \bibitem{indigo-gitbook}
-\emph{INDIGO--DataCloud GitBook Documentation repository},
+\emph{INDIGO--DataCloud GitBook Documentation repository} [Online]. Available: 
 https://www.gitbook.com/@indigo-dc
@@
 \bibitem{indigo-puppet}
-\emph{INDIGO--DataCloud PuppetForge repository},
+\emph{INDIGO--DataCloud PuppetForge repository} [Online]. Available: 
 https://forge.puppet.com/indigodc
@@
 \bibitem{indigo-galaxy}
-\emph{INDIGO--DataCloud Ansible Galaxy repository},
+\emph{INDIGO--DataCloud Ansible Galaxy repository} [Online]. Available: 
 https://galaxy.ansible.com/indigo-dc/
@@
 \bibitem{indigo-ansible}
-\emph{INDIGO--DataCloud Ansible Galaxy repository},
+\emph{INDIGO--DataCloud Ansible Galaxy repository} [Online]. Available: 
 https://galaxy.ansible.com/indigo-dc/
@@
 \bibitem{grimoirelab}
-\emph{GrimoireLab}, http://grimoirelab.github.io/
+\emph{GrimoireLab} [Online]. Available: http://grimoirelab.github.io/
 
 \bibitem{indigo-pkg-repo}
-\emph{INDIGO--DataCloud Linux Package repository}, https://jenkins.indigo-datacloud.eu/
+\emph{INDIGO--DataCloud Linux Package repository} [Online]. Available: https://jenkins.indigo-datacloud.eu/
 
 \bibitem{ggus}
-\emph{Global Grid User Support (GGUS)}, https://www.ggus.eu/
+\emph{Global Grid User Support (GGUS)} [Online]. Available: https://www.ggus.eu/
 

### 2018-02-08 paper-ieee-trans.tex
good layout of images
minor fix

@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.45\textwidth, height=50mm]{images/confman.png}
+\caption{Trend line showing the adoption of Configuration Management tools throughout the project lifetime.}
+\label{fig:fig_confman}
+\end{figure}
+
 \subsection{Software Quality Procedures}
@@
 
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=\textwidth]{images/devops.png}
+\caption{Continuous Delivery workflow for Docker images.}
+\label{fig:fig_CD}
+\end{figure*}
+
 \subsubsection{Continuous Integration (CI)}
@@
 the CI pipeline so to have a per-change trend evolution of the common source code 
-related metrics, such as the source lines of code (SLOC) or the cyclomatic complexity.
+related metrics, such as SLOC or the cyclomatic complexity.
 
@@
 automatically triggered by a change in the source code.
-
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=\textwidth]{images/devops.png}
-\caption{Continuous Delivery workflow for Docker images.}
-\label{fig:fig_CD}
-\end{figure*}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.45\textwidth, height=50mm]{images/confman.png}
-\caption{Trend line showing the adoption of Configuration Management tools throughout the project lifetime.}
-\label{fig:fig_confman}
-\end{figure}
 

### 2018-02-08 paper-ieee-trans.tex
last changes

@@
 by the particle physics scientific community, built up to satisfy the computing needs
-of processing, storing and analysing the huge quantity of data coming from the Large 
-Hadron Collider particle accelerator. A significant number of innovation projects and
+of processing, storing and analysing the huge quantity of data coming from the 
+LHC particle accelerator. A significant number of innovation projects and
 technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
@@
 Major developments were done throughout the years, mainly focused on providing new
-functionalities, leaving aside at early times the adoption of software engineering methodologies.
+functionalities, leaving aside in the early times the adoption of software engineering methodologies.
 The software delivered was then less tested, resulting in constant
@@
 
-Analyzing the European Commission (EC)-funded software projects in the last decade
+By analyzing the European Commission (EC)-funded software projects in the last decade
 we have observed a continuous increase of the prominence and robustness of
@@
 latest projects are the most committed in terms of software reliability awareness.
-Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting 
+Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting both
 the commonalities and specific features on software reliability.
@@
 The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
-had as main goal to develop the software to provide basic Grid functionality
-and associated management tools for a large scale testbed for demonstration projects in three
-specific areas of science, such as particle physics, bioinformatics and earth observation.
-%was devoted to develop the first European infrastructure for Grid computing. 
-The project brought together 21 academic and industry partners, from 15 different
-countries. 
-The project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
+had as the main goal to provide scientific communities, including physics, biology and 
+earth sciences, with intensive computing and large-scale dataset analysis capabilites.
+The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}. 
+Grid technology was the emerging technology to be used in order to address these requirements,
+thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
 based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
@@
 
-Agile methodologies were being introduced at the time
-through the Agile manifesto \cite{agile-manifesto}, as such the project's
+The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, thus not
+considered by the project, and as such the project's
 development design suffered from the methods, tools, techniques and best
@@
 \cite{agile}. Nonetheless, the project focused on the experiences and procedures
-used by different open source software projects, such as Linux and Apache, trying to reach a
-higher maturity level \cite{cmm}.
+used by diverse collaborative open source software projects, such as Linux and Apache, trying to 
+reach a higher maturity level \cite{cmm}.
 
@@
 The source code was available at a private, centralized version control system.
-The code was passed through a manual certification procedure at the time of the release. 
+The code passed through a manual certification procedure at the time of the release. 
 This procedure tried to improve the reliability of the software components by applying an
@@
 arose as the continuation of the {\sl EGEE--III} project, with the mission of establishing and
-maintaining a sustainable European Grid Infrastructure.
-
-{\sl EGI--InSPIRE} not only ensured the coordination of the operation of a secure, reliable 
-European-wide production Grid infrastructure federated from National Grid Initiatives (NGIs)
-that is integrated and interoperates with other Grids worldwide but also foreseed the 
+maintaining a sustainable European Grid Infrastructure, composed by a federation of National 
+Grid Initiatives and interoperable with other Grids worldwide.
+To accomplish this goal the project foreseed the 
 maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
-the {\sl EGI} Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
-coordination of the external provision of software required by {\sl EGI} to 
-form the production infrastructure. For the later it maintained a production--ready 
-Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution), and 
-defined general and component specific quality criteria to be applied to software 
-components made available through this repository.
-The role of {\sl UMD} was to enforce the fulfilment of this set of quality
-criteria \cite{egi-qc} in the software being delivered (i.e.,
-{\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
+the {\sl EGI} Helpdesk \cite{ggus} or the Grid configuration database (GOCDB) \cite{gocdb}, and the 
+management of the software provisioning process \cite{mario}, to validate and distribute the software that conformed
+the production infrastructure. As a result of this process, a production--ready 
+Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution) was
+maintained, comprised by roughly 250 software components supplied by diverse technology providers such 
+as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
+repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The 
+QC defines a set of requirements in different areas: documentation, deployment, security, information
+model and operations.
+{\sl UMD} is still being used and
 deployed in the European scientific e--Infrastructures under a follow--up
@@
 
-In the following section we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
+In the following sections we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
 software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
@@
 documentation completeness, etc. On approval, the candidate change is definitely merged into
-production. Human code reviewing is paramount in the software quality assessment, specially 
+the source code's production version. Human code reviewing is paramount in the software quality assessment, specially 
 important when the SQA testing requirements are fully automated. Secure code reviews are
@@
 tackled using a configuration management tool. These tools allow to have an 
-\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
-definitions. The definition files appear as additional source of documentation, 
-self-documenting code, as they sequentially guide the component's deployment process in multiple
+\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
+definitions. The definition files appear as additional source of documentation
+-- self-documenting code -- as they sequentially guide the component's deployment process in multiple
 platforms. In this regard, {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
@@
 
+\bibitem{agile-manifesto}
+\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org
+
 \bibitem{zhu}
@@
 Information Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/53665\_en.html
+
+\bibitem{gagliardi}
+Gagliardi F., Jones B., Reale M., Burke S., \emph{European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E-science Applications},
+in Calzarossa M.C., Tucci S. (eds) Performance Evaluation of Complex Systems: Techniques and Tools, Lecture Notes in Computer Science, vol 2459. Springer, 2002.
 
@@
 4-6 Dec. 2002.
-
-\bibitem{agile-manifesto}
-\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org
 
@@
 https://ggus.eu/
+
+\bibitem{egi-qc}
+\emph{EGI Quality Criteria} [Online]. Available: 
+https://egi-qc.github.io/
 

### 2018-02-09 paper-ieee-trans.tex
minor typos in DataGrid

@@
 The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}. 
-Grid technology was the emerging technology to be used in order to address these requirements,
+Grid was the emerging technology to be used in order to address these requirements,
 thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
-based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
+based on Globus middleware services \cite{globus}. The project was organized in 12 work packages, from 
 which 5 were devoted to software development coordinated by an 
@@
 
-The authors would like to thanks European Commission with the various funded
-projects.
+{\sl INDIGO--DataCloud} project has received funding from the European Research 
+Council (ERC) under the European Union's Horizon 2020 research and innovation 
+programme (grant agreement n° 777536).
 

### 2018-02-09 paper-ieee-trans.tex
revision corrections

@@
 of new tools and services to match the scientific community requirements has been
-addressed in Europe by a various number of projects dedicated to software development
+addressed in Europe by various projects dedicated to software development
 and e-Infrastructure creation, operation and management.
@@
 first project in 2001. In particular, we will summarize the techniques and procedures
-applied to deliver quality software, hightlighting the main challenges and barriers confronted
-at the time, and how they were totally or partially overcomed. The base of knowledge collected
-throughout these years, sustained by the advances in the area of software engineering,
+applied to deliver quality software, highlighting the main challenges and barriers confronted
+at the time, and how they were partially or totally overcome. The base of knowledge collected
+throughout the years, sustained by the advances in the area of software engineering,
 definitely conformed a significant breakthrough in the reliability of software delivered in the
-European research arena. Latest software development projects founded by the European 
+European research arena. Recent software development projects funded by the European 
 Commission are good evidences of such insights, where the enforcement of Quality Assurance
@@
 \IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
-which supports an unified access to large-scale computing and intense data analysis
+supporting an unified access to large-scale computing and intense data analysis
 applications, has been driven by the needs of scientific communities as part of their
-pan-European research collaborations. One of the well-known examples is the Worldwide 
-Large Hadron Collider (LHC) Computing Grid infrastructure, powered 
-by the particle physics scientific community, built up to satisfy the computing needs
-of processing, storing and analysing the huge quantity of data coming from the 
+pan-European research collaborations. One such example is the Worldwide 
+Large Hadron Collider (LHC) Computing Grid infrastructure, motivated 
+by the particle physics scientific community, build to satisfy the computing needs
+of processing, storing and analysis of the very large amount of data produced by the
 LHC particle accelerator. A significant number of innovation projects and
-technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
-and ultimately by the Horizon 2020 programme \cite{h2020} -- are being held to coordinate and 
-develop distributed solutions that provide a seamless and transparent access to those
-computing and data resources, available for the researchers.
+technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
+and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
+develop software solutions that provide a seamless and transparent access to distributed
+computing and data resources by the researchers.
 
-Major developments were done throughout the years, mainly focused on providing new
-functionalities, leaving aside in the early times the adoption of software engineering methodologies.
-The software delivered was then less tested, resulting in constant
+Major developments were achieved throughout the years focused on providing new
+functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
+The released software was less tested, resulting in a high rate of
 rollbacks and patches applied to production systems. Thereby, the stability and
@@
 
-Soon it become apparent that a balance was needed between new features addition
-and the reliability of the software produced. As a consequence, a substantial part of the
-software lifecycle would have to be devoted to deal with the quality of software being
-delivered. At this point, evolving software engineering practices were
-considered and gradually adopted to define the quality procedures, teams
-organization and pilot testbeds.
+Soon it become apparent the need for a better balance between the addition of new features
+and the reliability of the released software. Consequently, a substantial part of the
+software lifecycle would have to be devoted to the quality of the software being
+delivered. The evolution of software engineering practices were
+considered and gradually adopted to define and implement quality procedures, the organization of teams
+and deployment of pilot testbeds.
 
-By analyzing the European Commission (EC)-funded software projects in the last decade
-we have observed a continuous increase of the prominence and robustness of
-testing and validation procedures. One of many reasons is related to the
+Analysing the European Commission (EC)-funded software projects over the last decade,
+one can observe a continuous increase in the prominence and robustness of software
+testing and validation procedures. One of many reasons, is related to the
 evolution of the software engineering methodologies (creation of the Agile
@@
 
-In this paper, we introduce a set of EC-funded projects
-that go from 2001 up to now, in which the authors of this summary were involved.
-For each project, we describe the challenges arised, the methodologies used and
-the achievements towards software reliability. The reminder of this paper is
+In this paper, we outline a set of EC-funded projects
+ranging from 2001 to now, in which the authors were involved.
+For each project, the challenges, the methodologies and
+the achievements towards software reliability are described.
+
+The reminder of this paper is
 organized as follows. Section \ref{sec:ev} details the evolution of research on
@@
 information about new research trends in software reliability based upon the 
-experiences from the recent {\sl INDIGO--DataCloud} project. Finally, Section \ref{sec:con} concludes.
+experiences from the recent {\sl INDIGO--DataCloud} project. Conclusions are drawn in Section \ref{sec:con}.
 
@@
 
-The EC projects considered in this text are listed chronologically in Table
-\ref{tab:eup}. In most cases, each project paved the way of the subsequent projects,
-following a logic evolution in which the quality and reliability of software have
-increasingly had a more significant role. Experiences gathered during the 
-course of each project support this fact and, as it is shown in Table \ref{tab:feat},
-latest projects are the most committed in terms of software reliability awareness.
-Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting both
+The EC projects considered in this paper are listed chronologically in Table
+\ref{tab:eup}. Each project paved the way of the subsequent projects,
+following a logic evolution in which the role of the quality and reliability of the software have
+consistently increased. The experience gained during the 
+course of each project support this fact as shown in Table \ref{tab:feat}.
+The most recent projects are more committed and aware to software reliability practices.
+Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
 the commonalities and specific features on software reliability.
@@
 The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
-had as the main goal to provide scientific communities, including physics, biology and 
-earth sciences, with intensive computing and large-scale dataset analysis capabilites.
+main goal was to provide scientific communities, including physics, biology and 
+earth sciences, with intensive computing and large-scale dataset analysis capabilities.
 The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}. 
@@
 thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
-based on Globus middleware services \cite{globus}. The project was organized in 12 work packages, from 
-which 5 were devoted to software development coordinated by an 
+based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from 
+which 5 were devoted to software development and coordinated by an 
 Architecture Task Force, supervising the overall design and technical consistency 
@@
 
-Due to having no previous experience in such widely
-collaborative projects, a big effort was spent trying to devise solutions for several
+Without previous experience in such widely
+collaborative projects, a big effort was spent in devising solutions for several
 challenges \cite{datagrid}, namely the 1) communication overhead, as a
-result of the large geographical separation of the parties involved in the
-development tasks, 2) the evolution of the requirements coming from the user
+result of the large geographical separation of the involved parties in the
+development tasks, 2) the fast evolution of the requirements from the user
 communities (50 use cases from the three scientific area), and 3) the lack of a body 
@@
 practices emerging from the new discipline of software engineering
-\cite{agile}. Nonetheless, the project focused on the experiences and procedures
+\cite{agile}. Nevertheless, the project focused on the experiences and procedures
 used by diverse collaborative open source software projects, such as Linux and Apache, trying to 
-reach a higher maturity level \cite{cmm}.
+achieve a higher maturity level \cite{cmm}.
 
@@
 \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
-scientists and engineers from more than 240 institutions in 45 countries to
+scientists and engineers from over 240 institutions in 45 countries to
 provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
-featured the internationalization of the project, embracing worldwide research
+featured the internationalization (outside Europe) of the project, embracing worldwide research
 institutions and user communities. The software to sustain the increasing
@@
 
-The {\sl gLite} middleware \cite{glite} was the ultimate
+The {\sl gLite} middleware \cite{glite} was the
 official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
-re--engineering efforts to converge with {\sl LHC} Computing Grid (LCG--2), Virtual
-Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
+re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
+Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
 development team was comprised of more that 80 people from 12 academic and
-industrial partners, that issued more than 10 thousand bug fixes, 1.7 thousand patches and
+industrial partners, fixing more than 10 thousand bug, 1.7 thousand patches and
 defined over 300 development tasks tracked by the use of bug/task management tools.
@@
 The code passed through a manual certification procedure at the time of the release. 
-This procedure tried to improve the reliability of the software components by applying an
+This procedure aimed to improve the reliability of the software components by applying an
 acceptance criteria plan at each stage of the software release process 
@@
 Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
-arose as the continuation of the {\sl EGEE--III} project, with the mission of establishing and
+was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
 maintaining a sustainable European Grid Infrastructure, composed by a federation of National 
 Grid Initiatives and interoperable with other Grids worldwide.
-To accomplish this goal the project foreseed the 
-maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
-the {\sl EGI} Helpdesk \cite{ggus} or the Grid configuration database (GOCDB) \cite{gocdb}, and the 
-management of the software provisioning process \cite{mario}, to validate and distribute the software that conformed
-the production infrastructure. As a result of this process, a production--ready 
-Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution) was
-maintained, comprised by roughly 250 software components supplied by diverse technology providers such 
+This goal would be accomplished through the
+development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops}, 
+the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the 
+management of the software provisioning process \cite{mario}, to validate and distribute the software to
+the production infrastructure. As a result, this process led to a production--ready 
+Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). The {\sl UMD}
+was a stack of about 250 software components from several technology providers such 
 as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
@@
 {\sl UMD} is still being used and
-deployed in the European scientific e--Infrastructures under a follow--up
-project, the {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
+deployed in the European scientific e--Infrastructures under the follow--up
+project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
 Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
@@
 explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
-project of software development considered in this review. At the time of writing
+project of software development considered in this paper. At the time of writing
 this paper, the project is succeeding in providing solutions to address the existing gaps in
@@
 
-{\sl INDIGO--DataCloud} is the most recent project included in this paper thus it
+{\sl INDIGO--DataCloud} is the most recent project included in this paper thus, it
 emerged from the lessons learned in the previous experiences described in
@@
 success of the solutions implemented. All the above with the additional complexity of
-being framed into large collaborative projects with an extensive list of partners.
+large collaborative projects with an extensive list of partners.
 
-In the following sections we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
+In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the 
 software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
-and metrics gathering definition} will provide guidelines to develop quality 
+and metrics gathering definition}, will provide guidelines to develop quality 
 software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
@@
 the foregoing change-based development process and set up advanced delivery mechanisms.
-The last key pillar is constituted by a \textit{two-factor validation} carried out over the project's pilot testbeds and
+The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
 e-Infrastructure resource provider's early-adoption.
@@
 The software quality policies were initially defined in the first deliverable document of the
-project \cite{indigo-d31} and periodically reviewed to be endowed with applicable insights of 
-software engineering practices, responses to evolving needs emanated from the user communities and 
-feedback obtained from development teams. These policies cover the 1) identification
+project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
+from the user communities and feedback from the development teams. These policies cover the 1) identification
 and description of the \emph{SQA requirements} that the produced software need 
 to comply with, and the 2) \emph{quality metrics}, to monitor each software product's 
-behaviour throughout the development, release and post-release stages.
+behaviour throughout the development, release and post--release stages.
 
@@
 
-The SQA criteria is a collection of conventions and recommendations that pave the way for
-the adequate development, timely deliver and reliable operation of the software components
-produced. It emphasizes the quality requirements and best practices to be applied at the
+The SQA criteria is a set of conventions and recommendations that pave the way for
+the adequate development, timely deliver and reliable operation of the produced software components.
+It emphasizes the quality requirements and best practices to be applied at the
 development phase, such as style and testing compliance or human code reviews, to protect the
-production source code versions. Besides, the criteria promotes the adoption of the software 
+production source code versions. Furthermore, the criteria promotes the adoption of the software 
 by providing minimum requirements of documentation content and stressing the usage of automated
@@
 choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
-over the {\sl INDIGO--DataCloud} software stack.
+in the {\sl INDIGO--DataCloud} software stack.
 
 \item \textit{Unit and Functional testing}.
-Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
-testing} in this context is done enforcing the successful execution of the previous
-functionality tests already present in the code, so to uncover any malfunction introduced by
-the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
+Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
+testing} in this context is accomplished by enforcing the successful execution of the previous
+functionality tests already present in the code to uncover any malfunction introduced by
+the changes. \textit{Unit testing} completes the source code testing evaluation by
 focusing on the code's internal design. The SQA criteria sets a recommended threshold of
 70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
-for each software component over the two major {\sl INDIGO--DataCloud} releases, denoting an incremental 
-trend over the course of the project. As depicted in the Figure, 50\% of the components were over the
+for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental 
+trend over the course of the project. It can be seen that 50\% of the components were over the
 70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.
 
-\item \textit{Integration testing}. Software components usually interact with other services while on
-operation state. Integration testing deals with the interactions among coupled software components or
-parts of a system that cooperates to achieve a given functionality. This type of testing might be 
-complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
-guarantee the overall operation of the component with regards to the services it interfaces with,
+\item \textit{Integration testing}. Software components usually interact with other services during
+operation. Integration testing deals with the interactions among coupled software components or
+parts of a system that cooperate to achieve a given functionality. This type of testing might be 
+complex and, based on the project's experience, difficult to be implemented in an automatic way. The aim is to 
+guarantee the overall operation of the component with regard to the services it interfaces with,
 whenever new functionalities are introduced.
@@
 successfully passed through the testing methods described previously. It implies the human-based
-revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
+revision of the proposed change to discuss its adequacy in terms of scope, objective fulfilment, 
 documentation completeness, etc. On approval, the candidate change is definitely merged into
@@
 The SQA criteria sets the path for the adoption of the developed software by
-defining the documentation content, according to the target audience. These requirements also 
-promote the automation, both in terms of documentation creation and service deployment. The
+defining the documentation content, according to the target audience. This requirement also 
+promotes the automation, both in terms of documentation creation and service deployment. The
 \textit{documentation is treated as code}, using a markup language, automatically rendered and
@@
 \item \textit{Automated Deployment}.
-In order to lower the barriers of software adoption, the SQA criteria requires the automated 
+To lower the barriers of software adoption, the SQA criteria requires the automated 
 deployment of the products delivered as part of the catalogue. Automation in this context is 
@@
 -- self-documenting code -- as they sequentially guide the component's deployment process in multiple
-platforms. In this regard, {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
+platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
 as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
@@
 
-\textit{Release metric} sources are mainly the online repository servers, namely the 
+\textit{Release metric} sources are the online repository servers, namely the 
 Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins 
-server contains as well valuable release data as it is the service where the packages
+server also contains valuable release data as it is the service where the packages
 and containers are being built before being uploaded to the online repositories. Release
-KPIs primarly focus on the frequency and effiency -- mainly rollbacks -- statistics.
+KPIs primarily focus on the frequency and efficiency -- mainly rollbacks -- statistics.
 
 \textit{Maintenance and user support metrics} are key for continuously improving the response
-to issues reported by external users. Feeback is collected from outside helpdesks, such as 
+to issues reported by external users. Feedback is collected from outside helpdesks, such as 
 EGI's GGUS \cite{ggus}, and the GitHub Issues tracker.
@@
 
-The DevOps methods emphasize on exercising SQA techniques to avoid infrastructure
+The DevOps methods emphasize the SQA techniques to avoid infrastructure
 disruption whenever new developments are deployed into production systems. The 
 following DevOps approaches have been progressively adopted throughout the project's
-lifetime, being one of its major outcomes in terms of sofware reliability.
+lifetime, being one of its major outcomes in terms of software reliability.
 
@@
 The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
-for any piece of software produced, the testing requirements defined in the SQA criteria,
+for any piece of produced software, the testing requirements defined in the SQA criteria,
 as described in Section \ref{subsec:sqa}. Such environment requires an automation
@@
 constructs that could lead to security risks. Metrics gathering also takes part of
-the CI pipeline so to have a per-change trend evolution of the common source code 
+the CI pipeline to have a per-change trend evolution of the common source code 
 related metrics, such as SLOC or the cyclomatic complexity.
@@
 metric tracking checks on PR creation or update. The set of required checks contained
-in the CI pipeline report their exit status back to GitHub, preventing the change to be
+in the CI pipeline, report their exit status back to GitHub preventing the change to be
 approved if they are not successfully executed. As each change is validated, the chances
-of early detection of defects augments. Within this scenario, the cost of defect solving
+of early detection of defects increases. Within this scenario, the cost of defect solving
 is dramatically reduced and the reliability of the software solutions improved, as any
-bug or design issue is likely to be detected and subsequently corrected at this phase.
+bug or design issue is likely to be detected and subsequently corrected in this phase.
 
@@
 
-The experience gathered throughout the project with regard to the adoption of
+The experience gathered throughout the project regarding the adoption of
 different DevOps practices is not only useful and suitable for the software related
@@
 The application is then executed and the results compared with a set of reference outputs.
-Thus this pipeline implementation goes a step forward by testing the application
+This pipeline implementation goes a step forward by testing the application
 execution for the latest available Docker image in the catalogue.

### 2018-02-13 paper-ieee-trans.tex

@@
 Large Hadron Collider (LHC) Computing Grid infrastructure, motivated
-by the particle physics scientific community, build to satisfy the computing needs
+by the particle physics scientific community, built to satisfy the computing needs
 of processing, storing and analysis of the very large amount of data produced by the
@@
 software lifecycle would have to be devoted to the quality of the software being
-delivered. The evolution of software engineering practices were
+delivered. At this point, evolving software engineering practices were
 considered and gradually adopted to define and implement quality procedures, the organization of teams
@@
 one can observe a continuous increase in the prominence and robustness of software
-testing and validation procedures. One of many reasons, is related to the
+testing and validation procedures. The
 evolution of the software engineering methodologies (creation of the Agile
@@
 field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization). All of which
+operating--system--level virtualization known as containerization) definitely motivated this trend. All of which
 complemented by the emergence of automation and event-response tools.
@@
 In this paper, we outline a set of EC-funded projects
-ranging from 2001 to now, in which the authors were involved.
+ranging from 2001 to nowadays, in which the authors were involved.
 For each project, the challenges, the methodologies and
@@
 following a logic evolution in which the role of the quality and reliability of the software have
-consistently increased. The experience gained during the 
-course of each project support this fact as shown in Table \ref{tab:feat}.
-The most recent projects are more committed and aware to software reliability practices.
+consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed and aware to software reliability practices.
 Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
@@
 development team was comprised of more that 80 people from 12 academic and
-industrial partners, fixing more than 10 thousand bug, 1.7 thousand patches and
+industrial partners, which issued more than 10 thousand bug, 1.7 thousand patches and
 defined over 300 development tasks tracked by the use of bug/task management tools.
@@
 the production infrastructure. As a result, this process led to a production--ready
-Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). The {\sl UMD}
+Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
 was a stack of about 250 software components from several technology providers such
@@
 
-{\sl INDIGO--DataCloud} is the most recent project included in this paper thus, it
-emerged from the lessons learned in the previous experiences described in
+{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
+leveraged the lessons learned from the previous experiences described in
 Section \ref{sec:ev}. The expertise gathered throughout these years was highly
@@
 into practice, and the compilation and analysis of the appropriate set of metrics to measure the
-success of the solutions implemented. All the above with the additional complexity of
+success of the solutions implemented. All the above with the additional complexity of being framed into
 large collaborative projects with an extensive list of partners.
@@
 project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
-from the user communities and feedback from the development teams. These policies cover the 1) identification
+from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies cover the 1) identification
 and description of the \emph{SQA requirements} that the produced software need

### 2018-02-13 paper-ieee-trans.tex

@@
 pan-European research collaborations. One such example is the Worldwide
-Large Hadron Collider (LHC) Computing Grid infrastructure, motivated
+Large Hadron Collider (LHC) Grid computing infrastructure, motivated
 by the particle physics scientific community, built to satisfy the computing needs
@@
 Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
-development team was comprised of more that 80 people from 12 academic and
+development team was comprised of more than 80 people from 12 academic and
 industrial partners, which issued more than 10 thousand bug, 1.7 thousand patches and
@@
 development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops},
-the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
+the {\sl EGI} Helpdesk \cite{ggus}, or the Grid Operations Configuration DataBase (GOCDB) \cite{gocdb}, and the
 management of the software provisioning process \cite{mario}, to validate and distribute the software to
@@
 Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
-was a stack of about 250 software components from several technology providers such
+was a stack of about 250 software components from several technology providers, such
 as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
-repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
-QC defines a set of requirements in different areas: documentation, deployment, security, information
+repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. QC defines a set of requirements in different areas: documentation, deployment, security, information
 model and operations.
@@
 into practice, and the compilation and analysis of the appropriate set of metrics to measure the
-success of the solutions implemented. All the above with the additional complexity of being framed into
+success of the solutions implemented. All the above achievements were with the additional complexity of being framed into
 large collaborative projects with an extensive list of partners.
 
-In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the
+In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
 software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
-and metrics gathering definition}, will provide guidelines to develop quality
+and metrics gathering definition} provide guidelines to develop quality
 software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
 by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
-process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
+process. A \textit{bottom-up adoption of DevOps principles} promotes automation to enable
 the foregoing change-based development process and set up advanced delivery mechanisms.
@@
 successfully passed through the testing methods described previously. It implies the human-based
-revision of the proposed change to discuss its adequacy in terms of scope, objective fulfilment,
-documentation completeness, etc. On approval, the candidate change is definitely merged into
+revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
+and documentation completeness. On approval, the candidate change is definitely merged into
 the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
@@
 The practical implementation of the CI scenario leveraged on tightly integrated open
-source tools such as GitHub \cite{github} -- as the online source code repository --,
-Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
+source tools such as GitHub \cite{github} as the online source code repository,
+Jenkins \cite{jenkins}, the event-response CI system, and Docker container
 provisioning to provide instant computing power needed for executing the pipeline jobs.
@@
 As DevOps suggests, frequent releases positively affect the reliability of the software as
-they involve fewer changes and usually more correlated. The software updates of
+they involve fewer changes and usually are more correlated. The software updates of
 {\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
@@
 source code before merging and the creation of a new versioned Docker image,
-automatically available in the {\sl INDIGO--DataCloud}’s application repository.
+automatically available in the {\sl INDIGO--DataCloud}'s application repository.
 
@@
 
-After more that a decade of experiences collected throughout the reviewed
+After more than a decade of experiences collected throughout the reviewed
 EU-funded software development and e-Infrastructure management projects, it
@@
 through the preview testbeds or resource centers early adoption -- handled by the
-Staged--rollout process --, alleviates the likelihood of discovering uncovered
+Staged--rollout process -- alleviates the likelihood of discovering uncovered
 issues at production.

### 2018-02-13 paper-ieee-trans.tex
cris comments

--- 2018-02-13 paper-ieee-trans.tex
@@
 applied to deliver quality software, highlighting the main challenges and barriers confronted
-at the time, and how they were partially or totally overcome. The base of knowledge collected
+at the time, and how they were partially or totally overcome. The knowledge base established
 throughout the years, sustained by the advances in the area of software engineering,
@@
 pan-European research collaborations. One such example is the Worldwide
-Large Hadron Collider (LHC) Computing Grid infrastructure, motivated
+Large Hadron Collider (LHC) Computing Grid infrastructure, driven
 by the particle physics scientific community, built to satisfy the computing needs
@@
 LHC particle accelerator. A significant number of innovation projects and
-technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
-and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
+technology transfer activities -- from the European Union's funded 5th, 6th and 7th Framework Programmes,
+and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
 develop software solutions that provide a seamless and transparent access to distributed
@@
 
-Soon it become apparent the need for a better balance between the addition of new features
+Soon it became apparent the need for a better balance between the addition of new features
 and the reliability of the released software. Consequently, a substantial part of the
-software lifecycle would have to be devoted to the quality of the software being
+software lifecycle would have to be devoted to the improvement of the quality of the software being
 delivered. At this point, evolving software engineering practices were
-considered and gradually adopted to define and implement quality procedures, the organization of teams
+considered and gradually adopted to define and implement quality procedures, the organization of development teams
 and deployment of pilot testbeds.
@@
 following a logic evolution in which the role of the quality and reliability of the software have
-consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed and aware to software reliability practices.
+consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
 Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
@@
 The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
-Grid was the emerging technology to be used in order to address these requirements,
+Grid was the emerging technology to be used in order to address their requirements,
 thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
@@
 
-The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, thus not
+The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, not yet
 considered by the project, and as such the project's
-development design suffered from the methods, tools, techniques and best
-practices emerging from the new discipline of software engineering
+development design suffered from the lack of the methods, tools, techniques and best
+practices emerging from these new discipline of software engineering
 \cite{agile}. Nevertheless, the project focused on the experiences and procedures
@@
 \cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
-scientists and engineers from over 240 institutions in 45 countries to
+scientists and engineers from over 240 institutions in 45 countries aiming to
 provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
@@
 institutions and user communities. The software to sustain the increasing
-requirements from the diverse scientific communities would need to develop a
+requirements coming from the diverse scientific communities would need to develop a
 rich set of new services while maintaining a sustainable infrastructure for
@@
 development team was comprised of more that 80 people from 12 academic and
-industrial partners, which issued more than 10 thousand bug, 1.7 thousand patches and
-defined over 300 development tasks tracked by the use of bug/task management tools.
+industrial partners, which issued more than 10 thousand bugs fixes, 1.7 thousand patches and
+defined over 300 development tasks tracked using bug/task management tools.
 
@@
 The code passed through a manual certification procedure at the time of the release.
-This procedure aimed to improve the reliability of the software components by applying an
-acceptance criteria plan at each stage of the software release process
+This procedure aimed to improve the reliability of the software components by applying
+acceptance criteria checks at each stage of the software development lifecycle
 \cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and
@@
 Grid Initiatives and interoperable with other Grids worldwide.
-This goal would be accomplished through the
-development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops},
+This goal would had been accomplished through the
+development and maintenance of some operational tools, such as the Operations Portal \cite{egi-ops},
 the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
-management of the software provisioning process \cite{mario}, to validate and distribute the software to
-the production infrastructure. As a result, this process led to a production--ready
+management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
+the production infrastructures. As a result, this process led to a production--ready
 Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
@@
 as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
-repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
-QC defines a set of requirements in different areas: documentation, deployment, security, information
+repositories, the software had to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The
+QC defined a set of requirements in different areas: documentation, deployment, security, information
 model and operations.
@@
 Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
-helping developers, e--Infrastructures and scientific communities to exploit
+helping developers, e--Infrastructures administrators and scientific communities to exploit
 Cloud computing benefits.
@@
 into practice, and the compilation and analysis of the appropriate set of metrics to measure the
-success of the solutions implemented. All the above with the additional complexity of being framed into
-large collaborative projects with an extensive list of partners.
+quality of the solutions implemented. All the above with the additional complexity of being framed into a
+large collaborative project with an extensive list of partners.
 
@@
 and metrics gathering definition}, will provide guidelines to develop quality
-software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
+software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
 by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
@@
 %quality and reliability aspects throughout the past scientific European
-%experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
+%experiences. Nevertheless (I would remove this), the project???s Quality Assurance procedures are also
 %highly influenced by the insights of current big worldwide collaborations of
@@
 
-The software quality policies were initially defined in the first deliverable document of the
+The initial software quality policies were defined in the first deliverable document of the
 project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
@@
 development phase, such as style and testing compliance or human code reviews, to protect the
-production source code versions. Furthermore, the criteria promotes the adoption of the software
-by providing minimum requirements of documentation content and stressing the usage of automated
+production source code versions. Furthermore, there are provided the minimum requirements on documentation content and it is stresed the importance of the usage of automated
 solutions for software deployment.
@@
 Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
-testing} in this context is accomplished by enforcing the successful execution of the previous
-functionality tests already present in the code to uncover any malfunction introduced by
-the changes. \textit{Unit testing} completes the source code testing evaluation by
+testing} in this context is accomplished by enforcing the definition, and periodic execution, of tests for the fixed issues that ensure these issues are not reintroduced during the development activities.
+ \textit{Unit testing} completes the source code testing evaluation by
 focusing on the code's internal design. The SQA criteria sets a recommended threshold of
@@
 the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
-important when the SQA testing requirements are fully automated. Secure code reviews are
+important when the SQA testing requirements are fully automated. Secure code reviews are also 
 done at this stage, assessing common vulnerabilities from inputs coming from automated linters
@@
 \ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
-It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
+It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
 of the two major releases.
@@
 framework \cite{grimoirelab}. Per-component weekly reports, including the SQA requirement
-completeness, are issued and individually discussed through the different communication channels
+fulfilment, are issued and individually discussed through the different communication channels
 with the development teams. Issue tracking metrics and KPIs are an essential piece of information
@@
 \textit{Release metric} sources are the online repository servers, namely the
-Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins
+Linux packages \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins
 server also contains valuable release data as it is the service where the packages
@@
 source code before merging and the creation of a new versioned Docker image,
-automatically available in the {\sl INDIGO--DataCloud}’s application repository.
+automatically available in the {\sl INDIGO--DataCloud}???s application repository.
 
@@
 The released software is tested in production environments through the
-Staged--Rollout process, before inclusion in the EGI's Cloud Middleware Distribution (CMD).
+Staged--Rollout process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
 Selected resource providers are requested to install
@@
 Council (ERC) under the European Union's Horizon 2020 research and innovation
-programme (grant agreement n° 777536).
+programme (grant agreement n?? 777536).
 
@@
 Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2,
-pp. 115–128, 1997.
+pp. 115???128, 1997.
 
@@
 Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1),
-pp. 33–45, Scientific Publishers OWN, 2006.
+pp. 33???45, Scientific Publishers OWN, 2006.
 
@@
 D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the Grid}, in Grid
-Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63–70,
+Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63???70,
 Eds. John Wiley \& Sons Inc., 2002.
@@
 \bibitem{etics}
-A. Di Meglio, M.-E.Bégin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
+A. Di Meglio, M.-E.B??gin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
 the international software engineering service for the Grid}, in Journal of
@@
 M. David et al, \emph{Validation of Grid Middleware for the European Grid
-Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543–558,
+Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543???558,
 Springer, 2014.
@@
 DisVis: quantifying and visualizing the accessible interaction space of distance
-restrained biomolecular complexes, Bioinformatics 31 (2015) 3222–3224
+restrained biomolecular complexes, Bioinformatics 31 (2015) 3222???3224
 
@@
 Fast and sensitive rigid-body fitting into cryo-EM density maps with PowerFit,
-AIMS Biophys. 2 (2015) 73–87.
+AIMS Biophys. 2 (2015) 73???87.
 

### 2018-02-14 paper-ieee-trans.tex
Update paper-ieee-trans.tex

@@
 DisVis: quantifying and visualizing the accessible interaction space of distance
-restrained biomolecular complexes, Bioinformatics 31 (2015) 3222???3224
+restrained biomolecular complexes, Bioinformatics 31 (2015) 3222--3224
 
@@
 Fast and sensitive rigid-body fitting into cryo-EM density maps with PowerFit,
-AIMS Biophys. 2 (2015) 73???87.
+AIMS Biophys. 2 (2015) 73--87.
 

### 2018-02-14 paper-ieee-trans.tex
Fix reference

--- 2018-02-13 paper-ieee-trans.tex
@@
 branch. Before that, once the software packages have been created, they are uploaded to a
-testing branch. The available automated deployment solution -- see \cite{subsec:sqa} -- checks
+testing branch. The available automated deployment solution -- see Section \ref{subsec:sqa} -- checks
 the installation and configuration of the component relying on the testing repository. On

### 2018-02-16 paper-ieee-trans.tex
added ack

--- 2018-02-14 paper-ieee-trans.tex
@@
 information about new research trends in software reliability based upon the
-experiences from the recent {\sl INDIGO--DataCloud} project. Conclusions are drawn in Section \ref{sec:con}.
+experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section \ref{sec:con}.
 
@@
 consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
-Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
+Table ~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
 the commonalities and specific features on software reliability.
@@
 Grid Initiatives and interoperable with other Grids worldwide.
-This goal would had been accomplished through the
+This has been accomplished through the
 development and maintenance of some operational tools, such as the Operations Portal \cite{egi-ops},
@@
 Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). At that time the {\sl UMD}
-was a stack of about 250 software components from several technology providers such
+was a stack of about 250 software components from several technology providers, such
 as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
@@
 
-In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the
+In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
 software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
-and metrics gathering definition}, will provide guidelines to develop quality
+and metrics gathering definition}, provides guidelines to develop quality
 software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
 by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
-process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
+process. A \textit{bottom-up adoption of DevOps principles} promotes automation to enable
 the foregoing change-based development process and set up advanced delivery mechanisms.
@@
 The practical implementation of the CI scenario leveraged on tightly integrated open
-source tools such as GitHub \cite{github} -- as the online source code repository --,
-Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
+source tools such as GitHub \cite{github}, as the online source code repository,
+Jenkins \cite{jenkins}, the event-response CI system, and Docker container
 provisioning to provide instant computing power needed for executing the pipeline jobs.
@@
 source code before merging and the creation of a new versioned Docker image,
-automatically available in the {\sl INDIGO--DataCloud}???s application repository.
+automatically available in the {\sl INDIGO--DataCloud} application repository.
 
@@
 through the preview testbeds or resource centers early adoption -- handled by the
-Staged--rollout process --, alleviates the likelihood of discovering uncovered
+Staged--rollout process, alleviates the likelihood of discovering uncovered
 issues at production.
@@
 
+{\sl DataGrid - Research and Technological Development for an International Data Grid} project 
+has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
+FP5-IST - Programme for research, technological development and demonstration on a "User-friendly information society, 1998-2002" (IST--2000--25182E)
+
+{\sl EGEE - Enabling grids For E-science} project
+has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
+ FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)
+
+{\sl EGEE - Enabling grids For E-science--II} project
+ has received funding from the European Commission - Community Research and Development Information System (CORDIS) under 
+ FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high-capacity and high-speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)
+
+{\sl EGEE - Enabling grids For E-science--III} project
+  has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
+  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e-Science Grid infrastructures", (INFSO--RI--222667)
+
 {\sl INDIGO--DataCloud} project has received funding from the European Research
 Council (ERC) under the European Union's Horizon 2020 research and innovation
-programme (grant agreement n?? 777536).
+programme (GA--777536).
 

### 2018-02-16 paper-ieee-trans.tex

@@
 source tools such as GitHub \cite{github}, as the online source code repository,
+
 Jenkins \cite{jenkins}, the event-response CI system, and Docker container
@@
 source code before merging and the creation of a new versioned Docker image,
-automatically available in the {\sl INDIGO--DataCloud} application repository.
+automatically available in the {\sl INDIGO--DataCloud} applications repository.
 

### 2018-02-20 paper-ieee-trans.tex
Apply agreed changes

@@
 LHC particle accelerator. A significant number of innovation projects and
-technology transfer activities -- from the European Union's funded 5th, 6th and 7th Framework Programmes,
+technology transfer activities -- from the European Commission (EC)-funded 5th, 6th and 7th Framework Programmes,
 and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
-develop software solutions that provide a seamless and transparent access to distributed
-computing and data resources by the researchers.
+develop software solutions that provide a seamless and transparent access of researchers to distributed
+computing and data resources.
 
@@
 delivered. At this point, evolving software engineering practices were
-considered and gradually adopted to define and implement quality procedures, the organization of development teams
+considered and gradually adopted to define and implement quality procedures, the organization of cross-functional teams
 and deployment of pilot testbeds.
 
-Analysing the European Commission (EC)-funded software projects over the last decade,
+Analysing the EC-funded software projects over the last decade,
 one can observe a continuous increase in the prominence and robustness of software
@@
 consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
-Table ~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
+Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
 the commonalities and specific features on software reliability.
@@
 development design suffered from the lack of the methods, tools, techniques and best
-practices emerging from these new discipline of software engineering
+practices emerging from this new discipline of software engineering
 \cite{agile}. Nevertheless, the project focused on the experiences and procedures
@@
 Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
-development team was comprised of more that 80 people from 12 academic and
-industrial partners, which issued more than 10 thousand bugs fixes, 1.7 thousand patches and
+development team was comprised of more than 80 people from 12 academic and
+industrial partners, which issued more than 10 thousand bug fixes, 1.7 thousand patches and
 defined over 300 development tasks tracked using bug/task management tools.
@@
 This procedure aimed to improve the reliability of the software components by applying
-acceptance criteria checks at each stage of the software development lifecycle
+acceptance criteria checks at the pre-release stage
 \cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and
@@
 Grid Initiatives and interoperable with other Grids worldwide.
-This has been accomplished through the
-development and maintenance of some operational tools, such as the Operations Portal \cite{egi-ops},
+This goal would be accomplished through the
+development and maintenance of various operational tools, such as the Operations Portal \cite{egi-ops},
 the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
@@
 Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
-helping developers, e--Infrastructures administrators and scientific communities to exploit
+helping developers, e--Infrastructure administrators and scientific communities to exploit
 Cloud computing benefits.
@@
 into practice, and the compilation and analysis of the appropriate set of metrics to measure the
-quality of the solutions implemented. All the above with the additional complexity of being framed into a
-large collaborative project with an extensive list of partners.
+quality of the solutions implemented. All the above with the additional complexity of being framed into
+large collaborative projects with an extensive list of partners.
 
@@
 %quality and reliability aspects throughout the past scientific European
-%experiences. Nevertheless (I would remove this), the project???s Quality Assurance procedures are also
+%experiences. Nevertheless (I would remove this), the project's Quality Assurance procedures are also
 %highly influenced by the insights of current big worldwide collaborations of
@@
 
-The initial software quality policies were defined in the first deliverable document of the
+The software quality policies were initially defined in the first deliverable document of the
 project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
@@
 development phase, such as style and testing compliance or human code reviews, to protect the
-production source code versions. Furthermore, there are provided the minimum requirements on documentation content and it is stresed the importance of the usage of automated
+production source code versions. Furthermore, the criteria promotes the adoption of the software 
+by providing minimum requirements of documentation content and stressing the usage of automated
 solutions for software deployment.
@@
 Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
-testing} in this context is accomplished by enforcing the definition, and periodic execution, of tests for the fixed issues that ensure these issues are not reintroduced during the development activities.
+testing} in this context is accomplished by enforcing the definition and periodic execution
+of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development activities.
+
  \textit{Unit testing} completes the source code testing evaluation by
@@
 revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
-and documentation completeness. On approval, the candidate change is definitely merged into
+and documentation fulfilment. On approval, the candidate change is definitely merged into
 the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
@@
 \textit{Release metric} sources are the online repository servers, namely the
-Linux packages \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins
+Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins
 server also contains valuable release data as it is the service where the packages
@@
 source code before merging and the creation of a new versioned Docker image,
-automatically available in the {\sl INDIGO--DataCloud} applications repository.
+automatically available in the {\sl INDIGO--DataCloud}'s application repository.
 
@@
 through the preview testbeds or resource centers early adoption -- handled by the
-Staged--rollout process, alleviates the likelihood of discovering uncovered
+Staged--rollout process -- alleviates the likelihood of discovering uncovered
 issues at production.
@@
 Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2,
-pp. 115???128, 1997.
+pp. 115--128, 1997.
 
@@
 Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1),
-pp. 33???45, Scientific Publishers OWN, 2006.
+pp. 33--45, Scientific Publishers OWN, 2006.
 
@@
 D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the Grid}, in Grid
-Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63???70,
+Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63--70,
 Eds. John Wiley \& Sons Inc., 2002.
@@
 M. David et al, \emph{Validation of Grid Middleware for the European Grid
-Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543???558,
+Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543--558,
 Springer, 2014.

### 2018-02-20 paper-ieee-trans.tex
homogeneize hyphens

@@
 addressed in Europe by various projects dedicated to software development
-and e-Infrastructure creation, operation and management.
+and e--Infrastructure creation, operation and management.
 % the creation of software development and e-Infrastructure coordination dedicated projects.
 This work examines the evolution of the software engineering methodologies used in past and
-present European Commission-funded projects to illustrate how the research on software
+present European Commission--funded projects to illustrate how the research on software
 reliability has progressed in Europe over the last 15 years following the foundation of the
@@
 
-\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
-supporting an unified access to large-scale computing and intense data analysis
+\IEEEPARstart{D}uring the last 15 years the provision of an European e--Infrastructure,
+supporting an unified access to large--scale computing and intense data analysis
 applications, has been driven by the needs of scientific communities as part of their
-pan-European research collaborations. One such example is the Worldwide
+pan--European research collaborations. One such example is the Worldwide
 Large Hadron Collider (LHC) Computing Grid infrastructure, driven
@@
 rollbacks and patches applied to production systems. Thereby, the stability and
-reliability of the e-Infrastructures was often compromised, affecting
+reliability of the e--Infrastructures was often compromised, affecting
 negatively their exploitation by the scientific communities.
@@
 delivered. At this point, evolving software engineering practices were
-considered and gradually adopted to define and implement quality procedures, the organization of cross-functional teams
+considered and gradually adopted to define and implement quality procedures, the organization of cross--functional teams
 and deployment of pilot testbeds.
 
-Analysing the EC-funded software projects over the last decade,
+Analysing the EC--funded software projects over the last decade,
 one can observe a continuous increase in the prominence and robustness of software
@@
 operating--system--level virtualization known as containerization) definitely motivated this trend. All of which
-complemented by the emergence of automation and event-response tools.
+complemented by the emergence of automation and event--response tools.
 
-In this paper, we outline a set of EC-funded projects
+In this paper, we outline a set of EC--funded projects
 ranging from 2001 to nowadays, in which the authors were involved.
@@
 organized as follows. Section \ref{sec:ev} details the evolution of research on
-software reliability in EC-projects. Section \ref{sec:ntsr} provides
+software reliability in EC--projects. Section \ref{sec:ntsr} provides
 information about new research trends in software reliability based upon the
@@
 
-\section{Evolution of research on software reliability in EC-projects}
+\section{Evolution of research on software reliability in EC--projects}
 \label{sec:ev}
@@
 % \extrarowheight as needed to properly center the text within the cells
-\caption{List of EC-projects}
+\caption{List of EC--projects}
 \label{tab:eup}
@@
      & EGEE I, II, III &
-Enabling Grids for E-sciencE & 2004--2010\\
+Enabling Grids for E--sciencE & 2004--2010\\
 \begin{minipage}{.3\textwidth}
@@
 \end{minipage}
-     & EGI-Inspire &
-Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
+     & EGI--Inspire &
+Integrated Sustainable Pan--European Infrastructure for Researchers in Europe
  & 2010--2014\\
@@
 main goal was to provide scientific communities, including physics, biology and
-earth sciences, with intensive computing and large-scale dataset analysis capabilities.
+earth sciences, with intensive computing and large--scale dataset analysis capabilities.
 The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
@@
 % \extrarowheight as needed to properly center the text within the cells
-\caption{Features on software reliability in EC-projects}
+\caption{Features on software reliability in EC--projects}
 \label{tab:feat}
@@
 Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
-Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Auto--generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
 DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
@@
 This procedure aimed to improve the reliability of the software components by applying
-acceptance criteria checks at the pre-release stage
+acceptance criteria checks at the pre--release stage
 \cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and
@@
 integrated different technologies and tools in order to provide automated configuration,
-build and testing capabilities, as well as auto-generated documentation and
+build and testing capabilities, as well as auto--generated documentation and
 software metrics gathering such as Source Lines Of Code (SLOC), complexity and
@@
 by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
-process. A \textit{bottom-up adoption of DevOps principles} promotes automation to enable
-the foregoing change-based development process and set up advanced delivery mechanisms.
-The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
-e-Infrastructure resource provider's early-adoption.
+process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
+the foregoing change--based development process and set up advanced delivery mechanisms.
+The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
+e--Infrastructure resource provider's early--adoption.
 
@@
 project. Code style standards are enforced for every software component in the project stack. Community
-most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
+most--adopted or \textit{de--facto standards} are initially recommended, however the development teams eventually
 choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
@@
 Represents the last step in the change management pipeline, once the candidate change has
-successfully passed through the testing methods described previously. It implies the human-based
+successfully passed through the testing methods described previously. It implies the human--based
 revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
-and documentation fulfilment. On approval, the candidate change is definitely merged into
+and documentation completeness. On approval, the candidate change is definitely merged into
 the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
@@
 definitions. The definition files appear as additional source of documentation
--- self-documenting code -- as they sequentially guide the component's deployment process in multiple
+-- self--documenting code -- as they sequentially guide the component's deployment process in multiple
 platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
@@
 
-\textit{Development metrics} are obtained programatically from several sources, namely GitHub
+\textit{Development metrics} are obtained programmatically from several sources, namely GitHub
 API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
-framework \cite{grimoirelab}. Per-component weekly reports, including the SQA requirement
+framework \cite{grimoirelab}. Per--component weekly reports, including the SQA requirement
 fulfilment, are issued and individually discussed through the different communication channels
@@
 constructs that could lead to security risks. Metrics gathering also takes part of
-the CI pipeline to have a per-change trend evolution of the common source code
+the CI pipeline to have a per--change trend evolution of the common source code
 related metrics, such as SLOC or the cyclomatic complexity.
@@
 
-Jenkins \cite{jenkins}, the event-response CI system, and Docker container
+Jenkins \cite{jenkins}, the event--response CI system, and Docker container
 provisioning to provide instant computing power needed for executing the pipeline jobs.
@@
 a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
-successful execution of the CI pipeline previously described. Consequently, CI-validated
+successful execution of the CI pipeline previously described. Consequently, CI--validated
 software components are automatically delivered in online repositories without the need of
@@
 and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
-package pipeline} intentionally delivers the packages produced in a pre-production -- preview --
+package pipeline} intentionally delivers the packages produced in a pre--production -- preview --
 branch. Before that, once the software packages have been created, they are uploaded to a
@@
 After more than a decade of experiences collected throughout the reviewed
-EU-funded software development and e-Infrastructure management projects, it
+EU--funded software development and e--Infrastructure management projects, it
 becomes apparent that the European software produced for scientific purposes is
@@
 delivery of packages. Thereby, safe and frequent software releases are doable, reducing
-the impact on production systems. In this regard, a pre-production validation,
+the impact on production systems. In this regard, a pre--production validation,
 through the preview testbeds or resource centers early adoption -- handled by the
@@
 process that permits timely reactions to discover and fix defects throughout the
-development, release and post-release or maintenance stages. Conscientious metric
+development, release and post--release or maintenance stages. Conscientious metric
 tracking and response sits at the core of an appropriate software maturity, as
@@
 has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-FP5-IST - Programme for research, technological development and demonstration on a "User-friendly information society, 1998-2002" (IST--2000--25182E)
+FP5-IST - Programme for research, technological development and demonstration on a "User--friendly information society, 1998-2002" (IST--2000--25182E)
 
-{\sl EGEE - Enabling grids For E-science} project
+{\sl EGEE - Enabling grids For E--science} project
 has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
@@
 
-{\sl EGEE - Enabling grids For E-science--II} project
+{\sl EGEE - Enabling grids For E--science--II} project
  has received funding from the European Commission - Community Research and Development Information System (CORDIS) under 
- FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high-capacity and high-speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)
+ FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high--capacity and high--speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)
 
-{\sl EGEE - Enabling grids For E-science--III} project
+{\sl EGEE - Enabling grids For E--science--III} project
   has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e-Science Grid infrastructures", (INFSO--RI--222667)
+  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e--Science Grid infrastructures", (INFSO--RI--222667)
 
@@
 \bibitem{zhu}
-L. Zhu and L. Bass and G. Champlin-Scharff, \emph{DevOps and Its Practices},
-IEEE Software, vol. 33, no. 3, pp. 32--34, May-June 2016.
+L. Zhu and L. Bass and G. Champlin--Scharff, \emph{DevOps and Its Practices},
+IEEE Software, vol. 33, no. 3, pp. 32--34, May--June 2016.
 
@@
 \bibitem{gagliardi}
-Gagliardi F., Jones B., Reale M., Burke S., \emph{European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E-science Applications},
+Gagliardi F., Jones B., Reale M., Burke S., \emph{European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E--science Applications},
 in Calzarossa M.C., Tucci S. (eds) Performance Evaluation of Complex Systems: Techniques and Tools, Lecture Notes in Computer Science, vol 2459. Springer, 2002.
@@
 \bibitem{datagrid}
-L. Momtahan and A. Martin, \emph{e-Science Experiences: Software Engineering
-Practice and the EU DataGrid}, in Proc. Asia-Pacific Software Engineering
+L. Momtahan and A. Martin, \emph{e--Science Experiences: Software Engineering
+Practice and the EU DataGrid}, in Proc. Asia--Pacific Software Engineering
 Conference, Gold Coast, Queensland, Australia, pp. 269-275, IEEE Press,
@@
 Paulk et al., \emph{Capability maturity model for software}, Software
-Engineering Institute, report CMU/SEI-93-TR-24, ESC-TR-93-177, Feb. 1993.
+Engineering Institute, report CMU/SEI-93-TR-24, ESC--TR-93-177, Feb. 1993.
 
 \bibitem{cordis:egee}
-\emph{Enabling Grids for E-sciencE (EGEE)} project, European Community
+\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community
 Research and Development Information Service (CORDIS) [Online]. Available:
@@
 \bibitem{cordis:egee2}
-\emph{Enabling Grids for E-sciencE-II (EGEE-II)} project, European Community
+\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community
 Research and Development Information Service (CORDIS) [Online]. Available:
@@
 \bibitem{cordis:egee3}
-\emph{Enabling Grids for E-sciencE-III (EGEE-III)} project, European Community
+\emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
 Research and Development Information Service (CORDIS) [Online]. Available:
@@
 \bibitem{egee:acceptance-criteria}
-\emph{Definition and Documentation of the Revised Software Life-Cycle Process},
-Milestone MSA3.4.2, 2010, EGEE-III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1\_4.pdf
+\emph{Definition and Documentation of the Revised Software Life--Cycle Process},
+Milestone MSA3.4.2, 2010, EGEE--III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE--III--MSA3.4.2-1062487-v1\_4.pdf
 
@@
 \bibitem{cordis:etics}
-\emph{E-Infrastructure for Testing, Integration and Configuration of Software
+\emph{E--Infrastructure for Testing, Integration and Configuration of Software
 (ETICS)} project, European Community Research and Development Information
@@
 \bibitem{cordis:etics2}
-\emph{E-Infrastructure for Testing, Integration and Configuration of Software -
+\emph{E--Infrastructure for Testing, Integration and Configuration of Software -
 Phase 2 (ETICS 2)} project, European Community Research and Development
@@
 \bibitem{cordis:egi-inspire}
-\emph{European Grid Initiative: Integrated Sustainable Pan-European
-Infrastructure for Researchers in Europe (EGI-InSPIRE)} project, European
+\emph{European Grid Initiative: Integrated Sustainable Pan--European
+Infrastructure for Researchers in Europe (EGI--InSPIRE)} project, European
 Community Research and Development Information Service (CORDIS) [Online]. Available:
@@
 \bibitem{cordis:egi-engage}
-\emph{Engaging the EGI Community towards an Open Science Commons (EGI-ENGAGE)}
+\emph{Engaging the EGI Community towards an Open Science Commons (EGI--ENGAGE)}
 project, European Community Research and Development Information Service
@@
 \bibitem{powerfit} G.C.P. Van Zundert, A.M.J.J. Bonvin,
-Fast and sensitive rigid-body fitting into cryo-EM density maps with PowerFit,
+Fast and sensitive rigid--body fitting into cryo--EM density maps with PowerFit,
 AIMS Biophys. 2 (2015) 73--87.
@@
 assurance tasks within different European advanced computing research projects
-such as INDIGO--DataCloud and EGI-Engage. Currently involved in EOSC-hub project
-and participating in the software quality work packages of DEEP-HybridDataCloud
-and eXtreme-DataCloud.
+such as INDIGO--DataCloud and EGI--Engage. Currently involved in EOSC--hub project
+and participating in the software quality work packages of DEEP--HybridDataCloud
+and eXtreme--DataCloud.
 \end{IEEEbiography}
@@
 is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
-Member of DEEP-HybridDataCloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC-hub project and
+Member of DEEP--HybridDataCloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC--hub project and
 of the Portuguese National Computing Distributed Infrastructure.
@@
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/cris_foto_good}}]{Doina Cristina Duma}
-is the leader of the Distributed Systems group at the INFN National Center (CNAF), providing the core operational support for the INFN-wide Grid infrastructure and for the CNAF Cloud infrastructure.
-She has a long experience in managing distributed e-infrastructures, being involved since 2003 in major European projects such as DataGrid, EGEE (I, II, III), EGI-Inspire and EGI-Engage.
+is the leader of the Distributed Systems group at the INFN National Center (CNAF), providing the core operational support for the INFN--wide Grid infrastructure and for the CNAF Cloud infrastructure.
+She has a long experience in managing distributed e--infrastructures, being involved since 2003 in major European projects such as DataGrid, EGEE (I, II, III), EGI--Inspire and EGI--Engage.
 She was the Release Manager of the EMI (European Middleware Initiative) and INDIGO - DataCloud European project project and having as main responsibilities to manage all release activities, including release planning, building, deploying, reporting, establishing and maintaining the overall release timeline and milestones.

### 2018-02-20 paper-ieee-trans.tex

@@
 
-{\sl DataGrid - Research and Technological Development for an International Data Grid} project 
-has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-FP5-IST - Programme for research, technological development and demonstration on a "User--friendly information society, 1998-2002" (IST--2000--25182E)
+%{\sl DataGrid - Research and Technological Development for an International Data Grid} project 
+%has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
+%FP5-IST - Programme for research, technological development and demonstration on a "User--friendly information society, 1998-2002" (IST--2000--25182E)
+%{\sl EGEE - Enabling grids For E--science} project
+%has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
+% FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)
+%{\sl EGEE - Enabling grids For E--science--II} project
+% has received funding from the European Commission - Community Research and Development Information System (CORDIS) under 
+% FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high--capacity and high--speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)
+%{\sl EGEE - Enabling grids For E--science--III} project has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
+%  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e--Science Grid infrastructures", (INFSO--RI--222667)
 
-{\sl EGEE - Enabling grids For E--science} project
-has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
- FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)
+{\sl DataGrid - Research and Technological Development for an International Data Grid} project has received funding from the European Union's Fifth Framework Programme under grant agreement IST--2000--25182E.
 
-{\sl EGEE - Enabling grids For E--science--II} project
- has received funding from the European Commission - Community Research and Development Information System (CORDIS) under 
- FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high--capacity and high--speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)
+{\sl EGEE - Enabling grids For E--science} project has received funding from the European Union's Sixth Framework Programme under grant agreement INFSO--RI--508833.
 
-{\sl EGEE - Enabling grids For E--science--III} project
-  has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e--Science Grid infrastructures", (INFSO--RI--222667)
+{\sl EGEE - Enabling grids For E--science--II} project has received funding from the European Union's Sixth Framework Programme under grant agreement INFSO--RI--031688.
 
-{\sl INDIGO--DataCloud} project has received funding from the European Research
-Council (ERC) under the European Union's Horizon 2020 research and innovation
-programme (GA--777536).
+{\sl EGEE - Enabling grids For E--science--III} project has received funding from the European Union's Seventh Framework Programme under grant agreement INFSO--RI--222667.
+
+{\sl ETICS - E--Infrastructure for Testing, Integration and Configuration of Software} project has received funding from the European Union's Sixth Framework Programme under grant agreement INFSO--RI--026753.
+
+{\sl ETICS - E--Infrastructure for Testing, Integration and Configuration of Software - Phase 2} project has received funding from the European Union's Seventh Framework Programme under grant agreement INFSO--RI--223782.
+
+{\sl EGI--InSPIRE - European Grid Initiative: Integrated Sustainable Pan--European Infrastructure for Researchers in Europe} project has received funding from the European Union's Seventh Framework Programme under grant agreement INFSO--RI--261323.
+
+{\sl EGI--Engage} project has received funding from the European Union's Horizon 2020 research and innovation programme under grant agreement RIA--654142.
+
+{\sl INDIGO--DataCloud} project has received funding from the European Union's Horizon 2020 research and innovation programme under grant agreement RIA--653549.
 

### 2018-05-17 paper-ieee-trans.tex
added part of the English corrections

@@
 
-Major developments were achieved throughout the years focused on providing new
+Major developments were carried out throughout the years focused on providing new
 functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
@@
 and the reliability of the released software. Consequently, a substantial part of the
-software lifecycle would have to be devoted to the improvement of the quality of the software being
+software lifecycle was devoted to the improvement of the quality of the software being
 delivered. At this point, evolving software engineering practices were
@@
 The EC projects considered in this paper are listed chronologically in Table
-\ref{tab:eup}. Each project paved the way of the subsequent projects,
-following a logic evolution in which the role of the quality and reliability of the software have
+\ref{tab:eup}. Each project paved the way for the following ones, according to a logic evolution in which the role of the quality and reliability of the software have
 consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
@@
 
-\begin{table}[!h]
+\begin{table*}[!h]
 %% increase table row spacing, adjust to taste
@@
 %% than the plain LaTeX2e tabular which is used here.
-\begin{adjustbox}{max width=0.5\textwidth}
+%\begin{adjustbox}{max width=0.5\textwidth}
 \begin{tabular}{llllllll}
@@
 \end{tabular}
-\end{adjustbox}
-\end{table}
+%\end{adjustbox}
+\end{table*}
 
@@
 institutions and user communities. The software to sustain the increasing
-requirements coming from the diverse scientific communities would need to develop a
+requirements coming from the diverse scientific communities needed developing a
 rich set of new services while maintaining a sustainable infrastructure for
@@
 into practice, and the compilation and analysis of the appropriate set of metrics to measure the
-quality of the solutions implemented. All the above with the additional complexity of being framed into
+quality of the implemented solutions. All the above with the additional complexity of being framed into
 large collaborative projects with an extensive list of partners.
@@
 software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
-and metrics gathering definition}, provides guidelines to develop quality
+and metrics gathering definition} provides guidelines to develop quality
 software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
-by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
+by a set of metrics, allowing prompt reactions to detected malfunctions throughout the
 process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
@@
 The software quality policies were initially defined in the first deliverable document of the
-project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
-from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies cover the 1) identification
+project \cite{indigo-d31}. The policies were reviewed periodically taking into account the requirements
+from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies covered the 1) identification
 and description of the \emph{SQA requirements} that the produced software need
@@
 
-The SQA criteria is a set of conventions and recommendations that pave the way for
+The SQA criteria are a set of conventions and recommendations that pave the way for
 the adequate development, timely deliver and reliable operation of the produced software components.
@@
 development phase, such as style and testing compliance or human code reviews, to protect the
-production source code versions. Furthermore, the criteria promotes the adoption of the software 
+production source code versions. Furthermore, the criteria promote the adoption of the software 
 by providing minimum requirements of documentation content and stressing the usage of automated
@@
  \textit{Unit testing} completes the source code testing evaluation by
-focusing on the code's internal design. The SQA criteria sets a recommended threshold of
+focusing on the code's internal design. The SQA criteria set a recommended threshold of
 70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
@@
 \item \textit{Documentation}.
-The SQA criteria sets the path for the adoption of the developed software by
+The SQA criteria set the path for the adoption of the developed software by
 defining the documentation content, according to the target audience. This requirement also
@@
 \item \textit{Automated Deployment}.
-To lower the barriers of software adoption, the SQA criteria requires the automated
+To lower the barriers of software adoption, the SQA criteria require the automated
 deployment of the products delivered as part of the catalogue. Automation in this context is
@@
 \label{subsec:ci}
-The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
+The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforced,
 for any piece of produced software, the testing requirements defined in the SQA criteria,
@@
 
-The \textit{Docker container pipeline} uses instead the automated deployment solution to
+The \textit{Docker container pipeline} uses the automated deployment solution to
 actually install and configure the software component in the container image. Once this is

### 2018-05-19 figure2.R
added new file to produce figure2

@@ -1,0 +1,27 @@
+datafn<-read.csv('data/figure2n.csv', header=TRUE)
+names(datafn)
+
+val1<-datafn[as.character(datafn$Release)==unique(as.character(datafn$Release))[1],]
+val1$Avg
+
+val2<-datafn[as.character(datafn$Release)==unique(as.character(datafn$Release))[2],]
+val2$Avg
+       
+windows()
+ggplot(datafn,aes(x = Product,y = Value)) + 
+  geom_bar(aes(fill = Release),stat = "identity",position = "dodge") +
+  geom_line() + 
+  geom_hline(yintercept = val1$Avg[1], color="#CC6666", size=1) +
+  geom_text(aes(0,val1$Avg[1],label = val1$Avg[1], hjust= -17, vjust = +1)) + 
+  geom_hline(yintercept = val2$Avg[1], color="#9999CC", size=1) +
+  geom_text(aes(0,val2$Avg[1],label = val2$Avg[1], hjust= -17, vjust = +1)) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 60, hjust=1),
+        axis.line = element_line(colour = "black"),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        panel.border = element_blank(),
+        panel.background = element_blank(),
+        legend.position="top") + 
+  labs(x='Products',y='Unit Testing Coverage')+
+  scale_fill_manual('Releases', values=c("#CC6666", "#9999CC")) 

### 2018-05-23 figure2.R
redo fig 2

@@ -1,27 +1,48 @@
 datafn<-read.csv('data/figure2n.csv', header=TRUE)
 names(datafn)
 
+rel1<- as.character(datafn$Release)[1]
+rel1
 val1<-datafn[as.character(datafn$Release)==unique(as.character(datafn$Release))[1],]
 val1$Avg
 
+rel2<-as.character(datafn$Release)[2]
+rel2
 val2<-datafn[as.character(datafn$Release)==unique(as.character(datafn$Release))[2],]
 val2$Avg
-       
-windows()
-ggplot(datafn,aes(x = Product,y = Value)) + 
-  geom_bar(aes(fill = Release),stat = "identity",position = "dodge") +
-  geom_line() + 
-  geom_hline(yintercept = val1$Avg[1], color="#CC6666", size=1) +
-  geom_text(aes(0,val1$Avg[1],label = val1$Avg[1], hjust= -17, vjust = +1)) + 
-  geom_hline(yintercept = val2$Avg[1], color="#9999CC", size=1) +
-  geom_text(aes(0,val2$Avg[1],label = val2$Avg[1], hjust= -17, vjust = +1)) +
-  theme_bw() +
-  theme(axis.text.x = element_text(angle = 60, hjust=1),
+    
+Avg<-c(rel1,rel2)   
+library(ggplot2)
+
+ggplot(datafn, aes(fill=Release , x=Product, y=Value)) +
+  geom_bar(stat="identity", position="dodge",  size=1, width=0.5) +
+  xlab("Products") +
+  ylab("Unite Test Coverage %") +
+  scale_y_continuous(limits = c(0, 100))+
+  geom_hline(aes(yintercept=val1$Avg[1],  
+                 linetype='MidnightBlue'),
+                colour='red',
+                 size=1, alpha=1) +
+  geom_hline(aes(yintercept=val2$Avg[1],  
+                 linetype='ElectricIndigo'), 
+             colour='blue',
+                 size=1, alpha=1)  +
+    scale_fill_manual(name='Releases', values=c("blue", "red")) +
+  scale_linetype_manual(name="Avg", 
+                     values=c(2, 3),
+                     guide = guide_legend(override.aes = 
+                                            list(color = c("blue", "red"))),
+                     labels = c(paste(val2$Avg[1], '% INDIGO-2',sep=''),
+                                paste(val1$Avg[1], '% INDIGO-1',sep=''))) +
+  theme(legend.key = element_rect(fill = "white", colour = "white"),
+        axis.text=element_text(size=12),
[…]

### 2018-05-23 figure2.R
added changes

@@ -6,34 +6,41 @@ rel1
 val1<-datafn[as.character(datafn$Release)==unique(as.character(datafn$Release))[1],]
 val1$Avg
 
+rel1avg<-paste(val1$Avg[1], '% INDIGO-1',sep='')
+
 rel2<-as.character(datafn$Release)[2]
 rel2
 val2<-datafn[as.character(datafn$Release)==unique(as.character(datafn$Release))[2],]
 val2$Avg
-    
+
+rel2avg<-paste(val2$Avg[1], '% INDIGO-2',sep='')
+
 Avg<-c(rel1,rel2)   
 library(ggplot2)
 
 ggplot(datafn, aes(fill=Release , x=Product, y=Value)) +
   geom_bar(stat="identity", position="dodge",  size=1, width=0.5) +
   xlab("Products") +
-  ylab("Unite Test Coverage %") +
+  ylab("Unit Test Coverage %") +
   scale_y_continuous(limits = c(0, 100))+
   geom_hline(aes(yintercept=val1$Avg[1],  
                  linetype='MidnightBlue'),
-                colour='red',
+                 colour='darkred',
                  size=1, alpha=1) +
   geom_hline(aes(yintercept=val2$Avg[1],  
                  linetype='ElectricIndigo'), 
-             colour='blue',
+                 colour='blue',
                  size=1, alpha=1)  +
-    scale_fill_manual(name='Releases', values=c("blue", "red")) +
+  scale_fill_manual(name='Releases', values=c("blue", "darkred"),
+                    breaks = c(rel1,
+                               rel2)) +
   scale_linetype_manual(name="Avg", 
-                     values=c(2, 3),
+                     values=c(2, 2),
                      guide = guide_legend(override.aes = 
-                                            list(color = c("blue", "red"))),
-                     labels = c(paste(val2$Avg[1], '% INDIGO-2',sep=''),
-                                paste(val1$Avg[1], '% INDIGO-1',sep=''))) +
+                                            list(color = c("darkred", 
+                                                           "blue"))),
+                     labels = c(rel1avg,
+                                rel2avg)) +
   theme(legend.key = element_rect(fill = "white", colour = "white"),
         axis.text=element_text(size=12),
         axis.title=element_text(size=12),

### 2018-05-23 figure3.R

@@ -1,0 +1,58 @@
+datafn<-read.csv('data/figure3.csv', header=TRUE)
+names(datafn)
+str(datafn)
+
+i1<-datafn[datafn$Release=='INDIGO-1',]
+lm_fit1 <- lm(Number~Week, data=i1)
+summary(lm_fit1)
+
+predicted_df1 <- data.frame(number_pred = predict(lm_fit1, i1))
+predicted_df1
+
+i2<-datafn[datafn$Release=='INDIGO-2',]
+lm_fit2 <- lm(Number~Week, data=i2)
+summary(lm_fit2)
+
+predicted_df2 <- data.frame(number_pred = predict(lm_fit2, i2))
+predicted_df2
+
+predicted_df <- rbind(predicted_df1, predicted_df2)
+datafn$Lm <-rbind(predicted_df1, predicted_df2)
+
+names(datafn)
+str(datafn)
+
+library(ggplot2)
+
+ggplot(datafn, aes(x=Week, y=Number)) +
+  geom_point(data=datafn[datafn$Release=='INDIGO-1',],
+             aes(x=Week, y=Number, color='MidnightBlue'), 
+             size=3) +
+  geom_point(data=datafn[datafn$Release=='INDIGO-2',],
+             aes(x=Week, y=Number, color='ElectricIndigo'), 
+             size=3) +
+  geom_line(data=datafn[datafn$Release=='INDIGO-1',],
+             aes(x=Week, y=Lm, color='MidnightBlue'), 
+             size=1) +
+  geom_line(data=datafn[datafn$Release=='INDIGO-2',],
+             aes(x=Week, y=Lm, color='ElectricIndigo'), 
+             size=1) +
+  xlab("Weeks") +
+  ylab("Number of Software Products") +
+  scale_y_continuous(limits = c(0, 25)) +
+  scale_x_continuous(limits = c(0, 34), 
+                     breaks=c(0, 5, 10, 15, 20, 25, 30)) +
+  scale_colour_manual(name='Releases', values=c(MidnightBlue="blue",
+                                                ElectricIndigo="darkred"),
+                      breaks=c('MidnightBlue',
+                              'ElectricIndigo')) +
+  theme(legend.key = element_rect(fill = "white", colour = "white"),
+        axis.text=element_text(size=12),
+        axis.title=element_text(size=12),
+        axis.line = element_line(colour = "black"),
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
[…]

### 2018-05-23 figure3.R
added changes

@@ -42,8 +42,8 @@ ggplot(datafn, aes(x=Week, y=Number)) +
   scale_y_continuous(limits = c(0, 25)) +
   scale_x_continuous(limits = c(0, 34), 
                      breaks=c(0, 5, 10, 15, 20, 25, 30)) +
-  scale_colour_manual(name='Releases', values=c(MidnightBlue="blue",
-                                                ElectricIndigo="darkred"),
+  scale_colour_manual(name='Releases', values=c(MidnightBlue="darkred",
+                                                ElectricIndigo="blue"),
                       breaks=c('MidnightBlue',
                               'ElectricIndigo')) +
   theme(legend.key = element_rect(fill = "white", colour = "white"),

### 2018-05-25 paper-ieee-trans.tex
added new file to produce figure2
redo fig 2
change title, (re)move biography to separate file

@@
+% trial .tex file %
+
 \documentclass[journal]{IEEEtran}
+
 \usepackage{graphicx}
@@
 \usepackage{adjustbox}
-
+\usepackage{listings}
 \ifCLASSINFOpdf
@@
 \begin{document}
-\title{Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis}
+\title{Evolution of Methodologies in Delivering Reliable Software in the European Scientific Arena}
+%\title{Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis}
 %\title{15 years of European scientific software development: evolution and methodologies}
@@
 \centering
-\includegraphics[width=\textwidth]{images/unittest.png}
+\includegraphics[width=\textwidth]{images/fig2.png}
 \caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
@@
 \centering
-\includegraphics[width=0.45\textwidth, height=50mm]{images/confman.png}
+\includegraphics[width=0.5\textwidth, height=50mm]{images/fig3.png}
 \caption{Trend line showing the adoption of Configuration Management tools throughout the project lifetime.}
@@
 
-\begin{IEEEbiography} [{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/Ronchieri}}]{Elisabetta Ronchieri}
-is a Computer Science Engineer at INFN (Istituto Nazionale di Fisica Nucleare - National Institute of Nuclear Physics) CNAF.
-She holds a PhD in Automation, Robotics and Bioengineering from the University of Pisa.
-Her main area of interest include software engineering issues, such as software quality and software prediction model.
-She is also interested in (big and open) data analysis by using techniques coming from various theories, such as Machine Learning.
-\end{IEEEbiography}
-
-% if you will not have a photo at all:
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/pablo-orviz}}]{Pablo Orviz Fernandez}
-is a computing researcher at CSIC (Consejo Superior de Investigaciones
-Cientificas), holding an MSc in Scientific Computing from the University of
-Cantabria. During the past years he has been devoted to software quality
-assurance tasks within different European advanced computing research projects
-such as INDIGO--DataCloud and EGI--Engage. Currently involved in EOSC--hub project
-and participating in the software quality work packages of DEEP--HybridDataCloud
-and eXtreme--DataCloud.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/mario-david}}]{Mario David}
-is a research associate at LIP. He holds a PhD in Experimental Particle Physics from the University of Lisbon.
-Member of DEEP--HybridDataCloud project as Work Package leader of Testbed and integration with EOSC services. Member of EOSC--hub project and
-of the Portuguese National Computing Distributed Infrastructure.
-He held a research associate position at Institut de Physique du Globe de Paris (IPGP/CNRS) as Scientific Software Developer for the VERCE project, in particular in the data intensive use cases for seismology.
-He was been actively involved in the Validation, Quality Assurance and testing of middleware, in regional and global operations.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/cris_foto_good}}]{Doina Cristina Duma}
-is the leader of the Distributed Systems group at the INFN National Center (CNAF), providing the core operational support for the INFN--wide Grid infrastructure and for the CNAF Cloud infrastructure.
-She has a long experience in managing distributed e--infrastructures, being involved since 2003 in major European projects such as DataGrid, EGEE (I, II, III), EGI--Inspire and EGI--Engage.
-She was the Release Manager of the EMI (European Middleware Initiative) and INDIGO - DataCloud European project project and having as main responsibilities to manage all release activities, including release planning, building, deploying, reporting, establishing and maintaining the overall release timeline and milestones.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/jorge-gomes}}]{Jorge Gomes}
-is a computing researcher at LIP.
-He worked in the development of advanced data acquisition systems at CERN, and participated in pioneering projects in the domain of digital satellite data communications, IP over ATM, and advanced videoconferencing over IP networks.
-Since 2001 he has participated in numerous projects regarding distributed computing, networks and security in Europe and Latin America.
-He is the head of the LIP Advanced Computing and Digital Infrastructures Group and technical coordinator of the Portuguese National Grid Infrastructure, representative of Portugal in the Council of the European Grid Infrastructure (EGI) and responsible for the Portuguese participation in IBERGRID, that joins Portuguese and Spanish distributed computing infrastructures.
-\end{IEEEbiography}
-
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/DavideSalomoni}}]{Davide Salomoni}
-is Director of Technology at the Italian National Institute for Nuclear Physics (INFN).
-He has 27 years of international experience in private and public environments related to distributed computing and communication technologies.
-He currently leads the Software Development and Distributed Systems department at CNAF, the INFN National Center dedicated to research and development on IT technologies, located in Bologna, Italy.
-He was Project Coordinator of INDIGO--DataCloud, a project funded by the EC Horizon2020, that developed innovative open source computing and storage solutions.
-\end{IEEEbiography}
-
 \end{document}

### 2018-05-28 paper-ieee-trans.tex

@@
 \IEEEPARstart{D}uring the last 15 years the provision of an European e--Infrastructure,
-supporting an unified access to large--scale computing and intense data analysis
+supporting unified access to large--scale computing and intense data analysis
 applications, has been driven by the needs of scientific communities as part of their
@@
 of processing, storing and analysis of the very large amount of data produced by the
+% Comment from reviewer
+%(Update: I see this is discussed later... so it would be good just to signpost at the start that a) the focus is really on "grids", and b) that specific examples will be given.
 LHC particle accelerator. A significant number of innovation projects and
@@
 
+% Comment
+%%top of second column:  think about saying this more gently
 Major developments were carried out throughout the years focused on providing new
@@
 
+% Comment
+%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
 Analysing the EC--funded software projects over the last decade,
@@
 testing and validation procedures. The
+% Comments
+%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
 evolution of the software engineering methodologies (creation of the Agile
@@
 field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization) definitely motivated this trend. All of which
+operating--system--level virtualization known as containerization) motivated this trend. All of which
 complemented by the emergence of automation and event--response tools.
@@
 \hline
-Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Source code inspection&&&&$\surd$&&$\surd$\\
-Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
-Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
+Architecture Task Force            &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Communication Handling             &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Requirements Handling              &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Agile Methodologies                &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Source code inspection             &       &       &       &$\surd$&       &$\surd$\\
+Build/Testing Management Procedure &$\surd$&$\surd$&$\surd$&$\surd$&       &$\surd$\\
+Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+Quality Criteria Definition        &       &       &$\surd$&$\surd$&$\surd$&$\surd$\\
 %Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
-Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
-Auto--generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
+Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
+Auto--generated Documentation      &       &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
+DevOps practices adoption (CI, CD) &       &       &       &       &       &$\surd$\\
 \hline
@@
 \subsection{Enabling Grids for E--sciencE}
-
+% Comment
+%%P2 C2 L24: Typographical: It would be good to merge the three citations into one bracket/\cite
 The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
-\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
+\cite{cordis:egee,cordis:egee2,cordis:egee3} projects brought together
 scientists and engineers from over 240 institutions in 45 countries aiming to
@@
 explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
-project of software development considered in this paper. At the time of writing
-this paper, the project is succeeding in providing solutions to address the existing gaps in
+project of software development considered in this paper. At the time of writing,
+the project is succeeding in providing solutions to address the existing gaps in
 Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
@@
 
-{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
+{\sl INDIGO--DataCloud} is the most recent project included in this paper, it
 leveraged the lessons learned from the previous experiences described in
@@
 profitable in terms of the evaluation and application of innovative software engineering
-methodologies, the management of new technologies to put those methodologies
-into practice, and the compilation and analysis of the appropriate set of metrics to measure the
+methodologies, management of new technologies to put those methodologies
+into practice, compilation and analysis of an appropriate set of metrics to measure the
+% Comment
+%%L26-29 Missing verb
+%%(not a sentence—no verb) 
 quality of the implemented solutions. All the above with the additional complexity of being framed into
@@
 In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
+% Comment
+%%P3 C2 L29: This introductory paragraph goes into a bit too much jargony detail, I think, especially given that it is followed by in-depth coverage of each "pillar".
+I suggest a clearer identification of each pillar, maybe via bullets or definition paragraphs, avoiding or with general-audience definition of jargon like DevOps, delivery mechanisms, two-factor validation, pilot testbeds, and early adoption.
+I appreciate this may all be highly meaningful to someone steeped in this operational area, but to an interested outsider it reads much like management jargon -- particuarly the final sentence: is this a long-winded way of saying "used during the commissioning period"?
 software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
@@
 \includegraphics[width=\textwidth]{images/fig2.png}
+%\includegraphics[width=0.8\textwidth]{images/fig2a.pdf}
 \caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
@@
 most--adopted or \textit{de--facto standards} are initially recommended, however the development teams eventually
-choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
+choose the set of guidelines to comply with. Fig.~\ref{fig:fig_codestyle} highlights the code style standards and
+their popularity
 in the {\sl INDIGO--DataCloud} software stack.
@@
 
+%% Comment from reviewer
+%% These percentages appear as vague approximations by eye, please add the precise numbers rounded.
+
  \textit{Unit testing} completes the source code testing evaluation by
 focusing on the code's internal design. The SQA criteria set a recommended threshold of
-70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
+70\% coverage for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared
 for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
@@
 \item \textit{Code review}.
+% Comment
+%%Represents …   (not a sentence) 
 Represents the last step in the change management pipeline, once the candidate change has
@@
 such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
-the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
-\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
+the Ansible Galaxy portal \cite{indigo-galaxy}. Fig.~\ref{fig:fig_confman}
+shows the number of products that offer an automated means for deployment.
 It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
@@
 \label{sec:devops}
-
+% Comment
+%P5 C2 L21: Define DevOps!
 The DevOps methods emphasize the SQA techniques to avoid infrastructure
@@
 
+% Comment
+%%not a sentence, again
 Jenkins \cite{jenkins}, the event--response CI system, and Docker container
@@
 
-Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the
+Fig.~\ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the
 {\sl INDIGO--DataCloud}'s first and second releases, of the required test types builds since the
@@
 \subsubsection{Continuous Delivery}
+% Comment from reviewer
+%Please add a reference.
 As DevOps suggests, frequent releases positively affect the reliability of the software as
 they involve fewer changes and usually are more correlated. The software updates of
-{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
+{\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
 a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
@@
 done the recently created image is uploaded to the production DockerHub repository, tagged
-with the label corresponding to the current {\sl INDIGO--DataCloud} release. Figure
-\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
+with the label corresponding to the current {\sl INDIGO--DataCloud} release.
+Fig.~\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
 automatically triggered by a change in the source code.
@@
 integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
-Figure \ref{fig:fig_disvis} shows the pipeline for the Disvis application.
+Fig.~\ref{fig:fig_disvis} shows the pipeline for the Disvis application.
 
@@
 possible to production ones. A map of the Pilot Preview
-infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
+infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}, it shows the
 resources providers and the components or services they deployed and supported.
@@
 
+% Comment
+%%P8 C1 L26: What is the demonstration that reliability has been improved significantly with the addition of stricter methologies?
+Is there some quantitative evidence of uptime, or bug reports per user, or similar?
+Or some evidence of speed of defect detection & fixing, which you cite as greatly improved?
+I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
 Reliability of the software has been substantially improved by applying the SQA
@@
 
+% Comment
+%%L42-44 not clear, please rephrase
 Software lifecycle demands as well from a continuous measurement and analysis
@@
 \bibitem{h2020}
-\emph{HORIZON 2020}, The EU Framework Programme for Research and Innovation [Online]. Available: https://ec.europa.eu/programmes/horizon2020/
+\emph{HORIZON 2020}, The EU Framework Programme for Research and Innovation [Online]. Available: https://ec.europa.eu/programmes/horizon2020/ .
 
 \bibitem{agile-manifesto}
-\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org
+\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org .
 
@@
 \emph{DataGrid project}, European Community Research and Development
-Information Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/53665\_en.html
+Information Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/53665\_en.html .
 
@@
 Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/80149\_en.html
+http://cordis.europa.eu/project/rcn/80149\_en.html .
 
@@
 Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/99189\_en.html
+http://cordis.europa.eu/project/rcn/99189\_en.html .
 
@@
 Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/87264\_en.html
+http://cordis.europa.eu/project/rcn/87264\_en.html .
 
@@
 \emph{Definition and Documentation of the Revised Software Life--Cycle Process},
-Milestone MSA3.4.2, 2010, EGEE--III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE--III--MSA3.4.2-1062487-v1\_4.pdf
+Milestone MSA3.4.2, 2010, EGEE--III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE--III--MSA3.4.2-1062487-v1\_4.pdf .
 
@@
 (ETICS)} project, European Community Research and Development Information
-Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/80138\_en.html
+Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/80138\_en.html .
 
@@
 Phase 2 (ETICS 2)} project, European Community Research and Development
-Information Service (CORDIS), [Online]. Available: http://cordis.europa.eu/project/rcn/86604\_en.html
+Information Service (CORDIS), [Online]. Available: http://cordis.europa.eu/project/rcn/86604\_en.html .
 
@@
 Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/95311\_en.html
+http://cordis.europa.eu/project/rcn/95311\_en.html .
 
@@
 \emph{ISO/IEC 9126 Software Engineering - Product Quality}, International
-Organization for Standarization [Online]. Available: https://www.iso.org/standard/22749.html
+Organization for Standarization [Online]. Available: https://www.iso.org/standard/22749.html .
 
 \bibitem{emi-metrics}
-\emph{EMI Metrics Specification} [Online]. Available: https://goo.gl/CCtY7x
+\emph{EMI Metrics Specification} [Online]. Available: https://goo.gl/CCtY7x .
 
 \bibitem{emi-quality-model}
-\emph{EMI Quality Model} [Online]. Available: https://goo.gl/LdS6fL
+\emph{EMI Quality Model} [Online]. Available: https://goo.gl/LdS6fL .
 
@@
 Community Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/95923\_en.html
+http://cordis.europa.eu/project/rcn/95923\_en.html .
 
@@
 \emph{EGI Operations Portal} [Online]. Available:
-https://operations-portal.egi.eu/
+https://operations-portal.egi.eu/ .
 
@@
 \emph{GGUS} EGI Helpdesk [Online]. Available:
-https://ggus.eu/
+https://ggus.eu/ .
 
@@
 \emph{GOCDB} EGI Grid Configuration Repository [Online]. Available:
-https://ggus.eu/
+https://goc.egi.eu/ .
 
@@
 \emph{EGI Quality Criteria} [Online]. Available:
-https://egi-qc.github.io/
+https://egi-qc.github.io/ .
 
@@
 \emph{GITHUB} [Online]. Available:
-https://github.com/
+https://github.com/ .
 
@@
 \emph{Jenkins} Continuous Integration System [Online]. Available:
-https://jenkins-ci.org/
+https://jenkins-ci.org/ .
 
 \bibitem{egi-qc}
-\emph{EGI Quality Criteria} [Online]. Available: http://egi-qc.github.io/
+\emph{EGI Quality Criteria} [Online]. Available: http://egi-qc.github.io/ .
 
@@
 project, European Community Research and Development Information Service
-(CORDIS), [Online]. Available:  http://cordis.europa.eu/project/rcn/194937\_en.html
+(CORDIS), [Online]. Available:  http://cordis.europa.eu/project/rcn/194937\_en.html .
 
@@
 Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/194882\_en.html
+http://cordis.europa.eu/project/rcn/194882\_en.html .
 
@@
 \emph{INDIGO--DataCloud DockerHub repository} [Online]. Available:
-https://hub.docker.com/u/indigodatacloud
+https://hub.docker.com/u/indigodatacloud .
 
@@
 \emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1 [Online]. Available:
-https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31
+https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31 .
 
@@
 \emph{INDIGO--DataCloud Jenkins CI service} [Online]. Available:
-https://jenkins.indigo-datacloud.eu:8080/
+https://jenkins.indigo-datacloud.eu:8080/ .
 
@@
 \emph{INDIGO--DataCloud GitBook Documentation repository} [Online]. Available:
-https://www.gitbook.com/@indigo-dc
+https://www.gitbook.com/@indigo-dc .
 
@@
 \emph{INDIGO--DataCloud PuppetForge repository} [Online]. Available:
-https://forge.puppet.com/indigodc
+https://forge.puppet.com/indigodc .
 
@@
 \emph{INDIGO--DataCloud Ansible Galaxy repository} [Online]. Available:
-https://galaxy.ansible.com/indigo-dc/
+https://galaxy.ansible.com/indigo-dc/ .
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+% Duplicate
 \bibitem{indigo-ansible}
 \emph{INDIGO--DataCloud Ansible Galaxy repository} [Online]. Available:
-https://galaxy.ansible.com/indigo-dc/
+https://galaxy.ansible.com/indigo-dc/ .
 
 \bibitem{grimoirelab}
-\emph{GrimoireLab} [Online]. Available: http://grimoirelab.github.io/
+\emph{GrimoireLab} [Online]. Available: http://grimoirelab.github.io/ .
 
 \bibitem{indigo-pkg-repo}
-\emph{INDIGO--DataCloud Linux Package repository} [Online]. Available: https://jenkins.indigo-datacloud.eu/
+\emph{INDIGO--DataCloud Linux Package repository} [Online]. Available: https://jenkins.indigo-datacloud.eu/ .
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
+% Duplicate
 \bibitem{ggus}
@@
 DisVis: quantifying and visualizing the accessible interaction space of distance
-restrained biomolecular complexes, Bioinformatics 31 (2015) 3222--3224
+restrained biomolecular complexes, Bioinformatics 31 (2015) 3222--3224 .
 

### 2018-06-01 paper-ieee-trans.tex

@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
 \centering
-\includegraphics[width=0.5\textwidth, height=50mm]{images/fig3.png}
+\includegraphics[width=0.8\textwidth]{images/fig3.png}
 \caption{Trend line showing the adoption of Configuration Management tools throughout the project lifetime.}
 \label{fig:fig_confman}
-\end{figure}
+\end{figure*}
 
@@
 platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
-as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
+as Ansible \cite{indigo-galaxy} and Puppet \cite{indigo-puppet}. A representative example of
 such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-% Duplicate
-\bibitem{indigo-ansible}
-\emph{INDIGO--DataCloud Ansible Galaxy repository} [Online]. Available:
-https://galaxy.ansible.com/indigo-dc/ .
-
 \bibitem{grimoirelab}
@@
 \emph{INDIGO--DataCloud Linux Package repository} [Online]. Available: https://jenkins.indigo-datacloud.eu/ .
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
-% Duplicate
-\bibitem{ggus}
-\emph{Global Grid User Support (GGUS)} [Online]. Available: https://www.ggus.eu/
 

### 2018-06-01 fig3.R
update fig3, and Rscript

@@ -1,0 +1,66 @@
+datafn<-read.csv('data/figure3.csv', header=TRUE)
+names(datafn)
+str(datafn)
+
+i1<-datafn[datafn$Release=='INDIGO-1',]
+lm_fit1 <- lm(Number~Week, data=i1)
+summary(lm_fit1)
+
+predicted_df1 <- data.frame(number_pred = predict(lm_fit1, i1))
+predicted_df1
+
+i2<-datafn[datafn$Release=='INDIGO-2',]
+lm_fit2 <- lm(Number~Week, data=i2)
+summary(lm_fit2)
+
+predicted_df2 <- data.frame(number_pred = predict(lm_fit2, i2))
+predicted_df2
+
+predicted_df <- rbind(predicted_df1, predicted_df2)
+datafn$Lm <-rbind(predicted_df1, predicted_df2)
+
+names(datafn)
+str(datafn)
+
+library(ggplot2)
+
+ggplot(datafn, aes(x=Week, y=Number)) +
+  geom_point(data=datafn[datafn$Release=='INDIGO-1',],
+             aes(x=Week, y=Number, color='MidnightBlue'), 
+             size=5) +
+  geom_point(data=datafn[datafn$Release=='INDIGO-2',],
+             aes(x=Week, y=Number, color='ElectricIndigo'), 
+             size=5) +
+  geom_line(data=datafn[datafn$Release=='INDIGO-1',],
+            aes(x=Week, y=Lm, color='MidnightBlue'), 
+            size=2) +
+  geom_line(data=datafn[datafn$Release=='INDIGO-2',],
+            aes(x=Week, y=Lm, color='ElectricIndigo'), 
+            size=2) +
+  xlab("Weeks") +
+  ylab("Number of Software Products") +
+  scale_y_continuous(limits = c(0, 25)) +
+  scale_x_continuous(limits = c(0, 34), 
+                     breaks=c(0, 5, 10, 15, 20, 25, 30)) +
+  scale_colour_manual(name='Releases', values=c(MidnightBlue="darkblue",
+                                                ElectricIndigo="burlywood3"),
+                      breaks=c('MidnightBlue',
+                               'ElectricIndigo')) +
+  geom_vline(aes(xintercept=8),
+             colour='darkblue',
+             size=1, alpha=1) +
+  geom_vline(aes(xintercept=28), 
+             colour='burlywood3',
+             size=1, alpha=1)  +
[…]

### 2018-06-01 paper-ieee-trans.tex
update fig3, and Rscript

@@
 \includegraphics[width=\textwidth]{images/fig2.png}
-%\includegraphics[width=0.8\textwidth]{images/fig2a.pdf}
 \caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
-\includegraphics[width=0.8\textwidth]{images/fig3.png}
-\caption{Trend line showing the adoption of Configuration Management tools throughout the project lifetime.}
+\includegraphics[width=0.5\textwidth]{images/fig3.png}
+\caption{Adoption of Configuration Management tools throughout the project lifetime. Also shown the trend lines
+leading to the first release (dark blue points and lines), and to the second release (light cream points and lines).}
 \label{fig:fig_confman}
-\end{figure*}
+\end{figure}
 
@@
 deployment of the products delivered as part of the catalogue. Automation in this context is
-tackled using a configuration management tool. These tools allow to have an
+tackled using a configuration management tool. These tools allow the adoption of
 \textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
@@
 as Ansible \cite{indigo-galaxy} and Puppet \cite{indigo-puppet}. A representative example of
-such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
+such contribution are the \textit{50 roles} developed from scratch and currently hosted in
 the Ansible Galaxy portal \cite{indigo-galaxy}. Fig.~\ref{fig:fig_confman}
 shows the number of products that offer an automated means for deployment.
-It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
-of the two major releases.
+It shows an increase in the adoption of such tools leading to the {\sl INDIGO--1} release
+(dark blue points and trend line), as well as to the {\sl INDIGO--2} release
+(light cream points and trend line). The rate of adoption is lower during the weeks
+before the second release because a significant fraction of the products had already adopted it
+previously to the first release.
 \end{itemize}
@@
 \emph{ISO/IEC 9126 Software Engineering - Product Quality}, International
-Organization for Standarization [Online]. Available: https://www.iso.org/standard/22749.html .
+Organization for Standardization [Online]. Available: https://www.iso.org/standard/22749.html .
 
@@
 \emph{INDIGO--DataCloud GitHub Source Code repository} [Online]. Available:
-https://www.github.com/indigo-dc
+https://www.github.com/indigo-dc .
 

### 2018-06-01 paper-ieee-trans.tex
reviers comments

@@
 \usepackage{listings}
+\usepackage{url}
+\usepackage{cite}
 \ifCLASSINFOpdf
@@
 
-\IEEEPARstart{D}uring the last 15 years the provision of an European e--Infrastructure,
+\IEEEPARstart{D}{uring} the last 15 years the provision of an European e--Infrastructure,
 supporting unified access to large--scale computing and intense data analysis
@@
 \includegraphics[width=\textwidth]{images/fig2.png}
-\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud} software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
+\caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud}
+software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
 \label{fig:fig_unittest}
@@
 
-\bibitem{egi-qc}
-\emph{EGI Quality Criteria} [Online]. Available: http://egi-qc.github.io/ .
-
 \bibitem{mario}

### 2018-06-04 paper-ieee-trans.tex
Add exact values for code coverage.

@@
 testing} in this context is accomplished by enforcing the definition and periodic execution
-of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development activities.
-
-%% Comment from reviewer
-%% These percentages appear as vague approximations by eye, please add the precise numbers rounded.
-
- \textit{Unit testing} completes the source code testing evaluation by
-focusing on the code's internal design. The SQA criteria set a recommended threshold of
-70\% coverage for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared
-for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
-trend over the course of the project. It can be seen that 50\% of the components were over the
-70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.
+of the tests, covering past fixed issues, to ensure that they are not reintroduced during the development
+activities. \textit{Unit testing} completes the source code testing evaluation by focusing on the code's 
+internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage 
+for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software 
+component between the two major releases, denoting an incremental trend over the course of the project: 
+code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
+the project's first (INDIGO-1) and second (INDIGO-2) releases. By INDIGO-2, half of the software components
+(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation 
+(70\%), while three-quarters (24 out of 32) of the product stack exceeded 50\% coverage.
 

### 2018-06-04 paper-ieee-trans.tex
Referring to the next section in the intro

@@
 %(Update: I see this is discussed later... so it would be good just to signpost at the start that a) the focus is really on "grids", and b) that specific examples will be given.
-LHC particle accelerator. A significant number of innovation projects and
-technology transfer activities -- from the European Commission (EC)-funded 5th, 6th and 7th Framework Programmes,
-and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
-develop software solutions that provide a seamless and transparent access of researchers to distributed
-computing and data resources.
+LHC particle accelerator. A significant number of innovation projects and technology transfer
+activities -- later discussed in Section \ref{sec:ev} -- from the European Commission
+(EC)-funded 5th, 6th and 7th Framework Programmes, and recently from the Horizon 2020 programme
+\cite{h2020} had as their main goal to develop software solutions that provide a seamless and
+transparent access of researchers to distributed computing and data resources.
 

### 2018-06-04 figure3.R
improved fig

@@ -25,41 +25,38 @@ str(datafn)
 library(ggplot2)
 
 ggplot(datafn, aes(x=Week, y=Number)) +
+  geom_rect(data=NULL,aes(xmin=0.25,xmax=8.25,ymin=-Inf,ymax=Inf),
+                    fill="papayawhip")+
+  geom_rect(data=NULL,aes(xmin=20.25,xmax=33.75,ymin=-Inf,ymax=Inf),
+                    fill="aliceblue")+
   geom_point(data=datafn[datafn$Release=='INDIGO-1',],
-             aes(x=Week, y=Number, color='MidnightBlue'), 
-             size=3) +
+             aes(x=Week, y=Number, color='MidnightBlue (INDIGO-1)'), 
+             size=5) +
   geom_point(data=datafn[datafn$Release=='INDIGO-2',],
-             aes(x=Week, y=Number, color='ElectricIndigo'), 
-             size=3) +
+             aes(x=Week, y=Number, color='ElectricIndigo (INDIGO-2)'), 
+             size=5) +
   geom_line(data=datafn[datafn$Release=='INDIGO-1',],
-            aes(x=Week, y=Lm, color='MidnightBlue'), 
-            size=1) +
+             aes(x=Week, y=Lm, color='MidnightBlue (INDIGO-1)'), 
+             size=2) +
   geom_line(data=datafn[datafn$Release=='INDIGO-2',],
-            aes(x=Week, y=Lm, color='ElectricIndigo'), 
-            size=1) +
+             aes(x=Week, y=Lm, color='ElectricIndigo (INDIGO-2)'), 
+             size=2) +
   xlab("Weeks") +
   ylab("Number of Software Products") +
   scale_y_continuous(limits = c(0, 25)) +
   scale_x_continuous(limits = c(0, 34), 
                      breaks=c(0, 5, 10, 15, 20, 25, 30)) +
-  scale_colour_manual(name='Releases', values=c(MidnightBlue="darkblue",
-                                                ElectricIndigo="red"),
-                      breaks=c('MidnightBlue',
-                               'ElectricIndigo')) +
-  geom_vline(aes(xintercept=8),
-             colour='darkblue',
-             size=1, alpha=1) +
-  geom_vline(aes(xintercept=28), 
-             colour='red',
-             size=1, alpha=1)  +
+  scale_colour_manual(name='Releases', values=c('MidnightBlue (INDIGO-1)'="burlywood1",
[…]

### 2018-06-04 figure2.R
added changes

@@ -26,22 +26,22 @@ ggplot(datafn, aes(fill=Release , x=Product, y=Value)) +
                      expand=c(0,1))+
   geom_hline(aes(yintercept=val1$Avg[1],  
                  linetype='MidnightBlue'),
-             colour='red',
-             size=1, alpha=1) +
+                 colour='darkred',
+                 size=1, alpha=1) +
   geom_hline(aes(yintercept=val2$Avg[1],  
                  linetype='ElectricIndigo'), 
-             colour='darkblue',
-             size=1, alpha=1)  +
-  scale_fill_manual(name='Releases', values=c("blue", "burlywood3"),
+                 colour='blue',
+                 size=1, alpha=1)  +
+  scale_fill_manual(name='Releases', values=c("blue", "burlywood1"),
                     breaks = c(rel1,
                                rel2)) +
   scale_linetype_manual(name="Avg", 
-                        values=c(2, 2),
-                        guide = guide_legend(override.aes = 
-                                               list(color = c("red", 
-                                                              "darkblue"))),
-                        labels = c(rel1avg,
-                                   rel2avg)) +
+                     values=c(2, 2),
+                     guide = guide_legend(override.aes = 
+                                            list(color = c("burlywood1", 
+                                                           "blue"))),
+                     labels = c(rel1avg,
+                                rel2avg)) +
   coord_flip()+
   theme(legend.key = element_rect(fill = "white", colour = "white"),
         axis.text=element_text(size=14),

### 2018-06-04 figure3.R
added changes for back

@@ -25,9 +25,9 @@ str(datafn)
 library(ggplot2)
 
 ggplot(datafn, aes(x=Week, y=Number)) +
-  geom_rect(data=NULL,aes(xmin=0.25,xmax=8.25,ymin=-Inf,ymax=Inf),
+  geom_rect(data=NULL,aes(xmin=0.25,xmax=8.75,ymin=-Inf,ymax=Inf),
                     fill="papayawhip")+
-  geom_rect(data=NULL,aes(xmin=20.25,xmax=33.75,ymin=-Inf,ymax=Inf),
+  geom_rect(data=NULL,aes(xmin=8.76,xmax=33.75,ymin=-Inf,ymax=Inf),
                     fill="aliceblue")+
   geom_point(data=datafn[datafn$Release=='INDIGO-1',],
              aes(x=Week, y=Number, color='MidnightBlue (INDIGO-1)'), 

### 2018-06-04 figure3.R
added change

@@ -25,7 +25,7 @@ str(datafn)
 library(ggplot2)
 
 ggplot(datafn, aes(x=Week, y=Number)) +
-  geom_rect(data=NULL,aes(xmin=0.25,xmax=8.75,ymin=-Inf,ymax=Inf),
+  geom_rect(data=NULL,aes(xmin=0,xmax=8.75,ymin=-Inf,ymax=Inf),
                     fill="papayawhip")+
   geom_rect(data=NULL,aes(xmin=8.76,xmax=33.75,ymin=-Inf,ymax=Inf),
                     fill="aliceblue")+
@@ -43,9 +43,11 @@ ggplot(datafn, aes(x=Week, y=Number)) +
              size=2) +
   xlab("Weeks") +
   ylab("Number of Software Products") +
-  scale_y_continuous(limits = c(0, 25)) +
+  scale_y_continuous(limits = c(0, 25), 
+                     expand = c(0, 0)) +
   scale_x_continuous(limits = c(0, 34), 
-                     breaks=c(0, 5, 10, 15, 20, 25, 30)) +
+                     breaks=c(0, 5, 10, 15, 20, 25, 30),
+                     expand = c(0, 0)) +
   scale_colour_manual(name='Releases', values=c('MidnightBlue (INDIGO-1)'="burlywood1",
                                                 'ElectricIndigo (INDIGO-2)'="blue"),
                       breaks=c('MidnightBlue (INDIGO-1)',

### 2018-06-04 paper-ieee-trans.tex
Join the two paragraphs to fix review#2 comment (P1 C2 L24) about the emergence of the need of giving importance to SQA

@@
 rollbacks and patches applied to production systems. Thereby, the stability and
+% Comment
+%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
 reliability of the e--Infrastructures was often compromised, affecting
 negatively their exploitation by the scientific communities.
+% Comments
+%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
+It became then apparent the need for a better balance between the addition of new features
+and the reliability of the released software. Consequently, incoming projects progressively
+devoted a substantial part of the software lifecycle to the improvement of the quality of the
+software being delivered. At this point, evolving software engineering practices were considered
+and gradually adopted to define and implement quality procedures, the organization of
+cross--functional teams and deployment of pilot testbeds.
 
-Soon it became apparent the need for a better balance between the addition of new features
-and the reliability of the released software. Consequently, a substantial part of the
-software lifecycle was devoted to the improvement of the quality of the software being
-delivered. At this point, evolving software engineering practices were
-considered and gradually adopted to define and implement quality procedures, the organization of cross--functional teams
-and deployment of pilot testbeds.
-
-% Comment
-%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
 Analysing the EC--funded software projects over the last decade,
@@
 testing and validation procedures. The
-% Comments
-%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
 evolution of the software engineering methodologies (creation of the Agile

### 2018-06-04 paper-ieee-trans.tex
Add the comment back
Explaination of drop of activity in fig4

@@
 \item \textit{Unit and Functional testing}.
+%% Comment from reviewer       
+%% These percentages appear as vague approximations by eye, please add the precise numbers rounded.
 Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
@@
 code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
-the project's first (INDIGO-1) and second (INDIGO-2) releases. By INDIGO-2, half of the software components
+the project's first (INDIGO-1) and second (INDIGO-2) releases. By INDIGO-2, 53\% of the software components
 (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation 
-(70\%), while three-quarters (24 out of 32) of the product stack exceeded 50\% coverage.
+(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
 
@@
 implementation of the CI infrastructure. Automated functional testing coverage were not
-available for all the software stack, thus the associated number of builds are fewer.
+available for all the software stack, thus the associated number of builds are fewer. Towards the
+end of the project, the software development activity slowed the pace but not completely ceased 
+as the CI infrastructure is still maintained after the project's end of life.
 

### 2018-06-05 paper-ieee-trans.tex

@@
-% trial .tex file %
-
 \documentclass[journal]{IEEEtran}
@@
 
-The E--Infrastructure for Testing, Integration and Configuration of Software
-\cite{cordis:etics, cordis:etics2} ({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
+The E--Infrastructure for Testing, Integration and Configuration of Software~\cite{etics}
+({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
 at addressing the challenges in producing quality software in distributed,
@@
 metrics and Key Performance Indicators (KPIs) were identified and defined in
-detail in the {\sl EMI} Metrics Specification \cite{emi-metrics}. The project
+detail in the {\sl EMI} Metrics Specification \cite{emi-quality-model}. The project
 leveraged the {\sl ETICS} service for the development, continuous integration and release management,
@@
 \textit{documentation is treated as code}, using a markup language, automatically rendered and
-uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and
+uploaded to online repositories\footnote{\url{https://www.gitbook.com/@indigo-dc}}.
+Thus documentation is portable and
 human--reviewed, using the same workflow as the code does, validating the changes before being
@@
 platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
-as Ansible \cite{indigo-galaxy} and Puppet \cite{indigo-puppet}. A representative example of
+as Ansible
+and Puppet\footnote{\url{https://forge.puppet.com/indigodc}}. A representative example of
 such contribution are the \textit{50 roles} developed from scratch and currently hosted in
-the Ansible Galaxy portal \cite{indigo-galaxy}. Fig.~\ref{fig:fig_confman}
+the Ansible Galaxy portal\footnote{\url{https://galaxy.ansible.com/indigo-dc/}}. Fig.~\ref{fig:fig_confman}
 shows the number of products that offer an automated means for deployment.
@@
 \textit{Development metrics} are obtained programmatically from several sources, namely GitHub
-API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
-framework \cite{grimoirelab}. Per--component weekly reports, including the SQA requirement
+API\footnote{\url{https://developer.github.com/v3/}.}
+and Jenkins\footnote{\url{https://jenkins.io/}} service, and graphically displayed as GitHub
+pages using GrimoireLab framework\footnote{\url{http://grimoirelab.github.io/}}.
+Per--component weekly reports, including the SQA requirement
 fulfilment, are issued and individually discussed through the different communication channels
@@
 \textit{Release metric} sources are the online repository servers, namely the
-Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins
+Linux package\footnote{\url{https://jenkins.indigo-datacloud.eu/}}
+and DockerHub\footnote{\url{https://hub.docker.com/u/indigodatacloud}} repositories. Jenkins
 server also contains valuable release data as it is the service where the packages
@@
 to issues reported by external users. Feedback is collected from outside helpdesks, such as
-EGI's GGUS \cite{ggus}, and the GitHub Issues tracker.
+EGI's GGUS~\cite{ggus}, and the GitHub Issues tracker.
 
@@
 The practical implementation of the CI scenario leveraged on tightly integrated open
-source tools such as GitHub \cite{github}, as the online source code repository,
+source tools such as GitHub\footnote{\url{https://github.com/}}, as the online source code repository,
 
@@
 %%not a sentence, again
-Jenkins \cite{jenkins}, the event--response CI system, and Docker container
+Jenkins, the event--response CI system, and Docker container
 provisioning to provide instant computing power needed for executing the pipeline jobs.
@@
 \bibitem{h2020}
-\emph{HORIZON 2020}, The EU Framework Programme for Research and Innovation [Online]. Available: https://ec.europa.eu/programmes/horizon2020/ .
+\emph{Horizon 2020} -- European Commission.
+[Online] Available: \url{https://ec.europa.eu/programmes/horizon2020/}
+[Accessed 4 Jun. 2018].
 
 \bibitem{agile-manifesto}
-\emph{Manifesto for Agile Software Development} [Online]. Available: http://agilemanifesto.org .
+Beck, K., Beedle, et al., (2001).
+Manifesto for Agile Software Development.
+[Online] Available: \url{http://www.agilemanifesto.org/}
+[Accessed 4 Jun. 2018].
 
 \bibitem{zhu}
-L. Zhu and L. Bass and G. Champlin--Scharff, \emph{DevOps and Its Practices},
-IEEE Software, vol. 33, no. 3, pp. 32--34, May--June 2016.
+L. Zhu and L. Bass and G. Champlin--Scharff,
+DevOps and Its Practices,
+\emph{IEEE Software},
+vol. 33, n. 3, pp. 32--34, 2016.
 
 \bibitem{cordis:datagrid}
-\emph{DataGrid project}, European Community Research and Development
-Information Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/53665\_en.html .
+P. Kunszt,
+European DataGrid project: status and plans,
+\emph{Nucl. Instr. Meth. Phys. Res. A},
+vol. 502, n. 2, pp. 376--381, 2003.
+%\emph{DataGrid project}, European Community Research and Development Information Service (CORDIS).
+%[Online] Available: \url{http://cordis.europa.eu/project/rcn/53665\_en.html}
+%[Accessed 4 Jun. 2018].
 
 \bibitem{gagliardi}
-Gagliardi F., Jones B., Reale M., Burke S., \emph{European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E--science Applications},
-in Calzarossa M.C., Tucci S. (eds) Performance Evaluation of Complex Systems: Techniques and Tools, Lecture Notes in Computer Science, vol 2459. Springer, 2002.
+Gagliardi F., Jones B., Reale M., Burke S.,
+European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E--science Applications,
+in \emph{Performance Evaluation of Complex Systems: Techniques and Tools, LNCS},
+vol. 2459, 2002.
 
 \bibitem{globus}
-I. Foster and C. Kesselman, \emph{Globus: a Metacomputing Infrastructure
-Toolkit}, International Journal of Supercomputer Applications, vol. 11, no. 2,
-pp. 115--128, 1997.
+I. Foster and C. Kesselman,
+Globus: a Metacomputing Infrastructure Toolkit,
+\emph{Int. J. Supercomput. Appl.},
+vol. 11, no. 2, pp. 115--128, 1997.
 
 \bibitem{datagrid}
-L. Momtahan and A. Martin, \emph{e--Science Experiences: Software Engineering
-Practice and the EU DataGrid}, in Proc. Asia--Pacific Software Engineering
-Conference, Gold Coast, Queensland, Australia, pp. 269-275, IEEE Press,
-4-6 Dec. 2002.
+L. Momtahan and A. Martin,
+e--Science Experiences: Software Engineering Practice and the EU DataGrid,
+in \emph{Ninth Asia--Pacific Software Engineering Conference},
+pp. 269-275, 2002.
 
 \bibitem{agile}
-T. Dingsoyr, \emph{A decade of agile methodologies: Towards explaining agile
-software development} in The Journal of Systems and Software, Elsevier Inc,
-2012.
+Torgeir Dingsoyr, Sridhar Nerur, VenuGopal Balijepally, Nils Brede Moe,
+A decade of agile methodologies: Towards explaining agile software development,
+\emph{J. Syst. Softw.},
+vol. 85, Issue 6, pp. 1213--1221, June 2012.
 
 \bibitem{cmm}
-Paulk et al., \emph{Capability maturity model for software}, Software
-Engineering Institute, report CMU/SEI-93-TR-24, ESC--TR-93-177, Feb. 1993.
+Paulk M., Curtis B., Chrissis M., and Weber V. C.,
+Capability maturity model for software,
+\emph{Software Engineering Institute},
+report CMU/SEI-93-TR-24, ESC--TR-93-177, 1993.
 
 \bibitem{cordis:egee}
-\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community
-Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/80149\_en.html .
+F. Gagliardi and M. E. B\'{e}gin,
+EGEE -- providing a production quality grid for e-science,
+in \emph{2005 IEEE Inter. Symp. on Mass Storage Systems and Technology},
+pp. 88-92, 2005.
+%\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community Research and
+%Development Information Service (CORDIS). 
+%[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}
+%[Accessed 4 Jun. 2018].
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% TOBE checked if there are papers on these
 \bibitem{cordis:egee2}
-\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community
-Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/99189\_en.html .
+\emph{Enabling Grids for E--sciencE--II (EGEE--II)} project, European Community Research and
+Development Information Service (CORDIS).
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189\_en.html}
+[Accessed 4 Jun. 2018].
 
@@
 \emph{Enabling Grids for E--sciencE--III (EGEE--III)} project, European Community
-Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/87264\_en.html .
+Research and Development Information Service (CORDIS).
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264\_en.html}
+[Accessed 4 Jun. 2018].
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \emph{Definition and Documentation of the Revised Software Life--Cycle Process},
-Milestone MSA3.4.2, 2010, EGEE--III project [Online]. Available: https://edms.cern.ch/ui/file/1062487/2/EGEE--III--MSA3.4.2-1062487-v1\_4.pdf .
+Milestone MSA3.4.2, 2010, EGEE--III project.
+[Online] Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}
+[Accessed 4 Jun. 2018].
 
 \bibitem{glite}
-E. Laure et al, \emph{Programming the Grid with gLite}, in Jin, H., Reed, D.A.,
-Jiang, W. (eds.) Computational Methods in Science and Technology, vol. 12(1),
-pp. 33--45, Scientific Publishers OWN, 2006.
+E. Laure et al.,
+Programming the Grid with gLite,
+\emph{Computational Methods in Science and Technology},
+vol. 12(1), pp. 33--45, 2006.
 
 \bibitem{condor}
-D. Thain, T. Tannenbaum, M. Livny, \emph{Condor and the Grid}, in Grid
-Computing: Making the Global Infrastructure a Reality, Chapter 11, pp. 63--70,
-Eds. John Wiley \& Sons Inc., 2002.
+D. Thain, T. Tannenbaum, M. Livny,
+Condor and the Grid,
+in \emph{Grid Computing: Making the Global Infrastructure a Reality},
+Chapter 11, pp. 63--70, 2002.
 
-\bibitem{cordis:etics}
-\emph{E--Infrastructure for Testing, Integration and Configuration of Software
-(ETICS)} project, European Community Research and Development Information
-Service (CORDIS) [Online]. Available: http://cordis.europa.eu/project/rcn/80138\_en.html .
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% TOBE checked if there are papers on these
+%\bibitem{cordis:etics}
+%\emph{E--Infrastructure for Testing, Integration and Configuration of Software (ETICS)}
+%project, European Community Research and Development Information Service (CORDIS).
+%[Online] Available: \url{http://cordis.europa.eu/project/rcn/80138\_en.html}
+%[Accessed 4 Jun. 2018].
 
-\bibitem{cordis:etics2}
-\emph{E--Infrastructure for Testing, Integration and Configuration of Software -
-Phase 2 (ETICS 2)} project, European Community Research and Development
-Information Service (CORDIS), [Online]. Available: http://cordis.europa.eu/project/rcn/86604\_en.html .
+%\bibitem{cordis:etics2}
+%\emph{E--Infrastructure for Testing, Integration and Configuration of Software -
+%Phase 2 (ETICS 2)} project, European Community Research and Development
+%Information Service (CORDIS).
+%[Online]. Available: \url{http://cordis.europa.eu/project/rcn/86604\_en.html}
+%[Accessed 4 Jun. 2018].
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \bibitem{etics}
-A. Di Meglio, M.-E.B??gin, P. Couvares, E. Ronchieri, E. Takacs, \emph{ETICS:
-the international software engineering service for the Grid}, in Journal of
-Physics: Conference Series, Vol. 119, N. 4, IOP Publishing Ltd, 2008.
+A. Di Meglio, et al.,
+ETICS: the international software engineering service for the grid,
+in \emph{J. Phys.: Conf. Ser.},
+vol. 119, n. 4, 042010, 2008.
 
 \bibitem{cordis:emi}
-\emph{European Middleware Initiative (EMI)} project, European Community
-Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/95311\_en.html .
+C. Aiftimiei, et al.,
+Towards next generations of software for distributed infrastructures: The European Middleware Initiative,
+in \emph{2012 IEEE 8th Inter. Conf. on E-Science},
+pp. 1--10, 2012.
+%\emph{European Middleware Initiative (EMI)} project, European Community
+%Research and Development Information Service (CORDIS).
+%[Online] Available: \url{http://cordis.europa.eu/project/rcn/95311\_en.html}
+%[Accessed 4 Jun. 2018].
 
 \bibitem{iso-9126}
-\emph{ISO/IEC 9126 Software Engineering - Product Quality}, International
-Organization for Standardization [Online]. Available: https://www.iso.org/standard/22749.html .
-
-\bibitem{emi-metrics}
-\emph{EMI Metrics Specification} [Online]. Available: https://goo.gl/CCtY7x .
+\emph{ISO/IEC 9126 Software Engineering - Product Quality},
+International Organization for Standardization.
+[Online] Available: \url{https://www.iso.org/standard/22749.html}
+[Accessed 4 Jun. 2018].
 
 \bibitem{emi-quality-model}
-\emph{EMI Quality Model} [Online]. Available: https://goo.gl/LdS6fL .
+M Alandes, et al.,
+Experiences with Software Quality Metrics in the EMI middleware,
+in \emph{J. Phys.: Conf. Ser.}
+vol. 396, n. 5, 052003, 2012.
 
 \bibitem{cordis:egi-inspire}
-\emph{European Grid Initiative: Integrated Sustainable Pan--European
-Infrastructure for Researchers in Europe (EGI--InSPIRE)} project, European
-Community Research and Development Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/95923\_en.html .
+I. Campos Plasencia,
+EGI.eu the European grid initiative.
+in \emph{4 th Iberian Grid Infrastructure Conf. Proc.},
+pp. 5--15, 2010.
+%\emph{European Grid Initiative: Integrated Sustainable Pan--European
+%Infrastructure for Researchers in Europe (EGI--InSPIRE)} project, European
+%Community Research and Development Information Service (CORDIS).
+%[Online] Available: \url{http://cordis.europa.eu/project/rcn/95923\_en.html}
+%[Accessed 4 Jun. 2018].
 
 \bibitem{egi-ops}
-\emph{EGI Operations Portal} [Online]. Available:
-https://operations-portal.egi.eu/ .
+H. Cordier, et al.,
+From EGEE Operations Portal towards EGI Operations Portal,
+in \emph{Data Driven e-Science (ISGC2010)},
+pp. 129--140, 2011.
+%\emph{EGI Operations Portal}.
+%[Online] Available: \url{https://operations-portal.egi.eu/}
+%[Accessed 4 Jun. 2018].
 
 \bibitem{ggus}
-\emph{GGUS} EGI Helpdesk [Online]. Available:
-https://ggus.eu/ .
+T. Antoni, et al.,
+Global grid user support—building a worldwide distributed user support infrastructure,
+\emph{J. Phys.: Conf. Ser.},
+vol. 119, n. 5, 052002, 2008.
+%\emph{GGUS} EGI Helpdesk.
+%[Online] Available: \url{https://ggus.eu/}
+%[Accessed 4 Jun. 2018].
 
 \bibitem{gocdb}
-\emph{GOCDB} EGI Grid Configuration Repository [Online]. Available:
-https://goc.egi.eu/ .
+G. Mathieu and J. Casson,
+GOCDB4, a New Architecture for the European Grid Infrastructure,
+in \emph{Data Driven e-Science (ISGC2010)},
+pp. 163--174, 2011.
+%\emph{GOCDB} EGI Grid Configuration Repository.
+%[Online] Available: \url{https://goc.egi.eu/}
+%[Accessed 4 Jun. 2018].
 
 \bibitem{egi-qc}
-\emph{EGI Quality Criteria} [Online]. Available:
-https://egi-qc.github.io/ .
-
-\bibitem{github}
-\emph{GITHUB} [Online]. Available:
-https://github.com/ .
-
-\bibitem{jenkins}
-\emph{Jenkins} Continuous Integration System [Online]. Available:
-https://jenkins-ci.org/ .
+\emph{EGI Quality Criteria}.
+[Online] Available: \url{https://egi-qc.github.io/}
+[Accessed 4 Jun. 2018].
 
 \bibitem{mario}
-M. David et al, \emph{Validation of Grid Middleware for the European Grid
-Infrastructure}, in Journal of Grid Computing, vol. 12, issue 3, pp. 543--558,
-Springer, 2014.
+M. David et al.,
+Validation of Grid Middleware for the European Grid Infrastructure,
+\emph{Journal of Grid Computing},
+vol. 12, issue 3, pp. 543--558, 2014.
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% TOBE checked if there are papers on these
 \bibitem{cordis:egi-engage}
 \emph{Engaging the EGI Community towards an Open Science Commons (EGI--ENGAGE)}
-project, European Community Research and Development Information Service
-(CORDIS), [Online]. Available:  http://cordis.europa.eu/project/rcn/194937\_en.html .
+project, European Community Research and Development Information Service (CORDIS).
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/194937\_en.html}
+[Accessed 4 Jun. 2018].
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \bibitem{cordis:indigo}
-\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
-(INDIGO--DataCloud)} project, European Community Research and Development
-Information Service (CORDIS) [Online]. Available:
-http://cordis.europa.eu/project/rcn/194882\_en.html .
-
-\bibitem{indigo-dockerhub}
-\emph{INDIGO--DataCloud DockerHub repository} [Online]. Available:
-https://hub.docker.com/u/indigodatacloud .
+Davide Salomoni, et al.,
+INDIGO-DataCloud: {A} data and computing platform to facilitate seamless access to e-infrastructures,
+\emph{CoRR},
+vol. abs/1711.01981, 2018,
+[Online] Available: \url{http://arxiv.org/abs/1711.01981}.
+%\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
+%(INDIGO--DataCloud)} project, European Community Research and Development
+%Information Service (CORDIS).
+%[Online] Available: \url{http://cordis.europa.eu/project/rcn/194882\_en.html}
+%[Accessed 4 Jun. 2018].
 
 \bibitem{indigo-d31}
-\emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1 [Online]. Available:
-https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31 .
+J. Gomes, et al.,
+\emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1.
+[Online] Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}
+[Accessed 4 Jun. 2018].
 
-\bibitem{indigo-jenkins}
-\emph{INDIGO--DataCloud Jenkins CI service} [Online]. Available:
-https://jenkins.indigo-datacloud.eu:8080/ .
+\bibitem{disvis}
+G.C.P. Van Zundert, A.M.J.J. Bonvin,
+DisVis: quantifying and visualizing the accessible interaction space of distance
+restrained biomolecular complexes,
+\emph{Bioinformatics}, vol. 31, n. 19, pp. 3222--3224, 2015.
 
-\bibitem{indigo-github}
-\emph{INDIGO--DataCloud GitHub Source Code repository} [Online]. Available:
-https://www.github.com/indigo-dc .
-
-\bibitem{indigo-gitbook}
-\emph{INDIGO--DataCloud GitBook Documentation repository} [Online]. Available:
-https://www.gitbook.com/@indigo-dc .
-
-\bibitem{indigo-puppet}
-\emph{INDIGO--DataCloud PuppetForge repository} [Online]. Available:
-https://forge.puppet.com/indigodc .
-
-\bibitem{indigo-galaxy}
-\emph{INDIGO--DataCloud Ansible Galaxy repository} [Online]. Available:
-https://galaxy.ansible.com/indigo-dc/ .
-
-\bibitem{grimoirelab}
-\emph{GrimoireLab} [Online]. Available: http://grimoirelab.github.io/ .
-
-\bibitem{indigo-pkg-repo}
-\emph{INDIGO--DataCloud Linux Package repository} [Online]. Available: https://jenkins.indigo-datacloud.eu/ .
-
-\bibitem{disvis} G.C.P. Van Zundert, A.M.J.J. Bonvin,
-DisVis: quantifying and visualizing the accessible interaction space of distance
-restrained biomolecular complexes, Bioinformatics 31 (2015) 3222--3224 .
-
-\bibitem{powerfit} G.C.P. Van Zundert, A.M.J.J. Bonvin,
+\bibitem{powerfit}
+G.C.P. Van Zundert, A.M.J.J. Bonvin,
 Fast and sensitive rigid--body fitting into cryo--EM density maps with PowerFit,
-AIMS Biophys. 2 (2015) 73--87.
+\emph{AIMS Biophys.}, vol. 2, n. 20150273, 73--87, 2015.
 

### 2018-06-11 paper-ieee-trans.tex
MI/5 page 6 from reviewer#1: devops CD rephrase and referencing

@@
 \usepackage{cite}
-\ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
 
 \begin{document}
-\title{Evolution of Methodologies in Delivering Reliable Software in the European Scientific Arena}
-%\title{Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis}
-%\title{15 years of European scientific software development: evolution and methodologies}
+\title{Evolution of Processes in Delivering Reliable Software in the European Scientific Arena}
 
@@
 
-% The paper headers
-%\markboth{IEEEtran for IEEE Journals}
-
 \maketitle
@@
 \begin{abstract}
+
 From the advent of Grid technology -- as the new paradigm of distributed
@@
 and e--Infrastructure creation, operation and management.
-% the creation of software development and e-Infrastructure coordination dedicated projects.
 This work examines the evolution of the software engineering methodologies used in past and
@@
 procedures is present since the very early stages of the software development process.
+
 \end{abstract}
@@
 \IEEEpeerreviewmaketitle
-
 \section{Introduction}
@@
 of processing, storing and analysis of the very large amount of data produced by the
-% Comment from reviewer
-%(Update: I see this is discussed later... so it would be good just to signpost at the start that a) the focus is really on "grids", and b) that specific examples will be given.
 LHC particle accelerator. A significant number of innovation projects and technology transfer
-activities -- later discussed in Section \ref{sec:ev} -- from the European Commission
+activities -- later discussed in Section~\ref{sec:ev} -- from the European Commission
 (EC)-funded 5th, 6th and 7th Framework Programmes, and recently from the Horizon 2020 programme
@@
 
-% Comment
-%%top of second column:  think about saying this more gently
-Major developments were carried out throughout the years focused on providing new
-functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
-The released software was less tested, resulting in a high rate of
-rollbacks and patches applied to production systems. Thereby, the stability and
+Major developments carried throughout the years where focused on providing new
+functionalities, while in the early days, the adoption of software engineering methodologies had a lower priority.
+The released software was less tested, resulting in a higher rate of
+rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
 % Comment
@@
 reliability of the e--Infrastructures was often compromised, affecting
-negatively their exploitation by the scientific communities.
+negatively their exploitation by the scientific communities \cite{aiftimiei}.
 % Comments
@@
 one can observe a continuous increase in the prominence and robustness of software
-testing and validation procedures. The
+testing and validation procedures \cite{aiftimiei}. The
 evolution of the software engineering methodologies (creation of the Agile
@@
 field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization) motivated this trend. All of which
-complemented by the emergence of automation and event--response tools.
+operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend. All of which
+are complemented by the emergence of automation and event--response tools.
 
@@
 The reminder of this paper is
-organized as follows. Section \ref{sec:ev} details the evolution of research on
-software reliability in EC--projects. Section \ref{sec:ntsr} provides
+organized as follows. Section~\ref{sec:ev} details the evolution of research on
+software reliability in EC--projects. Section~\ref{sec:ntsr} provides
 information about new research trends in software reliability based upon the
-experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section \ref{sec:con}.
+experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
 
@@
 
-The EC projects considered in this paper are listed chronologically in Table
-\ref{tab:eup}. Each project paved the way for the following ones, according to a logic evolution in which the role of the quality and reliability of the software have
-consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
-Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
+The EC projects considered in this paper are listed chronologically in Table~\ref{tab:eup}.
+Each project paved the way for the following ones, according to a logic evolution in which
+the role of the quality and reliability of the software have
+consistently increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
+committed to and aware of software reliability practices.
+Table~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
 the commonalities and specific features on software reliability.
@@
 \begin{table}[!h]
-%% increase table row spacing, adjust to taste
 \renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
 \caption{List of EC--projects}
@@
 \centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
 \begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
@@
 \begin{table*}[!h]
-%% increase table row spacing, adjust to taste
 \renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
 \caption{Features on software reliability in EC--projects}
@@
 \centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{adjustbox}{max width=0.5\textwidth}
 \begin{tabular}{llllllll}
@@
 Quality Criteria Definition        &       &       &$\surd$&$\surd$&$\surd$&$\surd$\\
-%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
 Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
@@
 \end{tabular}
-%\end{adjustbox}
 \end{table*}
@@
 \subsection{Enabling Grids for E--sciencE}
-% Comment
-%%P2 C2 L24: Typographical: It would be good to merge the three citations into one bracket/\cite
+
 The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
@@
 featured the internationalization (outside Europe) of the project, embracing worldwide research
-institutions and user communities. The software to sustain the increasing
+institutions and user communities~\cite{egee}. The software to sustain the increasing
 requirements coming from the diverse scientific communities needed developing a
@@
 
-\subsection{INtegrating Distributed data Infrastructures for Global
-explOitation}
+\subsection{INtegrating Distributed data Infrastructures for Global explOitation}
 
@@
 leveraged the lessons learned from the previous experiences described in
-Section \ref{sec:ev}. The expertise gathered throughout these years was highly
-profitable in terms of the evaluation and application of innovative software engineering
+Section~\ref{sec:ev}. The expertise gathered throughout these years was highly
+profitable regarding the evaluation and application of innovative software engineering
 methodologies, management of new technologies to put those methodologies
 into practice, compilation and analysis of an appropriate set of metrics to measure the
-% Comment
-%%L26-29 Missing verb
-%%(not a sentence—no verb) 
-quality of the implemented solutions. All the above with the additional complexity of being framed into
-large collaborative projects with an extensive list of partners.
+quality of the implemented solutions.
+An additional level of complexity, comes from the fact that these large collaborative projects
+have an extensive list of partners.
 
@@
 e--Infrastructure resource provider's early--adoption.
-
-%The {\sl INDIGO--DataCloud} project reflects the progress made in software
-%quality and reliability aspects throughout the past scientific European
-%experiences. Nevertheless (I would remove this), the project's Quality Assurance procedures are also
-%highly influenced by the insights of current big worldwide collaborations of
-%software development. These collaborations have continuously inspired the use of new
-%technologies and procedures to increase the robustness and quality of the
-%software being delivered. In this scenario, the implementation of the DevOps culture
-%is one of the major outcomes, as this paradigm has been progressively adopted by the
-%project.
 
@@
 project \cite{indigo-d31}. The policies were reviewed periodically taking into account the requirements
-from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies covered the 1) identification
+from the user communities, the feedback from the development teams and the insights of software
+engineering practices. These policies covered the 1) identification
 and description of the \emph{SQA requirements} that the produced software need
@@
 \item \textit{Code style}.
-The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
+The ultimate goal for code style assessment is to improve the readability and reusability of the
+source code produced under the scope of the
 project. Code style standards are enforced for every software component in the project stack. Community
@@
 \item \textit{Unit and Functional testing}.
-%% Comment from reviewer       
-%% These percentages appear as vague approximations by eye, please add the precise numbers rounded.
 Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
@@
 \item \textit{Code review}.
-% Comment
-%%Represents …   (not a sentence) 
-Represents the last step in the change management pipeline, once the candidate change has
+This phase is the last step in the change management pipeline, once the candidate change has
 successfully passed through the testing methods described previously. It implies the human--based
@@
 \label{subsec:ci}
+
 The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforced,
@@
 
-% Comment
-%%not a sentence, again
-Jenkins, the event--response CI system, and Docker container
-provisioning to provide instant computing power needed for executing the pipeline jobs.
+Jenkins is an event--response CI and Docker container
+provisioning system that provides instant computing power needed for executing the pipeline jobs.
 The project defined a source code contribution workflow, based on GitHub Pull Requests
@@
 As DevOps suggests, frequent releases positively affect the reliability of the software as
-they involve fewer changes and usually are more correlated. The software updates of
+they reduce the time in which a given defect is exposed or has an impact, allowing the 
+development teams to act promptly based on the regular user feedback \cite{chen2015}. The software updates of
 {\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
@@
 
-% Comment
-%%L42-44 not clear, please rephrase
-Software lifecycle demands as well from a continuous measurement and analysis
-process that permits timely reactions to discover and fix defects throughout the
-development, release and post--release or maintenance stages. Conscientious metric
+The implementation of a Software lifecycle phases, implies a continuous measurement and analysis
+process that allows timely reactions to discover and fix defects throughout the
+development, release and maintenance stages. Conscientious metric
 tracking and response sits at the core of an appropriate software maturity, as
@@
 
-%{\sl DataGrid - Research and Technological Development for an International Data Grid} project
-%has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-%FP5-IST - Programme for research, technological development and demonstration on a "User--friendly information society, 1998-2002" (IST--2000--25182E)
-%{\sl EGEE - Enabling grids For E--science} project
-%has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-% FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)
-%{\sl EGEE - Enabling grids For E--science--II} project
-% has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-% FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high--capacity and high--speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)
-%{\sl EGEE - Enabling grids For E--science--III} project has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
-%  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e--Science Grid infrastructures", (INFSO--RI--222667)
+{\sl DataGrid - Research and Technological Development for an International Data Grid}
+project has received funding from the European Union's Fifth Framework Programme under
+grant agreement IST--2000--25182E.
 
-{\sl DataGrid - Research and Technological Development for an International Data Grid} project has received funding from the European Union's Fifth Framework Programme under grant agreement IST--2000--25182E.
+{\sl EGEE - Enabling grids For E--science} project has received funding from the European
+Union's Sixth Framework Programme under grant agreement INFSO--RI--508833.
 
-{\sl EGEE - Enabling grids For E--science} project has received funding from the European Union's Sixth Framework Programme under grant agreement INFSO--RI--508833.
+{\sl EGEE - Enabling grids For E--science--II} project has received funding from the
+European Union's Sixth Framework Programme under grant agreement INFSO--RI--031688.
 
-{\sl EGEE - Enabling grids For E--science--II} project has received funding from the European Union's Sixth Framework Programme under grant agreement INFSO--RI--031688.
+{\sl EGEE - Enabling grids For E--science--III} project has received funding from the
+European Union's Seventh Framework Programme under grant agreement INFSO--RI--222667.
 
-{\sl EGEE - Enabling grids For E--science--III} project has received funding from the European Union's Seventh Framework Programme under grant agreement INFSO--RI--222667.
+{\sl ETICS - E--Infrastructure for Testing, Integration and Configuration of Software}
+project has received funding from the European Union's Sixth Framework Programme under
+grant agreement INFSO--RI--026753.
 
-{\sl ETICS - E--Infrastructure for Testing, Integration and Configuration of Software} project has received funding from the European Union's Sixth Framework Programme under grant agreement INFSO--RI--026753.
+{\sl ETICS - E--Infrastructure for Testing, Integration and Configuration of Software - Phase 2}
+project has received funding from the European Union's Seventh Framework Programme under grant
+agreement INFSO--RI--223782.
 
-{\sl ETICS - E--Infrastructure for Testing, Integration and Configuration of Software - Phase 2} project has received funding from the European Union's Seventh Framework Programme under grant agreement INFSO--RI--223782.
+{\sl EGI--InSPIRE - European Grid Initiative: Integrated Sustainable Pan--European
+Infrastructure for Researchers in Europe} project has received funding from the European
+Union's Seventh Framework Programme under grant agreement INFSO--RI--261323.
 
-{\sl EGI--InSPIRE - European Grid Initiative: Integrated Sustainable Pan--European Infrastructure for Researchers in Europe} project has received funding from the European Union's Seventh Framework Programme under grant agreement INFSO--RI--261323.
+{\sl EGI--Engage} project has received funding from the European Union's Horizon 2020
+research and innovation programme under grant agreement RIA--654142.
 
-{\sl EGI--Engage} project has received funding from the European Union's Horizon 2020 research and innovation programme under grant agreement RIA--654142.
-
-{\sl INDIGO--DataCloud} project has received funding from the European Union's Horizon 2020 research and innovation programme under grant agreement RIA--653549.
+{\sl INDIGO--DataCloud} project has received funding from the European Union's Horizon
+2020 research and innovation programme under grant agreement RIA--653549.
 
@@
 [Accessed 4 Jun. 2018].
+
+\bibitem{aiftimiei}
+A. Aiftimiei, and D. Dongiovanni, et al.,
+Improving the quality of EMI Releases by leveraging the
+EMI Testing Infrastructure.
+\emph{Journal of Physics: Conference Series},
+vol. 396, 2012.
 
@@
 
+\bibitem{soltesz}
+S. Soltesz, and H. P\"{o}tzl, et. al.,
+Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
+\emph{SIGOPS Oper. Syst. Rev.},
+vol. 41, n. 3, pp. 275--287, 2007.
+
 \bibitem{cordis:datagrid}
@@
 vol. 502, n. 2, pp. 376--381, 2003.
-%\emph{DataGrid project}, European Community Research and Development Information Service (CORDIS).
-%[Online] Available: \url{http://cordis.europa.eu/project/rcn/53665\_en.html}
-%[Accessed 4 Jun. 2018].
 
@@
 \bibitem{cordis:egee}
-F. Gagliardi and M. E. B\'{e}gin,
-EGEE -- providing a production quality grid for e-science,
-in \emph{2005 IEEE Inter. Symp. on Mass Storage Systems and Technology},
-pp. 88-92, 2005.
-%\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community Research and
-%Development Information Service (CORDIS). 
-%[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}
-%[Accessed 4 Jun. 2018].
+\emph{Enabling Grids for E--sciencE (EGEE)} project, European Community Research and
+Development Information Service (CORDIS).
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}
+[Accessed 4 Jun. 2018].
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% TOBE checked if there are papers on these
 \bibitem{cordis:egee2}
@@
 [Accessed 4 Jun. 2018].
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\bibitem{egee}
+F. Gagliardi and M. E. B\'{e}gin,
+EGEE -- providing a production quality grid for e-science,
+in \emph{2005 IEEE Inter. Symp. on Mass Storage Systems and Technology},
+pp. 88-92, 2005.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% TOBE checked if there are papers on these
-%\bibitem{cordis:etics}
-%\emph{E--Infrastructure for Testing, Integration and Configuration of Software (ETICS)}
-%project, European Community Research and Development Information Service (CORDIS).
-%[Online] Available: \url{http://cordis.europa.eu/project/rcn/80138\_en.html}
-%[Accessed 4 Jun. 2018].
-
-%\bibitem{cordis:etics2}
-%\emph{E--Infrastructure for Testing, Integration and Configuration of Software -
-%Phase 2 (ETICS 2)} project, European Community Research and Development
-%Information Service (CORDIS).
-%[Online]. Available: \url{http://cordis.europa.eu/project/rcn/86604\_en.html}
-%[Accessed 4 Jun. 2018].
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \bibitem{etics}
@@
 pp. 1--10, 2012.
-%\emph{European Middleware Initiative (EMI)} project, European Community
-%Research and Development Information Service (CORDIS).
-%[Online] Available: \url{http://cordis.europa.eu/project/rcn/95311\_en.html}
-%[Accessed 4 Jun. 2018].
 
@@
 pp. 5--15, 2010.
-%\emph{European Grid Initiative: Integrated Sustainable Pan--European
-%Infrastructure for Researchers in Europe (EGI--InSPIRE)} project, European
-%Community Research and Development Information Service (CORDIS).
-%[Online] Available: \url{http://cordis.europa.eu/project/rcn/95923\_en.html}
-%[Accessed 4 Jun. 2018].
 
@@
 pp. 129--140, 2011.
-%\emph{EGI Operations Portal}.
-%[Online] Available: \url{https://operations-portal.egi.eu/}
-%[Accessed 4 Jun. 2018].
 
@@
 vol. 119, n. 5, 052002, 2008.
-%\emph{GGUS} EGI Helpdesk.
-%[Online] Available: \url{https://ggus.eu/}
-%[Accessed 4 Jun. 2018].
 
@@
 pp. 163--174, 2011.
-%\emph{GOCDB} EGI Grid Configuration Repository.
-%[Online] Available: \url{https://goc.egi.eu/}
-%[Accessed 4 Jun. 2018].
 
@@
 [Accessed 4 Jun. 2018].
+
+\bibitem{chen2015}
+L. Chen,
+Continuous delivery: Huge benefits, but challenges too,
+\emph{IEEE Software},
+vol. 32, no 2, pp. 50--54, 2015.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% TOBE checked if there are papers on these
 \bibitem{cordis:egi-engage}
@@
 [Accessed 4 Jun. 2018].
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 [Online] Available: \url{http://arxiv.org/abs/1711.01981}.
-%\emph{INtegrating Distributed data Infrastructures for Global ExplOitation
-%(INDIGO--DataCloud)} project, European Community Research and Development
-%Information Service (CORDIS).
-%[Online] Available: \url{http://cordis.europa.eu/project/rcn/194882\_en.html}
-%[Accessed 4 Jun. 2018].
 

### 2018-06-11 paper-ieee-trans.tex
P3 C2 L29 reviewer#1 jargon removed

@@
 I appreciate this may all be highly meaningful to someone steeped in this operational area, but to an interested outsider it reads much like management jargon -- particuarly the final sentence: is this a long-winded way of saying "used during the commissioning period"?
-software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
-and metrics gathering definition} provides guidelines to develop quality
+software lifecycle:
+\begin{itemize}
+    \item A solid \textit{Software Quality Assurance (SQA) criteria and metrics gathering definition} provides guidelines to develop quality
     software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
     by a set of metrics, allowing prompt reactions to detected malfunctions throughout the
-process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
-the foregoing change--based development process and set up advanced delivery mechanisms.
-The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
-e--Infrastructure resource provider's early--adoption.
+    process. 
+    \item The \textit{promotion of automation} to enable the foregoing change--based validation approach 
+    and to accelerate the delivery process of new software versions.
+    \item A post-release validation carried out over the project's \textit{preview testbeds}, where the new product versions
+    are deployed as part of the integration validation, and the \textit{testing in production environments} by external resource providers.
+\end{itemize}
 

### 2018-06-11 paper-ieee-trans.tex
changed abstract

--- 2018-06-11 paper-ieee-trans.tex
@@
 \begin{document}
-\title{Evolution of Processes in Delivering Reliable Software in the European Scientific Arena}
+\title{Evolution of Methodologies in Delivering Reliable Software in the European Scientific Arena}
 
@@
 
-From the advent of Grid technology -- as the new paradigm of distributed
-computing -- to the current days of Cloud computing models, the continuous need
-of new tools and services to match the scientific community requirements has been
-addressed in Europe by various projects dedicated to software development
-and e--Infrastructure creation, operation and management.
+In the recent decades the development of new technologies in the distributed computing field (i.e. Grid and Cloud computing) have contributed to
+providing new tools and services able to match the scientific community requirements.
+In Europe various projects have been designed in order to develop and maintain software able to support and manage computing infrastructure.
+In this context, activities have been performed to deliver reliable software in the European research arena by using innovative software engineering solutions.
+
 This work examines the evolution of the software engineering methodologies used in past and
-present European Commission--funded projects to illustrate how the research on software
-reliability has progressed in Europe over the last 15 years following the foundation of the
-first project in 2001. In particular, we will summarize the techniques and procedures
+current European Commission--funded projects to illustrate how the research on software
+reliability has progressed over the last 15 years since the
+first project in 2001. We summarize the techniques and procedures
 applied to deliver quality software, highlighting the main challenges and barriers confronted
-at the time, and how they were partially or totally overcome. The knowledge base established
-throughout the years, sustained by the advances in the area of software engineering,
-definitely conformed a significant breakthrough in the reliability of software delivered in the
-European research arena. Recent software development projects funded by the European
-Commission are good evidences of such insights, where the enforcement of Quality Assurance
+at the time, and how they were partially or totally overcome. 
+Recent software development projects show good evidences of such insights, where the enforcement of quality assurance
 procedures is present since the very early stages of the software development process.
@@
 reliability of the e--Infrastructures was often compromised, affecting
-negatively their exploitation by the scientific communities \cite{aiftimiei}.
+negatively their exploitation by the scientific communities.
 % Comments
@@
 one can observe a continuous increase in the prominence and robustness of software
-testing and validation procedures \cite{aiftimiei}. The
+testing and validation procedures. The
 evolution of the software engineering methodologies (creation of the Agile
@@
 field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend. All of which
+operating--system--level virtualization known as containerization) motivated this trend. All of which
 are complemented by the emergence of automation and event--response tools.
@@
 experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
+
+\section{Background}
+\label{sec:b}
+
+\section{Research methodology}
+\label{sec:rm}
 
@@
 As DevOps suggests, frequent releases positively affect the reliability of the software as
-they reduce the time in which a given defect is exposed or has an impact, allowing the 
-development teams to act promptly based on the regular user feedback \cite{chen2015}. The software updates of
+they involve fewer changes and usually are more correlated. The software updates of
 {\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
@@
 
-\bibitem{aiftimiei}
-A. Aiftimiei, and D. Dongiovanni, et al.,
-Improving the quality of EMI Releases by leveraging the
-EMI Testing Infrastructure.
-\emph{Journal of Physics: Conference Series},
-vol. 396, 2012.
-
 \bibitem{agile-manifesto}
@@
 vol. 33, n. 3, pp. 32--34, 2016.
-
-\bibitem{soltesz}
-S. Soltesz, and H. P\"{o}tzl, et. al.,
-Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
-\emph{SIGOPS Oper. Syst. Rev.},
-vol. 41, n. 3, pp. 275--287, 2007.
 
@@
 
-\bibitem{chen2015}
-L. Chen,
-Continuous delivery: Huge benefits, but challenges too,
-\emph{IEEE Software},
-vol. 32, no 2, pp. 50--54, 2015.
-
 \bibitem{mario}

### 2018-06-11 paper-ieee-trans.tex
applied changes in introduction; added two references (aiftimiei and soltesz)

@@
 
-\IEEEPARstart{D}{uring} the last 15 years the provision of an European e--Infrastructure,
-supporting unified access to large--scale computing and intense data analysis
-applications, has been driven by the needs of scientific communities as part of their
-pan--European research collaborations. One such example is the Worldwide
-Large Hadron Collider (LHC) Computing Grid infrastructure, driven
+\IEEEPARstart{D}{uring} the last 15 years scientific communities have worked together to provide an European e--Infrastructure
+able to support unified access to large--scale computing and intense data analysis
+applications.
+One such example is the Worldwide
+Large Hadron Collider (LHC) Computing Grid infrastructure, which is driven
 by the particle physics scientific community, built to satisfy the computing needs
 of processing, storing and analysis of the very large amount of data produced by the
-LHC particle accelerator. A significant number of innovation projects and technology transfer
-activities -- later discussed in Section~\ref{sec:ev} -- from the European Commission
-(EC)-funded 5th, 6th and 7th Framework Programmes, and recently from the Horizon 2020 programme
-\cite{h2020} had as their main goal to develop software solutions that provide a seamless and
+LHC particle accelerator. 
+In the European Commission
+(EC)-funded 5th, 6th and 7th Framework Programmes and recently in the Horizon 2020 programme
+\cite{h2020}, a significant number of innovation projects and technology transfer
+activities have to develop software solutions that provide a seamless and
 transparent access of researchers to distributed computing and data resources.
 
-Major developments carried throughout the years where focused on providing new
-functionalities, while in the early days, the adoption of software engineering methodologies had a lower priority.
-The released software was less tested, resulting in a higher rate of
-rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
+Throughout the years, major developments have been focused on providing new
+functionalities, while in the early days the adoption of software engineering procedures have had a higher priority.
+The released software was less tested, causing a higher rate of
+rollbacks and patches applied to production systems, compared to the software developed and maintained in the current projects. Thereby, the stability and
 % Comment
@@
 %P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
-It became then apparent the need for a better balance between the addition of new features
+Therefore, it has become clear the need for a better balance between the addition of new features
 and the reliability of the released software. Consequently, incoming projects progressively
 devoted a substantial part of the software lifecycle to the improvement of the quality of the
-software being delivered. At this point, evolving software engineering practices were considered
-and gradually adopted to define and implement quality procedures, the organization of
+software being delivered. At this point, evolving software engineering practices have been considered
+so far and gradually adopted to define and implement quality procedures, the organization of
 cross--functional teams and deployment of pilot testbeds.
@@
 one can observe a continuous increase in the prominence and robustness of software
-testing and validation procedures. The
-evolution of the software engineering methodologies (creation of the Agile
-Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
+testing and validation procedures \cite{aiftimiei}. The
+evolution of the software engineering methodologies (e.g., the creation of the Agile
+Manifesto \cite{agile-manifesto} and the rise of DevOps practices \cite{zhu}) together
 with the parallel advancements in the information and communication technology
-field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization) motivated this trend. All of which
+field (e.g, the enabling technology for Cloud computing and the
+operating--system--level virtualization through containerization \cite{soltesz}) motivated this trend. All of which
 are complemented by the emergence of automation and event--response tools.
@@
 
+\bibitem{aiftimiei}
+Aiftimiei, A., Dongiovanni, D., et al.,  
+Improving the quality of EMI Releases by leveraging the
+EMI Testing Infrastructure.
+\emph{Journal of Physics: Conference Series},
+vol. 396, 2012.
+
 \bibitem{agile-manifesto}
@@
 vol. 33, n. 3, pp. 32--34, 2016.
+
+\bibitem{soltesz}
+Soltesz, S., P\"{o}tzl, H., et. al.,
+Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
+\emph{SIGOPS Oper. Syst. Rev.},
+vol. 41, n. 3, pp. 275--287, 2007.
 

### 2018-06-11 paper-ieee-trans.tex
added reference
changed title

@@
 \begin{document}
-\title{Evolution of Methodologies in Delivering Reliable Software in the European Scientific Arena}
+\title{Evolution of Processes in Delivering Reliable Software in the European Scientific Arena}
 

### 2018-06-11 paper-ieee-trans.tex

--- 2018-06-11 paper-ieee-trans.tex
@@
 
-In the recent decades the development of new technologies in the distributed computing field (i.e. Grid and Cloud computing) have contributed to
-providing new tools and services able to match the scientific community requirements.
-In Europe various projects have been designed in order to develop and maintain software able to support and manage computing infrastructure.
-In this context, activities have been performed to deliver reliable software in the European research arena by using innovative software engineering solutions.
-
+From the advent of Grid technology -- as the new paradigm of distributed
+computing -- to the current days of Cloud computing models, the continuous need
+of new tools and services to match the scientific community requirements has been
+addressed in Europe by various projects dedicated to software development
+and e--Infrastructure creation, operation and management.
 This work examines the evolution of the software engineering methodologies used in past and
-current European Commission--funded projects to illustrate how the research on software
-reliability has progressed over the last 15 years since the
-first project in 2001. We summarize the techniques and procedures
+present European Commission--funded projects to illustrate how the research on software
+reliability has progressed in Europe over the last 15 years following the foundation of the
+first project in 2001. In particular, we will summarize the techniques and procedures
 applied to deliver quality software, highlighting the main challenges and barriers confronted
-at the time, and how they were partially or totally overcome. 
-Recent software development projects show good evidences of such insights, where the enforcement of quality assurance
+at the time, and how they were partially or totally overcome. The knowledge base established
+throughout the years, sustained by the advances in the area of software engineering,
+definitely conformed a significant breakthrough in the reliability of software delivered in the
+European research arena. Recent software development projects funded by the European
+Commission are good evidences of such insights, where the enforcement of Quality Assurance
 procedures is present since the very early stages of the software development process.
@@
 field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization) motivated this trend. All of which
+operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend. All of which
 are complemented by the emergence of automation and event--response tools.
@@
 experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
-
-\section{Background}
-\label{sec:b}
-
-\section{Research methodology}
-\label{sec:rm}
 
@@
 
+\bibitem{aiftimiei}
+A. Aiftimiei, and D. Dongiovanni, et al.,
+Improving the quality of EMI Releases by leveraging the
+EMI Testing Infrastructure.
+\emph{Journal of Physics: Conference Series},
+vol. 396, 2012.
+
 \bibitem{agile-manifesto}
@@
 vol. 33, n. 3, pp. 32--34, 2016.
+
+\bibitem{soltesz}
+S. Soltesz, and H. P\"{o}tzl, et. al.,
+Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
+\emph{SIGOPS Oper. Syst. Rev.},
+vol. 41, n. 3, pp. 275--287, 2007.
 

### 2018-06-11 paper-ieee-trans.tex
Defining DevOps and rephrasing 3 pillars

@@
 I appreciate this may all be highly meaningful to someone steeped in this operational area, but to an interested outsider it reads much like management jargon -- particuarly the final sentence: is this a long-winded way of saying "used during the commissioning period"?
-software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
-and metrics gathering definition} provides guidelines to develop quality
-software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
-by a set of metrics, allowing prompt reactions to detected malfunctions throughout the
-process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
-the foregoing change--based development process and set up advanced delivery mechanisms.
-The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
-e--Infrastructure resource provider's early--adoption.
+software lifecycle:
+\begin{enumerate}
+    \item The definition of a \textit{Software Quality Assurance (SQA) criteria and a metrics gathering procedure} to provide guidelines
+    aiming at developing quality software, validating each change in the code, to facilitate its adoption. The software lifecycle is
+    continuously monitored by a set of metrics, allowing prompt reactions to detected malfunctions that may occur throughout the process. 
+    \item The \textit{promotion of automation} to enable the foregoing change--based validation approach and to accelerate the delivery
+    process of new software versions.
+    \item A post-release validation to be carried out over the project's \textit{preview testbeds}, where the new product versions
+    are deployed as part of the integration validation, and the \textit{testing in production environments} by external resource providers.
+\end{enumerate}
 
@@
 
-\subsection{DevOps practices}
+\subsection{Software quality validation and delivery automation}
 \label{sec:devops}
@@
 %P5 C2 L21: Define DevOps!
-The DevOps methods emphasize the SQA techniques to avoid infrastructure
-disruption whenever new developments are deployed into production systems. The
-following DevOps approaches have been progressively adopted throughout the project's
-lifetime, being one of its major outcomes in terms of software reliability.
+Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
+is hardly achieved without the aid of automation. The introduction of automation to validate the quality
+requirements increases the overall reliability of the produced software: automated testing is more 
+time-efficient, leading to higher coverages and increased defect detection, when compared with a manual
+approach \cite{rafi2012}.
+
+As any new change, involving a new feature or fix, passes through the automated
+SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
+between the continouous release of new features -- pushed by the development teams -- and the maintenance of
+stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
+theorizes and provides practical solutions that aim at unifying both software development (dev) and software
+operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
+developments are deployed into production systems. The following DevOps approaches have been progressively
+adopted throughout the project's lifetime, being one of its major outcomes in terms of software reliability.
 
@@
 
+\bibitem{rafi2012}
+D. M. Rafi el al.,
+Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,
+\emph{Proceedings of the 7th International Workshop on Automation of Software Test},
+36--42, 2012.
+
+\bibitem{chen2015}
+L. Chen,
+Continuous delivery: Huge benefits, but challenges too,
+\emph{IEEE Software},
+vol. 32, no 2, pp. 50--54, 2015.
+
 \bibitem{mario}

### 2018-06-13 paper-ieee-trans.tex
started adding background section

@@
 
-Major developments carried throughout the years where focused on providing new
-functionalities, while in the early days, the adoption of software engineering methodologies had a lower priority.
+Major developments carried throughout the years were focused on providing new
+functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
 The released software was less tested, resulting in a higher rate of
@@
 The reminder of this paper is
-organized as follows. Section~\ref{sec:ev} details the evolution of research on
+organized as follows. Section~\ref{sec:b} summaries the essential context of the research problem. Section~\ref{sec:ev} details the evolution of research on
 software reliability in EC--projects. Section~\ref{sec:ntsr} provides
@@
 experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
+
+\section{Background}
+\label{sec:b}
+
+In this section, we detail the level of knowledge in delivering reliable software before and after the EC-founded projects.
+We divide literature into four different domains that intersect this work.
+Firstly, we describe the software development process showing how the various phases reflect on software quality attributes.
+Next, we describe the impact of software requirements on software quality.
+In the last, we introduce metrics and measurements.
+In the last, we talk about software testing and code coverage.
+
+\emph{Metrics and measurements.} Metrics and measurements have an important role in the software development process \cite{fenton, capers}.
+Metrics are used to improve the ability to control, identify and measure the essential parameters of software  during  its  development.
+leading to improve the results of the software product \cite{goodman}.
+With a correct usage of  software  metrics,  it  is possible to determine which parts of software require to be changed or modified.
+In literature different types of metrics exist able to provide information on various software aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}.
+Often they have similar definitions, as a consequence they can be implemented in different ways according to the interpretation of their definition.
+This is the reason why it has been always a challenge to select a proper software metrics tool.
+
+\emph{Testing.} Testing is an important factor for software quality and reliability improvement.
+In literature there are many types of software testing \cite{myers} that aim at finding errors, bugs and faults during the software development process, possibly in the early stages.
+Test criteria focus on measuring the coverage of the test suite
 
@@
 requirements increases the overall reliability of the produced software: automated testing is more 
-time-efficient, leading to higher coverages and increased defect detection, when compared with a manual
+time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
 approach \cite{rafi2012}.
@@
 operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
-developments are deployed into production systems. The following DevOps approaches have been progressively
-adopted throughout the project's lifetime, being one of its major outcomes in terms of software reliability.
+developments are deployed into production systems. 
+The incremental adoption of DevOps approaches throughout the project's lifetime was a key player towards
+software reliability. Starting with a CI scenario at the first stages of the project, the software delivered
+was eventually validated and distributed using CD pipelines.
 
@@
 they reduce the time in which a given defect is exposed or has an impact, allowing the 
-development teams to act promptly based on the regular user feedback \cite{chen2015}. The software updates of
-{\sl INDIGO--DataCloud} products, that has been taking place since the second major release, are passing through
+development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
+{\sl INDIGO--DataCloud} products, that have been taking place since the second major release, are passing through
 a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
@@
 
+\bibitem{fenton}
+N.  Fenton,  and  S.  L.  Pfleeger.
+Software  Metrics:  A  Rigorous  and  Practical Approach,
+\emph{International Thomson Computer
+Press}, London, UK, second edition, 1997.
+
+\bibitem{capers}
+J. T. Capers.
+Applied Software Measurement
+\emph{McGrawHill}, 1991.
+
+\bibitem{goodman}
+P. Goodman.
+Practical Implementation of Software Metrics, 
+\emph{McGraw Hill}, London, 1993.
+
+\bibitem{mccabe}
+T. J. McCabe.
+A   Complexity   Measure,
+\emph{IEEE   Transactions   on   Software Engineering},
+Vol. 2, No. 4, pp. 308-320, 1976.
+
+\bibitem{chidamber}
+S.  R.  Chidamber  and  C.  E  Kemerer.
+A  metrics  suite  for  Object--oriented  design,
+\emph{IEEE Trans. Software Engineering}, 
+June 1994.
+
+\bibitem{lorenz}
+M. Lorenz and J. Kidd.
+Object-Oriented Software Metrics: A Practical Approach,
+\emph{IEEE publications}.
+
+\bibitem{myers}
+G. J. Myers and C. Sandler and T. Badgett.
+The art of software testing.
+\emph{John Wiley \& Sons}, 2011.
+
 \bibitem{cordis:datagrid}

### 2018-06-14 paper-ieee-trans.tex
Address the largely developed/mature components.

@@
 experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
+
+\section{Background}
+\label{sec:b}
+
+In this section, we detail the level of knowledge requested in delivering reliable software in order 
+to understand the effort employed in the EC-funded software projects over time. We divide literature 
+into four different domains that intersect this work: 1) software requirements, 2) software coding,
+3) metrics and measurements, 4) testing.
+
+\emph{Software requirements.} They are defined at various stages of a software development process
+as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
+analysis is a process for studying user needs to arrive at a definition of system, hardware or software
+requirements. Characteristics of requirements that can assess software reliability are consistency and
+correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable
+results, and the latter usually refers to the satisfaction of certain goals.
+
+\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer
+program in a programming language. To meet specific requirements, a program can be developed by using
+high-level or low-level programming languages that envelope different characteristics and may affect the
+result and quality of software. Furthermore, software is generally written by geographically distributed
+teams whose developers can change over time, therefore it is extremely important for source code to be
+understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
+development process to easily adapt software \cite{buse, baggen}: programming features that influence the
+code readability and maintainability and standardized code that promotes the comparability of results.
+
+\emph{Metrics and measurements.} According to IEEE \cite{radatz}, metrics is a function, with input as 
+the software data, and output is a value which could decide on how the given attribute affects the software.
+Metrics and measurements have an important role in the development process \cite{fenton, capers}. Metrics
+are used to improve the ability to control, identify and measure the essential parameters of software 
+during its development, leading to improve the results of the software product \cite{goodman}. With a correct
+usage of metrics, it is possible to determine which parts of the software require to be changed or modified.
+In the literature, different types of metrics exist to be able to provide information on various software
+aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}. Often they have 
+similar definitions, so as a consequence they can be implemented in different ways according to the
+interpretation of their definition. This is the reason why it has been always a challenge to select the
+proper software metric tools.
+
+\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component
+under specified conditions, observing or recording the results, and making an evaluation of some aspect of
+system or component. Testing is an important factor for software quality and reliability improvement. This
+activity is quite complicated due to the variety of programming languages, operating systems and hardware
+platforms that evolve over time. In the current literature, there are many types of software
+testing methods \cite{myers} that aim at finding errors, bugs and faults during the software development
+process, possibly in the early stages. Over time, testing metrics have been defined to assess software
+quality \cite{athanasiou}: code coverage is the most frequently used metric for test software reliability
+and there exist many tools for dynamic code coverage estimation according to the programming language
+\cite{horgan}. A dedicated effort have to be taken into account in order to integrate testing solutions
+during the software development process.
 
@@
 code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
-the project's first (INDIGO-1) and second (INDIGO-2) releases. By INDIGO-2, 53\% of the software components
+the project's first (INDIGO--1) and second (INDIGO--2) releases. By INDIGO--2, 53\% of the software components
 (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
 (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage. 
+The indiviual records followed also a growing trend (only 6 out of 32 components had lower values in 
+INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments 
+implemented from scratch within the project. Software tools and libraries from 
+external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
+by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
+but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
+analysis. However the values given apply to the entire codebase, even for the latter case of products. These 
+products challenged the application of the SQA policies described in this section, by two means:
+\begin{itemize}
+    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
+    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
+\end{itemize}
 
@@
 
+\bibitem{radatz}
+J. Radatz and A. Geraci and F. Katki.
+IEEE Standard glossary of softare engineering terminology.
+\emph{IEEE Std. 610.12-1990}, 1990.
+
+\bibitem{sommerville}
+I. Sommerville.
+Integrated requiremetns envineering: A tutorial.
+\emph{IEEE Software},
+vol. 14, issue 1, pp. 16--23, 2005.
+
+\bibitem{tashtoush}
+Y. Tashtoush and Z. Odat, et al.
+Impact of programming features on code readability.
+\emph{International Journal of Software Engineering and its Applications}, 
+vol. 7, issue 6, pp. 441--458, 2013.
+
+\bibitem{buse}
+R. P. Buse and W. R. Weimer.
+Learning a metric for code readability.
+\emph{IEEE Transactions on Software Engineering},
+vol. 36, issue 4, pp. 546--558, 2010.
+
+\bibitem{baggen}
+R. Baggen and J. P. Correia, et al.
+Standardized code quality benchmarking for improving software maintainability.
+\emph{Software Quality Journal},
+vol. 20, issue 2, pp. 287--307, 2012.
+
+\bibitem{fenton}
+N.  Fenton,  and  S.  L.  Pfleeger.
+Software  Metrics:  A  Rigorous  and  Practical Approach,
+\emph{International Thomson Computer
+Press}, London, UK, second edition, 1997.
+
+\bibitem{capers}
+J. T. Capers.
+Applied Software Measurement
+\emph{McGrawHill}, 1991.
+
+\bibitem{goodman}
+P. Goodman.
+Practical Implementation of Software Metrics, 
+\emph{McGraw Hill}, London, 1993.
+
+\bibitem{mccabe}
+T. J. McCabe.
+A   Complexity   Measure,
+\emph{IEEE   Transactions   on   Software Engineering},
+Vol. 2, No. 4, pp. 308-320, 1976.
+
+\bibitem{chidamber}
+S.  R.  Chidamber  and  C.  E  Kemerer.
+A  metrics  suite  for  Object--oriented  design,
+\emph{IEEE Trans. Software Engineering}, 
+June 1994.
+
+\bibitem{lorenz}
+M. Lorenz and J. Kidd.
+Object-Oriented Software Metrics: A Practical Approach,
+\emph{IEEE publications}.
+
+\bibitem{myers}
+G. J. Myers and C. Sandler and T. Badgett.
+The art of software testing.
+\emph{John Wiley \& Sons}, 2011.
+
+\bibitem{athanasiou}
+D. Athanasiou and A. Nugroho and J. Visser and A. Zaidman.
+Test code quality and its relation to issue handling performance.
+\emph{IEEE Transactions on Software Engineering}, 
+vol. 40, issue 7, pp. 1100--1125, 2014.
+
+\bibitem{horgan}
+J. R. Horgan and S. London and M. R. Lyu.
+Achieving software quality with testing coverage measuring.
+\emph{IEEE Computer},
+vol. 27, issue 9, pp. 60--69, 1994.
+
 \bibitem{cordis:datagrid}

### 2018-06-15 paper-ieee-trans.tex
Add a reference to SQA baseline document

@@
 The SQA criteria are a set of conventions and recommendations that pave the way for
-the adequate development, timely deliver and reliable operation of the produced software components.
+the adequate development, timely delivery and reliable operation of the produced software components.
 It emphasizes the quality requirements and best practices to be applied at the
@@
 by providing minimum requirements of documentation content and stressing the usage of automated
-solutions for software deployment.
+solutions for software deployment. The SQA criteria is continuously evolving \cite{sqa-baseline}, aiming at serving
+as a reference for future software development projects within the European research ecosystem.
 
@@
 
+\bibitem{sqa-baseline}
+Pablo Orviz et al.,
+A set of common software quality assurance baseline criteria for research projects,
+2017.
+[Online] Available: \url{http://digital.csic.es/bitstream/10261/160086/4/CommonSQA-v2.pdf}
+
 \bibitem{disvis}

### 2018-06-15 paper-ieee-trans.tex
added background and references

--- 2018-06-13 paper-ieee-trans.tex
@@
 experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
+
+\section{Background}
+\label{sec:b}
+
+In this section, we detail the level of knowledge requested in delivering reliable software in order to understand the effort employed in the EC-funded software projects over time.
+We divide literature into four different domains that intersect this work: 1. software requirements, 2. software coding 3. metrics and measurements, 4. testing.
+
+\emph{Software requirements.} They are defined at various stages of a software development process as a specification of what should be implemented.
+According to IEEE \cite{radatz}, requirement analysis is a process for studying user needs to arrive at a definition of system, hardware or software requirements.
+Characteristics of requirements that can assess software reliability are consistency and correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable results, and the latter usually refers to the satisfaction of certain goals.
+
+\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer program in a programming language.
+To meet specific requirements, a program can be developed by using high-level or low-level programming languages that envelope different characteristics and may affect result and quality of software.
+Furthermore, software is generally written by geographically distributed teams whose developers can change over time, therefore it is extremely important for source code to be understandable \cite{tashtoush}.
+There are various instruments that can be adopted during the software development process to easily adapt software \cite{buse, baggen}: programming features that influence the code readability and maintainability and standardized code that promotes the comparability of results.
+
+\emph{Metrics and measurements.} According to IEEE \cite{radatz}, metrics   is   a
+function,  with  input  as  the  software  data,  and  output  is  a
+value  which  could  decide  on  how  the  given  attribute  affect
+the software. Metrics and measurements have an important role in the development process \cite{fenton, capers}.
+Metrics are used to improve the ability to control, identify and measure the essential parameters of software  during  its  development.
+leading to improve the results of the software product \cite{goodman}.
+With a correct usage of  software  metrics,  it  is possible to determine which parts of software require to be changed or modified.
+In literature different types of metrics exist able to provide information on various software aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}.
+Often they have similar definitions, as a consequence they can be implemented in different ways according to the interpretation of their definition.
+This is the reason why it has been always a challenge to select a proper software metrics tool.
+
+\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component under specified conditions, observing or recording the results, and making an evaluation of some aspect of system or component.
+Testing is an important factor for software quality and reliability improvement.
+This activity is quite complicated due to the variety of programming languages, operating systems and hardware platforms that evolve over time \cite{myers}.
+In literature there are many types of software testing \cite{myers} that aim at finding errors, bugs and faults during the software development process, possibly in the early stages.
+Over time testing metrics have been defined to assess software quality \cite{athanasiou}: code coverage is the most frequently used metric for test software reliability and there exist many tools for dynamic code coverage estimation according to the programming language \cite{horgan}.
+A dedicated effort have to be taken into account in order to integrate testing solutions during the software development process.
 

### 2018-06-15 paper-ieee-trans.tex
Adding some text in the conclusion about what happens when successful projects end

--- 2018-06-15 paper-ieee-trans.tex
@@
 
+Lessons learned showed that sustainability must be a key pursuit. The achievements, in terms of
+processes, knowledge, software and infrastructures, shall be profited from former successful 
+completed projects. In this regard, the more recent project described in this paper, {\sl INDIGO-DataCloud},
+may have set the path for future reference in the European software development research ecosystem.
+The SQA criteria, describing the quality requirements that might be considered during the entire
+software lifecycle, has been published as open access and it is periodically reviewed to incorporate
+new recommendations and best practices. Successful software products created during the lifetime of the 
+project are maintained leveraging the same source code repositories and CI infrastructure. Seamless 
+knowledge transfer among software development initiatives is key for the good progress of delivering
+reliable software.
+
 \section*{Acknowledgment}

### 2018-06-15 paper-ieee-trans.tex
fix minor typos and nesting

--- 2018-06-15 paper-ieee-trans.tex
@@
 The SQA criteria are a set of conventions and recommendations that pave the way for
-the adequate development, timely delivery and reliable operation of the produced software components.
+the adequate development, timely deliver and reliable operation of the produced software components.
 It emphasizes the quality requirements and best practices to be applied at the
@@
 by providing minimum requirements of documentation content and stressing the usage of automated
-solutions for software deployment. The SQA criteria is continuously evolving \cite{sqa-baseline}, aiming at serving
-as a reference for future software development projects within the European research ecosystem.
+solutions for software deployment.
 
@@
 code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
-the project's first (INDIGO--1) and second (INDIGO--2) releases. By INDIGO--2, 53\% of the software components
+the project's first (INDIGO-1) and second (INDIGO-2) releases. By INDIGO-2, 53\% of the software components
 (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
 (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
-The indiviual records followed also a growing trend (only 6 out of 32 components had lower values in 
-INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments 
-implemented from scratch within the project. Software tools and libraries from 
-external open--source projects -- such as {\sl tosca-parser} or {\sl heat-translator} --, contributed upstream
-by the project, and well--established products, involved in {\sl INDIGO--DataCloud}
-but not contributing with the 100\% of their codebases -- such as {\sl dCache} --, are also considered in the
-analysis. However the values given apply to the entire codebase, even for the latter case of products. These 
-products challenged the application of the SQA policies described in this section, by two means:
-\begin{itemize}
-    \item Previously developed code was not refactored since it was not \textit{owned} by the project.
-    \item An agreement on prevailing SQA policies for such cases where quality practices were already in place.
-\end{itemize}
 
@@
 
-\bibitem{sqa-baseline}
-Pablo Orviz et al.,
-A set of common software quality assurance baseline criteria for research projects,
-2017.
-[Online] Available: \url{http://digital.csic.es/bitstream/10261/160086/4/CommonSQA-v2.pdf}
-
 \bibitem{disvis}

### 2018-06-15 paper-ieee-trans.tex
fix
fix criteria (plural)

--- 2018-06-15 paper-ieee-trans.tex
@@
 by providing minimum requirements of documentation content and stressing the usage of automated
-solutions for software deployment. The SQA criteria is continuously evolving \cite{sqa-baseline}, aiming at serving
+solutions for software deployment. The SQA criteria are continuously evolving \cite{sqa-baseline}, aiming at serving
 as a reference for future software development projects within the European research ecosystem.
@@
 
-Lessons learned showed that sustainability must be a key pursuit. The achievements, in terms of
+Lessons learned show that sustainability is a key pursuit. The achievements, in terms of
 processes, knowledge, software and infrastructures, shall be profited from former successful 
@@
 The SQA criteria, describing the quality requirements that might be considered during the entire
-software lifecycle, has been published as open access and it is periodically reviewed to incorporate
+software lifecycle, have been published as open access and it is periodically reviewed to incorporate
 new recommendations and best practices. Successful software products created during the lifetime of the 

### 2018-06-15 paper-ieee-trans.tex
better link last sentence
fix

@@
 The SQA criteria, describing the quality requirements that might be considered during the entire
-software lifecycle, have been published as open access and it is periodically reviewed to incorporate
+software lifecycle, have been published as open access and periodically reviewed to incorporate
 new recommendations and best practices. Successful software products created during the lifetime of the 
-project are maintained leveraging the same source code repositories and CI infrastructure. Seamless 
-knowledge transfer among software development initiatives is key for the good progress of delivering
-reliable software.
+project are maintained leveraging the same source code repositories and CI infrastructure. A seamless 
+knowledge transfer among subsequent software development initiatives is key for the good progress of
+delivering reliable software.
 

### 2018-06-25 paper-ieee-trans.tex
all reviwer 3 comments done, remove some latex comments

@@
 Section~\ref{sec:ev}. The expertise gathered throughout these years was highly
-profitable regarding the evaluation and application of innovative software engineering
+profitable regarding evaluation and application of innovative software engineering
 methodologies, management of new technologies to put those methodologies
@@
 
-In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
-% Comment
-%%P3 C2 L29: This introductory paragraph goes into a bit too much jargony detail, I think, especially given that it is followed by in-depth coverage of each "pillar".
-I suggest a clearer identification of each pillar, maybe via bullets or definition paragraphs, avoiding or with general-audience definition of jargon like DevOps, delivery mechanisms, two-factor validation, pilot testbeds, and early adoption.
-I appreciate this may all be highly meaningful to someone steeped in this operational area, but to an interested outsider it reads much like management jargon -- particuarly the final sentence: is this a long-winded way of saying "used during the commissioning period"?
-software lifecycle:
+In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the software lifecycle:
+
 \begin{enumerate}
@@
 \label{sec:devops}
-% Comment
-%P5 C2 L21: Define DevOps!
 Tackling the granularity of the former SQA requirements -- most of which accounted in a per--change basis --
@@
 (PRs), where each change automatically triggers in Jenkins the associated quality and
-metric tracking checks on PR creation or update. The set of required checks contained
-in the CI pipeline, report their exit status back to GitHub preventing the change to be
-approved if they are not successfully executed. As each change is validated, the chances
+metric tracking checks on PR creation or update.
+
+The CI pipeline is composed from a set of required checks, their exit status
+is reported back to GitHub. The changes can be prevented if those tests are
+not successfully executed.
+As each change is validated, the chances
 of early detection of defects increases. Within this scenario, the cost of defect solving
@@
 \subsubsection{Continuous Delivery}
-% Comment from reviewer
-%Please add a reference.
 As DevOps suggests, frequent releases positively affect the reliability of the software as
@@
 the installation and configuration of the component relying on the testing repository. On
-successful completion, the packages are then automatically signed and subsequently moved to
-the preview branch, from where they finally be made available in the production branch after
-the release manager supervises the process.
+successful completion, the packages are automatically signed and subsequently moved to
+the preview branch.
+In the last step, the release manager supervises the process to move the  packages to the
+production branch.
 
@@
 resulted in the implementation of CD pipelines to extend automation to the
-delivery of packages. Thereby, safe and frequent software releases are doable, reducing
+delivery of packages. Thereby, safe and frequent software releases are possible, reducing
 the impact on production systems. In this regard, a pre--production validation,

### 2018-06-25 paper-ieee-trans.tex
all reviwer 4 comments done, except 1 OP

@@
 throughout the years, sustained by the advances in the area of software engineering,
-definitely conformed a significant breakthrough in the reliability of software delivered in the
+definitely constituted a significant breakthrough in the reliability of software delivered in the
 European research arena. Recent software development projects funded by the European
-Commission are good evidences of such insights, where the enforcement of Quality Assurance
+Commission are good evidence of such insights, where the enforcement of Quality Assurance
 procedures is present since the very early stages of the software development process.
@@
 \cite{h2020} had as their main goal to develop software solutions that provide a seamless and
-transparent access of researchers to distributed computing and data resources.
+transparent access by researchers to distributed computing and data resources.
 
@@
 field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend. All of which
-are complemented by the emergence of automation and event--response tools.
+operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend.
+This evolution was complemented by the emergence of automation and event--response tools.
 
@@
 
-The reminder of this paper is
-organized as follows. Section~\ref{sec:b} summaries the essential context of the research problem. Section~\ref{sec:ev} details the evolution of research on
+The remainder of this paper is
+organized as follows. Section~\ref{sec:b} summarizes the essential context of the research problem.
+Section~\ref{sec:ev} details the evolution of research on
 software reliability in EC--projects. Section~\ref{sec:ntsr} provides
 information about new research trends in software reliability based upon the
-experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
+experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC).
+Conclusions are drawn in Section~\ref{sec:con}.
 
@@
 Each project paved the way for the following ones, according to a logic evolution in which
-the role of the quality and reliability of the software have
+the role of the quality and reliability of the software has
 consistently increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
@@
 collaborative projects, a big effort was spent in devising solutions for several
-challenges \cite{datagrid}, namely the 1) communication overhead, as a
+challenges \cite{datagrid}, namely 1) the communication overhead, as a
 result of the large geographical separation of the involved parties in the
@@
 from the user communities, the feedback from the development teams and the insights of software
-engineering practices. These policies covered the 1) identification
-and description of the \emph{SQA requirements} that the produced software need
-to comply with, and the 2) \emph{quality metrics}, to monitor each software product's
+engineering practices. These policies covered 1) the identification
+and description of the \emph{SQA requirements} that the produced software needs
+to comply with, and 2) the \emph{quality metrics}, to monitor each software product's
 behaviour throughout the development, release and post--release stages.
@@
 \textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
-definitions. The definition files appear as additional source of documentation
--- self--documenting code -- as they sequentially guide the component's deployment process in multiple
+definitions. The definition files appear as an additional source of documentation
+-- self--documenting code -- as they sequentially guide the component's deployment process on multiple
 platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
@@
 and Puppet\footnote{\url{https://forge.puppet.com/indigodc}}. A representative example of
-such contribution are the \textit{50 roles} developed from scratch and currently hosted in
+such contributions are the \textit{50 roles} developed from scratch and currently hosted in
 the Ansible Galaxy portal\footnote{\url{https://galaxy.ansible.com/indigo-dc/}}. Fig.~\ref{fig:fig_confman}
@@
 and Jenkins\footnote{\url{https://jenkins.io/}} service, and graphically displayed as GitHub
-pages using GrimoireLab framework\footnote{\url{http://grimoirelab.github.io/}}.
+pages using the GrimoireLab framework\footnote{\url{http://grimoirelab.github.io/}}.
 Per--component weekly reports, including the SQA requirement
@@
 for any piece of produced software, the testing requirements defined in the SQA criteria,
-as described in Section \ref{subsec:sqa}. Such environment requires an automation
+as described in Section \ref{subsec:sqa}. Such an environment requires an automation
 ready--to--go infrastructure where the different services and technologies involved interact with each
@@
 integration tests for the software components where automation of this type of testing
-is applicable, or security linters for the code static analysis of suspicious
-constructs that could lead to security risks. Metrics gathering also takes part of
+is applicable, or security linters for the static code analysis of suspicious
+constructs that could lead to security risks. Metrics gathering is also a part of
 the CI pipeline to have a per--change trend evolution of the common source code
@@
 
-The practical implementation of the CI scenario leveraged on tightly integrated open
+The practical implementation of the CI scenario leverage from a tightly integrated open
 source tools such as GitHub\footnote{\url{https://github.com/}}, as the online source code repository,
@@
 successful execution of the CI pipeline previously described. Consequently, CI--validated
-software components are automatically delivered in online repositories without the need of
-any human interaction but ultimately requiring expert supervision to validate the results.
-
-{\sl INDIGO--DataCloud} software stack is delivered in form of Linux software packages (rpms, debs)
+software components are automatically delivered in online repositories. Nonetheless,
+expert supervision to validate the results, is required.
+
+{\sl INDIGO--DataCloud} software stack is delivered in the form of Linux software packages (rpms, debs)
 and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
@@
 possible to production ones. A map of the Pilot Preview
-infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}, it shows the
-resources providers and the components or services they deployed and supported.
+infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}. It shows the
+resource providers and the components or services they deployed and supported.
 
@@
 the impact on production systems. In this regard, a pre--production validation,
-through the preview testbeds or resource centers early adoption -- handled by the
+through the preview testbeds or resource center’s early adoption -- handled by the
 Staged--rollout process -- alleviates the likelihood of discovering uncovered
@@
 
-The implementation of a Software lifecycle phases, implies a continuous measurement and analysis
-process that allows timely reactions to discover and fix defects throughout the
+The software lifecycle demands as well a continuous measurement and analysis
+process, that allows timely reactions to discover and fix defects throughout the
 development, release and maintenance stages. Conscientious metric

### 2018-06-26 paper-ieee-trans.tex
paragraph break, check reviewer 1 comments
fix two issues sent by Eli

@@
 of processing, storing and analysis of the very large amount of data produced by the
-LHC particle accelerator. A significant number of innovation projects and technology transfer
+LHC particle accelerator.
+
+A significant number of innovation projects and technology transfer
 activities -- later discussed in Section~\ref{sec:ev} -- from the European Commission
@@
 consistently increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
-committed to and aware of software reliability practices.
+committed to, and aware, of software reliability practices.
 Table~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
@@
 As each change is validated, the chances
-of early detection of defects increases. Within this scenario, the cost of defect solving
+of early detection of defects increase. Within this scenario, the cost of defect solving
 is dramatically reduced and the reliability of the software solutions improved, as any
@@
 software components are automatically delivered in online repositories. Nonetheless,
-expert supervision to validate the results, is required.
+expert supervision to validate the results is required.
 

### 2018-06-26 paper-ieee-trans.tex
answer to Pablo comment, fix some refs

@@
 consistently increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
-committed to, and aware, of software reliability practices.
+committed to, and aware of, software reliability practices.
 Table~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
@@
 \bibitem{aiftimiei}
-A. Aiftimiei, and D. Dongiovanni, et al.,
+A. Aiftimiei, et al.,
 Improving the quality of EMI Releases by leveraging the
 EMI Testing Infrastructure.
-\emph{Journal of Physics: Conference Series},
+\emph{J. Phys. Conf. Ser.},
 vol. 396, 2012.
@@
 \bibitem{agile-manifesto}
-Beck, K., Beedle, et al., (2001).
+K. Beck, et al., (2001),
 Manifesto for Agile Software Development.
@@
 \bibitem{zhu}
-L. Zhu and L. Bass and G. Champlin--Scharff,
+L. Zhu, L. Bass and G. Champlin--Scharff,
 DevOps and Its Practices,
-\emph{IEEE Software},
+\emph{IEEE Softw.},
 vol. 33, n. 3, pp. 32--34, 2016.
@@
 \bibitem{soltesz}
-S. Soltesz, and H. P\"{o}tzl, et. al.,
+S. Soltesz, et. al.,
 Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
@@
 \bibitem{radatz}
-J. Radatz and A. Geraci and F. Katki.
+J. Radatz, A. Geraci and F. Katki,
 IEEE Standard glossary of softare engineering terminology.
-\emph{IEEE Std. 610.12-1990}, 1990.
+\emph{IEEE Std. 610.12--1990}, 1990.
 
 \bibitem{sommerville}
-I. Sommerville.
-Integrated requiremetns envineering: A tutorial.
-\emph{IEEE Software},
+I. Sommerville,
+Integrated requirements engineering: A tutorial.
+\emph{IEEE Softw.},
 vol. 14, issue 1, pp. 16--23, 2005.
@@
 \bibitem{tashtoush}
-Y. Tashtoush and Z. Odat, et al.
+Y. Tashtoush, et al.,
 Impact of programming features on code readability.
-\emph{International Journal of Software Engineering and its Applications}, 
+\emph{Int. J. Softw. Eng. Appl.}, 
 vol. 7, issue 6, pp. 441--458, 2013.
@@
 \bibitem{buse}
-R. P. Buse and W. R. Weimer.
+R. P. Buse and W. R. Weimer,
 Learning a metric for code readability.
-\emph{IEEE Transactions on Software Engineering},
+\emph{IEEE Trans. Softw. Eng.},
 vol. 36, issue 4, pp. 546--558, 2010.
@@
 \bibitem{baggen}
-R. Baggen and J. P. Correia, et al.
+R. Baggen, et al.,
 Standardized code quality benchmarking for improving software maintainability.
-\emph{Software Quality Journal},
+\emph{Softw. Qual. J.},
 vol. 20, issue 2, pp. 287--307, 2012.

### 2018-06-26 paper-ieee-trans.tex
halfway on the refs

@@
 \emph{Horizon 2020} -- European Commission.
-[Online] Available: \url{https://ec.europa.eu/programmes/horizon2020/}
+[Online]. Available: \url{https://ec.europa.eu/programmes/horizon2020/}.
 [Accessed 4 Jun. 2018].
@@
 \bibitem{aiftimiei}
-A. Aiftimiei, et al.,
-Improving the quality of EMI Releases by leveraging the
-EMI Testing Infrastructure.
+C Aiftimiei, A Ceccanti, D Dongiovanni, A Di Meglio and F Giacomini,
+``Improving the quality of EMI Releases by leveraging the EMI Testing Infrastructure,''
 \emph{J. Phys. Conf. Ser.},
-vol. 396, 2012.
+vol. 396, no. 5, 2012.
 
 \bibitem{agile-manifesto}
-K. Beck, et al., (2001),
-Manifesto for Agile Software Development.
-[Online] Available: \url{http://www.agilemanifesto.org/}
+K. Beck \emph{et al.},
+``Manifesto for Agile Software Development,'' 2012.
+[Online]. Available: \url{http://www.agilemanifesto.org/}.
 [Accessed 4 Jun. 2018].
@@
 L. Zhu, L. Bass and G. Champlin--Scharff,
-DevOps and Its Practices,
+``DevOps and Its Practices,''
 \emph{IEEE Softw.},
@@
 \bibitem{soltesz}
-S. Soltesz, et. al.,
-Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,
+S. Soltesz \emph{et al.},
+``Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,''
 \emph{SIGOPS Oper. Syst. Rev.},
-vol. 41, n. 3, pp. 275--287, 2007.
+vol. 41, no. 3, pp. 275--287, Mar. 2007.
 
 \bibitem{radatz}
-J. Radatz, A. Geraci and F. Katki,
-IEEE Standard glossary of softare engineering terminology.
-\emph{IEEE Std. 610.12--1990}, 1990.
+``IEEE Standard Glossary of Software Engineering Terminology,''
+\emph{IEEE Std. 610.12--1990}, pp.1--84, Dec. 31 1990.
 
@@
 I. Sommerville,
-Integrated requirements engineering: A tutorial.
-\emph{IEEE Softw.},
-vol. 14, issue 1, pp. 16--23, 2005.
+``Integrated requirements engineering: A tutorial,''
+\emph{IEEE Softw.}, vol. 22, no. 1, pp. 16--23, Jan.--Feb. 2005.
 
 \bibitem{tashtoush}
-Y. Tashtoush, et al.,
-Impact of programming features on code readability.
+Y. Tashtoush, Z. Odat, I. Alsmadi, M. Yatim,
+``Impact of programming features on code readability,''
 \emph{Int. J. Softw. Eng. Appl.}, 
-vol. 7, issue 6, pp. 441--458, 2013.
+vol. 7, no. 6, pp. 441--458, Nov. 2013.
 
@@
 R. P. Buse and W. R. Weimer,
-Learning a metric for code readability.
+``Learning a metric for code readability,''
 \emph{IEEE Trans. Softw. Eng.},
-vol. 36, issue 4, pp. 546--558, 2010.
+vol. 36, no. 4, pp. 546--558, Jul. 2010.
 
 \bibitem{baggen}
-R. Baggen, et al.,
-Standardized code quality benchmarking for improving software maintainability.
+R. Baggen, J. P. Correia, K. Schill and J. Visser,
+``Standardized code quality benchmarking for improving software maintainability,''
 \emph{Softw. Qual. J.},
-vol. 20, issue 2, pp. 287--307, 2012.
+vol. 20, no. 2, pp. 287--307, Jun. 2012.
 
 \bibitem{fenton}
-N.  Fenton,  and  S.  L.  Pfleeger.
-Software  Metrics:  A  Rigorous  and  Practical Approach,
-\emph{International Thomson Computer
-Press}, London, UK, second edition, 1997.
+N. Fenton  and  S.  L.  Pfleeger,
+``Software  Metrics:  A  Rigorous  and  Practical Approach,''
+\emph{International Thomson Computer Press},
+London, UK, second edition, 1997.
 
 \bibitem{capers}
-J. T. Capers.
-Applied Software Measurement
-\emph{McGrawHill}, 1991.
+J. Capers,
+``Applied Software Measurement (2nd Ed.): Assuring Productivity and Quality,''
+\emph{McGraw Hill}, New York, NY, USA, 1997.
 
 \bibitem{goodman}
-P. Goodman.
-Practical Implementation of Software Metrics, 
-\emph{McGraw Hill}, London, 1993.
+P. Goodman,
+``Practical Implementation of Software Metrics,''
+\emph{McGraw Hill}, New York, NY, USA, 1993.
 
 \bibitem{mccabe}
-T. J. McCabe.
-A   Complexity   Measure,
-\emph{IEEE   Transactions   on   Software Engineering},
-Vol. 2, No. 4, pp. 308-320, 1976.
+T. J. McCabe,
+``A Complexity Measure,''
+\emph{IEEE Trans. Softw. Eng.},
+vol. 2, no. 4, pp. 308--320, Dec. 1976.
 
 \bibitem{chidamber}
-S.  R.  Chidamber  and  C.  E  Kemerer.
-A  metrics  suite  for  Object--oriented  design,
-\emph{IEEE Trans. Software Engineering}, 
-June 1994.
+S. R. Chidamber and C. E. Kemerer,
+``A metrics suite for Object--oriented design,''
+\emph{IEEE Trans. Softw. Eng.},
+vol. 20, no. 6, pp. 476--493, Jun 1994.
 
 \bibitem{lorenz}
-M. Lorenz and J. Kidd.
-Object-Oriented Software Metrics: A Practical Approach,
-\emph{IEEE publications}.
+M. Lorenz and J. Kidd,
+\emph{Object--Oriented Software Metrics: A Practical Approach}.
+Prentice-Hall, Inc., Upper Saddle River, NJ, USA, 1994.
 
 \bibitem{myers}
-G. J. Myers and C. Sandler and T. Badgett.
-The art of software testing.
-\emph{John Wiley \& Sons}, 2011.
+G. J. Myers, C. Sandler and T. Badgett,
+\emph{The art of software testing, 3rd Ed.}.
+Wiley Publishing, Dec. 2011.
 
 \bibitem{athanasiou}
-D. Athanasiou and A. Nugroho and J. Visser and A. Zaidman.
-Test code quality and its relation to issue handling performance.
-\emph{IEEE Transactions on Software Engineering}, 
-vol. 40, issue 7, pp. 1100--1125, 2014.
+D. Athanasiou, A. Nugroho, J. Visser and A. Zaidman,
+``Test Code Quality and Its Relation to Issue Handling Performance,''
+\emph{IEEE Trans. Softw. Eng.},
+vol. 40, no. 11, pp. 1100--1125, Nov. 2014.
 
 \bibitem{horgan}
-J. R. Horgan and S. London and M. R. Lyu.
-Achieving software quality with testing coverage measuring.
-\emph{IEEE Computer},
-vol. 27, issue 9, pp. 60--69, 1994.
+J. R. Horgan, S. London and M. R. Lyu,
+``Achieving software quality with testing coverage measuring,''
+\emph{Comput.},
+vol. 27, no. 9, pp. 60--69, Sep. 1994.
 
@@
 P. Kunszt,
-European DataGrid project: status and plans,
+``European DataGrid project: status and plans,''
 \emph{Nucl. Instr. Meth. Phys. Res. A},
-vol. 502, n. 2, pp. 376--381, 2003.
+vol. 502, no. 2, pp. 376--381, Apr. 2003.
 
 \bibitem{gagliardi}
-Gagliardi F., Jones B., Reale M., Burke S.,
-European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E--science Applications,
-in \emph{Performance Evaluation of Complex Systems: Techniques and Tools, LNCS},
-vol. 2459, 2002.
+F. Gagliardi, B. Jones, M. Reale and S. Burke,
+``European DataGrid Project: Experiences of Deploying a Large Scale Testbed for E--science Applications,''
+in \emph{Performance Evaluation of Complex Systems: Techniques and Tools,}
+Performance 2002. LNCS, vol. 2459, pp. 480--499, 2002.
 
@@
 I. Foster and C. Kesselman,
-Globus: a Metacomputing Infrastructure Toolkit,
-\emph{Int. J. Supercomput. Appl.},
-vol. 11, no. 2, pp. 115--128, 1997.
+``Globus: a Metacomputing Infrastructure Toolkit,''
+\emph{Int. J. High Perfor. Comput. Appl.,}
+vol. 11, no. 2, pp. 115--128, Jun. 1997.
 
@@
 L. Momtahan and A. Martin,
-e--Science Experiences: Software Engineering Practice and the EU DataGrid,
+``e--Science Experiences: Software Engineering Practice and the EU DataGrid,''
 in \emph{Ninth Asia--Pacific Software Engineering Conference},
-pp. 269-275, 2002.
+pp. 269--275, Dec. 2002.
 
 \bibitem{agile}
-Torgeir Dingsoyr, Sridhar Nerur, VenuGopal Balijepally, Nils Brede Moe,
-A decade of agile methodologies: Towards explaining agile software development,
+T. Dingsoyr, S. Nerur, V. Balijepally and N. B. Moe,
+``A decade of agile methodologies: Towards explaining agile software development,''
 \emph{J. Syst. Softw.},
-vol. 85, Issue 6, pp. 1213--1221, June 2012.
+vol. 85, no. 6, pp. 1213--1221, Jun. 2012.
 

### 2018-06-26 paper-ieee-trans.tex
fixed all refs

@@
 \bibitem{cmm}
-Paulk M., Curtis B., Chrissis M., and Weber V. C.,
-Capability maturity model for software,
-\emph{Software Engineering Institute},
-report CMU/SEI-93-TR-24, ESC--TR-93-177, 1993.
+M. Paulk, B. Curtis, M. Chrissis and V. C. Weber,
+``Capability maturity model for software,''
+Softw. Eng. Inst.,
+Technical Report CMU/SEI-93-TR-024, ESC-TR-93-177, Feb. 1993.
+[Online]. Available: \url{https://resources.sei.cmu.edu/asset_files/TechnicalReport/1993_005_001_16211.pdf}.
+[Accessed 4 Jun. 2018].
 
@@
 Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}
+[Online]. Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}.
 [Accessed 4 Jun. 2018].
@@
 Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189\_en.html}
+[Online]. Available: \url{http://cordis.europa.eu/project/rcn/99189\_en.html}.
 [Accessed 4 Jun. 2018].
@@
 Research and Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264\_en.html}
+[Online]. Available: \url{http://cordis.europa.eu/project/rcn/87264\_en.html}.
 [Accessed 4 Jun. 2018].
@@
 \bibitem{egee}
-F. Gagliardi and M. E. B\'{e}gin,
-EGEE -- providing a production quality grid for e-science,
-in \emph{2005 IEEE Inter. Symp. on Mass Storage Systems and Technology},
-pp. 88-92, 2005.
+F. Gagliardi and M. E. Begin,
+``EGEE -- providing a production quality grid for e-science,''
+in \emph{2005 IEEE Inter. Symp. Mass Storage Syst. Technol.},
+pp. 88--92, Jun. 2005.
 
@@
 Milestone MSA3.4.2, 2010, EGEE--III project.
-[Online] Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}
+[Online]. Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}.
 [Accessed 4 Jun. 2018].
@@
 \bibitem{glite}
-E. Laure et al.,
-Programming the Grid with gLite,
-\emph{Computational Methods in Science and Technology},
+E. Laure \emph{et al.},
+``Programming the Grid with gLite,''
+\emph{Computational Meth. Sci. Technol.},
 vol. 12(1), pp. 33--45, 2006.
@@
 \bibitem{condor}
-D. Thain, T. Tannenbaum, M. Livny,
-Condor and the Grid,
+D. Thain, T. Tannenbaum and M. Livny,
+``Condor and the Grid,''
 in \emph{Grid Computing: Making the Global Infrastructure a Reality},
-Chapter 11, pp. 63--70, 2002.
+ch. 11, pp. 63--70, 2003.
 
 \bibitem{etics}
-A. Di Meglio, et al.,
-ETICS: the international software engineering service for the grid,
+A D Meglio, M-E Begin, P Couvares, E Ronchieri and E Takacs,
+``ETICS: the international software engineering service for the grid,''
 in \emph{J. Phys.: Conf. Ser.},
-vol. 119, n. 4, 042010, 2008.
+vol. 119, no. 4, 042010, 2008.
 
 \bibitem{cordis:emi}
-C. Aiftimiei, et al.,
-Towards next generations of software for distributed infrastructures: The European Middleware Initiative,
+C. Aiftimiei \emph{et al.},
+``Towards next generations of software for distributed infrastructures: The European Middleware Initiative,''
 in \emph{2012 IEEE 8th Inter. Conf. on E-Science},
-pp. 1--10, 2012.
+Chicago, IL, pp. 1--10, 2012.
 
@@
 International Organization for Standardization.
-[Online] Available: \url{https://www.iso.org/standard/22749.html}
+[Online]. Available: \url{https://www.iso.org/standard/22749.html}.
 [Accessed 4 Jun. 2018].
@@
 \bibitem{emi-quality-model}
-M Alandes, et al.,
-Experiences with Software Quality Metrics in the EMI middleware,
-in \emph{J. Phys.: Conf. Ser.}
-vol. 396, n. 5, 052003, 2012.
+M. Alandes \emph{et al.},
+``Experiences with Software Quality Metrics in the EMI middleware,''
+in \emph{J. Phys.: Conf. Ser.},
+vol. 396, no. 5, 052003, 2012.
 
 \bibitem{cordis:egi-inspire}
-I. Campos Plasencia,
-EGI.eu the European grid initiative.
-in \emph{4 th Iberian Grid Infrastructure Conf. Proc.},
+I. C. Plasencia,
+``EGI.eu the European grid initiative,''
+in \emph{4th Iberian Grid Infra. Conf. Proc.},
 pp. 5--15, 2010.
@@
 \bibitem{egi-ops}
-H. Cordier, et al.,
-From EGEE Operations Portal towards EGI Operations Portal,
+H. Cordier \emph{et al.},
+``From EGEE Operations Portal towards EGI Operations Portal,''
 in \emph{Data Driven e-Science (ISGC2010)},
@@
 \bibitem{ggus}
-T. Antoni, et al.,
-Global grid user support—building a worldwide distributed user support infrastructure,
+T. Antoni \emph{et al.},
+``Global grid user support—building a worldwide distributed user support infrastructure,''
 \emph{J. Phys.: Conf. Ser.},
-vol. 119, n. 5, 052002, 2008.
+vol. 119, no. 5, 052002, 2008.
 
@@
 G. Mathieu and J. Casson,
-GOCDB4, a New Architecture for the European Grid Infrastructure,
+``GOCDB4, a New Architecture for the European Grid Infrastructure,''
 in \emph{Data Driven e-Science (ISGC2010)},
@@
 \emph{EGI Quality Criteria}.
-[Online] Available: \url{https://egi-qc.github.io/}
+[Online]. Available: \url{https://egi-qc.github.io/}.
 [Accessed 4 Jun. 2018].
@@
 \bibitem{rafi2012}
-D. M. Rafi el al.,
-Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,
-\emph{Proceedings of the 7th International Workshop on Automation of Software Test},
+D. M. Rafi \emph{el al.},
+``Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,''
+in \emph{Proc, 7th Int. Workshop Automation Softw. Test},
 36--42, 2012.
@@
 L. Chen,
-Continuous delivery: Huge benefits, but challenges too,
-\emph{IEEE Software},
+``Continuous delivery: Huge benefits, but challenges too,''
+\emph{IEEE Softw.},
 vol. 32, no 2, pp. 50--54, 2015.
@@
 \bibitem{mario}
-M. David et al.,
-Validation of Grid Middleware for the European Grid Infrastructure,
-\emph{Journal of Grid Computing},
-vol. 12, issue 3, pp. 543--558, 2014.
+M. David \emph{et al.},
+``Validation of Grid Middleware for the European Grid Infrastructure,''
+\emph{J. Grid Comp.},
+vol. 12, no. 3, pp. 543--558, 2014.
 
@@
 project, European Community Research and Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/194937\_en.html}
+[Online]. Available: \url{http://cordis.europa.eu/project/rcn/194937\_en.html}.
 [Accessed 4 Jun. 2018].
@@
 \bibitem{cordis:indigo}
-Davide Salomoni, et al.,
-INDIGO-DataCloud: {A} data and computing platform to facilitate seamless access to e-infrastructures,
+D. Salomoni \emph{et al.},
+``INDIGO-DataCloud: A data and computing platform to facilitate seamless access to e-infrastructures,''
 \emph{CoRR},
-vol. abs/1711.01981, 2018,
-[Online] Available: \url{http://arxiv.org/abs/1711.01981}.
+vol. abs/1711.01981, 2018.
+[Online]. Available: \url{http://arxiv.org/abs/1711.01981}.
+[Accessed 4 Jun. 2018].
 
 \bibitem{indigo-d31}
-J. Gomes, et al.,
-\emph{Initial Plan for WP3} INDIGO--DataCloud Deliverable 3.1.
-[Online] Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}
+J. Gomes \emph{et al.},
+``Initial Plan for WP3,''
+\emph{INDIGO--DataCloud Deliverable 3.1}.
+[Online]. Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}.
 [Accessed 4 Jun. 2018].
@@
 \bibitem{sqa-baseline}
-Pablo Orviz et al.,
-A set of common software quality assurance baseline criteria for research projects,
+Pablo Orviz \emph{et al.},
+``A set of common software quality assurance baseline criteria for research projects,''
 2017.
-[Online] Available: \url{http://digital.csic.es/bitstream/10261/160086/4/CommonSQA-v2.pdf}
+[Online]. Available: \url{http://digital.csic.es/bitstream/10261/160086/4/CommonSQA-v2.pdf}.
+[Accessed 4 Jun. 2018].
 
 \bibitem{disvis}
-G.C.P. Van Zundert, A.M.J.J. Bonvin,
-DisVis: quantifying and visualizing the accessible interaction space of distance
-restrained biomolecular complexes,
-\emph{Bioinformatics}, vol. 31, n. 19, pp. 3222--3224, 2015.
+G.C.P. Van Zundert and A.M.J.J. Bonvin,
+``DisVis: quantifying and visualizing the accessible interaction space of distance
+restrained biomolecular complexes,''
+\emph{Bioinformatics}, vol. 31, no. 19, pp. 3222--3224, 2015.
 
 \bibitem{powerfit}
-G.C.P. Van Zundert, A.M.J.J. Bonvin,
-Fast and sensitive rigid--body fitting into cryo--EM density maps with PowerFit,
-\emph{AIMS Biophys.}, vol. 2, n. 20150273, 73--87, 2015.
+G.C.P. Van Zundert and A.M.J.J. Bonvin,
+``Fast and sensitive rigid--body fitting into cryo--EM density maps with PowerFit,''
+\emph{AIMS Biophys.}, vol. 2, no. 20150273, 73--87, 2015.
 

### 2018-06-26 paper-ieee-trans.tex
fix typo in biblio

--- 2018-06-26 paper-ieee-trans.tex
@@
 \caption{Adoption of Configuration Management tools throughout the project lifetime. Also shown the trend lines
-leading to the first release (dark blue points and lines), and to the second release (light cream points and lines).}
+leading to the first release (light cream points and lines), and to the second release (dark blue points and lines).}
 \label{fig:fig_confman}
@@
 T. Antoni, et al.,
-Global grid user support—building a worldwide distributed user support infrastructure,
+Global grid user support--building a worldwide distributed user support infrastructure,
 \emph{J. Phys.: Conf. Ser.},

### 2018-06-27 paper-ieee-trans.tex
Remove backslash from urls

@@
 Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149_en.html}
 [Accessed 4 Jun. 2018].
@@
 Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189\_en.html}
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189_en.html}
 [Accessed 4 Jun. 2018].
@@
 Research and Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264\_en.html}
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264_en.html}
 [Accessed 4 Jun. 2018].
@@
 project, European Community Research and Development Information Service (CORDIS).
-[Online] Available: \url{http://cordis.europa.eu/project/rcn/194937\_en.html}
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/194937_en.html}
 [Accessed 4 Jun. 2018].

### 2018-06-27 paper-ieee-trans.tex

--- 2018-06-26 paper-ieee-trans.tex
@@
 Development Information Service (CORDIS).
-[Online]. Available: \url{http://cordis.europa.eu/project/rcn/80149\_en.html}.
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/80149_en.html}
 [Accessed 4 Jun. 2018].
@@
 Development Information Service (CORDIS).
-[Online]. Available: \url{http://cordis.europa.eu/project/rcn/99189\_en.html}.
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/99189_en.html}
 [Accessed 4 Jun. 2018].
@@
 Research and Development Information Service (CORDIS).
-[Online]. Available: \url{http://cordis.europa.eu/project/rcn/87264\_en.html}.
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/87264_en.html}
 [Accessed 4 Jun. 2018].
@@
 \bibitem{ggus}
-T. Antoni \emph{et al.},
-``Global grid user support—building a worldwide distributed user support infrastructure,''
+T. Antoni, et al.,
+Global grid user support--building a worldwide distributed user support infrastructure,
 \emph{J. Phys.: Conf. Ser.},
@@
 project, European Community Research and Development Information Service (CORDIS).
-[Online]. Available: \url{http://cordis.europa.eu/project/rcn/194937\_en.html}.
+[Online] Available: \url{http://cordis.europa.eu/project/rcn/194937_en.html}
 [Accessed 4 Jun. 2018].

### 2018-06-28 paper-ieee-trans.tex
Remove plural on fig 3 caption
Address reviewrs comments 2.4 and 2.5

@@
 Major developments carried throughout the years were focused on providing new
-functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
-The released software was less tested, resulting in a higher rate of
+functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority 
+\cite{datagrid}. The released software was less tested, resulting in a higher rate of
 rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
@@
 %P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
-reliability of the e--Infrastructures was often compromised, affecting
-negatively their exploitation by the scientific communities \cite{aiftimiei}.
+reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
+in the participant resource centers. Back in the days when Grid middleware was actively developed, researchers 
+reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, eventually 
+developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.
+
 % Comments
 %P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
-It became then apparent the need for a better balance between the addition of new features
-and the reliability of the released software. Consequently, incoming projects progressively
-devoted a substantial part of the software lifecycle to the improvement of the quality of the
+In an effort to improve the research community's experience, it became apparent the need for a better
+balance between the addition of new features and the reliability of the released software. Consequently,
+incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the
 software being delivered. At this point, evolving software engineering practices were considered
@@
 \caption{Adoption of Configuration Management tools throughout the project lifetime. Also shown the trend lines
-leading to the first release (light cream points and lines), and to the second release (dark blue points and lines).}
+leading to the first release (light cream points and line), and to the second release (dark blue points and line).}
 \label{fig:fig_confman}
@@
 
+\bibitem{lingrand-egee}
+D. Lingrand, J. Montagnat, J. Martyniak and D. Colling,
+``Analyzing the EGEE production grid workload: application to jobs submission optimization,''
+\emph{Workshop on Job Scheduling Strategies for Parallel Processing},
+pp. 37--58, May. 2009.
+
+\bibitem{campana-atlas}
+S. Campana  \emph{et al.},
+``Analysis of the ATLAS Rome production experience on the LHC computing grid,''
+\emph{IEEE First International Conference of e-Science and Grid Computing},
+pp. 8-pp, 2005.
+
+\bibitem{kindermann}
+S. Kindermann, 
+``Climate Data Analysis and Grid Infrastructures: Experiences and Perspectives,''
+\emph{Grid-Enabling Legacy Applications and Supporting End Users Workshop (GELA)},
+vol. 20. 2006.
+
+\bibitem{mendez-lorenzo}
+P. Mendez-Lorenzo, J. T. Moscicki and A. Ribon,
+``Experiences in the gridification of the Geant4 toolkit in the WLCG/EGEE environment,''
+\emph{IEEE Nuclear Science Symposium Conference Record},
+vol. 2, 2006.
+
 \bibitem{aiftimiei}

### 2018-06-28 paper-ieee-trans.tex
fix refs

@@
 Major developments carried throughout the years were focused on providing new
-functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority 
-\cite{datagrid}. The released software was less tested, resulting in a higher rate of
+functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority. 
+The released software was less tested, resulting in a higher rate of
 rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
@@
 ``Analysis of the ATLAS Rome production experience on the LHC computing grid,''
-\emph{IEEE First International Conference of e-Science and Grid Computing},
+\emph{IEEE 1st International Conference of e-Science and Grid Computing},
 pp. 8-pp, 2005.
@@
 ``Experiences in the gridification of the Geant4 toolkit in the WLCG/EGEE environment,''
-\emph{IEEE Nuclear Science Symposium Conference Record},
+\emph{IEEE Nucl. Sci. Symp. Conf. Rec.},
 vol. 2, 2006.

### 2018-06-28 paper-ieee-trans.tex

@@
 
-\subsection{DataGrid}
-
-The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
-main goal was to provide scientific communities, including physics, biology and
-earth sciences, with intensive computing and large--scale dataset analysis capabilities.
-The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
-Grid was the emerging technology to be used in order to address their requirements,
-thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
-based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
-which 5 were devoted to software development and coordinated by an
-Architecture Task Force, supervising the overall design and technical consistency
-of the developments.
-
-Without previous experience in such widely
-collaborative projects, a big effort was spent in devising solutions for several
-challenges \cite{datagrid}, namely 1) the communication overhead, as a
-result of the large geographical separation of the involved parties in the
-development tasks, 2) the fast evolution of the requirements from the user
-communities (50 use cases from the three scientific area), and 3) the lack of a body
-of knowledge for academic software engineering.
-
-The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, not yet
-considered by the project, and as such the project's
-development design suffered from the lack of the methods, tools, techniques and best
-practices emerging from this new discipline of software engineering
-\cite{agile}. Nevertheless, the project focused on the experiences and procedures
-used by diverse collaborative open source software projects, such as Linux and Apache, trying to
-achieve a higher maturity level \cite{cmm}.
-
 \begin{table*}[!h]
@@
 
+\subsection{DataGrid}
+
+The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
+main goal was to provide scientific communities, including physics, biology and
+earth sciences, with intensive computing and large--scale dataset analysis capabilities.
+The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
+Grid was the emerging technology to be used in order to address their requirements,
+thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
+based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
+which 5 were devoted to software development and coordinated by an
+Architecture Task Force, supervising the overall design and technical consistency
+of the developments.
+
+Without previous experience in such widely
+collaborative projects, a big effort was spent in devising solutions for several
+challenges \cite{datagrid}, namely 1) the communication overhead, as a
+result of the large geographical separation of the involved parties in the
+development tasks, 2) the fast evolution of the requirements from the user
+communities (50 use cases from the three scientific area), and 3) the lack of a body
+of knowledge for academic software engineering.
+
+The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, not yet
+considered by the project, and as such the project's
+development design suffered from the lack of the methods, tools, techniques and best
+practices emerging from this new discipline of software engineering
+\cite{agile}. Nevertheless, the project focused on the experiences and procedures
+used by diverse collaborative open source software projects, such as Linux and Apache, trying to
+achieve a higher maturity level \cite{cmm}.
+
 \subsection{Enabling Grids for E--sciencE}
@@
 ``Analysis of the ATLAS Rome production experience on the LHC computing grid,''
-\emph{IEEE 1st International Conference of e-Science and Grid Computing},
+\emph{IEEE 1st Int. Conf. of e-Science and Grid Computing},
 pp. 8-pp, 2005.
@@
 ``e--Science Experiences: Software Engineering Practice and the EU DataGrid,''
-in \emph{Ninth Asia--Pacific Software Engineering Conference},
+in \emph{9th Asia--Pacific Softw. Eng. Conf.},
 pp. 269--275, Dec. 2002.

### 2018-06-29 paper-ieee-trans.tex
Minor fix in caption text for Fig3
Minor addition

@@
 \includegraphics[width=0.5\textwidth]{images/fig3.png}
-\caption{Adoption of Configuration Management tools throughout the project lifetime. Also shown the trend lines
-leading to the first release (light cream points and line), and to the second release (dark blue points and line).}
+\caption{Adoption of Configuration Management tools throughout the project lifetime. The figure shows the trend lines
+leading to the first (light cream points and line) and second release (dark blue points and line).}
 \label{fig:fig_confman}
@@
 end of the project, the software development activity slowed the pace but not completely ceased
-as the CI infrastructure is still maintained after the project's end of life.
+as the CI infrastructure is still maintained and used after the project's end of life.
 

### 2018-06-29 paper-ieee-trans.tex
Add Cris comment

@@
 and e--Infrastructure creation, operation and management.
-This work examines the evolution of the software engineering methodologies used in past and
+This work examines the evolution of the software engineering processes used in past and
 present European Commission--funded projects to illustrate how the research on software
@@
 A significant number of innovation projects and technology transfer
-activities -- later discussed in Section~\ref{sec:ev} -- from the European Commission
-(EC)-funded 5th, 6th and 7th Framework Programmes, and recently from the Horizon 2020 programme
-\cite{h2020} had as their main goal to develop software solutions that provide a seamless and
-transparent access by researchers to distributed computing and data resources.
+activities -- later discussed in Section~\ref{sec:ev} -- 
+were funded by the 5th, 6th and 7th Framework Programmes of the European Commission (EC) 
+and recently by the Horizon 2020 programme \cite{h2020}. Their main goal is to develop 
+software solutions that provide a seamless and transparent access by researchers to 
+distributed computing and data resources.
 
@@
 to understand the effort employed in the EC-funded software projects over time. We divide literature 
-into four different domains that intersect this work: 1) software requirements, 2) software coding,
+into four different domains that intersect this work: 1) software requirements analysis, 2) software
+coding,
 3) metrics and measurements, 4) testing.
 
-\emph{Software requirements.} They are defined at various stages of a software development process
+\emph{Software requirements analysis.} They are defined at various stages of a software development process
 as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
@@
 understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
-development process to easily adapt software \cite{buse, baggen}: programming features that influence the
-code readability and maintainability and standardized code that promotes the comparability of results.
+development process to easily adapt software: programming features that influence the
+code readability and maintainability \cite{buse} and standardized code that promotes the comparability of results \cite{baggen}.
 
@@
 under specified conditions, observing or recording the results, and making an evaluation of some aspect of
-system or component. Testing is an important factor for software quality and reliability improvement. This
+the system or component. Testing is an important factor for software quality and reliability improvement. This
 activity is quite complicated due to the variety of programming languages, operating systems and hardware
@@
 process, possibly in the early stages. Over time, testing metrics have been defined to assess software
-quality \cite{athanasiou}: code coverage is the most frequently used metric for test software reliability
+quality \cite{athanasiou}: code coverage is the most frequently used metric for testing software reliability
 and there exist many tools for dynamic code coverage estimation according to the programming language
-\cite{horgan}. A dedicated effort have to be taken into account in order to integrate testing solutions
+\cite{horgan}. A dedicated effort has to be taken into account in order to integrate testing solutions
 during the software development process.
@@
 The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, not yet
-considered by the project, and as such the project's
+considered by the project. As such, the project's
 development design suffered from the lack of the methods, tools, techniques and best
@@
 requirements coming from the diverse scientific communities needed developing a
-rich set of new services while maintaining a sustainable infrastructure for
-Grid computing, eventually used by more than 15 thousand researchers and deployed in
-over 250 institutions.
+rich set of new services, while maintaining a sustainable infrastructure for
+Grid computing. This e-Infrastructure was eventually used by more than 15 thousand researchers
+and deployed in over 250 institutions.
 
@@
 build and testing capabilities, as well as auto--generated documentation and
-software metrics gathering such as Source Lines Of Code (SLOC), complexity and
-number of defects/bugs \cite{etics}. The {\sl ETICS} framework was the first automated
+software metrics gathering -- such as Source Lines Of Code (SLOC), complexity and
+number of defects/bugs \cite{etics} --. The {\sl ETICS} framework was the first automated
 service for delivering quality software products in distributed environments like
@@
 as well as for metric tracking, making queries on the collected data to display
-them through the chart generation framework.
+them through a chart generation framework.
 
@@
 This goal would be accomplished through the
-development and maintenance of various operational tools, such as the Operations Portal \cite{egi-ops},
-the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the
+development and maintenance of various operational tools -- such as the Operations Portal \cite{egi-ops},
+the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb} -- and the
 management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
@@
 the adequate development, timely delivery and reliable operation of the produced software components.
-It emphasizes the quality requirements and best practices to be applied at the
+They emphasize the quality requirements and best practices to be applied at the
 development phase, such as style and testing compliance or human code reviews, to protect the
@@
 \item \textit{Unit and Functional testing}.
-Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
+Changes involving the addition of new functionalities are required to be tested. \textit{Regression
 testing} in this context is accomplished by enforcing the definition and periodic execution
@@
 (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage. 
-The indiviual records followed also a growing trend (only 6 out of 32 components had lower values in 
+The individual records followed also a growing trend (only 6 out of 32 components had lower values in 
 INDIGO--2). It is worth noting that Fig.~\ref{fig:fig_unittest} not only includes new software developments 
@@
 deployment of the products delivered as part of the catalogue. Automation in this context is
-tackled using a configuration management tool. These tools allow the adoption of
-\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
+tackled using configuration management tools. These tools allow the adoption of
+\textit{Infrastructure as Code} (IaC) practice, managing the component's deployment through declarative
 definitions. The definition files appear as an additional source of documentation
@@
 It shows an increase in the adoption of such tools leading to the {\sl INDIGO--1} release
-(dark blue points and trend line), as well as to the {\sl INDIGO--2} release
-(light cream points and trend line). The rate of adoption is lower during the weeks
+(light cream points and trend line), as well as to the {\sl INDIGO--2} release
+(dark blue points and trend line). The rate of adoption is lower during the weeks
 before the second release because a significant fraction of the products had already adopted it
@@
 API\footnote{\url{https://developer.github.com/v3/}.}
-and Jenkins\footnote{\url{https://jenkins.io/}} service, and graphically displayed as GitHub
+and the Jenkins\footnote{\url{https://jenkins.io/}} service, and graphically displayed as GitHub
 pages using the GrimoireLab framework\footnote{\url{http://grimoirelab.github.io/}}.
@@
 SQA machinery, the software is always ready to be released. As a result, the often compromised harmonization
-between the continouous release of new features -- pushed by the development teams -- and the maintenance of
+between the continuous release of new features -- pushed by the development teams -- and the maintenance of
 stable production systems -- demanded by users and resource providers -- can be guaranteed. The DevOps culture
@@
 The incremental adoption of DevOps approaches throughout the project's lifetime was a key player towards
-software reliability. Starting with a CI scenario at the first stages of the project, the software delivered
-was eventually validated and distributed using CD pipelines.
+software reliability. Starting with a Continuous Integration (CI) scenario at the first stages of the project, the software delivered
+was eventually validated and distributed using Continuous Delivery (CD) pipelines.
 
@@
 implementation of the CI infrastructure. Automated functional testing coverage were not
-available for all the software stack, thus the associated number of builds are fewer. Towards the
-end of the project, the software development activity slowed the pace but not completely ceased
-as the CI infrastructure is still maintained and used after the project's end of life.
+available for all the software stack, thus the associated number of builds are fewer. 
+Towards the end of the project, the software development activity slowed the pace but not completely 
+ceased. The support of the CI infrastructure, once reached the project's end of life, allowed the 
+continued maintenance and development of the prevailing software. Some development teams deployed 
+parallel CI systems, taking advantage of the experience gained during the project.
 
@@
 development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
-{\sl INDIGO--DataCloud} products, that have been taking place since the second major release, are passing through
+{\sl INDIGO--DataCloud} products, which have been taking place since the second major release, are passing through
 a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
@@
 
-Two applications from "Definition of Support to Research Communities" work package,
+Two applications coming from the supported research communities,
 DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
 integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
-Fig.~\ref{fig:fig_disvis} shows the pipeline for the Disvis application.
+Fig.~\ref{fig:fig_disvis} shows the pipeline for the DisVis application.
 
@@
 Once the application is packaged as a Docker image, and subsequently uploaded
-to the Dockerhub repository, it is instantiated in a new container to be validated.
+to the DockerHub repository, it is instantiated in a new container to be validated.
 The application is then executed and the results compared with a set of reference outputs.
@@
 level of integration between the components involved in the {\sl INDIGO--DataCloud}
-solutions and use cases validation, by deploying and executing the applications
+solutions and use cases validation. Within these testbeds, the applications are deployed
+and executed
 with the last stable version of the software in environments as similar as
@@
 the impact on production systems. In this regard, a pre--production validation,
-through the preview testbeds or resource center’s early adoption -- handled by the
+through the preview testbeds or resource center's early adoption -- handled by the
 Staged--rollout process -- alleviates the likelihood of discovering uncovered
@@
 The software lifecycle demands as well a continuous measurement and analysis
-process, that allows timely reactions to discover and fix defects throughout the
+process, which allows timely reactions to discover and fix defects throughout the
 development, release and maintenance stages. Conscientious metric
@@
 ``e--Science Experiences: Software Engineering Practice and the EU DataGrid,''
-in \emph{9th Asia--Pacific Softw. Eng. Conf.},
+in \emph{Proc. 9th Asia--Pacific Softw. Eng. Conf.},
 pp. 269--275, Dec. 2002.
@@
 ``EGI.eu the European grid initiative,''
-in \emph{4th Iberian Grid Infra. Conf. Proc.},
+in \emph{Proc. 4th Iberian Grid Infra. Conf.},
 pp. 5--15, 2010.
@@
 ``Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,''
-in \emph{Proc, 7th Int. Workshop Automation Softw. Test},
+in \emph{Proc. 7th Int. Workshop Automation Softw. Test},
 36--42, 2012.

### 2018-06-29 paper-ieee-trans.tex
minor fix

@@
 In this section, we detail the level of knowledge requested in delivering reliable software in order 
-to understand the effort employed in the EC-funded software projects over time. We divide literature 
+to understand the effort employed in the EC-funded software projects over time. We divide the literature 
 into four different domains that intersect this work: 1) software requirements analysis, 2) software

### 2018-06-29 paper-ieee-trans.tex

@@
 were funded by the 5th, 6th and 7th Framework Programmes of the European Commission (EC) 
-and recently by the Horizon 2020 programme \cite{h2020}. Their main goal is to develop 
+and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop 
 software solutions that provide a seamless and transparent access by researchers to 

### 2018-08-23 paper-ieee-trans.tex
Caption text elaborated with an explanation of zero-value coverage for products in INDIGO-1

@@
 \caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud}
-software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
+software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases. Products that
+were exclusively released during {\sl INDIGO--2} do not show data for the first period of
+{\sl INDIGO--1}.}
 \label{fig:fig_unittest}
@@
 \textit{documentation is treated as code}, using a markup language, automatically rendered and
-uploaded to online repositories\footnote{\url{https://www.gitbook.com/@indigo-dc}}.
+uploaded to online repositories \cite{foot1}.
 Thus documentation is portable and
@@
 as Ansible
-and Puppet\footnote{\url{https://forge.puppet.com/indigodc}}. A representative example of
+and Puppet \cite{foot2}. A representative example of
 such contributions are the \textit{50 roles} developed from scratch and currently hosted in
-the Ansible Galaxy portal\footnote{\url{https://galaxy.ansible.com/indigo-dc/}}. Fig.~\ref{fig:fig_confman}
+the Ansible Galaxy portal \cite{foot3}. Fig.~\ref{fig:fig_confman}
 shows the number of products that offer an automated means for deployment.
@@
 \textit{Development metrics} are obtained programmatically from several sources, namely GitHub
-API\footnote{\url{https://developer.github.com/v3/}.}
-and the Jenkins\footnote{\url{https://jenkins.io/}} service, and graphically displayed as GitHub
-pages using the GrimoireLab framework\footnote{\url{http://grimoirelab.github.io/}}.
+API \cite{foot4} and the Jenkins \cite{foot5} service, and graphically displayed as GitHub
+pages using the GrimoireLab framework \cite{foot6}.
 Per--component weekly reports, including the SQA requirement
@@
 \textit{Release metric} sources are the online repository servers, namely the
-Linux package\footnote{\url{https://jenkins.indigo-datacloud.eu/}}
-and DockerHub\footnote{\url{https://hub.docker.com/u/indigodatacloud}} repositories. Jenkins
+Linux package \cite{foot7} and DockerHub \cite{foot8} repositories. Jenkins
 server also contains valuable release data as it is the service where the packages
@@
 The practical implementation of the CI scenario leverage from a tightly integrated open
-source tools such as GitHub\footnote{\url{https://github.com/}}, as the online source code repository,
+source tools such as GitHub \cite{foot9}, as the online source code repository,
 
@@
 
+\bibitem{foot1}
+indigo-dc, ``Sign in to GitBook - GitBook,''
+2018.
+[Online]. Available: \url{https://www.gitbook.com/@indigo-dc}.
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot2}
+puppetforge, ``Modules by INDIGO Datacloud - Puppet Forge,''
+2018.
+[Online]. Available: \url{https://forge.puppet.com/indigodc}.
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot3}
+GALAXY, ``Ansible Galaxy,''
+2018.
+[Online]. Available: \url{https://galaxy.ansible.com/indigo-dc/}
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot4}
+GitHub Developer, ``GitHub API v3 | GitHub Developer Guide,''
+2018.
+[Online]. Available: \url{https://developer.github.com/v3/}
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot5}
+Jenkins, ``Jenkins,''
+2018.
+[Online]. Available: \url{https://jenkins.io/}
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot6}
+GRIMOIRELAB, ``GrimoireLab - Software Development and Community Analytics Platform,''
+2017.
+[Online]. Available: \url{http://grimoirelab.github.io/}
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot7}
+Jenkins Indigo-dc, ``Jenkins - Indigo-DataCloud,''
+2018.
+[Online]. Available: \url{https://jenkins.indigo-datacloud.eu/}
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot8}
+Indigo-dc, ``indigodatacloud - Docker Hub,''
+2018.
+[Online]. Available: \url{https://hub.docker.com/u/indigodatacloud}
+[Accessed 30 Aug. 2018].
+
+\bibitem{foot9}
+GitHub, ``Learn Git and GitHub without any code!,''
+2018.
+[Online]. Available: \url{https://github.com/}
+[Accessed 30 Aug. 2018].
+
 \bibitem{disvis}

### 2018-08-23 paper-ieee-trans.tex
Added paragraph to explain decreases in coverage values for some products in INDIGO-2

@@
 code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
-the project's first (INDIGO--1) and second (INDIGO--2) releases. By INDIGO--2, 53\% of the software components
-(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
-(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
+the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
+shows a decrease in the unit testing coverage values for several products. As unit tests cover low-level
+code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
+We have observed that this decrease was aligned with high-demanding periods of software development, as
+seen in the previous months of the INDIGO-2 major release. Regardless of the potentially higher defect
+density that this fact brought along, the overall performance kept progressing as the rest of the software
+stack substantially improved their own unit testing coverage statistics. By INDIGO--2, 53\% of the software
+components (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage
+recommendation (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage. 
 The individual records followed also a growing trend (only 6 out of 32 components had lower values in 
@@
 
+\bibitem{unit-test-frameworks}
+Paul Hamill,
+``Unit test frameworks: tools for high-quality software development,''
+\emph{O'Reilly Media, Inc.},
+2004.
+
 \bibitem{disvis}

### 2018-08-31 paper-ieee-trans.tex
added a possible answer to editor s comments for the lack of quantification

@@
 \begin{document}
-\title{Evolution of Processes in Delivering Reliable Software in the European Scientific Arena}
+\title{Evolution of Processes in Delivering Reliable Software within the European Scientific Arena}
 
@@
 The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
-main goal was to provide scientific communities, including physics, biology and
-earth sciences, with intensive computing and large--scale dataset analysis capabilities.
+main goal was to provide scientific communities (including physics, biology and
+earth sciences) with intensive computing and large--scale dataset analysis capabilities.
 The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
@@
 collaborative projects, a big effort was spent in devising solutions for several
-challenges \cite{datagrid}, namely 1) the communication overhead, as a
+challenges \cite{datagrid}, namely 1) the communication overhead (as a
 result of the large geographical separation of the involved parties in the
-development tasks, 2) the fast evolution of the requirements from the user
+development tasks), 2) the fast evolution of the requirements from the user
 communities (50 use cases from the three scientific area), and 3) the lack of a body
@@
 achieve a higher maturity level \cite{cmm}.
+
+On behalf of the quality assurance group, specific measures for the software development process and quality assurance were defined and introduced in the project.
+Through project monitoring and reporting mechanisms there was a successful control of the risks, efforts and document produced.
+The definition of a project developer's guide \cite{datagridguide} was able to provide developers with guidelines for the various software development phases (such as packaging, test and validation process, and style and naming convention) and to allow quality group to control software release via a procedure checklist.
+The specification of quality indicators \cite{datagridindicators} allowed project managers to have an update view of the quality of the services over months.
+Figure \ref{fig:efficiency} shows the efficiency performance indicator measured as the number of jobs successfully completed divided the total number of jobs submitted: generally the project managers observed a positive trend of efficiency over months with only one exception for the LCG1 experiment due to some errors on the dedicated resources.
+Figure \ref{fig:bugzilla} shows the Bugzilla anomalies follow up in the second release.
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.5\textwidth, height=50mm]{images/dgeff.png}
+\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted): results are provided for five experiments involved in the projects.}
+\label{fig:efficiency}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.5\textwidth, height=50mm]{images/dgbug.png}
+\caption{Bugzilla anomalies follow up: new and pending anomalies are reported.}
+\label{fig:bugzilla}
+\end{figure}
 
@@
 the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
-shows a decrease in the unit testing coverage values for several products. As unit tests cover low-level
+shows a decrease in the unit testing coverage values for 18\% of products. As unit tests cover low-level
 code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
@@
 
-The practical implementation of the CI scenario leverage from a tightly integrated open
-source tools such as GitHub \cite{foot9}, as the online source code repository,
-
-Jenkins is an event--response CI and Docker container
-provisioning system that provides instant computing power needed for executing the pipeline jobs.
+The practical implementation of the CI scenario leverage from tightly integrated open
+source tools such as:
+\begin{itemize}
+    \item GitHub \cite{foot9} as the online source code repository hosting service,
+    \item Jenkins as the event--response CI,
+    \item Docker container provisioning system, to provide instant computing power needed
+          for executing the pipeline jobs.
+\end{itemize}
 The project defined a source code contribution workflow, based on GitHub Pull Requests
@@
 
+\bibitem{datagridguide}
+Quality Assurance Group, ``DataGrid - European DataGrid Developers' Guide,''
+2003.
+[Online] Available: \url{https://edms.cern.ch/ui/file/358824/1.1/EDG-DevGuide-v1-2.pdf}
+[Accessed 30 Aug. 2018].
+
+\bibitem{datagridindicators}
+DataGrid, ``DataGrid Internal Document - Quality and Performance Indicators for DataGrid,''
+2003.
+[Online] Available: \url{https://edms.cern.ch/ui/file/386039/2/QIv0-3.pdf}
+[Accessed 30 Aug. 2018].
+
 \bibitem{cmm}
@@
 
+\bibitem{glite}
+E. Laure \emph{et al.},
+``Programming the Grid with gLite,''
+\emph{Computational Meth. Sci. Technol.},
+vol. 12(1), pp. 33--45, 2006.
+
+\bibitem{condor}
+D. Thain, T. Tannenbaum and M. Livny,
+``Condor and the Grid,''
+in \emph{Grid Computing: Making the Global Infrastructure a Reality},
+ch. 11, pp. 63--70, 2003.
+
 \bibitem{egee:acceptance-criteria}
@@
 
-\bibitem{glite}
-E. Laure \emph{et al.},
-``Programming the Grid with gLite,''
-\emph{Computational Meth. Sci. Technol.},
-vol. 12(1), pp. 33--45, 2006.
-
-\bibitem{condor}
-D. Thain, T. Tannenbaum and M. Livny,
-``Condor and the Grid,''
-in \emph{Grid Computing: Making the Global Infrastructure a Reality},
-ch. 11, pp. 63--70, 2003.
-
 \bibitem{etics}
@@
 
-\bibitem{egi-qc}
-\emph{EGI Quality Criteria}.
-[Online]. Available: \url{https://egi-qc.github.io/}.
-[Accessed 4 Jun. 2018].
-
-\bibitem{rafi2012}
-D. M. Rafi \emph{el al.},
-``Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,''
-in \emph{Proc. 7th Int. Workshop Automation Softw. Test},
-36--42, 2012.
-
-\bibitem{chen2015}
-L. Chen,
-``Continuous delivery: Huge benefits, but challenges too,''
-\emph{IEEE Softw.},
-vol. 32, no 2, pp. 50--54, 2015.
-
 \bibitem{mario}
@@
 vol. 12, no. 3, pp. 543--558, 2014.
+
+\bibitem{egi-qc}
+\emph{EGI Quality Criteria}.
+[Online]. Available: \url{https://egi-qc.github.io/}.
+[Accessed 4 Jun. 2018].
 
@@
 
+\bibitem{unit-test-frameworks}
+Paul Hamill,
+``Unit test frameworks: tools for high-quality software development,''
+\emph{O'Reilly Media, Inc.},
+2004.
+
 \bibitem{foot1}
@@
 
+\bibitem{rafi2012}
+D. M. Rafi \emph{el al.},
+``Benefits and limitations of automated software testing: Systematic literature review and practitioner survey,''
+in \emph{Proc. 7th Int. Workshop Automation Softw. Test},
+36--42, 2012.
+
 \bibitem{foot9}
@@
 
-\bibitem{unit-test-frameworks}
-Paul Hamill,
-``Unit test frameworks: tools for high-quality software development,''
-\emph{O'Reilly Media, Inc.},
-2004.
+\bibitem{chen2015}
+L. Chen,
+``Continuous delivery: Huge benefits, but challenges too,''
+\emph{IEEE Softw.},
+vol. 32, no 2, pp. 50--54, 2015.
 

### 2018-08-31 datagridbugzilla.R

@@ -1,0 +1,20 @@
+datafn<-read.csv('data/datagridbugzilla.csv', header=TRUE)
+names(datafn)
+str(datafn)
+
+datafn$Date<-as.Date(datafn$Date)
+
+library(ggplot2)
+library(scales)
+p<-ggplot(data=datafn, aes(x=Date,y=Anomalies, color=factor(Category))) +
+  geom_point(size=4) +
+  geom_line(size=2) +
+  labs(y='N. Anomalies',color='Category', x='Month') +
+  scale_x_date(labels=date_format('%m-%Y'), date_breaks='1 month') +
+  theme(legend.key = element_rect(fill = "white", colour = "white"),
+        axis.text=element_text(size=12),
+        axis.line = element_line(colour = "black"),
+        legend.position="top",
+        legend.text=element_text(size=12))
+  
+p

### 2018-08-31 datagridefficiency.R

@@ -1,0 +1,18 @@
+datafn<-read.csv('data/datagridefficiency.csv', header=TRUE)
+names(datafn)
+str(datafn)
+
+datafn$Date<-as.Date(datafn$Date)
+
+library(ggplot2)
+p<-ggplot(data=datafn, aes(x=Date,y=Efficiency/100, color=factor(Experiment))) +
+  geom_point(size=4) +
+  scale_y_continuous(limits=c(0,1),labels=scales::percent) +
+  labs(y='Efficiency %',color='Experiment') +
+  theme(legend.key = element_rect(fill = "white", colour = "white"),
+        axis.text=element_text(size=12),
+        axis.line = element_line(colour = "black"),
+        legend.position="top",
+        legend.text=element_text(size=12))
+  
+p

### 2018-08-31 paper-ieee-trans.tex
added a possible answer to editor s comments for the lack of quantification

@@
 Software Product Metrics           &$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
-Quality Criteria Definition        &       &       &$\surd$&$\surd$&$\surd$&$\surd$\\
+Quality Criteria Definition        &$\surd$&       &$\surd$&$\surd$&$\surd$&$\surd$\\
 Automatic Certification            &       &       &$\surd$&       &$\surd$&$\surd$\\
@@
 at addressing the challenges in producing quality software in distributed,
-collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
+collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The ETICS framework
 integrated different technologies and tools in order to provide automated configuration,
@@
 Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
-with the goal to maintain and evolve the middleware focusing on extending their
-interoperability and improving the reliability of the services. The
+with the goal of maintaining and evolving the different Grid middleware solutions, 
+focusing on extending their interoperability and improving the reliability of the services. The
 ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
@@
 leveraged the {\sl ETICS} service for the development, continuous integration and release management,
-as well as for metric tracking, making queries on the collected data to display
+as well as for metric tracking, making queries on the collected data in order to display
 them through a chart generation framework.
@@
 model and operations.
+
+The impact of the quality assurance activities was observed in the software provided by EMI to the EGI--InSPIRE project.
+Data reported in Table \ref{tab:emi} (see page 13 in \cite{emi-quality-model}) summarizes the EMI software quality evolution per project quarters (PQ) as evaluated by the EGI project with the help of the UMD QC.
+It is possible to observe the improvement of EMI releases over time that met the required quality criteria by considering the products that have passed the UMD QC, the products that have passed the EGI Stage Rollout (SR) phase (i.e.
+it consisted of testing the certified updates of the middleware by a subset of sites before made them available to all sites), and the products rejected because they did not meet the UMD QC.
+\begin{table}[!h]
+\renewcommand{\arraystretch}{1.3}
+\caption{EMI software quality evolution}
+\label{tab:emi}
+\centering
+\begin{tabular}{lp{1cm}p{1cm}p{1cm}p{1cm}}
+\hline
+\hline
+PQ & Release Products (RP) & N. RP Passed UMD QC & N. RP Passed SR & N. RP Failed UMD QC\\
+\hline
+\hline
+5 & 30 & 27 & 27 & 0\\
+6 & 30 & 28 & 26 & 2 \\
+7 & 27 & 26& 24& 2\\
+8 & 18 & 18 & 18 & 0\\
+\hline
+\hline
+\end{tabular}
+\end{table}
+
 {\sl UMD} is still being used and
@@
 The released software is tested in production environments through the
-Staged--Rollout process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
+SR process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
 Selected resource providers are requested to install

### 2018-09-14 paper-ieee-trans.tex

--- 2018-08-23 paper-ieee-trans.tex
@@
 \caption{Comparison of the unit testing coverage values for the {\sl INDIGO--DataCloud}
-software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases. Products that
-were exclusively released during {\sl INDIGO--2} do not show data for the first period of
-{\sl INDIGO--1}.}
+software stack over the {\sl INDIGO--1} and {\sl INDIGO--2} major releases.}
 \label{fig:fig_unittest}
@@
 code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
-the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
-shows a decrease in the unit testing coverage values for several products. As unit tests cover low-level
-code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
-We have observed that this decrease was aligned with high-demanding periods of software development, as
-seen in the previous months of the INDIGO-2 major release. Regardless of the potentially higher defect
-density that this fact brought along, the overall performance kept progressing as the rest of the software
-stack substantially improved their own unit testing coverage statistics. By INDIGO--2, 53\% of the software
-components (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage
-recommendation (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage. 
+the project's first (INDIGO--1) and second (INDIGO--2) releases. By INDIGO--2, 53\% of the software components
+(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
+(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
 The individual records followed also a growing trend (only 6 out of 32 components had lower values in
@@
 
-\bibitem{unit-test-frameworks}
-Paul Hamill,
-``Unit test frameworks: tools for high-quality software development,''
-\emph{O'Reilly Media, Inc.},
-2004.
-
 \bibitem{disvis}

### 2018-09-14 paper-ieee-trans.tex
remove end parenthesis

--- 2018-08-31 paper-ieee-trans.tex
@@
 \includegraphics[width=0.5\textwidth, height=50mm]{images/dgeff.png}
-\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted): results are provided for five experiments involved in the projects.}
+\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the projects.}
 \label{fig:efficiency}

### 2018-09-19 paper-ieee-trans.tex

--- 2018-09-14 paper-ieee-trans.tex
@@
 code coverage increased from an average value of 50.32\% to 63.34\%, outlined by the respective dates of
-the project's first (INDIGO--1) and second (INDIGO--2) releases. By INDIGO--2, 53\% of the software components
-(17 out of 32) were over the threshold of the aforementioned project's SQA code coverage recommendation
-(70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
+the project's first (INDIGO--1) and second (INDIGO--2) releases. Nevertheless, Fig.~\ref{fig:fig_unittest}
+shows a decrease in the unit testing coverage values for several products. As unit tests cover low-level
+code elements, they are best considered while the new code is being written \cite{unit-test-frameworks}.
+We have observed that this decrease was aligned with high-demanding periods of software development, as
+seen in the previous months of the INDIGO-2 major release. Regardless of the potentially higher defect
+density that this fact brought along, the overall performance kept progressing as the rest of the software
+stack substantially improved their own unit testing coverage statistics. By INDIGO--2, 53\% of the software
+components (17 out of 32) were over the threshold of the aforementioned project's SQA code coverage
+recommendation (70\%), while 75\% (24 out of 32) of the product stack exceeded 50\% coverage.
 The individual records followed also a growing trend (only 6 out of 32 components had lower values in
@@
 
-\bibitem{egee:acceptance-criteria}
-\emph{Definition and Documentation of the Revised Software Life--Cycle Process},
-Milestone MSA3.4.2, 2010, EGEE--III project.
-[Online]. Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}.
-[Accessed 4 Jun. 2018].
-
 \bibitem{glite}
@@
 
+\bibitem{egee:acceptance-criteria}
+\emph{Definition and Documentation of the Revised Software Life--Cycle Process},
+Milestone MSA3.4.2, 2010, EGEE--III project.
+[Online]. Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}.
+[Accessed 4 Jun. 2018].
+
 \bibitem{etics}
@@
 
+\bibitem{chen2015}
+L. Chen,
+``Continuous delivery: Huge benefits, but challenges too,''
+\emph{IEEE Softw.},
+vol. 32, no 2, pp. 50--54, 2015.
+
 \bibitem{disvis}

### 2018-09-25 paper-ieee-trans.tex
Remove bugzilla figure and rephrasing

@@
 achieve a higher maturity level \cite{cmm}.
+
+Project monitoring and reporting mechanisms allowed to assess the risks, efforts and documentation
+produced. Development guidelines \cite{datagridguide} were produced, covering the various phases, including
+1) packaging, 2) test and validation, and 3) style and naming convention. The specification of quality indicators
+\cite{datagridindicators} facilitated the detection of deviations in the collected metric data. Figure 
+\ref{fig:efficiency} shows the efficiency performance indicator, measured in production environments as the
+number of jobs successfully completed divided by the total number of jobs submitted. As it can be observed,
+there is a positive trend of efficiency over the months, with the unique exception of the LCG1 experiment.
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.5\textwidth, height=50mm]{images/dgeff.png}
+\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the project.}
+\label{fig:efficiency}
+\end{figure}
 
@@
 Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
-with the goal to maintain and evolve the middleware focusing on extending their
-interoperability and improving the reliability of the services. The
+with the goal to maintain and evolve the middleware focusing on extending their interoperability and improving the reliability of the services. The
 ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
@@
 model and operations.
+
+The impact of the quality assurance activities was observed once the EMI software contributed to the 
+EGI--InSPIRE project. The data reported in Table \ref{tab:emi} (see page 13 in \cite{emi-quality-model})
+summarizes the EMI software quality evolution per project quarters (PQ) as evaluated by the EGI project
+with the help of the UMD QC. We can observe the improvement in quality of the EMI releases over time, 
+measured by the number of products that both met the software distribution criteria, by means of the UMD QC
+definition, and the EGI Stage Rollout (SR) phase. This latter phase completed the criteria validation by 
+deploying the UMD QC--certified software in a set of candidate production sites.
+
+\begin{table}[!h]
+\renewcommand{\arraystretch}{1.3}
+\caption{EMI software quality evolution}
+\label{tab:emi}
+\centering
+\begin{tabular}{lp{1cm}p{1cm}p{1cm}p{1cm}}
+\hline
+\hline
+PQ & Release Products (RP) & N. RP Passed UMD QC & N. RP Passed SR & N. RP Failed UMD QC\\
+\hline
+\hline
+5 & 30 & 27 & 27 & 0\\
+6 & 30 & 28 & 26 & 2 \\
+7 & 27 & 26& 24& 2\\
+8 & 18 & 18 & 18 & 0\\
+\hline
+\hline
+\end{tabular}
+\end{table}
+
 {\sl UMD} is still being used and

### 2018-09-26 paper-ieee-trans.tex
improved table III, improved reference order, changed fig1 filename
added figure

@@
 there is a positive trend of efficiency over the months, with the unique exception of the LCG1 experiment.
-\begin{figure}[ht]
+\begin{figure}[h]
 \centering
-\includegraphics[width=0.5\textwidth, height=50mm]{images/dgeff.png}
+\includegraphics[width=0.5\textwidth, height=50mm]{images/datagridEff.png}
 \caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the project.}
@@
 \centering
-\begin{tabular}{lp{1cm}p{1cm}p{1cm}p{1cm}}
+\begin{tabular}{lp{1.6cm}p{1.6cm}p{1.3cm}p{1.5cm}}
 \hline
@@
 
+Figure \ref{fig:ingigobugs}
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.5\textwidth, height=50mm]{images/indigoBugs.png}
+\caption{Number of INDIGO bugs detected during the development and production phases.}
+\label{fig:indigobugs}
+\end{figure}
+
 \section*{Acknowledgment}
@@
 vol. 85, no. 6, pp. 1213--1221, Jun. 2012.
-
-\bibitem{datagridguide}
-Quality Assurance Group, ``DataGrid - European DataGrid Developers' Guide,''
-2003.
-[Online] Available: \url{https://edms.cern.ch/ui/file/358824/1.1/EDG-DevGuide-v1-2.pdf}
-[Accessed 30 Aug. 2018].
-
-\bibitem{datagridindicators}
-DataGrid, ``DataGrid Internal Document - Quality and Performance Indicators for DataGrid,''
-2003.
-[Online] Available: \url{https://edms.cern.ch/ui/file/386039/2/QIv0-3.pdf}
-[Accessed 30 Aug. 2018].
 
@@
 
+\bibitem{datagridguide}
+Quality Assurance Group, ``DataGrid - European DataGrid Developers' Guide,''
+2003.
+[Online] Available: \url{https://edms.cern.ch/ui/file/358824/1.1/EDG-DevGuide-v1-2.pdf}
+[Accessed 30 Aug. 2018].
+
+\bibitem{datagridindicators}
+DataGrid, ``DataGrid Internal Document - Quality and Performance Indicators for DataGrid,''
+2003.
+[Online] Available: \url{https://edms.cern.ch/ui/file/386039/2/QIv0-3.pdf}
+[Accessed 30 Aug. 2018].
+
 \bibitem{cordis:egee}

### 2018-09-26 paper-ieee-trans.tex
Added some text for figure 9

@@
 Staged--rollout process -- alleviates the likelihood of discovering uncovered
-issues at production.
+issues at production. To put it in numbers, as Fig.~\ref{fig:ingigobugs} showcases, the 
+total number of defects detected in the pre--production stages surpassed by a factor of 30
+the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud} 
+project.
 
@@
 
-Figure \ref{fig:ingigobugs}
 \begin{figure}[ht]
@@
 \includegraphics[width=0.5\textwidth, height=50mm]{images/indigoBugs.png}
-\caption{Number of INDIGO bugs detected during the development and production phases.}
+\caption{Number of defects detected in development and production stages over the lifetime of the
+{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project, 
+in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
+months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
+of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
+stages, thus not affecting the software released.}
 \label{fig:indigobugs}

### 2018-09-27 paper-ieee-trans.tex
Review #2

@@
 
-\subsubsection{Continuous Delivery}
-As DevOps suggests, frequent releases positively affect the reliability of the software as
-they reduce the time in which a given defect is exposed or has an impact, allowing the
-development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
-{\sl INDIGO--DataCloud} products, which have been taking place since the second major release, are passing through
-a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
-successful execution of the CI pipeline previously described. Consequently, CI--validated
-software components are automatically delivered in online repositories. Nonetheless,
-expert supervision to validate the results is required.
-
-{\sl INDIGO--DataCloud} software stack is delivered in the form of Linux software packages (rpms, debs)
-and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
-package pipeline} intentionally delivers the packages produced in a pre--production -- preview --
-branch. Before that, once the software packages have been created, they are uploaded to a
-testing branch. The available automated deployment solution -- see Section \ref{subsec:sqa} -- checks
-the installation and configuration of the component relying on the testing repository. On
-successful completion, the packages are automatically signed and subsequently moved to
-the preview branch.
-In the last step, the release manager supervises the process to move the  packages to the
-production branch.
-
-The \textit{Docker container pipeline} uses the automated deployment solution to
-actually install and configure the software component in the container image. Once this is
-done the recently created image is uploaded to the production DockerHub repository, tagged
-with the label corresponding to the current {\sl INDIGO--DataCloud} release.
-Fig.~\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
-automatically triggered by a change in the source code.
-
-\begin{figure*}
+\begin{figure*}[ht]
 	\centering
@@
 \end{figure*}
+
+\subsubsection{Continuous Delivery}
+As DevOps suggests, frequent releases positively affect the reliability of the software as
+they reduce the time in which a given defect is exposed or has an impact, allowing the
+development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
+{\sl INDIGO--DataCloud} products, which have been taking place since the second major release, are passing through
+a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
+successful execution of the CI pipeline previously described. Consequently, CI--validated
+software components are automatically delivered in online repositories. Nonetheless,
+expert supervision to validate the results is required.
+
+{\sl INDIGO--DataCloud} software stack is delivered in the form of Linux software packages (rpms, debs)
+and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
+package pipeline} intentionally delivers the packages produced in a pre--production -- preview --
+branch. Before that, once the software packages have been created, they are uploaded to a
+testing branch. The available automated deployment solution -- see Section \ref{subsec:sqa} -- checks
+the installation and configuration of the component relying on the testing repository. On
+successful completion, the packages are automatically signed and subsequently moved to
+the preview branch.
+In the last step, the release manager supervises the process to move the  packages to the
+production branch.
+
+The \textit{Docker container pipeline} uses the automated deployment solution to
+actually install and configure the software component in the container image. Once this is
+done the recently created image is uploaded to the production DockerHub repository, tagged
+with the label corresponding to the current {\sl INDIGO--DataCloud} release.
+Fig.~\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
+automatically triggered by a change in the source code.
 
@@
 Staged--rollout process -- alleviates the likelihood of discovering uncovered
-issues at production. To put it in numbers, as Fig.~\ref{fig:ingigobugs} showcases, the 
+issues at production. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
 total number of defects detected in the pre--production stages surpassed by a factor of 30
@@
 project.
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth, height=50mm]{images/indigoBugs.png}
+\caption{Number of defects detected in development and production stages over the lifetime of the
+{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
+in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
+months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
+of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
+stages, thus not affecting the software released.}
+\label{fig:indigobugs}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width=0.5\textwidth, height=50mm]{images/indigoBugs.png}
-\caption{Number of defects detected in development and production stages over the lifetime of the
-{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project, 
-in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
-months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
-of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
-stages, thus not affecting the software released.}
-\label{fig:indigobugs}
-\end{figure}
-
 \section*{Acknowledgment}

### 2019-02-11 paper-jgc.tex
added file for the new template

--- 2018-09-27 paper-ieee-trans.tex
@@
-\documentclass[journal]{IEEEtran}
-
+%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is a general template file for the LaTeX package SVJour3
+% for Springer journals.          Springer Heidelberg 2010/09/16
+%
+% Copy it to a new file with a new name and use it as the basis
+% for your article. Delete % signs as needed.
+%
+% This template includes a few options for different layouts and
+% content for various journals. Please consult a previous issue of
+% your journal as needed.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% First comes an example EPS file -- just ignore it and
+% proceed on the \documentclass line
+% your LaTeX will extract the file if required
+%\begin{filecontents*}{example.eps}
+%!PS-Adobe-3.0 EPSF-3.0
+%%BoundingBox: 19 19 221 221
+%%CreationDate: Mon Sep 29 1997
+%%Creator: programmed by hand (JK)
+%%EndComments
+%gsave
+%newpath
+%  20 20 moveto
+%  20 220 lineto
+%  220 220 lineto
+%  220 20 lineto
+%closepath
+%2 setlinewidth
+%gsave
+%  .4 setgray fill
+%grestore
+%stroke
+%grestore
+%\end{filecontents*}
+%
+\RequirePackage{fix-cm}
+%
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
 \usepackage{graphicx}
@@
 
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+% etc.
+%
+% please place your own definitions here and don't use \def but
+% \newcommand{}{}
+%
+% Insert the name of "your journal" with
+% \journalname{myjournal}
+%
 \begin{document}
-\title{Evolution of Processes in Delivering Reliable Software within the European Scientific Arena}
-
-\author{E. Ronchieri,
-        P. Orviz Fernandez,
-        M. David,
-        D. C. Duma,
-        J. Gomes,
+
+\title{Evolution of Processes in Delivering Reliable Software within the European Scientific Arena%\thanks{Grants or other notes
+%about the article that should go on the front page should be
+%placed here. General acknowledgments should be placed at the end of the article.}
+}
+%\subtitle{}
+
+%\titlerunning{Short form of title}        % if too long for running head
+
+\author{M. David \and
+        D. C. Duma \and
+        P. Orviz Fernandez \and
+        E. Ronchieri \and
+        J. Gomes \and
         D. Salomoni
-\thanks{D. C. Duma, E. Ronchieri, and D. Salomoni are with INFN - CNAF, Bologna, Italy.}
-\thanks{P. Orviz Fernandez, is with CSIC, Santander, Spain.}
-\thanks{M. David, and J. Gomes are with LIP, Lisbon, Portugal.}%
+%
+%First Author         \and
+%        Second Author %etc.
 }
 
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{M. David \at
+LIP, Lisbon, Portugal \\
+%Tel.: +123-45-678910\\
+%Fax: +123-45-678910\\
+\email{david@lip.pt}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+\and
+D. C. Duma \at
+INFN - CNAF, Bologna, Italy\\
+\email{cristina.aiftimiei@cnaf.infn.it}
+\and
+P. Orviz Fernandez \at
+CSIC, Santander, Spain \\
+\email{orviz@ifca.unican.es}
+\and
+J. Gomes \at
+LIP, Lisbon, Portugal \\
+%Tel.: +123-45-678910\\
+%Fax: +123-45-678910\\
+\email{jorge@lip.pt}
+\and
+E. Ronchieri \at
+INFN - CNAF, Bologna, Italy\\
+\email{elisabetta.ronchieri@cnaf.infn.it}
+\and
+D. Salomoni \at
+INFN - CNAF, Bologna, Italy\\
+\email{davide.salomoni@cnaf.infn.it}
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
+
 \maketitle
@@
 \begin{abstract}
-
 From the advent of Grid technology -- as the new paradigm of distributed
@@
 procedures is present since the very early stages of the software development process.
-
+\keywords{Software Reliability \and Quality Assurance \and Software Metrics \and Software Testing
+Techniques \and DevOps}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
 
-\begin{IEEEkeywords}
-Software Reliability, Quality Assurance, Software Metrics, Software Testing
-Techniques, DevOps
-\end{IEEEkeywords}
-
-\IEEEpeerreviewmaketitle
 \section{Introduction}
-
-\IEEEPARstart{D}{uring} the last 15 years the provision of an European e--Infrastructure,
+\label{intro}
+During the last 15 years the provision of an European e--Infrastructure,
 supporting unified access to large--scale computing and intense data analysis
@@
 activities -- later discussed in Section~\ref{sec:ev} --
-were funded by the 5th, 6th and 7th Framework Programmes of the European Commission (EC)
+were funded by the 5$^{th}$, 6$^{th}$ and 7$^{th}$ Framework Programmes of the European Commission (EC)
 and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop
@@
 rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
-% Comment
-%P1 C2 L18: Claiming that stability and reliability was regularly compromised is quite a bold statement -- it needs at least some citations, and preferably some explanation of what forms this compromise took for a few prominent cases.
 reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
@@
 
-% Comments
-%P1 C2 L24: Was this realisation imposed "from above" e.g. by coordinated EC/ERC oversight, or was it "emergent" from the demands of the independent scientific communities?
 In an effort to improve the research community's experience, it became apparent the need for a better
 balance between the addition of new features and the reliability of the released software. Consequently,
-incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the
-software being delivered. At this point, evolving software engineering practices were considered
+incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the software being delivered.
+At this point, evolving software engineering practices were considered
 and gradually adopted to define and implement quality procedures, the organization of
@@
 \centering
-\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
+\begin{tabular}{p{1.6cm}p{2cm}p{5cm}l}
 \hline
@@
 \centering
-\begin{tabular}{llllllll}
+\begin{tabular}{p{2.6cm}lllllll}
 \hline
@@
 \centering
-\includegraphics[width=0.5\textwidth, height=50mm]{images/datagridEff.png}
+%\includegraphics[width=0.8\textwidth, height=50mm]{images/datagridEff.png}
+\includegraphics[width=1\textwidth]{images/datagridEff.png}
 \caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the project.}
@@
 \centering
-\begin{tabular}{lp{1.6cm}p{1.6cm}p{1.3cm}p{1.5cm}}
+\begin{tabular}{lp{2.5cm}p{2.2cm}p{1.9cm}p{2.1cm}}
 \hline
@@
 \centering
-\includegraphics[width=0.5\textwidth]{images/codestyle.png}
+\includegraphics[width=0.8\textwidth]{images/codestyle.png}
 \caption{Code style standards followed by {\sl INDIGO--DataCloud}'s software products.}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{images/fig3.png}
+\includegraphics[width=0.8\textwidth]{images/fig3.png}
 \caption{Adoption of Configuration Management tools throughout the project lifetime. The figure shows the trend lines
@@
 \centering
-\includegraphics[width=0.5\textwidth, height=50mm]{images/jenkins_CI_builds.png}
+\includegraphics[width=0.8\textwidth, height=50mm]{images/jenkins_CI_builds.png}
 \caption{Evolution of the total number of testing builds triggered automatically as part of the Jenkins CI implementation.}
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \centering
@@
 \label{fig:fig_CD}
-\end{figure*}
+\end{figure}
 
@@
 \centering
-\includegraphics[width=0.5\textwidth, height=50mm]{images/indigoBugs.png}
+\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
 \caption{Number of defects detected in development and production stages over the lifetime of the
@@
 
+%
+%
+%\section{Section title}
+%\label{sec:1}
+%Text with citations \cite{RefB} and \cite{RefJ}.
+%\subsection{Subsection title}
+%\label{sec:2}
+%as required. Don't forget to give each section
+%and subsection a unique label (see Sect.~\ref{sec:1}).
+%\paragraph{Paragraph headings} Use paragraph headings as needed.
+%\begin{equation}
+%a^2+b^2=c^2
+%\end{equation}
+%
+%% For one-column wide figures use
+%\begin{figure}
+%% Use the relevant command to insert your figure file.
+%% For example, with the graphicx package use
+%  \includegraphics{example.eps}
+%% figure caption is below the figure
+%\caption{Please write your figure caption here}
+%\label{fig:1}       % Give a unique label
+%\end{figure}
+%%
+%% For two-column wide figures use
+%\begin{figure*}
+%% Use the relevant command to insert your figure file.
+%% For example, with the graphicx package use
+%  \includegraphics[width=0.75\textwidth]{example.eps}
+%% figure caption is below the figure
+%\caption{Please write your figure caption here}
+%\label{fig:2}       % Give a unique label
+%\end{figure*}
+%%
+%% For tables use
+%\begin{table}
+%% table caption is above the table
+%\caption{Please write your table caption here}
+%\label{tab:1}       % Give a unique label
+%% For LaTeX tables use
+%\begin{tabular}{lll}
+%\hline\noalign{\smallskip}
+%first & second & third  \\
+%\noalign{\smallskip}\hline\noalign{\smallskip}
+%number & number & number \\
+%number & number & number \\
+%\noalign{\smallskip}\hline
+%\end{tabular}
+%\end{table}
+
+%\begin{acknowledgements}
+%If you'd like to thank anyone, place your comments here
+%and remove the percent signs.
+%\end{acknowledgements}
+
+% BibTeX users please use one of
+%\bibliographystyle{spbasic}      % basic style, author-year citations
+%\bibliographystyle{spmpsci}      % mathematics and physical sciences
+%\bibliographystyle{spphys}       % APS-like style for physics
+%\bibliography{}   % name your BibTeX data base
+
+% Non-BibTeX users please use
+%\begin{thebibliography}{}
+%
+% and use \bibitem to create references. Consult the Instructions
+% for authors for reference list style.
+%
+%\bibitem{RefJ}
+% Format for Journal Reference
+%Author, Article title, Journal, Volume, page numbers (year)
+% Format for books
+%\bibitem{RefB}
+%Author, Book title, page numbers. Publisher, place (year)
+% etc
+%\end{thebibliography}
+
 \end{document}
+% end of file template.tex

### 2019-02-13 paper-jgc.tex
Added text&reference of last year's EGI UMD publication.

@@
 the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb} -- and the
-management of the software provisioning process \cite{mario}, dealing with the validation and distribution of the software to
+management of the software provisioning process (SWPP) \cite{mario}, dealing with the validation and distribution of the software to
 the production infrastructures. As a result, this process led to a production--ready
@@
 Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
-by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).
+by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution). The
+increase in the number of products that the new CMD distribution brought in,
+compromised the effectiveness of the SWPP realization, in particular the
+software validation phase. The modernization of the EGI QC validation was
+accomplished through the adoption of automation, by means of the programmatic
+evaluation of its fundamental quality requirements \cite{orviz2018umd}.
 
@@
 [Accessed 4 Jun. 2018].
+
+\bibitem{orviz2018umd}
+P. Orviz \emph{et al.},
+``umd-verification: Automation of Software Validation for the EGI Federated e-Infrastructure,''
+\emph{J. Grid Comp.},
+vol. 16, no. 4, pp. 683--696, 2018.
 

### 2019-02-13 paper-jgc.tex
Rephrase to refer INDIGO as a concluded project

@@
 explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
-project of software development considered in this paper. At the time of writing,
-the project is succeeding in providing solutions to address the existing gaps in
-Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
+project of software development considered in this paper. The project addressed
+the existing gaps at the Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
 helping developers, e--Infrastructure administrators and scientific communities to exploit

### 2019-02-13 paper-jgc.tex
Updated reference to INDIGO paper published in 2018

@@
 The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
-explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
+explOitation, Apr 2015 -- Sep 2017) \cite{salomoni2018indigo} is the last
 project of software development considered in this paper. The project addressed
@@
 
-\bibitem{cordis:indigo}
+\bibitem{salomoni2018indigo}
 D. Salomoni \emph{et al.},
-``INDIGO-DataCloud: A data and computing platform to facilitate seamless access to e-infrastructures,''
-\emph{CoRR},
-vol. abs/1711.01981, 2018.
-[Online]. Available: \url{http://arxiv.org/abs/1711.01981}.
-[Accessed 4 Jun. 2018].
+``Indigo-datacloud: a platform to facilitate seamless access to e-infrastructures,''
+\emph{J. Grid Comp.},
+vol. 16, no. 3, pp. 381--408, 2018.
 

### 2019-02-13 paper-jgc.tex
Small statement to justify the elaborated analysis of INDIGO with respect to the others

@@
 {\sl INDIGO--DataCloud} is the most recent project included in this paper, it
-leveraged the lessons learned from the previous experiences described in
-Section~\ref{sec:ev}. The expertise gathered throughout these years was highly
+leverages the lessons learned from the previous experiences described in
+Section~\ref{sec:ev} and, consequently, it will be thoroughly analyzed. The expertise gathered throughout these years was highly
 profitable regarding evaluation and application of innovative software engineering

### 2019-02-14 paper-jgc.tex
Take defects figure out from conclusions.
Add Pablo as first author
Use PID for sqa-baseline
Double-check online URLs and update access date

@@
 
-\author{M. David \and
+\author{P. Orviz Fernandez \and
+        M. David \and
         D. C. Duma \and
-        P. Orviz Fernandez \and
         E. Ronchieri \and
@@
 
-\institute{M. David \at
+\institute{
+P. Orviz Fernandez \at
+CSIC, Santander, Spain \\
+\email{orviz@ifca.unican.es}
+\and
+M. David \at
 LIP, Lisbon, Portugal \\
@@
 \email{cristina.aiftimiei@cnaf.infn.it}
-\and
-P. Orviz Fernandez \at
-CSIC, Santander, Spain \\
-\email{orviz@ifca.unican.es}
 \and
@@
 time-efficient, leading to higher code coverages and increased defect detection, when compared with a manual
-approach \cite{rafi2012}.
+approach \cite{rafi2012}. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
+total number of defects detected in the pre--production stages surpassed by a factor of 30
+the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
+project.
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
+\caption{Number of defects detected in development and production stages over the lifetime of the
+{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
+in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
+months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
+of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
+stages, thus not affecting the software released.}
+\label{fig:indigobugs}
+\end{figure}
 
@@
 Staged--rollout process -- alleviates the likelihood of discovering uncovered
-issues at production. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
-total number of defects detected in the pre--production stages surpassed by a factor of 30
-the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
-project.
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
-\caption{Number of defects detected in development and production stages over the lifetime of the
-{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
-in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
-months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
-of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
-stages, thus not affecting the software released.}
-\label{fig:indigobugs}
-\end{figure}
+issues at production.
 
@@
 [Online]. Available: \url{https://ec.europa.eu/programmes/horizon2020/}.
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{http://www.agilemanifesto.org/}.
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://resources.sei.cmu.edu/asset_files/TechnicalReport/1993_005_001_16211.pdf}.
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online] Available: \url{https://edms.cern.ch/ui/file/358824/1.1/EDG-DevGuide-v1-2.pdf}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online] Available: \url{https://edms.cern.ch/ui/file/386039/2/QIv0-3.pdf}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online] Available: \url{http://cordis.europa.eu/project/rcn/80149_en.html}
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online] Available: \url{http://cordis.europa.eu/project/rcn/99189_en.html}
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online] Available: \url{http://cordis.europa.eu/project/rcn/87264_en.html}
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://edms.cern.ch/ui/file/1062487/2/EGEE-III-MSA3.4.2-1062487-v1_4.pdf}.
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://www.iso.org/standard/22749.html}.
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://egi-qc.github.io/}.
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online] Available: \url{http://cordis.europa.eu/project/rcn/194937_en.html}
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}.
-[Accessed 4 Jun. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 2017.
-[Online]. Available: \url{http://digital.csic.es/bitstream/10261/160086/4/CommonSQA-v2.pdf}.
-[Accessed 4 Jun. 2018].
+[Online]. Available: \url{http://hdl.handle.net/10261/160086}.
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://www.gitbook.com/@indigo-dc}.
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://forge.puppet.com/indigodc}.
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://galaxy.ansible.com/indigo-dc/}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://developer.github.com/v3/}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://jenkins.io/}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{http://grimoirelab.github.io/}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 2018.
-[Online]. Available: \url{https://jenkins.indigo-datacloud.eu/}
-[Accessed 30 Aug. 2018].
+[Online]. Available: \url{https://jenkins.indigo-datacloud.eu:8080/}
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://hub.docker.com/u/indigodatacloud}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 
@@
 [Online]. Available: \url{https://github.com/}
-[Accessed 30 Aug. 2018].
+[Accessed 14 Feb. 2019].
 

### 2019-02-14 paper-jgc.tex
Add SQA baseline GitHub repo reference

@@
 by providing minimum requirements of documentation content and stressing the usage of automated
-solutions for software deployment. The SQA criteria are continuously evolving \cite{sqa-baseline}, aiming at serving
-as a reference for future software development projects within the European research ecosystem.
+solutions for software deployment. The SQA criteria are publicly available \cite{sqa-baseline} and
+it is continuously evolving by means of an open collaboration procedure \cite{sqa-baseline-repo}
+that aims at serving as a reference for quality assurance realization in research software.
 
@@
 [Online]. Available: \url{http://hdl.handle.net/10261/160086}.
+[Accessed 14 Feb. 2019].
+
+\bibitem{sqa-baseline-repo}
+Members of the INDIGO-DataCloud, DEEP Hybrid-DataCloud and eXtreme DataCloud collaborations, 2015-2020,
+``A set of Common Software Quality Assurance Baseline Criteria for Research Projects,''
+2018.
+[Online]. Available: \url{https://github.com/indigo-dc/sqa-baseline}
 [Accessed 14 Feb. 2019].

### 2019-08-22 paper-jgc.tex
Set title for indigo-dc@GitBook
Title, abstract, intro & conclusions

@@
 
-\title{Evolution of Processes in Delivering Reliable Software within the European Scientific Arena%\thanks{Grants or other notes
-%about the article that should go on the front page should be
-%placed here. General acknowledgments should be placed at the end of the article.}
-}
+%\title{Evolution of Processes in Delivering Reliable Software within the Distributed European Research Infrastructures%\thanks{Grants or other notes
+%%about the article that should go on the front page should be
+%%placed here. General acknowledgments should be placed at the end of the article.}
+%}
+\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
 %\subtitle{}
@@
 of new tools and services to match the scientific community requirements has been
-addressed in Europe by various projects dedicated to software development
-and e--Infrastructure creation, operation and management.
-This work examines the evolution of the software engineering processes used in past and
-present European Commission--funded projects to illustrate how the research on software
-reliability has progressed in Europe over the last 15 years following the foundation of the
-first project in 2001. In particular, we will summarize the techniques and procedures
-applied to deliver quality software, highlighting the main challenges and barriers confronted
-at the time, and how they were partially or totally overcome. The knowledge base established
-throughout the years, sustained by the advances in the area of software engineering,
-definitely constituted a significant breakthrough in the reliability of software delivered in the
-European research arena. Recent software development projects funded by the European
-Commission are good evidence of such insights, where the enforcement of Quality Assurance
-procedures is present since the very early stages of the software development process.
+addressed in Europe through dedicated software development projects for
+e--Infrastructure creation, operation and management.
+This work presents the most significant software quality breakthroughs obtained in
+one of such projects, {\sl INDIGO--DataCloud}, the main challenges and barriers 
+confronted throughout the lifespan of the project, and how they were partially or 
+totally overcome.
+The knowledge base established throughout the last 15 years 
+of diverse software development initiatives in Europe for sustaining distributed research 
+e-Infrastructures, supported by the advances in the area of software engineering, definitely
+%constituted a significant breakthrough in the reliability of the software delivered in the 
+contributed to improve the reliability of the software delivered, and consequently, the 
+operational stability of the European e--Infrastructures. {\sl INDIGO--DataCloud} project is 
+a good evidence of such insights, where, unlike the preceding trend found in past projects,
+the enforcement of Software Quality Assurance practices has been present since the very early 
+stages of the software lifecycle.
 \keywords{Software Reliability \and Quality Assurance \and Software Metrics \and Software Testing
@@
 \label{intro}
-During the last 15 years the provision of an European e--Infrastructure,
-supporting unified access to large--scale computing and intense data analysis
-applications, has been driven by the needs of scientific communities as part of their
-pan--European research collaborations. One such example is the Worldwide
-Large Hadron Collider (LHC) Computing Grid infrastructure, driven
-by the particle physics scientific community, built to satisfy the computing needs
-of processing, storing and analysis of the very large amount of data produced by the
-LHC particle accelerator.
-
-A significant number of innovation projects and technology transfer
-activities -- later discussed in Section~\ref{sec:ev} --
-were funded by the 5$^{th}$, 6$^{th}$ and 7$^{th}$ Framework Programmes of the European Commission (EC)
-and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop
-software solutions that provide a seamless and transparent access by researchers to
-distributed computing and data resources.
-
-Major developments carried throughout the years were focused on providing new
-functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
-The released software was less tested, resulting in a higher rate of
-rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
-reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
-in the participant resource centers. Back in the days when Grid middleware was actively developed, researchers
-reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, eventually
-developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.
-
-In an effort to improve the research community's experience, it became apparent the need for a better
-balance between the addition of new features and the reliability of the released software. Consequently,
-incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the software being delivered.
+
+Throughout the last 15 years, Europe has have continuously invested on software development initiatives to provide
+research e--Infrastructures with the capability of supporting unified access to large--scale computing and intense 
+data analysis facilities. The developed tools have enabled the federation of geographically distributed resource 
+providers, with the ultimate aim of supporting the growing needs of the scientific comumunities. According to their
+requirements, those research e--Infrastructures have been evolving, exploiting diverse distributed computing 
+technologies namely Grid, in the first place, and, eventually, Cloud.
+
+The urging needs of providing stable e--Infrastructures, which were based on novel distributed computing technologies, 
+initially drove the efforts spent in software development towards the provision of new functionalities, relegating the 
+adoption of software engineering methodologies to a second place. The released software was less tested, resulting in 
+a high rate of rollbacks and patches applied to production systems, when compared to the subsequent projects. 
+Thereby, the stability and reliability of the e--Infrastructures were often compromised \cite{lingrand-egee}, leading 
+to outages in the participant resource centers. Back in the days when Grid middleware was actively developed, 
+researchers reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, 
+eventually developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.
+
+In an effort to improve the operational experience of research communities, it soon became apparent the need for a better 
+balance between the addition of new features and the reliability of the released software. Consequently, incoming European 
+funding programmes and open calls progressivelly stressed the fact of devoting a more substantial part of the software 
+lifecycle to the improvement of the quality and sustainability of the products being delivered. At this point,
+evolving software engineering practices were considered and gradually adopted in order to define and implement the quality
+procedures, the organization of cross--functional teams and the deployment of pilot testbeds.
+
+The increasing prominence of software quality procedures observed in the analysis of the last decade of EU--funded
+projects related to the development of distributed computing solutions --as described in Section~\ref{sec:ev}--, is 
+pictured in the software project herein introduced: {\sl INDIGO--DataCloud}. As we will thoroughly describe in the next
+sections, the ground base laid by these preceding projects and the adoption of practices taken from the prevailing 
+software engineering methodologies --in particular the Agile Software Development \cite{agile-manifesto} and the DevOps
+culture \cite{zhu}--, were the catalyst to establish a set of Software Quality Assurance (SQA) criteria 
+-- as explained in Section~\ref{software-quality-procedures} -- that guided the software lifecycle of the software 
+produced.
+
+Following the DevOps principles, the SQA criteria requires the verification of the source code on each change in the 
+production version, defining the essential requirements to be processed. Such scenario is hardly accomplished without
+the aid of automation, and as a consequence -- as described in Section~\ref{software-automation} -- the testing and 
+validation solution adopted by {\sl INDIGO--DataCloud} relied on the great advancements that continuous integration 
+(CI) tools have undergone in the last years. Accordingly, both the validation and delivery of {\sl INDIGO--DataCloud} 
+software was tackled by automation means.
+
+The fundamental shift brought by the INDIGO--DC SQA procedures was 
+to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
+it was packaged within a pre-release stage, the source code itself is verified for each minor change. This strategy 
+allows to improve the sustainability of the software, by checking the compliance with code style standards, and its 
+reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
+presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
+significantly lower number of bugs detected by the end users. Consequently, bugs are more effortessly fixed since they 
+are early captured and end users interface with more reliable and stable software.
+
+The paper is organized as follows. Section~\ref{sec:b} summarizes the essential context of the research problem.
+Section~\ref{sec:ev} details the evolution of research on software reliability in the selected EC--projects. 
+Section~\ref{sec:ntsr} provides information about new research trends in software reliability based upon the
+experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%
+
+%In this paper, we outline the set of EC--funded projects, ranging from 2001 to nowadays,
+%that have raised the awareness and, at the same time, laid the ground work for the establishment of 
+%thorough SQA practices to guide the software development lifecycle within the {\sl INDIGO--DataCloud}
+%project. 
+%%
+%%For each project, the challenges, the methodologies and the achievements towards software 
+%%reliability are described.
+%%During the last 15 years the provision of European e--Infrastructures,
+%%supporting unified access to large--scale computing and intense data analysis
+%%applications have been driven by the needs of scientific communities as part of their
+%%pan--European research collaborations. 
+%%One such example is the Worldwide
+%%Large Hadron Collider (LHC) Computing Grid infrastructure, driven
+%%by the particle physics scientific community, built to satisfy the computing needs
+%%of processing, storing and analysis of the very large amount of data produced by the
+%%LHC particle accelerator.
+%
+%A significant number of the innovation projects and technology transfer
+%activities presented in Section~\ref{sec:ev}
+%were funded by the 5$^{th}$, 6$^{th}$ and 7$^{th}$ Framework Programmes of the European Commission (EC)
+%and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop
+%software solutions that provide a seamless and transparent access by researchers to
+%distributed computing and data resources.
+%
+%Major developments carried throughout the years were focused on providing new
+%functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
+%The released software was less tested, resulting in a higher rate of
+%rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
+%reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
+%in the participant resource centers. Back in the days when Grid middleware was actively developed, researchers
+%reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, eventually
+%developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.
+%
+%In an effort to improve the research community's experience, it became apparent the need for a better
+%balance between the addition of new features and the reliability of the released software. Consequently,
+%incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the software being delivered.
 At this point, evolving software engineering practices were considered
-and gradually adopted to define and implement quality procedures, the organization of
-cross--functional teams and deployment of pilot testbeds.
-
-Analysing the EC--funded software projects over the last decade,
-one can observe a continuous increase in the prominence and robustness of software
-testing and validation procedures \cite{aiftimiei}. The
-evolution of the software engineering methodologies (creation of the Agile
-Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
-with the parallel advancements in the information and communication technology
-field (such as virtualization, the enabling technology for Cloud computing and
-operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend.
-This evolution was complemented by the emergence of automation and event--response tools.
-
-In this paper, we outline a set of EC--funded projects
-ranging from 2001 to nowadays, in which the authors were involved.
-For each project, the challenges, the methodologies and
-the achievements towards software reliability are described.
-
-The remainder of this paper is
-organized as follows. Section~\ref{sec:b} summarizes the essential context of the research problem.
-Section~\ref{sec:ev} details the evolution of research on
-software reliability in EC--projects. Section~\ref{sec:ntsr} provides
-information about new research trends in software reliability based upon the
-experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC).
-Conclusions are drawn in Section~\ref{sec:con}.
+%and gradually adopted to define and implement quality procedures, the organization of
+%cross--functional teams and deployment of pilot testbeds.
+%
+%As a result of the analysis the aforementioned projects, 
+%one can observe a continuous increase in the prominence and robustness of software
+%testing and validation procedures \cite{aiftimiei}. The
+%evolution of the software engineering methodologies (creation of the Agile
+%Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
+%with the parallel advancements in the information and communication technology
+%field (such as virtualization, the enabling technology for Cloud computing and
+%operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend.
+%This evolution was complemented by the emergence of automation and event--response tools.
+%
+%The former know-how combined with the continuous advancements in sofware engineering 
+%have been the catalysts for drawing attention to the relevance of SQA ..
 
@@
 
-\subsection{Software Quality Procedures}
+\subsection{Software Quality Procedures}\label{software-quality-procedures}
 \label{subsec:sqa}
@@
 
-\subsection{Software quality validation and delivery automation}
+\subsection{Software quality validation and delivery automation}\label{software-automation}
 \label{sec:devops}
@@
 \includegraphics[width=0.8\textwidth, height=50mm]{images/indigoBugs.png}
-\caption{Number of defects detected in development and production stages over the lifetime of the
-{\sl INDIGO-DataCloud} project. The defects started to arise in the second year of the project,
-in parallel with the release of {\sl INDIGO--1} software, while the major activity was concentrated in the
-months that preceded the second major release of {\sl INDIGO--2}, in April 2017. The practical implementation
-of the SQA definition described throughout this paper was the key actor in uncovering defects while in pre--production
-stages, thus not affecting the software released.}
+\caption{Number of software bugs -- documentation typos are excluded -- detected in development 
+and production stages over the lifetime of the {\sl INDIGO-DataCloud} project. Production 
+defects correspond to software bugs filed by users of the EGI e--Infrastructure throughout 
+the next 2 years after the {\sl INDIGO--1} major release, in August 2016 (source: EGI's GGUS
+helpdesk). The development defects stem only from the pre--release testing phase, filed 
+exclusively as part of the SQA team work (source: GitHub Issues tracker). As a consequence, the 
+development defects started to arise in the previous months prior to the {\sl INDIGO--1} release, 
+while the major activity was concentrated in the months that preceded the second major release 
+of {\sl INDIGO--2}, in April 2017. As it can be observed, the offset between the advent of defects
+in development and production corresponds to nearly half a year, with the first production defect
+reported 3 months after the {\sl INDIGO--1} release.}
+
 \label{fig:indigobugs}
@@
 
-After more than a decade of experiences collected throughout the reviewed
-EU--funded software development and e--Infrastructure management projects, it
-becomes apparent that the European software produced for scientific purposes is
-evolving to a more sustainable model where the quality and reliability of
-software is being prioritized. The gradual adoption of software engineering insights, such as
-DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
-collaborative tools, extending their capabilities to bring along tight
-integrations amongst other services, matched the requirements of such practices
-and procedures, facilitating in a great manner their implementation in real scenarios.
+%After more than a decade of experiences collected throughout the reviewed
+%EU--funded software development and e--Infrastructure management projects, it
+%becomes apparent that the European software produced for scientific purposes is
+%evolving to a more sustainable model where the quality and reliability of
+%software is being prioritized. The gradual adoption of software engineering insights, such as
+%DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
+%collaborative tools, extending their capabilities to bring along tight
+%integrations amongst other services, matched the requirements of such practices
+%and procedures, facilitating in a great manner their implementation in real scenarios.
 
@@
 I am very sympathetic to the idea that systematic development techniques can help improve software quality (without getting into circular definitions, where "quality" means that methodologies or standards have been applied!) but there will also be a point at which the methodology gets too heavy and compromises the "agility" of the development team: how can you identify which processes are helping and which are not?
-Reliability of the software has been substantially improved by applying the SQA
-criteria in the development stage. Consequently, each change in the source code
-that is meant to be integrated in the production version, is validated according
-to the SQA testing requirements. Thus, defects are detected and corrected early in the
-software lifecycle, reducing the cost of fixing errors at later stages. Automation is
-key in this process, allowing the implementation of a CI scenario that enforces
-the SQA criteria compliance. The application of more advanced DevOps practices
-resulted in the implementation of CD pipelines to extend automation to the
-delivery of packages. Thereby, safe and frequent software releases are possible, reducing
-the impact on production systems. In this regard, a pre--production validation,
+Both reliability and sustainability of the {\sl INDIGO--DataCloud} software products 
+have been primarily improved by acting at the source code level. According to the 
+size of the project, which comprised more than 30 products and 200 GitHub repositories, 
+this strategy required a remarkable preliminary ground work. On the one hand, the 
+definition of a set of SQA requirements -- compiled in the described SQA criteria 
+-- that shall be fulfilled prior to release by the {\sl INDIGO--DataCloud} core 
+products. On the other hand, the need of setting up the testing infrastructure and
+the associated CI/CD pipelines to check that the changes in the source code matched 
+those requirements in an automated fashion. As a complement to the above strategy, a 
+final validation step combined the deployment of the pre--released software versions 
+in the project's preview testbed, and subsequently, most of the products were tested
+in selected production environments through the aid of EGI's staged--rollout process.
+
+The most noticeable metric that demonstrates a reliability improvement comes from the 
+fact that the number of software defects uncovered throughout the pre--release phase 
+surpassed by a factor of 30 the ones reported by end users in the EGI e-Infrastructure
+during the next two years after the first major release (INDIGO-1). Accordingly, the
+sustainability of the software was ...
+
+Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
+a progressive quality improvement throughout the lifespan of the project. Indeed,
+unit testing coverage dropped for about 18\% of the total products halfway through
+the project. This behavior was aligned with the high-demanding periods of software
+development, in particular, the previous months before the second major release 
+(INDIGO--2). This fact demonstrates that, even considered as best practice, test 
+cases are not commonly written right after, or even before, introducing a new unit
+or functionality in the code. High coverage values are generally advisable but they
+entail time--consuming work that is often neglected by the code owners. The effort 
+spent in testing should be oriented to those sections of the code that are more 
+likely to cause negative effects to end users. Consequently, a high coverage value 
+does not generally provide a linear benefit with respect to the effort spent -- the
+SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
+it does not indicate whether the key sections of the code are covered. However, unit
+testing is a valuable tool to detect defects at coding time, so whenever the test 
+cases accompany the changes done in the code, a high unit testing coverage value is 
+a good indicator of reliability in the software.
+%(related defect increase?, resulting from the INDIGO-2 release?)
+
+% products not developed from scratch challenged the application of SQA policies??
+
+%
+
+applying the SQA criteria in the development stage. As each change in 
+the source code is validated to match the SQA requirements, defects in the software 
+are more likely to be detected and fixed at the earliest stage. Consequently, costs
+are reduced, since solving bugs at later stages require greater efforts, and the end
+user obtains a much more reliable piece of software, as demonstrated by the low rate
+of defects found once it is in production. Automation is of paramount importance to 
+guarantee the {\sl INDIGO--DataCloud}'s SQA criteria compliance. Thus, through the
+application of DevOps practices, a CI/CD scenario was put in place to cover both the
+validation and delivery of the {\sl INDIGO--DataCloud} software products. The CI/CD 
+pipelines provide the agility to produce releases frequently enough to reduce the 
+impact on production systems, since the software gets progressively updated.
+
+In this regard, a pre--production validation,
 through the preview testbeds or resource center's early adoption -- handled by the
@@
 \bibitem{foot1}
-indigo-dc, ``Sign in to GitBook - GitBook,''
+indigo-dc, ``indigo-dc Spaces - GitBook,''
 2018.

### 2019-08-23 paper-jgc.tex
conclusions concluded

@@
 %}
-\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
+%\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
+\title{INDIGO-DataCloud: a converging evolution of software engineering practices to support European Research e-Infrastructures}
 %\subtitle{}
@@
 
-The fundamental shift brought by the INDIGO--DC SQA procedures was 
+The fundamental shift brought by the {\sl INDIGO--DataCloud} SQA procedures was 
 to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
@@
 
-The most noticeable metric that demonstrates a reliability improvement comes from the 
-fact that the number of software defects uncovered throughout the pre--release phase 
-surpassed by a factor of 30 the ones reported by end users in the EGI e-Infrastructure
-during the next two years after the first major release (INDIGO-1). Accordingly, the
-sustainability of the software was ...
+The most noticeable metric that demonstrates a reliability improvement is represented 
+by the ratio between the number of software defects uncovered throughout the pre--release
+phase of each of the major releases, {\sl INDIGO--1} and {\sl INDIGO--2}, and the ones 
+reported by end users within the EGI e--Infrastructure, where core products were 
+deployed. The number of bugs detected in the pre--release phase, filed by the SQA
+responsibles of the project, surpassed by a factor of 30 the ones reported by those end 
+users, which in absolute terms, this value reached no more than 200 bugs over the
+observed timeframe. Considering that these low rate of bugs were uncovered throughout 
+the testing phase in the project's preview testbed -- where the new releases were 
+deployed in order to validate their integration with other component's production 
+versions --, one can infer that the previous phase of code analysis -- running unit and
+functional test cases for each minor change -- considerably helped in preventing the 
+emergence of a greater quantity of software defects. As it was described, this latter
+phase of code analysis included as well the compliance with de--facto style standards, 
+improving the readability of the code and, consequently, building on the sustainability
+path of the software.
 
@@
 a good indicator of reliability in the software.
-%(related defect increase?, resulting from the INDIGO-2 release?)
-
-% products not developed from scratch challenged the application of SQA policies??
-
-%
-
-applying the SQA criteria in the development stage. As each change in 
-the source code is validated to match the SQA requirements, defects in the software 
-are more likely to be detected and fixed at the earliest stage. Consequently, costs
-are reduced, since solving bugs at later stages require greater efforts, and the end
-user obtains a much more reliable piece of software, as demonstrated by the low rate
-of defects found once it is in production. Automation is of paramount importance to 
-guarantee the {\sl INDIGO--DataCloud}'s SQA criteria compliance. Thus, through the
-application of DevOps practices, a CI/CD scenario was put in place to cover both the
-validation and delivery of the {\sl INDIGO--DataCloud} software products. The CI/CD 
-pipelines provide the agility to produce releases frequently enough to reduce the 
-impact on production systems, since the software gets progressively updated.
-
-In this regard, a pre--production validation,
-through the preview testbeds or resource center's early adoption -- handled by the
-Staged--rollout process -- alleviates the likelihood of discovering uncovered
-issues at production.
-
-The software lifecycle demands as well a continuous measurement and analysis
-process, which allows timely reactions to discover and fix defects throughout the
-development, release and maintenance stages. Conscientious metric
-tracking and response sits at the core of an appropriate software maturity, as
-ultimate reliability can only be achieved by enforcing and empowering the SQA
-criteria through each stage of the software lifecycle.
-
-Lessons learned show that sustainability is a key pursuit. The achievements, in terms of
-processes, knowledge, software and infrastructures, shall be profited from former successful
-completed projects. In this regard, the more recent project described in this paper, {\sl INDIGO-DataCloud},
-may have set the path for future reference in the European software development research ecosystem.
-The SQA criteria, describing the quality requirements that might be considered during the entire
-software lifecycle, have been published as open access and periodically reviewed to incorporate
-new recommendations and best practices. Successful software products created during the lifetime of the
-project are maintained leveraging the same source code repositories and CI infrastructure. A seamless
-knowledge transfer among subsequent software development initiatives is key for the good progress of
-delivering reliable software.
+% ?? (related defect increase?, resulting from the INDIGO-2 release?)
+
+All along the project we have observed that source code review is the cornerstone to
+improve the completeness of the unit test cases, as a commitment to strenghten the 
+structural quality of the software. Therefore, we move forward in avoiding the 
+aforementioned risks of having high coverage values obtained from meaningless test 
+cases, which execute a great amount of statements and lines of the code, but using
+incomplete test cases. Code reviewers analyse the content of each change in the 
+source code and the suitability -- in the above terms -- of the associated tests, 
+being the last checkpoint before the change is added in the production branch. Hence 
+the vital importance, as mentioned earlier, of providing the tests alongside the code
+for every given change. As we have seen, the production branches of 
+{\sl INDIGO--DataCloud} source code repositories were protected against direct pushes,
+requiring a manual assessment step before approval through peer code review.
+% ?? figure with code reviews or sth about PRs, external participation
+
+Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
+software lifecycle through the per--change execution of the functional tests provided 
+by the developers and the ultimate stage of integration testing in the aforementioned 
+preview testbed. As we have seen, the latter testing strategy was entirely performed 
+in a manual way, and appeared as a helpful tool to spot defects after exposing the 
+deployed software to interactions with external components. Additionally, the functional 
+tests, were primarily useful to check that the software met the design specifications, 
+and most importantly, that previously working functionalities are preserved when new 
+features are added. As mentioned, these type of testing is meant to be continuously 
+executed in an automated manner using the CI infrastructure. However, by the time of
+{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
+reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked for those 
+products that did not provide automated functional tests, tracking the fulfillment of 
+their fundamental functional requirements was tougher a change--based approach, 
+especially when assessing the regressions. The major issues observed that hindered the
+implementation of automated functional tests were releated to the unavailability, at
+the time, of libraries and frameworks for performing specific actions -- such as 
+testing graphical interfaces -- or due to the code owners' neglect -- insufficient
+knowledge of the tools -- or rejection -- time--consuming task --. As a result of this,
+those products eventually -- even before {\sl INDIGO--1} -- provided functional test
+reports prior to the release, thus being moved away from the CI scenario. We infer
+from this experience that automated functional tests should be a goal for modern
+software since they complement the structural quality checks by adding the validation
+of the software's identified functionalities to improve the effectiveness of the 
+DevOps approach.
+% figure with % of products with functional tests
+
+As a final remark, we want to highlight the practical experience resulted from the
+application of the formerly described software engineering practices to improve the 
+validation of a specific application, selected from the existing use cases in the 
+{\sl INDIGO--DataCloud} project, thus expanding the scope beyond the core products. 
+The outcome was a prototype CI/CD pipeline that analysed the code, packaged and 
+tested the application based on a pre--defined set of inputs and reference outputs. 
+The pipeline delivered then the new version only if all the stages were successful
+completed. Research communities are not usually experts in programming or in related 
+software engineering practices, so this fact was a real accomplishment in terms of 
+the quality and sustainability of their software.
+
+% ?? products not developed from scratch challenged the application of SQA policies??
+
+%
+
+%applying the SQA criteria in the development stage. As each change in 
+%the source code is validated to match the SQA requirements, defects in the software 
+%are more likely to be detected and fixed at the earliest stage. Consequently, costs
+%are reduced, since solving bugs at later stages require greater efforts, and the end
+%user obtains a much more reliable piece of software, as demonstrated by the low rate
+%of defects found once it is in production. Automation is of paramount importance to 
+%guarantee the {\sl INDIGO--DataCloud}'s SQA criteria compliance. Thus, through the
+%application of DevOps practices, a CI/CD scenario was put in place to cover both the
+%validation and delivery of the {\sl INDIGO--DataCloud} software products. The CI/CD 
+%pipelines provide the agility to produce releases frequently enough to reduce the 
+%impact on production systems, since the software gets progressively updated. 
+%
+%
+%In this regard, a pre--production validation,
+%through the preview testbeds or resource center's early adoption -- handled by the
+%Staged--rollout process -- alleviates the likelihood of discovering uncovered
+%issues at production.
+
+The software lifecycle demands from a continuous measurement and analysis process, 
+which allows timely reactions in order to discover and fix defects throughout the
+development, release and maintenance stages. Conscientious metric tracking and 
+response sits at the core of an appropriate software maturity, as ultimate reliability
+can only be achieved by enforcing and empowering the SQA criteria through each stage
+of the software lifecycle. Lessons learned show that sustainability is a key pursuit.
+The achievements, in terms of processes, knowledge, software and infrastructures, shall
+be profited from former successful completed projects. In this regard, 
+{\sl INDIGO-DataCloud} may have set a path for future reference in software development
+practices for the European research e--Infrastructures. The SQA criteria, describing 
+the quality requirements that might be considered during the entire software lifecycle,
+have been published as open access and periodically reviewed to incorporate new 
+recommendations and best practices. Successful software products created during the
+lifetime of the project are maintained leveraging the same source code repositories and
+CI infrastructure. A seamless knowledge transfer among subsequent software development
+initiatives is key for the good progress of delivering reliable software.
 

### 2019-09-11 paper-jgc.tex
Compacting conclusions

@@
 definition of a set of SQA requirements -- compiled in the described SQA criteria 
--- that shall be fulfilled prior to release by the {\sl INDIGO--DataCloud} core 
-products. On the other hand, the need of setting up the testing infrastructure and
-the associated CI/CD pipelines to check that the changes in the source code matched 
-those requirements in an automated fashion. As a complement to the above strategy, a 
-final validation step combined the deployment of the pre--released software versions 
-in the project's preview testbed, and subsequently, most of the products were tested
-in selected production environments through the aid of EGI's staged--rollout process.
+-- ought to be fulfilled by the set of core products before being
+released. On the other hand, the setup of a testing infrastructure and
+the associated CI/CD pipelines that guarantee that the changes done at the source code
+level match
+those requirements in an automated fashion. As a final step, the validation of the 
+resultant pre--released software versions firstly through integration testing in the
+project's preview testbed, and subsequently complemented by the EGI's staged--rollout
+process, where the brand new versions are adopted in the first place by candidate
+resource providers within the e-Infrastructure -- accessible for user evaluation --, 
+before being tagged as production--ready software, and consequently, distributed 
+through the EGI channels.
 
@@
 phase of each of the major releases, {\sl INDIGO--1} and {\sl INDIGO--2}, and the ones 
-reported by end users within the EGI e--Infrastructure, where core products were 
-deployed. The number of bugs detected in the pre--release phase, filed by the SQA
-responsibles of the project, surpassed by a factor of 30 the ones reported by those end 
-users, which in absolute terms, this value reached no more than 200 bugs over the
-observed timeframe. Considering that these low rate of bugs were uncovered throughout 
-the testing phase in the project's preview testbed -- where the new releases were 
-deployed in order to validate their integration with other component's production 
-versions --, one can infer that the previous phase of code analysis -- running unit and
-functional test cases for each minor change -- considerably helped in preventing the 
-emergence of a greater quantity of software defects. As it was described, this latter
-phase of code analysis included as well the compliance with de--facto style standards, 
-improving the readability of the code and, consequently, building on the sustainability
-path of the software.
+reported by the end users within the EGI e--Infrastructure, where the core products were 
+deployed. The number of bugs detected throughout the SQA process here described 
+surpassed by a factor of 30 the number of bugs reported by those end users, over a total
+of almost 200 bugs in the observed timeframe. Considering that this low rate of bugs -- 
+taking into account the amount of software products existing in {\sl INDIGO--DataCloud} -- 
+correspond to the outcomes of the integration 
+testing phase, it can be inferred that the preceding phase of code analysis -- enforcing 
+compliance with a style standard, and running
+unit and functional test cases for each minor change -- substantially contributed 
+to the robustness of the {\sl INDIGO--DataCloud} software, uncovering defects at earlier 
+stages, thus leaving the pre--released software with a low amount of errors.
+% graph with GitHub defects/issues in development
+% graph showing types of defects detected in the integration phase
 
 Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
-a progressive quality improvement throughout the lifespan of the project. Indeed,
+a progressive quality improvement throughout its lifespan. Indeed,
 unit testing coverage dropped for about 18\% of the total products halfway through
@@
 or functionality in the code. High coverage values are generally advisable but they
-entail time--consuming work that is often neglected by the code owners. The effort 
-spent in testing should be oriented to those sections of the code that are more 
-likely to cause negative effects to end users. Consequently, a high coverage value 
-does not generally provide a linear benefit with respect to the effort spent -- the
-SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
-it does not indicate whether the key sections of the code are covered. However, unit
-testing is a valuable tool to detect defects at coding time, so whenever the test 
-cases accompany the changes done in the code, a high unit testing coverage value is 
-a good indicator of reliability in the software.
+entail time--consuming work that is often neglected by the code owners. Rather than
+requiring a high coverage value, our experience showed that increasing the coverage
+of those sections in the code that are more likely to cause negative effects to end
+users is not only a more effective approach to improve software reliability but it 
+also keeps developers motivated in the task of test coding, as it brings more 
+noticeable benefits in the software operation by end users.
 % ?? (related defect increase?, resulting from the INDIGO-2 release?)
 
+%The effort 
+%spent in testing should be oriented to those sections of the code that are more 
+%likely to cause negative effects to end users. Consequently, a high coverage value 
+%does not generally provide a linear benefit with respect to the effort spent -- the
+%SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
+%it does not indicate whether the key sections of the code are covered. However, unit
+%testing is a valuable tool to detect defects at coding time, so whenever the test 
+%cases accompany the changes done in the code, a high unit testing coverage value is 
+%a good indicator of reliability in the software.
+
 All along the project we have observed that source code review is the cornerstone to
-improve the completeness of the unit test cases, as a commitment to strenghten the 
-structural quality of the software. Therefore, we move forward in avoiding the 
-aforementioned risks of having high coverage values obtained from meaningless test 
-cases, which execute a great amount of statements and lines of the code, but using
-incomplete test cases. Code reviewers analyse the content of each change in the 
-source code and the suitability -- in the above terms -- of the associated tests, 
-being the last checkpoint before the change is added in the production branch. Hence 
-the vital importance, as mentioned earlier, of providing the tests alongside the code
-for every given change. As we have seen, the production branches of 
-{\sl INDIGO--DataCloud} source code repositories were protected against direct pushes,
-requiring a manual assessment step before approval through peer code review.
+improve the effectiveness of the unit test cases. {\sl INDIGO--DataCloud} source code 
+repositories are protected against direct pushes so that the code review stage always
+takes place. Code reviewers analyse the content of each change in the source code and
+the suitability -- in the above terms -- of the associated tests, being the last
+checkpoint before the change is added into the production branch. Hence the vital
+importance of providing the tests alongside the code for every given change.
 % ?? figure with code reviews or sth about PRs, external participation

### 2019-09-16 paper-jgc.tex
Finishing conclusions & remove what it was called background

--- 2019-08-22 paper-jgc.tex
@@
 %}
-\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
+%\title{Delivering more Reliable Software within the INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
+\title{Software Quality Assurance in INDIGO-DataCloud project: a converging evolution of software engineering practices to support European Research e-Infrastructures}
 %\subtitle{}
@@
 
-Throughout the last 15 years, Europe has have continuously invested on software development initiatives to provide
+Throughout the last 15 years, the European Commission (EC) has continuously invested on software development initiatives to provide
 research e--Infrastructures with the capability of supporting unified access to large--scale computing and intense 
@@
 The urging needs of providing stable e--Infrastructures, which were based on novel distributed computing technologies, 
-initially drove the efforts spent in software development towards the provision of new functionalities, relegating the 
-adoption of software engineering methodologies to a second place. The released software was less tested, resulting in 
+initially drove the efforts spent in software development towards the provision of new functionalities, reducing the 
+adoption of software engineering methodologies to a marginal player. The released software was less tested, resulting in 
 a high rate of rollbacks and patches applied to production systems, when compared to the subsequent projects. 
@@
 projects related to the development of distributed computing solutions --as described in Section~\ref{sec:ev}--, is 
-pictured in the software project herein introduced: {\sl INDIGO--DataCloud}. As we will thoroughly describe in the next
+pictured in the software project herein described: {\sl INDIGO--DataCloud}. As we will thoroughly describe in the next
 sections, the ground base laid by these preceding projects and the adoption of practices taken from the prevailing 
@@
 
-The paper is organized as follows. Section~\ref{sec:b} summarizes the essential context of the research problem.
+The paper is organized as follows. %Section~\ref{sec:b} summarizes the essential context of the research problem.
 Section~\ref{sec:ev} details the evolution of research on software reliability in the selected EC--projects. 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-
-%In this paper, we outline the set of EC--funded projects, ranging from 2001 to nowadays,
-%that have raised the awareness and, at the same time, laid the ground work for the establishment of 
-%thorough SQA practices to guide the software development lifecycle within the {\sl INDIGO--DataCloud}
-%project. 
-%%
-%%For each project, the challenges, the methodologies and the achievements towards software 
-%%reliability are described.
-%%During the last 15 years the provision of European e--Infrastructures,
-%%supporting unified access to large--scale computing and intense data analysis
-%%applications have been driven by the needs of scientific communities as part of their
-%%pan--European research collaborations. 
-%%One such example is the Worldwide
-%%Large Hadron Collider (LHC) Computing Grid infrastructure, driven
-%%by the particle physics scientific community, built to satisfy the computing needs
-%%of processing, storing and analysis of the very large amount of data produced by the
-%%LHC particle accelerator.
-%
-%A significant number of the innovation projects and technology transfer
-%activities presented in Section~\ref{sec:ev}
-%were funded by the 5$^{th}$, 6$^{th}$ and 7$^{th}$ Framework Programmes of the European Commission (EC)
-%and recently by the Horizon 2020 programme \cite{h2020}. Their main goal was to develop
-%software solutions that provide a seamless and transparent access by researchers to
-%distributed computing and data resources.
-%
-%Major developments carried throughout the years were focused on providing new
-%functionalities, whereas in the early days the adoption of software engineering methodologies had a lower priority.
-%The released software was less tested, resulting in a higher rate of
-%rollbacks and patches applied to production systems, compared to the later projects. Thereby, the stability and
-%reliability of the e--Infrastructures was often compromised \cite{lingrand-egee}, leading to outages
-%in the participant resource centers. Back in the days when Grid middleware was actively developed, researchers
-%reported the need of improvements both in terms of reliability and performance \cite{campana-atlas}, eventually
-%developing their own custom solutions \cite{kindermann}\cite{mendez-lorenzo} on top of the existing services.
-%
-%In an effort to improve the research community's experience, it became apparent the need for a better
-%balance between the addition of new features and the reliability of the released software. Consequently,
-%incoming European funding programmes and open calls stressed the fact of devoting a substantial part of the software lifecycle to the improvement of the quality of the software being delivered.
-At this point, evolving software engineering practices were considered
-%and gradually adopted to define and implement quality procedures, the organization of
-%cross--functional teams and deployment of pilot testbeds.
-%
-%As a result of the analysis the aforementioned projects, 
-%one can observe a continuous increase in the prominence and robustness of software
-%testing and validation procedures \cite{aiftimiei}. The
-%evolution of the software engineering methodologies (creation of the Agile
-%Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
-%with the parallel advancements in the information and communication technology
-%field (such as virtualization, the enabling technology for Cloud computing and
-%operating--system--level virtualization known as containerization \cite{soltesz}) motivated this trend.
-%This evolution was complemented by the emergence of automation and event--response tools.
-%
-%The former know-how combined with the continuous advancements in sofware engineering 
-%have been the catalysts for drawing attention to the relevance of SQA ..
-
-\section{Background}
-\label{sec:b}
-
-In this section, we detail the level of knowledge requested in delivering reliable software in order
-to understand the effort employed in the EC-funded software projects over time. We divide the literature
-into four different domains that intersect this work: 1) software requirements analysis, 2) software
-coding,
-3) metrics and measurements, 4) testing.
-
-\emph{Software requirements analysis.} They are defined at various stages of a software development process
-as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
-analysis is a process for studying user needs to arrive at a definition of system, hardware or software
-requirements. Characteristics of requirements that can assess software reliability are consistency and
-correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable
-results, and the latter usually refers to the satisfaction of certain goals.
-
-\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer
-program in a programming language. To meet specific requirements, a program can be developed by using
-high-level or low-level programming languages that envelope different characteristics and may affect the
-result and quality of software. Furthermore, software is generally written by geographically distributed
-teams whose developers can change over time, therefore it is extremely important for source code to be
-understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
-development process to easily adapt software: programming features that influence the
-code readability and maintainability \cite{buse} and standardized code that promotes the comparability of results \cite{baggen}.
-
-\emph{Metrics and measurements.} According to IEEE \cite{radatz}, metrics is a function, with input as
-the software data, and output is a value which could decide on how the given attribute affects the software.
-Metrics and measurements have an important role in the development process \cite{fenton, capers}. Metrics
-are used to improve the ability to control, identify and measure the essential parameters of software
-during its development, leading to improve the results of the software product \cite{goodman}. With a correct
-usage of metrics, it is possible to determine which parts of the software require to be changed or modified.
-In the literature, different types of metrics exist to be able to provide information on various software
-aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}. Often they have
-similar definitions, so as a consequence they can be implemented in different ways according to the
-interpretation of their definition. This is the reason why it has been always a challenge to select the
-proper software metric tools.
-
-\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component
-under specified conditions, observing or recording the results, and making an evaluation of some aspect of
-the system or component. Testing is an important factor for software quality and reliability improvement. This
-activity is quite complicated due to the variety of programming languages, operating systems and hardware
-platforms that evolve over time. In the current literature, there are many types of software
-testing methods \cite{myers} that aim at finding errors, bugs and faults during the software development
-process, possibly in the early stages. Over time, testing metrics have been defined to assess software
-quality \cite{athanasiou}: code coverage is the most frequently used metric for testing software reliability
-and there exist many tools for dynamic code coverage estimation according to the programming language
-\cite{horgan}. A dedicated effort has to be taken into account in order to integrate testing solutions
-during the software development process.
-
-\section{Evolution of research on software reliability in EC--projects}
+%\section{Background}
+%\label{sec:b}
+%
+%In this section, we detail the level of knowledge requested in delivering reliable software in order
+%to understand the effort employed in the EC-funded software projects over time. We divide the literature
+%into four different domains that intersect this work: 1) software requirements analysis, 2) software
+%coding,
+%3) metrics and measurements, 4) testing.
+%
+%\emph{Software requirements analysis.} They are defined at various stages of a software development process
+%as a specification of what should be implemented. According to IEEE \cite{radatz}, requirement
+%analysis is a process for studying user needs to arrive at a definition of system, hardware or software
+%requirements. Characteristics of requirements that can assess software reliability are consistency and
+%correctness \cite{sommerville}: the former aims at providing predictable, maintainable and reliable
+%results, and the latter usually refers to the satisfaction of certain goals.
+%
+%\emph{Software coding.} According to IEEE \cite{radatz}, coding is the process of expressing a computer
+%program in a programming language. To meet specific requirements, a program can be developed by using
+%high-level or low-level programming languages that envelope different characteristics and may affect the
+%result and quality of software. Furthermore, software is generally written by geographically distributed
+%teams whose developers can change over time, therefore it is extremely important for source code to be
+%understandable \cite{tashtoush}. There are various instruments that can be adopted during the software
+%development process to easily adapt software: programming features that influence the
+%code readability and maintainability \cite{buse} and standardized code that promotes the comparability of results \cite{baggen}.
+%
+%\emph{Metrics and measurements.} According to IEEE \cite{radatz}, metrics is a function, with input as
+%the software data, and output is a value which could decide on how the given attribute affects the software.
+%Metrics and measurements have an important role in the development process \cite{fenton, capers}. Metrics
+%are used to improve the ability to control, identify and measure the essential parameters of software
+%during its development, leading to improve the results of the software product \cite{goodman}. With a correct
+%usage of metrics, it is possible to determine which parts of the software require to be changed or modified.
+%In the literature, different types of metrics exist to be able to provide information on various software
+%aspects, such as complexity \cite{mccabe} and object orientation \cite{chidamber, lorenz}. Often they have
+%similar definitions, so as a consequence they can be implemented in different ways according to the
+%interpretation of their definition. This is the reason why it has been always a challenge to select the
+%proper software metric tools.
+%
+%\emph{Testing.} According to IEEE \cite{radatz}, testing is a process of operating a system or component
+%under specified conditions, observing or recording the results, and making an evaluation of some aspect of
+%the system or component. Testing is an important factor for software quality and reliability improvement. This
+%activity is quite complicated due to the variety of programming languages, operating systems and hardware
+%platforms that evolve over time. In the current literature, there are many types of software
+%testing methods \cite{myers} that aim at finding errors, bugs and faults during the software development
+%process, possibly in the early stages. Over time, testing metrics have been defined to assess software
+%quality \cite{athanasiou}: code coverage is the most frequently used metric for testing software reliability
+%and there exist many tools for dynamic code coverage estimation according to the programming language
+%\cite{horgan}. A dedicated effort has to be taken into account in order to integrate testing solutions
+%during the software development process.
+%
+%
+%\section{Evolution of research on software reliability in EC--projects}
+\section{Background work on European e--Infrastructure projects}
 \label{sec:ev}
+
+One of the fundamental pillars of the {\sl INDIGO--DataCloud}'s SQA process has been established
+by the outcomes resulted from similar EC--funded e--Infrastructure--related projects over
+the last 15 years. In particular, in this section we are primarily interested in those projects that consisted in
+the collaborative development of software solutions, intended for improving the accessibility and exploitation 
+of distributed computing technologies by research communities within the European research
+landscape. We will highlight the most representative software engineering
+practices adopted within each of these projects to understand how they inspired the SQA process 
+definition and implementation of {\sl INDIGO--DataCloud}.
 
@@
 Each project paved the way for the following ones, according to a logic evolution in which
-the role of the quality and reliability of the software has
-consistently increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
+the quality and reliability of the software solutions have progressively increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
 committed to, and aware of, software reliability practices.
@@
 definition of a set of SQA requirements -- compiled in the described SQA criteria 
--- that shall be fulfilled prior to release by the {\sl INDIGO--DataCloud} core 
-products. On the other hand, the need of setting up the testing infrastructure and
-the associated CI/CD pipelines to check that the changes in the source code matched 
-those requirements in an automated fashion. As a complement to the above strategy, a 
-final validation step combined the deployment of the pre--released software versions 
-in the project's preview testbed, and subsequently, most of the products were tested
-in selected production environments through the aid of EGI's staged--rollout process.
-
-The most noticeable metric that demonstrates a reliability improvement comes from the 
-fact that the number of software defects uncovered throughout the pre--release phase 
-surpassed by a factor of 30 the ones reported by end users in the EGI e-Infrastructure
-during the next two years after the first major release (INDIGO-1). Accordingly, the
-sustainability of the software was ...
+-- that ought to be fulfilled by the set of core products before being
+released. On the other hand, the setup of a testing infrastructure and
+the associated CI/CD pipelines to guarantee that the changes done at the source code
+level match
+those requirements in an automated fashion. As a final step, the validation of the 
+resultant pre--released software versions, firstly tackled through integration testing in
+the project's preview testbed, and subsequently complemented by the EGI's 
+staged--rollout process, where the brand new versions are adopted by candidate resource
+providers within the e-Infrastructure -- accessible for user evaluation --, before being
+tagged as production--ready software, and consequently, distributed through the official
+EGI channels.
+
+The most noticeable metric that demonstrates a reliability improvement is represented 
+by the ratio between the number of software defects uncovered throughout the pre--release
+phase of each of the major releases, {\sl INDIGO--1} and {\sl INDIGO--2}, and the ones 
+reported by the end users within the EGI e--Infrastructure, where the core products were 
+deployed. The number of bugs detected throughout the SQA process here described 
+surpassed by a factor of 30 the number of bugs reported by those end users, over a total
+of almost 200 bugs in the observed timeframe. Considering that this low rate of bugs -- 
+taking into account the amount of software products existing in the project -- 
+correspond to the outcomes of the integration testing phase, we infer that the preceding
+phase of code analysis -- comprised by the style standard compliance, as well as the unit and
+functional test cases execution for each minor change -- had a substantial impact on the robustness
+of the resultant software.
+% graph with GitHub defects/issues in development
+% graph showing types of defects detected in the integration phase
 
 Nevertheless, the individual products of {\sl INDIGO--DataCloud} did not undergo 
-a progressive quality improvement throughout the lifespan of the project. Indeed,
+a progressive quality improvement throughout its lifespan. Indeed,
 unit testing coverage dropped for about 18\% of the total products halfway through
@@
 or functionality in the code. High coverage values are generally advisable but they
-entail time--consuming work that is often neglected by the code owners. The effort 
-spent in testing should be oriented to those sections of the code that are more 
-likely to cause negative effects to end users. Consequently, a high coverage value 
-does not generally provide a linear benefit with respect to the effort spent -- the
-SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
-it does not indicate whether the key sections of the code are covered. However, unit
-testing is a valuable tool to detect defects at coding time, so whenever the test 
-cases accompany the changes done in the code, a high unit testing coverage value is 
-a good indicator of reliability in the software.
-%(related defect increase?, resulting from the INDIGO-2 release?)
-
-% products not developed from scratch challenged the application of SQA policies??
-
-%
-
-applying the SQA criteria in the development stage. As each change in 
-the source code is validated to match the SQA requirements, defects in the software 
-are more likely to be detected and fixed at the earliest stage. Consequently, costs
-are reduced, since solving bugs at later stages require greater efforts, and the end
-user obtains a much more reliable piece of software, as demonstrated by the low rate
-of defects found once it is in production. Automation is of paramount importance to 
-guarantee the {\sl INDIGO--DataCloud}'s SQA criteria compliance. Thus, through the
-application of DevOps practices, a CI/CD scenario was put in place to cover both the
-validation and delivery of the {\sl INDIGO--DataCloud} software products. The CI/CD 
-pipelines provide the agility to produce releases frequently enough to reduce the 
-impact on production systems, since the software gets progressively updated.
-
-In this regard, a pre--production validation,
-through the preview testbeds or resource center's early adoption -- handled by the
-Staged--rollout process -- alleviates the likelihood of discovering uncovered
-issues at production.
-
-The software lifecycle demands as well a continuous measurement and analysis
-process, which allows timely reactions to discover and fix defects throughout the
-development, release and maintenance stages. Conscientious metric
-tracking and response sits at the core of an appropriate software maturity, as
-ultimate reliability can only be achieved by enforcing and empowering the SQA
-criteria through each stage of the software lifecycle.
-
-Lessons learned show that sustainability is a key pursuit. The achievements, in terms of
-processes, knowledge, software and infrastructures, shall be profited from former successful
-completed projects. In this regard, the more recent project described in this paper, {\sl INDIGO-DataCloud},
-may have set the path for future reference in the European software development research ecosystem.
-The SQA criteria, describing the quality requirements that might be considered during the entire
-software lifecycle, have been published as open access and periodically reviewed to incorporate
-new recommendations and best practices. Successful software products created during the lifetime of the
-project are maintained leveraging the same source code repositories and CI infrastructure. A seamless
-knowledge transfer among subsequent software development initiatives is key for the good progress of
-delivering reliable software.
+entail time--consuming work that is often neglected by the code owners. Rather than
+requiring a high coverage value, our experience showed that increasing the coverage
+of those sections in the code that are more likely to cause negative effects to end
+users is not only a more effective approach to improve software reliability, but it 
+also keeps developers motivated in the tedious task of test coding, as it brings more 
+noticeable benefits for the end user.
+% ?? (related defect increase?, resulting from the INDIGO-2 release?)
+
+%The effort 
+%spent in testing should be oriented to those sections of the code that are more 
+%likely to cause negative effects to end users. Consequently, a high coverage value 
+%does not generally provide a linear benefit with respect to the effort spent -- the
+%SQA criteria recommends 70\% of coverage as a good compromise -- and, in addition,
+%it does not indicate whether the key sections of the code are covered. However, unit
+%testing is a valuable tool to detect defects at coding time, so whenever the test 
+%cases accompany the changes done in the code, a high unit testing coverage value is 
+%a good indicator of reliability in the software.
+
+All along the project we have observed that source code review is the cornerstone to
+improve the effectiveness of the unit and functional -- when applicable -- test cases. 
+{\sl INDIGO--DataCloud} source code repositories are protected against direct pushes 
+so that the code review stage always takes place. Code reviewers analyse the content
+of each change in the source code and the suitability -- according to the SQA criteria -- of the
+associated tests, being the last checkpoint before the change is added into the
+production branch. Hence the vital importance of providing the tests alongside the
+code for every given change.
+% ?? figure with code reviews or sth about PRs, external participation
+%Functional quality, on the other hand, was assessed within the {\sl INDIGO--DataCloud} 
+%software lifecycle through the per--change execution of the functional tests provided 
+%by the developers and the ultimate stage of integration testing in the aforementioned 
+%preview testbed. As we have seen, the latter testing strategy was entirely performed 
+%in a manual way, and appeared as a helpful tool to spot defects after exposing the 
+%deployed software to interactions with external components. Additionally, the functional 
+%tests, were primarily useful to check that the software met the design specifications, 
+%and most importantly, that previously working functionalities are preserved when new 
+%features are added. As mentioned, these type of testing is meant to be continuously 
+%executed in an automated manner using the CI infrastructure. 
+However, by the time of
+{\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
+reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked in those cases
+where automated functional tests were not provided, the fulfillment of 
+their fundamental functional requirements was tougher to track, resulting in a non--viable
+solution for assessing minor changes, especially when analysing the regressions. The major issues observed that hindered developers from the 
+implementation of automated functional tests were releated to the unavailability
+-- at the time -- of libraries and frameworks for tackling specific tests -- such as 
+testing graphical interfaces --, or in some cases, the unawareness about their existence.
+We infer from this experience that automated functional tests should be a goal in modern
+software development as they complement the structural quality checks by adding the
+validation of the software's identified functionalities, key in the operational impact 
+and user acceptance. 
+% figure with % of products with functional tests
+% ?? products not developed from scratch challenged the application of SQA policies??
+
+The application of the formerly described SQA process was not only beneficial for the
+core products' quality and reliability, but it has proven to be equally effective for
+those applications uniquely supported by research communities with little background 
+on software engineering. This fact was demonstrated by the elaboration of an ad--hoc validation pipeline for one
+of the use cases supported in the project. The CI/CD pipeline extended the source code testing and delivery
+-- already existing in the pipelines for the core products -- with a basic system--like testing. 
+For the latter, the
+pre--packaged application was tested with a pre--defined set of inputs and expected
+outputs. Only if successfully validated, the source code could be merged into production 
+and the resultant pre--packaged application would be delivered. When this pipeline was
+integrated in the usual workflow of the application development, it was reported to have constituted a real 
+breakthrough, allowing a faster development and more reliable software, as it was programmatically tested
+before being distributed. Consequently, the 
+benefits of a continuous SQA improvement process are not only suitable for experienced
+software developers, but also to any type of software practitioner, such as the numerous ones
+commonly found within the research communities.
+
+The {\sl INDIGO--DataCloud}'s SQA process has persisted once the project concluded, being 
+extensively used by the supported software products. The SQA criteria is actively
+maintained to include new recommendations and best practices, not only targeted for
+experts but for any enthusiast on research software. Hence, they have established the foundations for 
+knowledge transfer capabilities, which we foresee it will be convenient for the rapid adoption of 
+SQA practices in the subsequent software development initiatives and projects for e--Infrastructure 
+building and management, taking a step forward in the overarching goal
+of delivering reliable software in research.
 

### 2019-09-17 paper-jgc.tex
remove background section

@@
 Section~\ref{sec:ntsr} provides information about new research trends in software reliability based upon the
-experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section~\ref{sec:con}.
+experiences from the recent {\sl INDIGO--DataCloud} project. Conclusions are drawn in Section~\ref{sec:con}.
 
@@
 by the outcomes resulted from similar EC--funded e--Infrastructure--related projects over
-the last 15 years. In particular, in this section we are primarily interested in those projects that consisted in
-the collaborative development of software solutions, intended for improving the accessibility and exploitation 
-of distributed computing technologies by research communities within the European research
-landscape. We will highlight the most representative software engineering
-practices adopted within each of these projects to understand how they inspired the SQA process 
-definition and implementation of {\sl INDIGO--DataCloud}.
+the last 15 years. In particular, in this section we are primarily interested in hightlighting the 
+collaborative
+software development efforts within the category of projects that improved the accessibility and
+exploitation of distributed computing technologies in the European research landscape. As such, 
+we will present the most representative software engineering practices adopted by those projects
+to understand how they ultimately inspired the SQA process definition and implementation in 
+{\sl INDIGO--DataCloud}, further discussed in Section~\ref{sec:ntsr}.
 
@@
 Each project paved the way for the following ones, according to a logic evolution in which
-the quality and reliability of the software solutions have progressively increased. As shown in Table~\ref{tab:feat}, the most recent projects are more
-committed to, and aware of, software reliability practices.
-Table~\ref{tab:feat} also shows the topics addressed in the following sections, highlighting
-the commonalities and specific features on software reliability.
+an increasing awareness in software quality and reliability has been showcased. As shown in 
+Table~\ref{tab:feat}, the most recent projects are more committed to, and aware of, software
+engineering practices. Here we include {\sl INDIGO--DataCloud} (INDIGO--DC) -- see 
+Section~\ref{sec:ntsr} -- as the reference project where all the listed features have been achieved.
 
@@
 number of jobs successfully completed divided by the total number of jobs submitted. As it can be observed,
-there is a positive trend of efficiency over the months, with the unique exception of the LCG1 experiment.
+there is a positive trend of efficiency over the months, with the unique exception of the LCG--1 prototype.
 \begin{figure}[h]
@@
 
-\subsection{INtegrating Distributed data Infrastructures for Global explOitation}
-
-The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
-explOitation, Apr 2015 -- Sep 2017) \cite{salomoni2018indigo} is the last
-project of software development considered in this paper. The project addressed
-the existing gaps at the Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
-helping developers, e--Infrastructure administrators and scientific communities to exploit
-Cloud computing benefits.
-
-\section{New trends in software reliability}
+%\subsection{INtegrating Distributed data Infrastructures for Global explOitation}
+%
+%The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
+%explOitation, Apr 2015 -- Sep 2017) \cite{salomoni2018indigo} is the last
+%project of software development considered in this paper. The project addressed
+%the existing gaps at the Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
+%helping developers, e--Infrastructure administrators and scientific communities to exploit
+%Cloud computing benefits.
+
+\section{INDIGO--DataCloud Software Quality Assurance Process}
 \label{sec:ntsr}
 
-{\sl INDIGO--DataCloud} is the most recent project included in this paper, it
+{\sl INDIGO--DataCloud} project \cite{salomoni2018indigo} %is the most recent project included in this paper, it
 leverages the lessons learned from the previous experiences described in
-Section~\ref{sec:ev} and, consequently, it will be thoroughly analyzed. The expertise gathered throughout these years was highly
-profitable regarding evaluation and application of innovative software engineering
-methodologies, management of new technologies to put those methodologies
-into practice, compilation and analysis of an appropriate set of metrics to measure the
-quality of the implemented solutions.
-An additional level of complexity, comes from the fact that these large collaborative projects
-have an extensive list of partners.
-
-In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the software lifecycle:
+Section~\ref{sec:ev}. The expertise gathered throughout these years was highly
+profitable regarding the adoption of state--of--the--art software engineering
+methodologies, the management of new technologies to put those methodologies
+into practice and the analysis of an appropriate set of metrics to measure the
+quality of the implemented solutions. All of which with the additional complexity of
+being framed under collaborative environments with an extensive list of partners.
+
+However, new challenges appear in the course of a software development project, which in some
+cases are tackled using novel software engineering approaches. In this regard, 
+{\sl INDIGO--DataCloud} faced analogous challenges as the ones reported by parallel EC--funded 
+software development initiatives \cite{casale2016current} that were addressed using similar 
+solutions, such as relying on DevOps practices for tackling the software lifecycle management. 
+The SQA process followed by {\sl INDIGO--DataCloud}
+combines these solutions with pragmatic approaches built upon external and first--hand 
+experiences. In the following sections, we describe the three main pillars that sustain the
+project's strategy for attaining quality in the software produced:
 
@@
 \end{enumerate}
+
+As we will describe hereafter, the {\sl INDIGO--DataCloud} SQA process is not restricted to software targeted to distributed
+computing environments, but instead oriented to the management of any form of collaboration driven software development, composed
+by multiple software projects that collaborate jointly. However, there are particularities, such
+as the stage--rollout phase, that are attached to an e--Infrastructure's operation, thus might not
+be applicable to all cases.
 
@@
 vol. 16, no. 3, pp. 381--408, 2018.
+
+\bibitem{casale2016current}
+G. Casale \emph{et al.},
+``Current and future challenges of software engineering for services and applications,''
+\emph{Procedia Computer Science},
+vol. 97, no. 3, pp. 34--42, 2016.
 

### 2019-09-25 paper-jgc.tex
reliability term usage

@@
 %constituted a significant breakthrough in the reliability of the software delivered in the 
-contributed to improve the reliability of the software delivered, and consequently, the 
+contributed to improve the quality and reliability of the software delivered, and consequently, the 
 operational stability of the European e--Infrastructures. {\sl INDIGO--DataCloud} project is 
@@
 data analysis facilities. The developed tools have enabled the federation of geographically distributed resource 
-providers, with the ultimate aim of supporting the growing needs of the scientific comumunities. According to their
+providers, with the ultimate aim of supporting the growing needs of the scientific communities. According to their
 requirements, those research e--Infrastructures have been evolving, exploiting diverse distributed computing 
@@
 a high rate of rollbacks and patches applied to production systems, when compared to the subsequent projects. 
-Thereby, the stability and reliability of the e--Infrastructures were often compromised \cite{lingrand-egee}, leading 
+Thereby, the stability of the e--Infrastructures were often compromised \cite{lingrand-egee}, leading 
 to outages in the participant resource centers. Back in the days when Grid middleware was actively developed, 
@@
 In an effort to improve the operational experience of research communities, it soon became apparent the need for a better 
-balance between the addition of new features and the reliability of the released software. Consequently, incoming European 
-funding programmes and open calls progressivelly stressed the fact of devoting a more substantial part of the software 
+balance between the addition of new features and the quality of the released software. Consequently, incoming European 
+funding programmes and open calls progressively stressed the fact of devoting a more substantial part of the software 
 lifecycle to the improvement of the quality and sustainability of the products being delivered. At this point,
@@
 
-Following the DevOps principles, the SQA criteria requires the verification of the source code on each change in the 
-production version, defining the essential requirements to be processed. Such scenario is hardly accomplished without
-the aid of automation, and as a consequence -- as described in Section~\ref{software-automation} -- the testing and 
-validation solution adopted by {\sl INDIGO--DataCloud} relied on the great advancements that continuous integration 
-(CI) tools have undergone in the last years. Accordingly, both the validation and delivery of {\sl INDIGO--DataCloud} 
-software was tackled by automation means.
-
-The fundamental shift brought by the {\sl INDIGO--DataCloud} SQA procedures was 
-to act at the early stages of the software development lifecycle. Thus, instead of validating the software product once 
-it was packaged within a pre-release stage, the source code itself is verified for each minor change. This strategy 
-allows to improve the sustainability of the software, by checking the compliance with code style standards, and its 
+Following the DevOps principles, the SQA criteria encourages a thorough verification of the source code, inspecting
+and testing each minor change meant to be included in the production version. Considering that 
+{\sl INDIGO--DataCloud} solution comprises more than 30 software products and 200 source code repositories, such 
+scenario is hardly accomplished without the aid of automation. Consequently, the SQA process builds on the great 
+advancements that the continuous integration (CI) tools have undergone over the last years, as described in 
+Section~\ref{software-automation}. Hence, acting at the early stages of the software development lifecycle brought
+a fundamental shift when compared with the reviewed {\sl INDIGO--DataCloud}'s preceding projects that dealt with
+the task of developing software solutions for Grid and Cloud--based e--Infrastructures.
+
+This strategy 
+allows to improve the overall quality of the software, both in terms of sustainability, by checking the compliance with code style standards, and 
 reliability, through the execution of functional and integration tests, ensuring the total absence of regressions. The results
 presented in this paper show a divergence between the high rate of bug fixes resolved at development time and the 
-significantly lower number of bugs detected by the end users. Consequently, bugs are more effortessly fixed since they 
+significantly lower number of bugs detected by the end users. Consequently, bugs are more effortlessly fixed since they 
 are early captured and end users interface with more reliable and stable software.
@@
 The paper is organized as follows. %Section~\ref{sec:b} summarizes the essential context of the research problem.
-Section~\ref{sec:ev} details the evolution of research on software reliability in the selected EC--projects. 
-Section~\ref{sec:ntsr} provides information about new research trends in software reliability based upon the
-experiences from the recent {\sl INDIGO--DataCloud} project. Conclusions are drawn in Section~\ref{sec:con}.
+Section~\ref{sec:ev} details the evolution of SQA processes in the preceding EC--funded projects that developed 
+software solutions for Grid and Cloud--based e--Infrastructures. 
+Section~\ref{sec:ntsr} thoroughly describes the new trends in software engineering conducted during the course of the recently finished
+{\sl INDIGO--DataCloud} project. Conclusions are drawn in Section~\ref{sec:con}.
 
@@
 by the outcomes resulted from similar EC--funded e--Infrastructure--related projects over
-the last 15 years. In particular, in this section we are primarily interested in hightlighting the 
+the last 15 years. In particular, in this section we are primarily interested in emphasizing the 
 collaborative
@@
 produced. Development guidelines \cite{datagridguide} were produced, covering the various phases, including
-1) packaging, 2) test and validation, and 3) style and naming convention. The specification of quality indicators
-\cite{datagridindicators} facilitated the detection of deviations in the collected metric data. Figure
-\ref{fig:efficiency} shows the efficiency performance indicator, measured in production environments as the
-number of jobs successfully completed divided by the total number of jobs submitted. As it can be observed,
-there is a positive trend of efficiency over the months, with the unique exception of the LCG--1 prototype.
+1) packaging, 2) test and validation, and 3) style and naming convention. A set of quality indicators
+\cite{datagridindicators} were defined to measure the efficiency of the system from the user perspective. 
+Due to the developmental nature of DataGrid project, they solely consider the \textit{crude efficiency} 
+metric as the relevant indicator to track the quality and performance of the core software, also known as
+\emph{middleware}
+in Grid terminology. This metric considered the success rate of all the jobs submitted by users using the 
+production testbed, so that all the failures of the middleware were included in the statistic \cite{avellino2004datagrid}. However, it
+also accounted for the failures in the remaining pieces in the job submission chain, such as the user application 
+software or the targeted datacenter deployment. Even so, as it can be observed in Fig.~\ref{fig:efficiency}, 
+the crude efficiency showed a positive trend over the months that followed, with the unique exception of the 
+LCG--1 middleware distribution.
 \begin{figure}[h]
@@
 \includegraphics[width=1\textwidth]{images/datagridEff.png}
-\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are provided for five experiments involved in the project.}
+\caption{Efficiency \% = Number of jobs successfully completed / Total Number of jobs submitted: results are 
+provided for the different DataGrid's EDG middleware distributions released throughout the project. Note that the figure
+includes metrics of the same release (EDGv1.4) for three Large Hadron Collider (LHC) experiments.}
 \label{fig:efficiency}
@@
 Grid computing. This e-Infrastructure was eventually used by more than 15 thousand researchers
-and deployed in over 250 institutions.
+and deployed in over 250 institutions \cite{ferrari2011resources}.
 
@@
 were exclusively released during {\sl INDIGO--2} do not show data for the first period of
-{\sl INDIGO--1}.}
+{\sl INDIGO--1}. Source code for all the {\sl INDIGO--DataCloud} components can be found
+in the GitHub's \emph{indigo--dc} organization \cite{foot9}}
 \label{fig:fig_unittest}
@@
 internal design. The SQA criteria of {\sl INDIGO--DataCloud} set a recommended threshold of 70\% coverage
-for unit testing. In Fig.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
+for unit testing. In Figure.~\ref{fig:fig_unittest}, unit testing coverage is compared for each software
 component between the two major releases, denoting an incremental trend over the course of the project:
@@
 approach \cite{rafi2012}. To put it in numbers, as Fig.~\ref{fig:indigobugs} showcases, the
-total number of defects detected in the pre--production stages surpassed by a factor of 30
+total number of defects detected in the pre--production stage, i.e. throughout the integration testing phase, surpassed by a factor of 30
 the ones reported by external users, over the lifetime of the {\sl INDIGO-DataCloud}
-project.
+project. An accurate estimation of the software bugs detected -- and fixed -- with the aid of 
+the CI/CD infrastructure, previously conducted at the source code analysis stage, cannot be
+given as those bugs are usually fixed on the fly without being tracked. Nevertheless, we can
+safely state that they exceed by far the total amount of 200 bugs that the Fig.~\ref{fig:indigobugs}
+shows.
 \begin{figure}[h]
@@
 operations (ops) teams, by emphasizing the SQA techniques to avoid infrastructure disruption whenever new
-developments are deployed into production systems.
-The incremental adoption of DevOps approaches throughout the project's lifetime was a key player towards
-software reliability. Starting with a Continuous Integration (CI) scenario at the first stages of the project, the software delivered
+developments are deployed into production systems. Starting with a Continuous Integration (CI) scenario at the first stages of the project, the software delivered
 was eventually validated and distributed using Continuous Delivery (CD) pipelines.
@@
 
-Two pilot infrastructures are at the disposal of developers and scientific
-communities involved in the project. The aim of these testbeds is to test the
-level of integration between the components involved in the {\sl INDIGO--DataCloud}
-solutions and use cases validation. Within these testbeds, the applications are deployed
-and executed
-with the last stable version of the software in environments as similar as
-possible to production ones. A map of the Pilot Preview
+Two pilot infrastructures, integration and preview, were at the disposal of developers and use cases, 
+respectively, involved in the project. The aim of these testbeds is to test the level of integration 
+between the core {\sl INDIGO--DataCloud}'s components and the user applications.
+
+The integration testbed was primarily targeted and used by the software developers. 
+There, unstable
+software releases were tested by the maintainers themselves in order to validate the interactions
+with coupled services. Computing and storage resources were occasionaly needed, meaning that no real
+testbed -- with all
+the {\sl INDIGO--DataCloud}'s components -- was needed.
+
+Conversely, the preview testbed was actively maintained, where the last stable version of
+the software was deployed by the SQA team. On top of that, the user applications were 
+deployed in order to test the integration with the {\sl INDIGO--DataCloud} solution. The
+preview testbed proved to be an effective tool to uncover bugs by means of testing the 
+basic functionalities of the core components. The bugs tagged as \emph{Dev} in
+Fig.~\ref{fig:indigobugs} were spotted this stage. A map of the Pilot Preview
 infrastructure is depicted in Fig.~\ref{fig:fig_pilotpreview}. It shows the
-resource providers and the components or services they deployed and supported.
-
-The released software is tested in production environments through the
-Staged--Rollout process, before its inclusion in the EGI's Cloud Middleware Distribution (CMD).
-Selected resource providers are requested to install
-the most updated stable versions of the software, and to give access to them to their users. The
-Staged--Rollout process is key to detect and mitigate issues that could only
-appear in production environments.
+resource providers and the services deployed. The number of resources
+steadily increased as the project progressed, due mainly to the progressive increase of 
+supported services and the occasional need for deploying more than one stable version of
+the same software component in different providers.
+
+Computing and storage resource consumption, required by the pilot testbeds and SQA services
+(metrics, CI/CD infrastructure), was never an issue for the project's resource providers.
+In the case of computing resources, the massive use of container--based virtualization 
+alleviated the need of resources as it allowed CPU and memory over--subscription, when no
+performance test was carried out. Most of the providers were already members of the EGI 
+FedCloud, providing medium--sized Infrastructure--as--a--Service Cloud frameworks with 
+enough computing and storage capacity to meet the deployment requirements of more than 30 
+products that comprised the {\sl INDIGO--DataCloud} solution and the related SQA services.
+Nevertheless, the pilot infrastructure coordination task distributed the deployments
+according each provider's capacity and effort received within the project. Consequently,
+Fig.~\ref{fig:fig_pilotpreview} shows an unbalanced distribution of services.
+
+As the final validation step, the released software was tested in production environments
+through the \emph{staged--rollout} process, before its inclusion in the EGI's Cloud 
+Middleware Distribution (CMD). This process selected resource providers from the EGI FedCloud
+infrastructure in order to install the last stable versions of the {\sl INDIGO--DataCloud}
+software, giving user access to them. The staged--rollout process is key to detect and 
+mitigate issues that could only appear in production environments.
 
@@
 solution for assessing minor changes, especially when analysing the regressions. The major issues observed that hindered developers from the 
-implementation of automated functional tests were releated to the unavailability
+implementation of automated functional tests were related to the unavailability
 -- at the time -- of libraries and frameworks for tackling specific tests -- such as 
@@
 
+\bibitem{avellino2004datagrid}
+G. Avellino \emph{et al.},
+``The DataGrid workload management system: Challenges and results,''
+\emph{J. Grid Comp.},
+vol. 2, no. 4, pp. 353--367, 2004.
+
 \bibitem{gagliardi}
@@
 
+\bibitem{ferrari2011resources}
+T. Ferrari \emph{et al.},
+``Resources and services of the EGEE production infrastructure,''
+\emph{J. Grid Comp.},
+vol. 9, no. 2, pp. 119--133, 2011.
+
 \bibitem{glite}
@@
 [Online]. Available: \url{https://hub.docker.com/u/indigodatacloud}
+[Accessed 14 Feb. 2019].
+
+\bibitem{foot9}
+GitHub's indigo-dc organization, ``indigo-dc,''
+2018.
+[Online]. Available: \url{https://github.com/indigo-dc}
 [Accessed 14 Feb. 2019].

### 2019-12-05 paper-jgc.tex
Add clarification for Grid jobs

@@
 \emph{middleware}
-in Grid terminology. This metric considered the success rate of all the jobs submitted by users using the 
+in Grid terminology. This metric considered the success rate of all the Grid jobs submitted by users using the 
 production testbed, so that all the failures of the middleware were included in the statistic \cite{avellino2004datagrid}. However, it

### 2019-12-08 paper-jgc.tex
minor revision changes

@@
 thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
-based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
-which 5 were devoted to software development and coordinated by an
-Architecture Task Force, supervising the overall design and technical consistency
-of the developments.
+based on Globus middleware components and services \cite{globus}.
 
@@
 they reduce the time in which a given defect is exposed or has an impact, allowing the
-development teams to act promptly based on the regular feedback \cite{chen2015}. The software updates of
-{\sl INDIGO--DataCloud} products, which have been taking place since the second major release, are passing through
+development teams to act promptly based on the regular feedback \cite{chen2015}. A total of 12 software
+updates --including 91 product releases \cite{indigo-d33}--, which have been taking place since {\sl INDIGO--1}, are passing through
 a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
@@
 {\sl INDIGO--1} only 30\% of the total of products provided automated functional tests,
-reaching almost 60\% in {\sl INDIGO--2}. Although test reports were asked in those cases
-where automated functional tests were not provided, the fulfillment of 
+reaching almost 60\% in {\sl INDIGO--2}. This fact was undoubtedly the main cause that challenged the full implementation of 
+the described DevOps approach. 
+Although testing reports were asked for those components that did not provide automated functional tests, the fulfillment of 
 their fundamental functional requirements was tougher to track, resulting in a non--viable
@@
 [Online]. Available: \url{https://www.indigo-datacloud.eu/documents/initial-plan-wp3-d31}.
+[Accessed 14 Feb. 2019].
+\bibitem{indigo-d33}
+J. Gomes \emph{et al.},
+``WP3 Final Report,''
+\emph{INDIGO--DataCloud Deliverable 3.3}.
+[Online]. Available: \url{https://www.indigo-datacloud.eu/documents/wp3-final-report-d33}.
 [Accessed 14 Feb. 2019].

### 2020-01-03 paper-jgc.tex
Comment unnecessary references

@@
 
-\bibitem{h2020}
-\emph{Horizon 2020} -- European Commission.
-[Online]. Available: \url{https://ec.europa.eu/programmes/horizon2020/}.
-[Accessed 14 Feb. 2019].
+%\bibitem{h2020}
+%\emph{Horizon 2020} -- European Commission.
+%[Online]. Available: \url{https://ec.europa.eu/programmes/horizon2020/}.
+%[Accessed 14 Feb. 2019].
 
@@
 
-\bibitem{aiftimiei}
-C Aiftimiei, A Ceccanti, D Dongiovanni, A Di Meglio and F Giacomini,
-``Improving the quality of EMI Releases by leveraging the EMI Testing Infrastructure,''
-\emph{J. Phys. Conf. Ser.},
-vol. 396, no. 5, 2012.
+%\bibitem{aiftimiei}
+%C Aiftimiei, A Ceccanti, D Dongiovanni, A Di Meglio and F Giacomini,
+%``Improving the quality of EMI Releases by leveraging the EMI Testing Infrastructure,''
+%\emph{J. Phys. Conf. Ser.},
+%vol. 396, no. 5, 2012.
 
@@
 
-\bibitem{soltesz}
-S. Soltesz \emph{et al.},
-``Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,''
-\emph{SIGOPS Oper. Syst. Rev.},
-vol. 41, no. 3, pp. 275--287, Mar. 2007.
-
-\bibitem{radatz}
-``IEEE Standard Glossary of Software Engineering Terminology,''
-\emph{IEEE Std. 610.12--1990}, pp.1--84, Dec. 31 1990.
-
-\bibitem{sommerville}
-I. Sommerville,
-``Integrated requirements engineering: A tutorial,''
-\emph{IEEE Softw.}, vol. 22, no. 1, pp. 16--23, Jan.--Feb. 2005.
-
-\bibitem{tashtoush}
-Y. Tashtoush, Z. Odat, I. Alsmadi, M. Yatim,
-``Impact of programming features on code readability,''
-\emph{Int. J. Softw. Eng. Appl.},
-vol. 7, no. 6, pp. 441--458, Nov. 2013.
-
-\bibitem{buse}
-R. P. Buse and W. R. Weimer,
-``Learning a metric for code readability,''
-\emph{IEEE Trans. Softw. Eng.},
-vol. 36, no. 4, pp. 546--558, Jul. 2010.
-
-\bibitem{baggen}
-R. Baggen, J. P. Correia, K. Schill and J. Visser,
-``Standardized code quality benchmarking for improving software maintainability,''
-\emph{Softw. Qual. J.},
-vol. 20, no. 2, pp. 287--307, Jun. 2012.
-
-\bibitem{fenton}
-N. Fenton  and  S.  L.  Pfleeger,
-``Software  Metrics:  A  Rigorous  and  Practical Approach,''
-\emph{International Thomson Computer Press},
-London, UK, second edition, 1997.
-
-\bibitem{capers}
-J. Capers,
-``Applied Software Measurement (2nd Ed.): Assuring Productivity and Quality,''
-\emph{McGraw Hill}, New York, NY, USA, 1997.
-
-\bibitem{goodman}
-P. Goodman,
-``Practical Implementation of Software Metrics,''
-\emph{McGraw Hill}, New York, NY, USA, 1993.
-
-\bibitem{mccabe}
-T. J. McCabe,
-``A Complexity Measure,''
-\emph{IEEE Trans. Softw. Eng.},
-vol. 2, no. 4, pp. 308--320, Dec. 1976.
-
-\bibitem{chidamber}
-S. R. Chidamber and C. E. Kemerer,
-``A metrics suite for Object--oriented design,''
-\emph{IEEE Trans. Softw. Eng.},
-vol. 20, no. 6, pp. 476--493, Jun 1994.
-
-\bibitem{lorenz}
-M. Lorenz and J. Kidd,
-\emph{Object--Oriented Software Metrics: A Practical Approach}.
-Prentice-Hall, Inc., Upper Saddle River, NJ, USA, 1994.
-
-\bibitem{myers}
-G. J. Myers, C. Sandler and T. Badgett,
-\emph{The art of software testing, 3rd Ed.}.
-Wiley Publishing, Dec. 2011.
-
-\bibitem{athanasiou}
-D. Athanasiou, A. Nugroho, J. Visser and A. Zaidman,
-``Test Code Quality and Its Relation to Issue Handling Performance,''
-\emph{IEEE Trans. Softw. Eng.},
-vol. 40, no. 11, pp. 1100--1125, Nov. 2014.
-
-\bibitem{horgan}
-J. R. Horgan, S. London and M. R. Lyu,
-``Achieving software quality with testing coverage measuring,''
-\emph{Comput.},
-vol. 27, no. 9, pp. 60--69, Sep. 1994.
+%\bibitem{soltesz}
+%S. Soltesz \emph{et al.},
+%``Container-based Operating System Virtualization: A Scalable, High-performance Alternative to Hypervisors,''
+%\emph{SIGOPS Oper. Syst. Rev.},
+%vol. 41, no. 3, pp. 275--287, Mar. 2007.
+
+%\bibitem{radatz}
+%``IEEE Standard Glossary of Software Engineering Terminology,''
+%\emph{IEEE Std. 610.12--1990}, pp.1--84, Dec. 31 1990.
+
+%\bibitem{sommerville}
+%I. Sommerville,
+%``Integrated requirements engineering: A tutorial,''
+%\emph{IEEE Softw.}, vol. 22, no. 1, pp. 16--23, Jan.--Feb. 2005.
+
+%\bibitem{tashtoush}
+%Y. Tashtoush, Z. Odat, I. Alsmadi, M. Yatim,
+%``Impact of programming features on code readability,''
+%\emph{Int. J. Softw. Eng. Appl.},
+%vol. 7, no. 6, pp. 441--458, Nov. 2013.
+
+%\bibitem{buse}
+%R. P. Buse and W. R. Weimer,
+%``Learning a metric for code readability,''
+%\emph{IEEE Trans. Softw. Eng.},
+%vol. 36, no. 4, pp. 546--558, Jul. 2010.
+
+%\bibitem{baggen}
+%R. Baggen, J. P. Correia, K. Schill and J. Visser,
+%``Standardized code quality benchmarking for improving software maintainability,''
+%\emph{Softw. Qual. J.},
+%vol. 20, no. 2, pp. 287--307, Jun. 2012.
+
+%\bibitem{fenton}
+%N. Fenton  and  S.  L.  Pfleeger,
+%``Software  Metrics:  A  Rigorous  and  Practical Approach,''
+%\emph{International Thomson Computer Press},
+%London, UK, second edition, 1997.
+
+%\bibitem{capers}
+%J. Capers,
+%``Applied Software Measurement (2nd Ed.): Assuring Productivity and Quality,''
+%\emph{McGraw Hill}, New York, NY, USA, 1997.
+
+%\bibitem{goodman}
+%P. Goodman,
+%``Practical Implementation of Software Metrics,''
+%\emph{McGraw Hill}, New York, NY, USA, 1993.
+
+%\bibitem{mccabe}
+%T. J. McCabe,
+%``A Complexity Measure,''
+%\emph{IEEE Trans. Softw. Eng.},
+%vol. 2, no. 4, pp. 308--320, Dec. 1976.
+
+%\bibitem{chidamber}
+%S. R. Chidamber and C. E. Kemerer,
+%``A metrics suite for Object--oriented design,''
+%\emph{IEEE Trans. Softw. Eng.},
+%vol. 20, no. 6, pp. 476--493, Jun 1994.
+
+%\bibitem{lorenz}
+%M. Lorenz and J. Kidd,
+%\emph{Object--Oriented Software Metrics: A Practical Approach}.
+%Prentice-Hall, Inc., Upper Saddle River, NJ, USA, 1994.
+
+%\bibitem{myers}
+%G. J. Myers, C. Sandler and T. Badgett,
+%\emph{The art of software testing, 3rd Ed.}.
+%Wiley Publishing, Dec. 2011.
+
+%\bibitem{athanasiou}
+%D. Athanasiou, A. Nugroho, J. Visser and A. Zaidman,
+%``Test Code Quality and Its Relation to Issue Handling Performance,''
+%\emph{IEEE Trans. Softw. Eng.},
+%vol. 40, no. 11, pp. 1100--1125, Nov. 2014.
+
+%\bibitem{horgan}
+%J. R. Horgan, S. London and M. R. Lyu,
+%``Achieving software quality with testing coverage measuring,''
+%\emph{Comput.},
+%vol. 27, no. 9, pp. 60--69, Sep. 1994.
 
